### 2017-11-24 thesis/contents/abstract.tex
Initial ShareLaTeX Import

Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches. TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear. Single-tensor diffusion tensor neuroimaging (DTI) studies of TN revealed diffusivity changes in the cistern CNV. However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers. Thereby the study of TN using diffusion imaging is time-consuming and error-prone due to the complete manual involvement in all steps of data process, placement of region-of-interest (ROI), and tractography parameter tuning, and data analysis. 
There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments. 2) to apply the method developed to the analysis of TN in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further. 
The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway. 
Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve. In Study II, we discover the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image. In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) software framework. Finally, in Study IV, we present an end-to-end machine-learning TN classification study that auto-discovers diffusivity disruptions in the cistern/REZ CNV and thalamo-cortico S1 pathway. 
In sum, this thesis presents a detailed roadmap of both the development and application of a novel end-to-end diffusion tractography machine learning classification. The application of this system to TN revealed specific diffusivity changes in deeper white matter structures, and pin-points the exact location of the diffusivity disruption at millimetre level.

### 2017-11-24 thesis/contents/acknowledgements.tex
Initial ShareLaTeX Import

\chapter{Acknowledgements}

The journey towards finalizing this thesis has been a long one. Much has transpired in progress, and I have met many amazing people along this trek. Looking back, I would not have imagined the various opportunities, and happenstances that revealed themselves to me, and I am forever grateful.

First I thank my parents. They have sacrificed their careers and endured much hardship by immigrating to Canada, to provide me with a better future. All my achievements are only possible after I stand on their shoulders.

I met my amazing wife Jidan Zhong in 2012, at Human Brain Mapping conference in Beijing. Little did I know that I had found my kindred spirit. Without her continue to support and persistent encouragement, this thesis would not be possible.

Importantly, I thank my mentor Dr. Mojgan Hodaie. She took a chance on me when I had zero experience in neuroscience, and graciously supported me while I was having a hard time being accepted into the graduate program. I also thank her incredible patience and support for during my thesis writing. Dr. Hodaie is the most accomplished and hard-working person I have ever known. She is a world-class neurosurgeon, while at the same time balances her time between graduate mentoring, grant writing, international neurosurgical education, and at the same time multiple projects in scientific research. Every time I feel dejected from the workload, I always remind myself that Moji is doing more and that I have no excuses. It is truly my fortune to have her as my PI, and words cannot express how thankful I am.

I would like to thank my graduate committee members Dr. Karen Davis and Dr. David Mikulis. Whose extensible experience and insights were invaluable for guiding me through this process. Dr. Davis' impressive scientific acumen, and Dr. Mikulis' enthusiasm will forever stay with me.

I would also like to thank the past and present graduate and post-graduate members of the Hodaie Lab, whom I have shared an unforgettable period of my life: Danielle DeSouza, Allie Rosen, David Hayes, Brendan Behan, Matt Walker and Peter Hung. I also thank Jessica Quan, who introduced me to Dr. Hodaie and spark my journey in Neuroscience.

I was fortunate and thankful to receive support from Multiple Sclerosis Society of Canada, James F. Crother's Family Fellowship in peripheral nerve damage, and Unilever/Lipton graduate fellowships.

I thank all the research assistants over the years that helped me to perform the thankless job of download and organize patient data: Philip Yee, Udi Blankstein, Tessa Burkholder, Kevin Liang, and Erika Wharton-Shukster.

I would like to thank the friends I made over the years at the University of Toronto Chinese Martial Arts Club. We had some good times. While the club is now defunct, the friendships made there will last a lifetime.

Lastly, I would like to thank my martial arts mentor Dr. James Guo, and the Jifeng crew. The training I received and allowed me to grow beyond my limits, and instilled in me values that made this possible.

### 2017-11-24 thesis/contents/ap_paper.tex
Initial ShareLaTeX Import

\graphicspath{{images/ap-reg/}}

\chapter{Study II: Diffusion Weighted Image to T1 Co-registration Without Reverse-blip: Investigation of Best Practices}
\chaptermark{Study II}
\section{Abstract}

\textbf{Purpose:} The registration or alignment of diffusion weighted images (DWI) with other imaging modalities is a critical step in neuroimaging analysis. Within-subject T1-DWI co-registration is particularly instrumental, although DWI acquisitions are susceptible to linear and non-linear field distortions, and it becomes even more important with increasing dataset size. DWI-derived scalar images are commonly used as intermediates for T1-DWI co-registration, and the resulting registration transforms are applied to all other scalar images for analysis. The ideal registration intermediate should register well to T1 and other multimodal images and be practically easy to obtain. It is however, currently unclear which DWI-derived scalar image serves as the best intermediate. We aim to determine the best, practical, intermediate for image co-registration.

\textbf{Methods:} T1 and DWI images were acquired from 20 healthy subjects. DWIs were acquired with 60 directions.  Six DWI-derived scalar images were compared including: 1) fractional anisotropy (FA); 2) generalized FA (GFA); 3) B0 images; 4) mean DWIs with the B0 image (MDWI); 5) anisotropic power (AP) images. Registrations were performed using Anatomical Normalization Tools (ANTs) with affine registration and symmetric diffeomorphic registrations. Pair-wise log joint histograms of the DWIs were plotted to examine possible group biases. Mutual information (MI) and cross-correlation (CC) scores were used to measure the similarities between the deformed DWIs and T1 images. Percent change in these similarity scores relative to scores derived from the identity (baseline) transformation were used for comparison, with higher values indicating better registration.

\textbf{Results:} AP showed the smallest variability in registration improvements across all the tested DWI derived scalar images, and show the highest average percent changes with CC registration cost function (CC=1.2\%, MI=15\%). In contrast, the FA and GFA transforms resulted in significantly poorer registration across DWI types.

\textbf{Conclusions:} The AP image was the DWI-derived scalar image that provided the most consistent registration to all other images. Practically, it is generated easily and so could be implemented in basic and clinical research pipelines currently using other intermediates. Given these findings, it is recommended that AP images be used for T1–DWI co-registration, and that FA and GFA images in particular be avoided.

\section{Introduction}

Diffusion magnetic resonance imaging (dMRI) allows for visualization and quantification of the brain’s white matter by measuring the anisotropy of water molecules \cite{Lenglet2009f}. The resulting diffusivity parameters are used to infer white matter directionality which are visualized using diffusion tensor images \cite{Basser2002}. These form the basis of study for a wide range of disease processes \cite{Chen2016a,Ciccarelli2008,Hodaie2012g}. DMRI has become an essential tool in the in vivo analysis of brain white matter, particularly when used at the group level \cite{Chen2016} to compare diffusion metrics and to model white matter tracts using tractography \cite{Chen2016,Chen2015c,Hodaie2010}. An increasing number of neuroimaging techniques, including structural connectivity analysis, also depend on dMRI to segment white matter regions \cite{Moayedi2012a,McGrath2013,Wiech2014}.

Image registration permits the transformation of different, individual, diffusion images into one brain template and allows for group analysis. This multi-step process involves the accurate co-registration of within-subject diffusion-weighted images (DWI) and anatomical T1-weighted (T1) images, followed by registration of between-subject T1s to create a common template. Individual T1 images typically serve as the intermediate space due to their high spatial resolution and low incidence of distortions \cite{Avants2009,Brown1992,Klein2009,Tustison2014}. The registration of T1s to DWIs is critical to ensure the validity, reliability, and interpretability of the final results (e.g., in correlating DWI data to other measures, and for use in group analyses e.g. \cite{Chen2016}. However, while T1 to DWI co-registration is a common procedure in neuroimaging studies, it is currently unclear which DWI-derived scalar image serves as the best intermediate.

The greatest challenge of T1-DWI co-registration is that DWI acquisitions are susceptible to both affine/linear (i.e. eddy-current and head motion) and non-linear echo planar image field distortions \cite{Rohde2004}. The most common strategies to account for such issues include: 1) correcting all DWI distortions before co-registration with a T1 image, and/or 2) using non-linear co-registration transformations to best warp the anatomical image to DWI space. Corrections for DWI distortions generally involve affine registration of each of the diffusion gradient images to a non-diffusion-weighted image (B0), followed by a rotational correction of the original diffusion gradient b-matrix \cite{Leemans2009}. This approach, however, does not account for non-linear distortions, such as those typically found in the brainstem and frontal and temporal cortices, that are due to MRI field inhomogeneity. Affine co-registration of the T1 image directly to DWI results in poor and highly variable overlaps – increasing the likelihood of both type I and type II errors.

One strategy is to calculate an “anti-distortion” image by acquiring either an extra set of B0 images, or a full DWI sequence with reversed phase-encoding, followed by the construction of a displacement field map which provides an estimate of the undistorted DWI using a least squares \cite{Andersson2003c}, diffeomorphic \cite{Irfanoglu2015}, or Gaussian approach for acquisitions with high b-values \cite{Andersson2015}. However, most prevailing MR datasets, particularly in the clinical domain, still use lower b-values, and lack reversed B0 acquisitions. In such cases, the most commonly used strategy to minimize errors is to non-linearly co-register the T1 image to an intermediate DWI-derived scalar image. Although it is also possible to indirectly perform T1-DWI co-registration through high resolution T2-weighted images, as with the other issues, clinical researchers rarely acquire such scans. As such, we have focused exclusively on direct within-subject DWI-T1 co-registration solutions. 
An additional important consideration is the ability of a registration transform to improve the registration for all relevant DWI-derived scalar images. In addition to the distortion issue, a reliable intermediate image must also be chosen for T1-DWI co-registration, from which the registration transform can be derived. Importantly, since MR image intensities can be inverse (e.g. T1-like and T2-like intensities),  registration improvements may not be uniform, and instead vary depending on the intensity profiles similarities. Despite these concerns, the co-registration utility of the most common, easily producible, DWI-derived intermediates has not yet been investigated.

\subsection{Image Intermediates}
Currently, the five most common DWI-derived scalar images, which are most likely to be of practical use as co-registration intermediates, include: 1) fractional anisotropy (FA) images \cite{Basser2002,Sboto-Frankenstein2013}; 2) generalized FA (GFA)  images; 3) non-diffusion-weighted B0 images (B0); 4) mean DWIs (MDWI) where the DWI acquisition, often represented as a 4D image, is averaged across the sequences;  5) and the anisotropic power (AP) image \cite{DellAcqua2014}, which is an anisotropy map derived from the spherical harmonics estimate coefficients of high angular resolution DWI images. AP directly derives a T1-like contrast image from the diffusion weighted image itself, and therefore shows promise in improving T1-DWI co-registration. Some images, such as B0 and mean DWI, are derived directly from the DWI acquisition volume, while others (FA, GFA, AP) are parametric image maps indirectly calculated from mathematical diffusion models.

The simplest scalar image is the B0 images as part of the DWI acquisition. B0 is used as part of the pipeline for grouped diffusion studies \cite{Gupta2016,Yeatman2012}and to register T1 anatomical to diffusion tensor images for structural connectivity studies \cite{Cao2013}. Since B0’s intensity profile matches well with T2 anatomical images, there also has been attempts to use inverse intensity to optimize registration with T1 using B0 \cite{Bhushan2015}.

The second set of scalar images is derived by taking the mean of the DWI dataset (MDWI) by collapsing the gradient dimension, which transforms the 4D dataset to 3D. MDWI is commonly used in multi-modal registration pipelines \cite{Peng2009b}.

Fractional anisotropy (FA) image \cite{Basser2011c} is also commonly used as a registration intermediate. Due to its high intensity in white matter regions, FA is commonly used with probabilistic tractography registration to standard T1 space \cite{Mansour2013,Salomons2012}. Research also shows that normalized cross-correlation and template matching can improve FA-T1 registration \cite{Malinsky2013}.

An alternative to FA is to use the generalized fractional anisotropy (GFA) image. Since FA shows low intensity in cross-fiber regions, it may adversely affect registration in those areas. GFA can properly characterize crossing-fiber intensities, and therefore may be a better candidate than FA. The GFA is computed from the SH coefficients of spherical harmonics based methods such as Q-ball, where the diffusion orientation distribution function (ODF) is denoted as  $\Psi(u)$ \cite{Tuch2004}, such that:
\begin{equation}
GFA = \sqrt{\frac{n \sum_{i=1}^{n}( \Psi(u_i) - \langle\Psi\rangle)^2} {(n-1) \sum_{i=1}^{n}\Psi(u_i)^2}}
\end{equation}

where $n$ is the number of gradient directions.

Another promising candidate is the anisotropic power (AP) image \cite{DellAcqua2014}. AP have similar visual intensity to T$_1$, and can be computed from HARDI acquisitions without any acquisition modifications. Since AP are derived from the SH model, it is also capable of proper cross-fiber region characterization. AP images are derived from the even harmonic orders ($l$) of SH coefficients \cite{Descoteaux2006,Frank2002} and are defined as
\begin{equation}
AP = \sum_{l=2,4,...} \frac{1}{2l+1}\sum_{m=-l}^{l} \lvert a_{l,m}\lvert^2
\end{equation}
where l is the harmonic order, m is the l-th order coefficient index, a is the corresponding SH coefficient. The AP values are then normalized to non-negative scale by the function:
\begin{equation}
\ln( \frac{AP}{AP_{ref}})
\end{equation}
where $\text{AP}_{\text{ref}}$ is the normalization constant. The resulting image will closely match T$_1$ intensities, and can better capture contrasts in the gray matter structures. AP values stabilize at 6 harmonic orders (28 directions), meaning that AP images can be reliably derived from HARDI sequences with more than or equal to 28 gradient directions.

\subsection{Quantification of Image Registration}

Image registration methodologies involve iterative processing aimed at minimizing the output of a cost function reflecting overlying image differences, thus maximizing the similarity between reference and moving images. Two commonly used cost functions for assessing similarity in medical imaging are cross-correlation (CC) and mutual information (MI) \cite{Liu2013,Mercier2012}. Mutual information is based on the Shannon entropy score, measuring the dispersion of values between two images \cite{Pluim2003}, and by extension represents the likelihood that two images share mutually dependent information \cite{Mattes2003}. Cross-correlation measures image similarities by the profile of their intensities \cite{Maintz1998a}. While CC appears to be a more sensitive measure of matching local features, MI is thought to be a more global measure of similarity \cite{Crum2004}. Furthermore, MI may be the optimal cost function for affine registration, while CC may be a better choice for non-linear registration where local feature-matching is of greater importance \cite{Avants2011}. Unlike T1 anatomical registration comparisons \cite{Klein2009}, where automated segmentation tools are available to serve as an independent metric of comparison, there is currently no automatic method available in DWI-T1 space that can be used to judge the quality of registration. As the best cost function to use for T1-DWI co-registration remain unclear, we use both CC and MI to independently assess the registrations.

There is currently no consensus in the literature about which intermediate image should be used for T1-DWI image registration purposes. This information is pertinent to ensure the increased validity, reliability, and interpretability of neuroimaging results. Our present study aims to quantitatively evaluate the use of each readily available DWI-derived scalar image type (i.e. B0, MDWI, FA, GFA and AP) in the registration process, using CC and MI cost functions as quantitative assessors, in order to determine the most efficient image type for registration and its ability to translate registration improvements to other image types.

\section{Methods}

\subsection{Acquisition and Preprocessing}
T$_1$ and DWIs were acquired from 20 healthy subjects (mean age $31.1\pm10.2$; 10 females). Ethics approval was granted by the University Health Network Research Ethics Board (Toronto, Canada), MR images were acquired at the Toronto Western Hospital, and all subjects gave their informed written consent. DWIs were acquired on a GE HDx 3 Tesla MRI scanner, 8-channels head coil, with 60 directions and 1 non-diffusion-weighted acquisition using the following parameters: 0.9375$\times$0.9375$\times$3 mm3 resolution, matrix=256$\times$256, b=1000 s/$mm\textsuperscript{2}$, field of view (FOV)=240mm, TE=86.4 ms, TR=17000 ms, flip angle=90 deg. T$_1$s were acquired with 0.9375$\times$0.9375$\times$1 $mm\textsuperscript{3}$ resolution, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256$\times$256. DWIs were initially corrected for eddy-current and  head motion with affine registration using FSL FLIRT \cite{Jenkinson2001}, and appropriate rotational corrections to gradient vectors \cite{Leemans2009} using in house software written in Python Numpy.

\subsection{DWI Derived Scalar Image Processing}
Brain masks were created for T$_1$ and DWI using FSL bet \cite{Smith2002}, with bet fraction=0.2. Brain masks were contracted by a 2 mm spherical kernel in FSL to minimize the high-intensity skull halo that occurs in image types such as FA that may interfere with registration. The brain mask is then applied to all DWI derived scalar images.

B0, ADWI, MDWI, FA, GFA and AP images were each generated (Figures 1, 2) as follows: B0, ADWI and MDWI images were created using FSL (Smith et al., 2004); FA images were created in 3D Slicer by calculating the scalars following diffusion tensor estimation \cite{Pieper2006}; GFA and AP images were created with Dipy software library \cite{Descoteaux2007a,Garyfallidis2014}.

Two sets of AP images were generated with an author-implemented AP algorithm in Dipy. The first (denoted as AP1) is created with matching normalization constant with the original method from Dell’Acqua et al \cite{DellAcqua2014} ($\text{AP}_{\text{ref}}$ = 10-5 , then zero all image values \textless 0). The second (denoted as AP2), is created with $\text{AP}_{\text{ref}}$ = 1, and then the entire image values are shifted to the positive range based on the image minimum. Pair-wise log joint histogram of the DWI derived scalar images was plotted \cite{Hunter2007} to evaluate image intensity orthogonality.

\subsection{T1 to DWI Co-registration}
Registrations were performed using Anatomical Normalization Tools (ANTs) \cite{Avants2008b} with affine and symmetric diffeomorphic registration. MI was used as the cost function for affine registration, while CC and MI were used separately as the SyN registration cost-functions. Parameters for affine registration: step size = 0.1, metric = mutual-information (MI), convergence = 10000x10000x10000x10000x10000, shrink factors = 5x4x3x2x1, smoothing sigmas = 4x3x2x1x0mm. Parameters for symmetric diffeomorphic registration (SyN): MI metric weight = 1, MI bins = 32; CC metric weight = 1, CC radius = 3; convergence = 50x35x15,1e-7, shrink-factors = 3x2x1, smoothing sigmas = 2x1x0mm, and use-histogram-matching = true.

The resulting registration transforms from each DWI derived scalar image were applied to all DWI derived scalar image types including themselves using 3rd order B-spline interpolation, where they were projected into T$_1$ space. For each DWI scalar image, an identity transformation ($T_{\text{Identity}}$), affine-only transform ($T_{\text{Affine}}$), and affine with SyN transforms ($T_{\text{Affine}}$+SyN) were applied separately. MI and CC similarity metrics between the transformed DWI scalar and T$_1$ images were obtained using ANTs. Percent change in both CC and MI similarity scores for each type of transformed images were calculated as:
\begin{equation}
\text{Percent Change} = \frac{\text{Transformed} - \text{Identity}}{\text{Identity}} \times 100
\end{equation}
The higher the percent change is, the more similar the transformed DWI scalar and T$_1$ images are after accounting for the identity transformation.
ANOVA and Tukey post-hoc tests were performed using R statistics software (R Core Team, 2014) to compare differences in percent changes between similarity scores of images deformed by their own $T_{\text{Affine}}$+SyN (autodeformation: $D_{\text{auto}}$; e.g. comparing pre- to post-FA scores when FA image was used as the intermediate) as well as similarity scores across groups ($D_{\text{all}}$; e.g. comparing scores resulting from using the FA intermediate to transform all other image types).

Additionally, we used a joint log-histogram of all the DWI derived scalar images to examine possible group biases in DWI scalar types. We intuited that the scalar maps can roughly be divided into two primary groups: those that are directly computed from the DWI directions (DSG), and share similar intensity profiles, such as B0, MDWI, ADWI; and those computed from anisotropic diffusion models (ASG), such as FA, GFA, and AP. It’s possible that there exists a degree of orthogonality between image types, such that there may be no single image type that can equally translate the registration improvements to other images with different intensity profiles. The joint log-histogram is to check for the existence of this type of orthogonality.

\subsection{Results}
Pair-wise joint log histogram (Figure 3) of the DWI derived scalar images revealed distinct correlation patterns in some pairs of images. FA–AP and FA–GFA showed distinct diagonal correlations patterns with each other; while B0 showed no clear pattern of correlation with other image types, with the exception of MDWI. The scalar images FA, GFA and AP are parameter maps derived from diffusion models, whereas B0 and MDWI are directly derived from the DWI image sequence. We thus grouped FA, AP, and GFA based on their intensity histogram correlations as the anisotropic scalar group (ASG), while B0 and MDWI form a directly derived scalar group (DSG).

\subsubsection{Differences in initial similarity scores affect registration accuracy}
Visual overlay of MDWI to T1 co-registrations (Figure 4) suggests that DWI derived scalar image with higher initial CC and MI scores (Figures 5) appear to result in more accurate registrations in the brainstem, insula, temporal cortices, and lateral ventricles (especially around the ventricular-caudate boundary where CSF/grey/white-matter are found in close proximity) compared to the other image types. Visually, B0, FA and GFA show registration inaccuracies fitting the areas surrounding the brainstem and the lateral ventricles (Figure 4), while MDWI and AP result in improved registration in these key areas. The results of MDWI and AP are visually very similar.

The initial T1–intermediate CC and MI similarity scores without any applied transformation ($T_{\text{Identity}}$) showed varied initial values. MDWI showed the highest initial CC scores, with FA images showing lowest initial CC (Figure 5); AP showed similar initial CC score with B0 and GFA. For MI scores (it is important to note that a more negative MI score implies higher similarity in this study), AP showed very similar initial MI score when compared to MDWI (best), while GFA resulted in the highest (worst) initial MI score.

\subsubsection{Registration improvements are not uniformly transferred when applied to other image types}
The progression of similarity scores following $T_{\text{Affine}}$ and $T_{\text{Affine}}$+SyN for each of the scalar images (B0, ADWI, MDWI, FA, GFA, and AP) showed similar trends between CC and MI cost-functions (Figures 5,  Supplementary 1; top rows). In general, the similarities of the images between T1 and transformed DWI scalar images increased marginally after $T_{\text{Affine}}$, and substantially after $T_{\text{Affine}}$+SyN for most of the co-registration transforms. An exception is FA transforms, where CC scores decreased when the transform was applied to other images (Figure 5; rows 1-2, column 4), with the exception of AP.

Improvement in CC and MI measures were not uniform for all DWI derived scalar images; this was better illustrated by percent changes (Figures 5; bottom rows). Different intermediate transforms differentially affected the scores in accordance with the degree of intensity correlations as measured in Figure 3. Therefore, DSG transforms preferentially improved DSG, and ASG transforms improved ASG images. FA transforms, however, are notable, where CC scores across all DWI derived scalar images diverged for both CC and MI cost functions (Figures 5, Supplementary 2; column 4). In fact, FA transforms substantially decreased DSG CC scores, and also negatively impacted DSG MI scores as well. FA transforms also showed exception to GFA, where it affected GFA negatively similar to DSG images. GFA behaved more like DSG images, where B0 and MDWI all favorably affected its CC and MI scores, while at the same time it was strongly affected by AP as well. However GFA seem to behave differently under autodeformation, as evident by its distinct MI changes (Figures 5; column 5). AP transforms showed the least amount of variability when applied to DWI derived scalar images, and consistently improved all the image scores.

\subsubsection{AP-derived transforms show the most consistent improvements when applied to other image types}

Between-group statistical comparisons of only $T_{\text{Affine}}$+SyN percent change revealed significant findings. Under the CC cost-function (Figure 6), percent change in $D_{\text{auto}}$ (Figure 6, bottom panel, column 3) showed that FA and GFA have the highest percent change, while AP has the lowest percent change. With MI scores, $D_{\text{auto}}$ showed that only GFA percent changes are significantly greater than others (Figure 6, bottom panel, column 4). GFA, however, also shows the greatest variability in distribution (Figure 6, top panel, column 4).  $D_{\text{all}}$ in contrast, showed that FA is an outlier where its percentage change is significantly lower than all other images; it is also notable that it shows a distinct bimodal distribution (Figure 6, top panel, column 1); GFA was shown to be significantly higher than B0, MDWI and FA, but not AP. MI percent changes are not significantly different from each other (Figure 6, bottom panel, column 2). MI cost-function results were similar to that of the CC cost-function (Supplementary 2).

\section{Discussion}

This study presents a comparison of the DWI derived scalar image to use for T1 to DWI co-registration. AP images showed the most consistent improvements in image scores across all of the intermediate scalar images tested, suggesting that AP can offer the most consistent T1-DWI co-registration improvements, while FA and GFA images are poorer choices. We observed, via CC and MI scores, that non-linear registration shows similar trends in registration performance. DWI derived image registration improvements show biases that depend on the intensity correlations with directly derived scalar image group (DSG) or anisotropic scalar image group (ASG) similarities. FA showed significantly lower MI and CC similarity scores that worsened registration results. GFA showed a significant increase in similarity scores, but also greater variability in percent increases when its transform was applied to other image types – suggesting that while this measure is better than FA alone, it is still highly unreliable across subjects.

\subsubsection{Registration improvements are biased towards similar image intensity groups}

The log joint histogram showed that there is a degree of intensity orthogonality between FA, AP, and GFA as a group (ASG) compared to B0, ADWI and MDWI (DSG), suggesting that there is some connection between co-registration performance and intensity similarity (Figure 3). The greater the intensity similarities between the two images, the greater the effect of the same set of registration transforms. Intensity similarity is not the only factor, however, as evident by the exception to this rule in FA and GFA deformation results (Figure 5). This suggests that intensity orthogonality is not a limiting factor in registration transfer across these scalar images. Therefore, our hypothesis that there exists a better DWI intermediate scalar image that can improve multimodal registration is correct.

\subsubsection{Autodeformation is not sufficient to assess registration intermediate performance}

Our study demonstrates clear differences in initial similarity scores (i.e. CC and MI) across different DWI scalar images when compared to T1, where MDWI had the highest initial CC and MI scores.  Moreover, the AP and MDWI images clearly outperformed the B0, FA, GFA and intermediates in visual inspections.

As expected, autodeformation (e.g. FA to T1 registration applied to the FA image, from greek autos: self) yielded the best metric of similarity compared to when the transform was applied to other image types. The optimal T1-DWI co-registration performance of an intermediate image could be identified by the image’s deformation by its own transform. However, we show that only comparing the autodeformation results is not ideal (Figure 7, 8). Specifically, while the FA autodeformation (using $T_{\text{Affine}}$+SyN) performed better compared to all other autodeformations, registrations of other scalar images based on FA transformations clearly resulted in the worst results. Therefore, future comparisons of candidate intermediate images should consider not only the autodeformation, but also the ability of a transform to deform other scalar images. It is also worth noting that affine transformations marginally improve MI scores – which is expected as the head orientations within MR scanners usually do not deviate very far between acquisitions, but most registration improvements stem from corrections to DWI distortions through non-linear registration.

The larger percent change improvements in FA and GFA can be explained by the ceiling effect of registration, where images that are initially more similar to the T1 anatomical image (e.g. MDWI or AP) would have less room for improvements compared to those that are less similar (e.g. FA or GFA). As such, the absolute difference in similarity scores should still be included in the final assessment of registration. It is also important to note that computational algorithms using CC and MI as the optimization metric in the registration process itself cannot fully characterize the goodness-of-fit of one set of images when matched with another. Therefore, our current use of these scores does not preclude the possibility that certain image sub-regions would be better registered using other scalar image types under specific conditions. Moreover, the method of estimating an undistorted DWI with reversed phase-encoding images may still be superior to the use of image intermediates, described here, when this approach is possible. Nonetheless, the post-processing use of intermediate images combined with reverse-phase corrected DWI may result in improved registrations.

As a new DWI derived scalar image, our study is the first to evaluate AP for the purpose of T1-DWI co-registration. AP images can consistently improve similarity outcomes for the various image types and shows more consistent similarity outcomes when compared to other modalities. Further research is needed to determine the optimum method for AP normalization.

\subsubsection{Limitations}

One important limitation is that the DWI voxel sizes are not isotropic. For this study, the DWI images were acquired at 0.94x0.94x3 $mm^3$ voxel resolution. The dataset was acquired on a 3T GE HDx MRI with 8-channels head coil, and therefore does not permit DWI scans of less than 2.6 mm isovoxel resolution at a clinically acceptable scanning time. As such, we compromised on an anisotropic voxel resolution to gain in-plane resolution. Although recent authors have suggested using isotropic up-sampling of MR images to improve image quality \cite{Dyrby2014}, we decided not to introduce up-sampling into our processing pipeline in order to avoid the possibility of additional interpolation-introduced artifacts that may bias towards certain image types.  Clinical datasets are often acquired under time and resource constraints, and therefore we believe our findings will be novel for the application of clinical T1-DWI co-registrations at less than ideal conditions.

\section{Conclusions}
This study demonstrated that for T1-DWI co-registration, using AP as the DWI derived scalar image type resulted in improved registration performance. AP offers more consistent registration across DWI derived scalar images. AP requires the calculations of spherical harmonic coefficients as a preprocessing step, and therefore stabilizes when there are greater than 28 gradient directions. It is thereby unsuitable for some legacy DWI datasets acquired with fewer than 28 directions. MDWI can be readily obtained from existing DWI sequences, and is therefore suitable for all existing datasets. We generally recommend the use of AP where available. In the case where calculation of AP images is not possible, MDWI is a viable alternatives as T1-DWI co-registration intermediates. FA and GFA were the poorest performers, and we recommend not using them for registration purposes.

\begin{figure}[H]
\includegraphics[width=\linewidth]{figure1}
\caption{Illustrated overview of the processing steps to measure DWI to T1 registration similarities.}
\centering
\label{fig:APfigure1}
\end{figure}

 \begin{figure}[H]
 \includegraphics[width=\linewidth]{figure2}
 \caption{DWI derived scalar images (B0, MDWI, FA, GFA, and AP) comparing to T1 (top left).}
 \centering
 \label{fig:APfigure2}
 \end{figure}

\begin{figure}[H]
\includegraphics[width=\linewidth]{figure3}
\caption{Pair-wise joint log histogram of the DWI derived scalar images.}
\centering
\label{fig:APfigure3}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\linewidth]{figure4}
\caption{Visual comparisons of MDWI to T1 co-registrations. 
Shown are examples of MDWI images from one subject deformed using different registration transforms derived from the different image intermediates. 
T1 is overlaid as thresholded outlines in red. 
Primary differences in registration quality can be found in the brainstem (left and middle columns), and in the lateral ventricles, especially in the ventricular-caudate boundary (right column). }
\centering
\label{fig:APfigure4}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\linewidth]{figure5}
\caption{Similarity scores between deformed dMRI scalar images and T1 following registrations with CC as cost function for SyN. 
Three steps of the registration process are plotted: Initial identity transformation (none), affine transformation (affine) and affine+SyN transformation (SyN).
Top panel shows similarity scores as measured by CC and MI, and bottom panel shows average percent change progressions of CC and MI across dMRI scalar images from \protect $T_{\text{Affine}}$ to \protect $T_{\text{Identity}}$, and from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Affine}}$. The error bars denote standard deviation across subjects.}
\centering
\label{fig:APfigure5}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\linewidth]{figure6}
\caption{
Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with CC as cost function. Top row shows the violin plot distribution of CC and MI percent change under All Deformations and Autodeformations. Bottom row plots the statistically significant pair-wise correlations of the different transforms. 
Under All Deformations (\protect $D_{\text{all}}$), CC percent change of FA is shown to be significantly lower (p<0.05) than that of all other transform types. GFA can be observed to have higher percent change than that of B0, MDWI, and FA. Under Autodeformations (\protect $D_{\text{auto}}$), CC percent change of FA and GFA are significantly higher than others, while AP has lower percent change. MI shows little percentage between all the images; GFA shows significantly differences under \protect $D_{\text{auto}}$, but also shows high variability.}
\centering
\label{fig:APfigure6}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\linewidth]{sup1}
\caption{Similarity scores between deformed dMRI scalar images and T1 following registrations with MI as cost function for SyN. 
Three steps of the registration process are plotted: Initial identity transformation (none), affine transformation (affine) and affine+SyN transformation (SyN).
Top panel shows similarity scores as measured by CC and MI, and bottom panel shows average percent change progressions of CC and MI across dMRI scalar images from \protect $T_{\text{Affine}}$ to \protect $T_{\text{Identity}}$, and from \protect $T_{\text{Affine}}$+SyN to\protect  $T_{\text{Affine}}$.}
\centering
\label{fig:APfigure7}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\linewidth]{sup2}
\caption{Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with MI as cost function. Top row shows the violin plot distribution of CC and MI percent change under All Deformations and Autodeformations. Bottom row plots the statistically significant pair-wise correlations of the different transforms.
Under all deformations ( \protect$D_{\text{all}}$), CC percent change of FA is shown to be significantly lower (p<0.05) from that of other transform types. In autodeformations (\protect$D_{\text{auto}}$). FA has higher percent change. MI percent changes show little differences, and once again highlights the high variability of GFA. }
\centering
\label{fig:APfigure8}
\end{figure}

### 2017-11-24 thesis/contents/discussions.tex
Initial ShareLaTeX Import

\graphicspath{{images/discussions/}}

\chapter{General Discussions}
\section{Tractography}
\subsection{Delineating the trigeminal nerve into the brainstem}
The primary motivation to visualize the trigeminal nerve via tractography, was due to the inability to discern both the brainstem trigeminal nerve body, nor the trigeminal nucleus in all known MRI modalities. Before tractography, the curvature of the trigeminal nerve could only be inferred from previous anatomical studies, or by postmortem dissection. 
For successful tractography of the CN V, both DWI sequence and the algorithm choice is paramount. Given a DWI sequence, at 25 gradient directions, cisternal CN V could be readily reconstructed with Gaussian tensors. Resolution of the DWI sequence would affect the output. Unlike major white matter bundles, cranial nerves diameters are in the milimeter range, and therefore voxel sizes greater than 2 mm will result in severe partial volume of cranial nerves. Achieving 2mm isovoxel on a 3T scanner requires 32 channel head coil. Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable. For this reason, the acquisition in-plane resolution can be tune to 1x1 mm, and slice thickness at 3mm. While the acquisition is not ideal, it sacrifices resolution in only the z-axis, the tractography visual fidelity is improved. 
Discerning brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers. A number of methods has been devised, including probabilistic tractography made popular by the FSL software package, which generates image-based results that represents the propagation probability of a seed region. Neurosurgeons however often need visualizations that represent anatomy geometrically. Therefore streamline-based tractography visualizations are of broader appeal. In our experience, both XST and sphere harmic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions. 
Another often overlooked area is the seeding pattern. When tracing the streamlines, an algorithm will need to generate a number of starting points in a defined volume. Common patterns in prepackaged software include in a regularly spaced grid, where seed spacing can make decisive differences in visuals, and random sampling. The seeding pattern is important in both the final visual style of the tracts, as well as if key anatomical landmarks can be properly tracked. Another consideration is the reproducibility of the desired anatomy, as randomly seeded bundles cannot be reproduced exactly. Grid seeding also can produce sparse spacing of the tract bundles when viewed from a certain angle, while random seeding can reduce this kind of visual artifact. A good balance might be to have a reproducible sampling scheme that also is robust to different viewing directions, for example a phyllotaxis pattern.

\subsection{Tractography group registration}
Tractography is a powerful methodology that reveals the morphology of brain white matter that were otherwise previously invisible. It's primary usage however has been limited to case studies and small scale anatomy comparative analysis. The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of large groups.

Group comparative studies requires the establishment of anatomical consensus, such that similar white matter structures between populations can be readily studied. Efforts in MRI medical image registration have paved the way to this end, especially in the fields of diffeomorphic registration. The difficulty in adapting registration to tractography, is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods. The best efforts to date applies affine transformation for between-tract registration \cite{Garyfallidis2015}, however true non-linear geometric registration is still a topic of future research. Our approach is to recast tractography as a brain segmentation method that is spatially-aware. Tractography then can be considered a point-based sub-sampling method that are represented in the same coordinate frame as the native DWI space. Any transforms applied to this DWI space is then applicable also to any tractography models that share its coordinate frame. Existing diffeomorphic registration solutions can then be used to non-linearly distort tractography, and circumvent the computational difficulty of geometry based non-linear registration. The advantage of our tractography registration strategy is that any new advances in the field of MR registration can be readily adopted to improve the output of our method.

Group registration of tractography towards a common template space requires two separate steps: T1---DWI co-registration, and T1---Template registration. Any registration errors in these two steps will become multiplicative in the end result. T1 registration is well studied, however T1---DWI registration has few literature. Therefore it was imperative that we investigated the best method to minimize registration errors that were compatible with our image data format.

\subsection{Measurements from tracts}
Aggregating along-the-measure statistics is a form of clustering that requires projection of tract vertices onto a structural skeleton, or centroid. Common clustering methods calculates a polyline centroid based on different distancing functions between streamlines. These methods however require tractography streamlines to have similar length, and will underperform when tractography structures contain sharp turns. We have explored a number of strategies, and considered adaptation of manifold-based methods, and point-cloud skeletal extraction method from computer graphics literature. These approaches however contain assumptions that do not suit our needs. For example, manifolds commonly consider that the latent structure of the data lie in a lower-dimensional surface, while tractography are common arranged in complex bundles; while point-cloud methods deals with point-cloud data obtained from 3D scanners, that are situation on a 3D shell, while tractography data are solid point clusters. White matter anatomy can have complex crossings and multi-way join and diverging structures that are difficult for existing 3D analytical methods that often need simpler assumptions to reduce the number of unknown parameters. A potentially promising future direction may be to automatically discover these more complex latent structures using deep learning methods.

\section{Trigeminal Neuralgia}

The findings in Study I \ref{section:study1} revealed clearly localized CNV tissue differences in MS-TN and idiopathic TN. The implication is that disruption along the primary trigeminal afferent may lead to similar TN-like pain syndromes. MS is a classic myelin inflammatory disease, and therefore these diseases may share similar disruptions in myelin ultrastructure. In Gamma-Knife radiosurgery treatment for TN, the target is often the Gessarian ganglion, which situates distally from the CNV nerve root, beyond the affect mid-cistern segment. The mechanism for GK treatment may function as a "volume control" for peripheral trigeminal afferents. Where some interaction between peripheral innocuous sensory conductance and the site of proximal CNV disruptions. These interaction may take the form of cross-talk between axons that carry discriminative sensory signals and pain fibers, or maladaptive convergences in the trigeminal nucleus. It's likely that these interaction are probabilistic in nature, due to the observation that TN pain is paroxysmal.

Study IV further suggested that the unaffected CNV diffusivity is also different from that of controls. The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain. In many studies it has also been used as a basis of comparision. This finding however is not as suprising as one may suspect. It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases. Moreover, in third degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space. It is reasonable that these observable macroscopic changes can effect CNS tissue changes in the unaffected side. Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases a more centrally involved changes on the trigeminal nucleus and its associated pathways is possible.

\section{Limitations}

One limitation across the studies in this thesis was the choice of anisotropic voxels of the DWI sequences. This was limited by the clinical 3-channel GE 3T HDx MR scanner, which places constraints both the acquisition resolution and time. Commonly, DWI voxel resolution on a 32-channel 3T research scanner can obtain up to 2mm iso-voxel resolution, on older 8-channel clinical scanners, maximum iso-voxel resolution is limitied at about 2.6mm iso-voxel. This is clearly insufficient for the study of cranial nerves such as CNV. Our resolution is limited to 1x1x3 mm voxel dimensions, in order to retain in-plane resolution at the sacrifice of slice thickness. Despite this, certain regions such as the TPT could not be reliably reconstructed in all of subjects. Anatomy featuring intricate but complex cross-sections, such as the thalamo-insula projections, which involves the claustrum, and medullary decussations of the spinal trigeminal fibers will remain out of reach of the current algorithms without substantial improvement in DWI resolution. Fortunately, SAGIT can adopt MR images of any resolution, so future studies in higher iso-voxel resolution can be immediately processed at soon as they become available.

The results of the thesis ultimately depend on the algorithm performances at each step. The SAGIT framework relies on the accurate image registrations to perform template, and tractography deformations. Therefore any inaccuracies in the multi-step registration pipeline will become cumulative. We went to great length to determine the best method in order to minimize registration error on a massive scale (See \ref{section:study2}), and we stress the importance to strive for registration accuracy.

\section{Future development}

The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers are not limited to the application of TN. The methods can generalize to any anatomy where consistent tracts can be obtained at a group level. The limitation of its examination power is often in the data, both in data volume, resolution, and quality.

An immediate application is to re-examine TPT tracts, both to establish its diffusivity baseline, and to confirm its asymmetry. This can be immediately deployed on the Human Connectome Project (HCP) dataset. Additional examine in a diseased population requires the availability of clinical MR images that at minimum matches the parameters of HCP.

For pain studies, detailed examination of the pain matrix, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest. Multi-modal and multi-variable study with function connectivity with this technique would also be possible.

The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal important insights in the MS population. Where focal lesion groups can be correlated with disease symptoms such as pain and motor disabilities.

Gaussian Process can be used as an regressor to model tissue parameters along merged white-matter bundles. Any MR scalar image can be embedded into the tractography streamlines, so tissue quantifications such as G-ratio, myelin, proton-density, kurtosis, magnetization transer, PET images can be examined. Since GP is a bayesian method, the derived GP model can be used to provide predictive confidence on centroid measures from single subjects. When combined with the end-to-end tractography extraction ability of SAGIT, this provides a basis to build deployable framework for disease inference on new patients.

\section{Conclusions}

In sum, this thesis provides novel evidence on the diffusivity tissue disruptions of the CNV in TN, and that there is evidence of bilateral CNV tissue changes in TN patients when compare to controls. The examination was also extended to the thalamo-cortical S1 pathway and have demonstrated localized white matter changes near the white-grey matter boundary. These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies as a whole. These result demonstrate that TN involve wide-spread changes at all levels of CNS white matter, and that the methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.

### 2017-11-24 thesis/contents/hypotheses.tex
Initial ShareLaTeX Import

\chapter{Aims/Hypotheses}

The general aim of this thesis is quantify the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways in order to identify specific diffusivity pattern that differentiates between TN and non-TN CN V tissue. In achieve this aim, a generalizable group diffusion imaging tractography software framework must be developed to overcome the technical barrier of large scale group tractography in arbitrary populations. A number of confounds on data registration and measure must be resolved in order to gain confidence in the result of this methodology. These include a) establish the feasibility of diffusion tractography of brainstem CN V originating from the trigeminal nucleus; b) establish the best way to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects; c) Make possible non-linear registration of tractography geometries, and to quantify the geometry registration performance at the group level; d) apply previously established methodology to the reconstruction and quantification of the trigeminal sensory pathway.

### 2017-11-24 thesis/contents/paper1.tex
Initial ShareLaTeX Import

\graphicspath{{images/mspaper/drafts/}}

\chapter{STUDY I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
\chaptermark{Study I}
\label{section:study1}

This study was published in Multiple Sclerosis Journal \textcopyright 2015 Chen et al. :

\bibentry{Chen2015a}

\section{Abstract}
\textbf{Background: }Trigeminal neuralgia secondary to multiple sclerosis (MS-TN) is a facial neuropathic pain syndrome similar to classic trigeminal neuralgia (TN). While TN is caused by neurovascular compression of the $5\textsuperscript{th}$ cranial nerve (CN V), how MS-related demyelination correlates with pain in MS-TN is not understood.

\textbf{Objectives: }We aim to examine diffusivities along CN V in MS-TN, TN and controls in order to reveal differential neuroimaging correlates across groups.

\textbf{Methods: }3T MR diffusion weighted, T$_{1}$, T$_{2}$ and FLAIR sequences were acquired for MS-TN, TN, and controls. Multi-tensor tractography was used to delineate CN V across cisternal, root entry zone (REZ), pontine, and peri-lesional segments. Diffusion metrics including FA, RD, AD and MD were measured from each segment.

\textbf{Results:} CN V segments showed distinctive diffusivity patterns. The TN group showed higher FA in the cisternal segment ipsilateral to the side of pain, and lower FA in the ipsilateral REZ segment. The MS-TN group showed lower FA in the ipsilateral peri-lesional segments,  suggesting differential microstructural changes along CN V in these conditions.

\textbf{Conclusions:} The study demonstrates objective differences in CN V microstrucuture in TN and MS-TN using non-invasive neuroimaging. This represents significant improvement in the methods currently available to study pain in MS.

\section{Introduction}
Trigeminal neuralgia is a neuropathic pain disorder characterized by severe, lancinating, facial pain with no major clinical sensory deficits. The pain occurs in one or more territories of the branches of the 5th cranial nerve (CN V) and is often triggered by innocuous stimuli, such as a light touch to the face. Classic trigeminal neuralgia is presumed to occur as a result of neurovascular compression at the root entry zone (REZ) of CN V (also referred as ‘idiopathic’; henceforth referred to simply as TN)(M Hodaie \& Coello, 2013). TN pain symptoms can also occur secondary to multiple sclerosis (MS-TN), where MS patients have a 20-fold risk of developing TN pain than the general population (up to 0.1\%)\cite{VanHecke2014}. 
While experientially similar, the pathophysiology of MS-TN differs from TN and involves CNS demyelination\cite{lazar1979trigeminal,Nurmikko2009}. The manner in which CNS demyelination results in pain is unclear. Brainstem plaques can be common in MS, but not all MS patient with brainstem plaques suffer from MS-TN. Furthermore, the presence of bilateral plaques does not clearly correlate with bilateral MS-TN pain\cite{DeSanti2011a,Love2001a}, highlighting the complex and uncertain relationship between the presence of plaques and pain\cite{DaSilva2005}. Importantly, this also highlights the fact that conventional MR plaque visualization does not fully correlate with MS clinical symptoms or their severity, and so improved neuroimaging methods are needed\cite{Seixas2014}. As classic TN is most commonly associated with vascular compression at the trigeminal nerve root entry zone, it is possible that in some MS patients neurovascular compression can also be present, adding further complexity to the diagnosis and treatment of trigeminal pain\cite{Sandell2010e}.

Understanding the microstructural anatomy of the trigeminal fibres is essential to determine how MS affects CN V brainstem fibres and how alterations to CN V in MS are related to the clinical presentation of pain. Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure 1A-B)\cite{Cruccu2009,DaSilva2005}. Conventional MR imaging cannot visualize in isolation the brainstem fibres of CN V. Furthermore conventional tractography cannot resolve the brainstem crossing fibres\cite{Farquharson2013}, which leads to unreliable CN V delineations within the brainstem, and inaccuracies in diffusivity measurements.

Diffusion tensor imaging (DTI) metrics, based on estimating the Gaussian diffusion of water in the neural tissue, can provide important insights to microstructural neural tissue changes in both pathological and normal states. Metrics including fractional anisotropy (FA), and radial (RD), axial (AD), and mean diffusivities (MD), reflect diffusivity change measured as a parameter of the estimated Gaussian diffusion tensor eigenvectors.  FA reflects the shape of the ellipsoid Gaussian tensor as a measure of anisotropy, and is correlated with both structural and cognitive changes that follow adverse brain changes such as traumatic brain injury, aging and Alzheimer’s disease\cite{Bendlin2008,Charlton2006,Kinnunen2011c,Mielke2012}. RD is a measure of the cross-sectional diffusion as an average of the second and third diffusion eigenvalues, and has been shown to positively correlate with axonal demyelination \cite{Song2002,Song2005}. AD is the greatest diffusion eigenvalue and has been positively correlated with axonal degeneration\cite{Budde2009,Song2002}, and MD is the mean of all three diffusion eigenvalues and is related to the increase in tissue inflammation and edema\cite{Beaulieu2002}. In MS studies, FA, RD and MD appear abnormal in or near MS lesions. The changes in diffusivity alter the Gaussian tensor profile and results in decreased FA, increased RD and increased MD\cite{Braley2012,Janve2013,Klawiter2011,Lin2007,Senda2012g}. While AD is generally not correlated with demyelination\cite{Song2002}, there have been findings associating decreases in AD with the presence of neurological damage in MS\cite{Budde2008,Kim2006}.

The major technical limitation of diffusion tractography to date has been the inability of resolving crossing fibres using a single tensor Gaussian diffusion model\cite{Farquharson2013}, which has proved unsuitable for the visualization of brainstem tracts. However, techniques that use high angular resolution imaging  with an increased number of gradient directions acquired during diffusion weighted imaging (DWI), and advances in multi-tensor tractography (MTT) have improved the ability to delineate fibres in brain regions that have dense fibre crossings, such as the brainstem \cite{Dell'Acqua2007,Descoteaux2009c,Fillard2011,Qazi2009}. The microstructure of CN V in TN has recently been shown to be abnormal in terms of DTI metrics in the trigeminal REZ\cite{Desouza2013}. DTI metrics, as correlates of microstructural anatomy, are also important in identifying the nerve in pain in TN\cite{Hodaie2009a}.

In this study of MS-TN, we hypothesize that A) myelination correlates of the trigeminal brainstem fibres, assessed with diffusion metrics, will reveal a pattern which is unique to MS-TN, despite inherent differences in precise lesion location, and B) diffusion metrics along the nerve will help differentiate between TN, MS-TN and healthy subjects, and may in future serve as a neuroimaging signature of pain in these populations. 
We used MTT and diffusivity metrics to identify specific neuroanatomical diffusivity differences between TN and MS-TN. Specifically, our aims were: 1) Visualize the isolated tracts of brainstem trigeminal fibres using MTT; 2) Sample and compare CN V diffusivities in MS-TN between symptomatic (+) and asymptomatic (-) sides of a) CN V at the REZ, b) brainstem CN V segments unaffected by MS plaques, and c) brainstem CN V segments with the presence of MS plaques; 3) Contrast diffusivity findings with CN V diffusivities in TN patients, MS-TN patients and healthy controls.

\section{Methods}

Magnetic resonance images were acquired using GE Signa HDx 3T scanner with an 8 channel head-coil. MR sequences were acquired from 3 groups (n=10 in each): unilateral MS-TN (4 males, 6 females; mean age  $53\pm8.6$ years) with no evidence of neurovascular compression of CN V at the REZ, unilateral TN (3 males, 7 females; mean age $57.1\pm8.5$ years), and healthy controls (3 males, 7 females; mean age $55.8\pm7.4$ years). Clinical pre-screening of MS patients and their associated lesion locations were performed using T$_{2}$ and FLAIR sequences. However clinical T$_{2}$ slice thickness was too large (\textgreater 3 mm) for proper assessment of 3D lesion volume. Therefore once the lesion was identified, the fast spoiled gradient echo (FSPGR) T$_{1}$ anatomical image was registered to the diffusion image in order to localize the MS lesion across multiple modalities for ROI placement. DWIs were acquired with 1 B$_{0}$ scan, 60 gradient directions, 3mm slice thickness and in-plane resolution of 0.9375$\times$0.9375 mm, b$_{0}$=1000 s/mm$\textsuperscript{2}$, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, matrix=128$\times$128. T$_{1}$ FSPGR anatomical scans were acquired with 1mm slice thickness and in-plane resolution of 0.9375$\times$0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, field of view (FOV)=240 deg, matrix=256$\times$256. T$_{2}$images were acquired with 4mm slice thickness, in-plane resolution=0.4297$\times$0.4297 mm, TE=94.14 ms, TR=5200 ms, flip angle=90 deg, FOV=170 deg, matrix=512$\times$512. FLAIR images were acquired with 4mm slice thickness, in-plane resolution=0.4297$\times$0.4297 mm, TE=141.4 ms, TR=8652 ms, flip angle=90 deg, FOV=220 deg, matrix=384$\times$224.

Images were analyzed with 3D Slicer (NA-MIC©, http://www.slicer.org)\cite{Pieper2004}. DWI sequences were corrected for eddy-current and motion distortions with appropriate corrections to gradient vectors, and imported into 3D Slicer for single-tensor diffusion tractography (SDT), where tracts were generated with initial seed spacing=0.5 mm, initial seeding FA threshold=0.2, stopping FA threshold=0.1, curvature threshold=0.8 and integration distance=0.1. Region-of-interests (ROI) seeding points were generated at a spacing of 0.25mm, or 64 points per voxel; MTT propagations were initiated at C1 threshold=0.2, tensor fraction=0.2, curve radius=0.8 rad, minimal length=10 mm, step size=1 mm. T$_{1}$ anatomical images were registered to the DWI using rigid transformation. To generate CN V tractography, ROIs were placed bilateral at the CN V at the REZ, and were used for both SDT and MTT.

Diffusion parameters (FA, AD, RD, MD) were measured from 4 sets of pre-defined bilateral ROIs (Figure~\ref{fig:MSfigure1}C-D). Cross-sectional areas were based on tractography delineations, and thickness of 2 voxels. Location of the ROIs were defined based on trigeminal anatomy: 1) The cisternal ROI is measured from the nerve segment situated within the cisternal space surrounding the pons, at the mid-point between the Gasserian ganglion and the root entry zone; 2) the REZ ROI is defined at the junction where the CN V crosses from the cisternal space into pontine brainstem. This is where the neural myelination transitions from peripheral myelination (Schwann cells) to central myelination (oligodendrocytes). The ROI must include the cisternal segment and pontine segment at the transitional zone; 3) the pontine ROI is defined in the intra-pontine CN V segment at the mid-point between the REZ and the trigeminal nucleus. In MS-TN patients it is defined as the mid-point between the peri-lesional ROI and the REZ along the nerve that is not visually affected by any MS lesions (and note that subjects with MS lesions that did not allow ROI placement with at least 2 voxel spacing from peri-lesional and REZ ROIs were excluded as confounders); and 4) the peri-lesional ROI is defined as the intra-brainstem CN V segment at the closest proximity to MS lesions in its vicinity. In TN and control subjects, the corresponding ROI was the nerve segment closest in proximity to the trigeminal nucleus. In these patient groups, the pontine and peri-lesional areas are referred to as distal and proximal brainstem ROIs, respectively.

Statistical analysis were performed using R statistics software suite(R Development Core Team, 2012). Segments and diffusivity measurements were analyzed separately. Patient group is considered an independent and between-subject factor; symptomatic and asymptomatic sides within the same patient are considered within-subject measures; in controls, the mean measures of left/right sides are used for comparison against other groups. A linear mixed model\cite{Pinheiro2011} is used to determine the statistical significance of within- and between-subject effects. Post hoc analyses were performed using pair-wise Student’s t-test, where multiple comparisons are corrected with false discovery rate correction\cite{Hothorn2008}. Statistical results were plotted\cite{Hunter2007} as diffusivity difference estimate matrices between anatomical regions and different subject groups. Values are mapped as red/blue based on mixed linear regression estimates. Positive estimates, where the group diffusivity parameter is greater in the vertical axis versus the horizontal axis are in red, and negative coefficients are in blue.

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure1.png}
\caption{Figure shows the MR images of a typical MS-TN patient at the level of pons. The presence of MS plaque in the CNS at the level of trigeminal nerve is characteristic of MS-TN (panel A). Panel B shows the expected course of the trigeminal nerve (CN V) as it synapses onto the principal trigeminal and motor trigeminal nuclei. Note that the exact anatomical relationship between the MS plaque and CN V in this instance cannot be clearly discerned in T2 due to the lack of contrast between the nerve and the surrounding tissue (panels A,B,C). Panels C and D show the placements of the ROIs along the CN V nerve bilaterally in order to measure diffusivity statistics, overlaid with the axial anatomical/ DTI image. The colors of the DTI image are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. Diffusion statistics were measured from four groups of regions: cisternal segment, root entry zone (REZ), pontine, and peri-lesional, as illustrated. Analogous regions in TN patients and controls were similarly placed. Care was taken to differentiate pontine and peri- lesional ROIs in order to measure lesioned and unlesioned regions.}
\centering
\label{fig:MSfigure1}
\end{figure}

\section{Results}
\subsection{MTT permits visualization of brainstem trigeminal fibres}
MTT allowed for the visualization of brainstem CN V fibres in isolation, by distinguishing them from the surrounding crossing cerebellar peduncular fibres. The SDT method could not adequately delineate the nerve (Figure 2). Comparison of tractography results between SDT and MTT showed that SDT was unable to differentiate fibres crossing the cerebellar peduncle fibres, while MTT delineated the brainstem course of CN V into the region of the trigeminal main sensory nucleus. Delineation of MS-TN CN V by MTT revealed that the tracts became visibly sparse as they came into proximity of the MS plaque (Figure 3A). FA decreases in proximity to the MS plaque could also be visualized by colour shift in the tractography model with more orange/yellow colour presented. The fibres of the unaffected CN V did not exhibit this change, and extended caudally into the trigeminal nuclei (Figure 3B).

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure2.png}
\caption{Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views. The reconstructed tracts are overlaid onto the axial DTI/anatomical images. The colors of the underlying scan are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. The colors of the tractography fibers represent the spectrum values of FA (0 to 1), as shown by the legend at the top right corner. Panel A shows that SDT tractography delineated the cerebellar peduncle fibers, but cannot distinguish the brainstem trigeminal fibers (yellow arrows). Red arrows denote the starting point of tract generation for both SDT and MTT methods. Panel B shows that MTT tractography visualized CN V as it coursed through the brainstem towards the trigeminal main sensory nuclei (yellow arrows).}
\centering
\label{fig:MSfigure2}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure3.png}
\caption{Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method. Images are shown in superior axial view. Panel A shows the MS-TN tractography delineations. There is focal decrease in CN V FA in the region of the MS lesion (yellow arrow). Panel B shows the TN tractography delineations, where there is no such FA decrease in the brainstem segment. Please refer to Figure 2 for explanations of color scales and figure annotations.}
\centering
\label{fig:MSfigure3}
\end{figure}

\subsection{TN and MS-TN are characterized by specific diffusivity signatures}
The resulting coefficient matrices demonstrated unique patterns of diffusion metrics across subject groups in each of the CN V segments measured. Each nerve segment is studied separately, and differences in regional diffusivities were examined in both intra-subject comparisons between symptomatic and asymptomatic sides, as well as inter-subject comparisons between subject groups. The details of the examinations are described as follows:

\subsubsection{Cisternal changes are unique in TN}
In the cistern segment (Figure 4), statistically significant differences are found only between the TN group and other groups. No significant differences were found in between segments of other subject groups. Specifically, the FA measure of symptomatic side (+) in TN patients is shown to be significantly greater compared to all of the other patient groups and segments, while the asymptomatic side (-) in TN patients does not show any significant differences compared to other groups. In RD, AD and MD measurements, the only significant differences are observed between TN(+) and TN(-), where TN(+) RD/AD/MD are lower than that of TN(-). No symptomatic/asymptomatic (+/-) or left/right differences were observed in MS-TN and control groups. These results suggest that the cisternal CN V diffusivity changes are unique to the TN group.

\subsubsection{REZ changes are found predominantly in TN}
In the REZ (Figure 5), statistically significant results are found within, but not unique to TN. TN(+) FA is shown to be significantly lower than TN(-) FA, while TN(+) RD and MD are shown to be significantly greater than TN(-); TN(+) MD and RD are shown to be significantly greater compared to all other groups and sides; while TN AD showed no significant differences. In contrast, MS-TN patients showed no significant differences between MS-TN(+) and MS-TN(-) sides in any of the diffusivity metrics, suggesting no intra-group differences. Comparisons between MS-TN(+/-) and TN(+/-) revealed no significant inter-group FA differences. However, both TN and MS-TN groups are shown to be different from controls. Both TN(+/-) and MS-TN(+/-) REZ have significantly lower FA than that of control REZ; control AD and RD are shown to be significantly lower than TN(+/-); while there is no significant difference between left and right sides in controls. 
\subsubsection{Pontine diffusivities showed no significant changes in all groups}
In the pontine tissue (Figure 6), there are no significant intra-group or inter-group differences in FA, RD and MD for all groups and sides. However MS-TN(-) AD is significantly higher than TN(+/-) and controls. No intra-group AD differences were found in MS-TN.
2D. Diffusivities in peri-lesional trigeminal fibres highlight MS-TN pain

In the peri-lesional region (Figure 7), the patterns shows significant intra-group and inter-group differences in MS-TN. MS-TN(+) FA is significantly lower than that of MS-TN(-), as well as significantly lower than TN(+/-)and Controls, while MS-TN(-) FA showed no inter-group differences with TN and Controls. RD, and MD showed only intra-group differences between MS-TN(+) and MS-TN(-),where MS-TN(+) RD and MD are significantly higher than that of MS-TN(-). AD showed no significant differences. The pattern of differences for peri-lesional region is unique to MS-TN.

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure4.png}
\caption{Top row: Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls. Colored squares indicate statistically significant differences (p\textless 0.05) after correcting for multiple comparisons with false discovery rate. The (+/-) sign indicate symptomatic (+) and asymptomatic (-) sides of the subject group. The matrix should be read as ‘Y axis in relation to X axis’—for example, A red square indicates that the vertical axis measure (e.g. TN+, the affected side in TN patients) is significantly higher than that of the horizontal axis measure (e.g. MS-TN+, the affected side in MS-TN patients); blue indicates that the measure is lower. The number within the square indicates relative estimated differences. Bottom row: Box-plots of the diffusivity distributions. Note that the control group does not have pain lateralities, therefore ipsilateral indicate the right side in the control group in this figure. The mean of the left-right diffusivities in controls were used for comparisons against other groups. The analysis highlights TN+ specific diffusivity differences in the Cisternal segment of the trigeminal nerve. Specifically TN+ has significantly higher FA than all the other groups, and that there are significant differences between TN+ and TN-. The points above and below the box plots are outliers. Significances are not indicated in the plot, for pair-wise significance, please refer to the estimate matrices above row.}
\centering
\label{fig:MSfigure4}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure5.png}
\caption{In the REZ, MS-TN, and TN patients have significantly different diffusivities when compared to controls. TN+ also shows significantly different FA, RD, and MD from TN-. Moreover TN+ shows significant different from MS-TN+ as well. Please refer to Figure 4 for explanations of layout and annotations.}
\centering
\label{fig:MSfigure5}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure6.png}
\caption{In the pontine segment unaffected by MS plaques, there are little differences between groups. However MS-TN- shows significantly different AD than TN- and controls. Please refer to Figure 4 for explanations of layout and annotations.}
\centering
\label{fig:MSfigure6}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure7.png}
\caption{The peri-lesional segment highlighting MS-TN+ specific diffusivity differences. Both MS-TN+ and MS-TN- show significantly lowered FA than all the other groups. RD of MS-TN- is also significantly lower than that of MS-TN+. Please refer to Figure 4 for explanations of layout and annotations.
http://msj.sagepu}
\centering
\label{fig:MSfigure7}
\end{figure}

\section{Discussion}

We demonstrate for the first time that diffusion MRI tractography is capable of anatomically distinguishing between TN and MS-TN. Analyses revealed unique, focal MS-TN diffusivity changes along CN V. These changes highlight the differences in diffusivities between MS-TN, TN and healthy controls, thereby permitting their use as neuroimaging signatures for distinguishing these conditions. 
Previous studies of the CN V with DTI tractography have been met with limitations of the single-tensor Gaussian diffusion model, and were incapable of resolving brainstem trigeminal tracts (Figure 2A). Recent improvements in multi-tensor imaging are now capable of more complex diffusion profile of crossing fibres in neural tissue, thereby permitting reliable reconstruction of the brainstem CN V (Figure 2B), and subsequently the study of CN V changes in MS patients (Figure 3).

In patients with MS-TN we expected no changes in the extra-axial CN V segment, and that adverse changes to the CN V would be found in the intra-axial segment, within close proximity to the MS lesions. Conversely in TN patients, we expected CN V changes to be found within or near the extra-axial segment due to neurovascular compression.

The results have agreed with our initial hypothesis. The coefficient matrices (Figures 4-7) revealed specific patterns demonstrating TN-specific diffusivity changes in the cistern and MS-TN-specific changes in the peri-lesional regions. In the TN REZ, FA changes are similar to the results found by DeSouza et al (Desouza, Moayedi, Chen, Davis, \& Hodaie, 2013; DeSouza et al., 2014), where the FA of the TN(+; pain-affected side) CN V is lower than that of TN(-; non-affected side). Additionally, while RD and MD changes in the TN REZ showed significant differences, no AD differences in the REZ were detected, suggesting that the nature of the diffusivity changes may be due to demyelination of the central myelin within the REZ. Interestingly, the cisternal TN(+) FA is demonstrably higher than the TN(-) segment, and also the cisternal FA of all other groups, suggesting that cisternal and REZ FA changes in TN are not homogenous along the nerve segment. Although there has been reports of hyperintensities in the cisternal CN V T$_{1}$ image with gadolinium contrast in MS \cite{VanderMeijs2002}, the lack of MS cistern findings in this study may be due to the limit of the sampled MS population. MS plaque locations are highly variable, and MS-TN patients were not filtered specifically for cistern region plaques.

Diffusivity disruptions of CN V in MS-TN patients are likely due to MS plaques at the regions proximal to the main sensory nucleus. While AD has been associated with neurological disruptions in MS\cite{Budde2009,Kim2006}, the present study showed a lack of change in AD. This may relate to the observation that we did not select subjects based on their disease severity, but based on their pain. Disease severity was in fact variable in this group. Future studies will be able to address this point in more detail, by taking into account other MS manifestations in addition to pain, and correlation with diffusion metrics. Lower REZ FA in MS-TN compared to controls suggests that global neural tissue diffusivity in MS-TN is significantly different from that of healthy controls. Similar findings can be seen in the pontine region, where only the MS-TN AD measures are distinctly different from the other groups. However while AD is commonly correlated with axonal integrity, the high concentration of crossing fibers in the pontine region may introduce confounds into these correlations. Improvements in diffusivity metric measurements in such tissues are needed to further understand these findings.

\subsection{Limitations and future directions}
The study aimed to demonstrate that diffusion MRI can detect localized CN V differences in MS-TN, and to evaluate the feasibility of this novel approach in a group of patients. Therefore we did not explore the course of MS disease progression and other details in the MS-TN patient group in this study. However it has demonstrated that by utilizing MTT to not only visualize, but also to assist in nerve segmentation and statistical analysis, more complete insights into the changes of white matter anatomy can be gained. We have found that peripheral CN V does not change homogeneously in pathology. Other white matter pathways are also likely to be differentially affected in varying segments in other pathological states such as MS.

With this approach, future longitudinal studies may reveal more about white matter changes and associated pain, cognitive, and motor-related measures across the disease’s progression. Moreover, this strategy may allow for pre- and post-treatment predictions in white matter changes, ultimately aiding in clinical prognostication and the identification of high-risk subjects. Therefore, this study paves the way for a more systematic approach to the understanding of pain and white matter anatomy using non-invasive neuroimaging techniques.

\section{Conclusion}
The study demonstrates that MS-TN and TN have differentially localized pathophysiology at the level of CN V in the brainstem. MS plaques disrupt the diffusivity of brainstem CN V fibres near the CN V nucleus, while TN diffusivity disruptions are focused on the cistern and REZ segments. The regions affected by MS and TN REZ suggest changes in myelination of the affected CN V segment. Using this imaging technique, TN and MS-TN can be distinguished by unique localized diffusivity changes in the different CN V segments.

### 2017-11-24 thesis/contents/reviews/dmri.tex
Initial ShareLaTeX Import

\section{Diffusion Magnetic Resonance Imaging}

Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain tissue in-vivo, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.

\subsection{Diffusion Weighted Image (DWI)}

The basis of dMRI is based on the observation that the location probability distribution of water molecules, which undergoes Brownian motion during diffusion, is a function of its suspension medium. The rate of diffusion can be formulated as \cite{Mukherjee2008}: 
\begin{equation}
<r^2>=6Dt
\end{equation}
where $<r^2>$ is the mean square displacement of the molecule after time $t$ with diffusion constant $D$. In an unhindered medium, such as in liquid water, its diffusion probability is can be described as an spherical volume, in other words it undergoes \textit{isotropic} diffusion. In biological tissue such as the neuron axonal bundle, its diffusion probability is then constrained to be more likely in a direction along the bundle, rather than parallel to the cross-section, its diffusion state is then described as \textit{anisotropic} (\ref{fig:diffusion}).

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=3.5in]{diffusion.pdf}
\end{tabular}
\caption{Isotropic diffusion: diffusion of molecules have the same probable displacement in all directions. Anisotropic diffusion displacement of modules are constrained to its diffusion medium.}
\label{fig:diffusion}
\end{center}
\end{figure}

To image this difference in tissue diffusion, a diffusion-weighted imaging (DWI) pulse sequence, which is termed Stejskal-Tanner encoding, is used by adding diffusion-sensitizing gradients to a T2-weighted spin-echo sequence, before and after the 180 degrees refocusing pulse. The gradient is parameterized by the gradient amplitude $G$ in $mT/m$, duration $\delta$ in $ms$, and refocusing delay interval $\Delta$ in $ms$.

The image contrast is determined by the amount molecular motion between the two gradients, where motion along the gradient axis results in a loss of image intensity. The contrast is defined as: 
\begin{equation}
S_i = S_0 \cdot e^{-b \cdot ADC_i}
\end{equation}
Here $S_i$ denote the image contrast at a voxel. $S_0$ the signal intensity without the gradients applied. $ADC_i$ being the apparent diffusion coefficient in the $i$ direction. $b$ is the diffusion weighted factor that is a function is the duration, strength and temporal spacing of the gradients: 
\begin{equation}
b = \gamma^2 G^2 \delta^2(\Delta - \delta/3)
\end{equation}
where $\gamma$ is the gyromagnetic ratio constant. The ADC is measured in $mm^2/s$, and denotes the average molecular diffusion area with a given acquisition time. The higher ADC values indicate higher molecular mobility. Clinical DWI sequences typically use a diffusion time of 10-50 ms, or a travel of about 10$\mu$m. Typical clinical $b$ values range from 600 to 1500, to balance signal contrast with acquisition time.

\subsection{Diffusion Tensor Imaging}

To properly characterize anisotropic diffusion, there must be more than 1 gradient directions available. In fact it is possible to describe voxel-wise diffusion using a Gaussian diffusion model, where the diffusion anisotropy can be visualized as an ellipsoidal tensor \cite{Jones2002}. This type of imaging methodology is termed diffusion tensor imaging (DTI). At least 6 non-collinear gradient directions must be captured in order to compute the DT. In practice up to 32 gradients directions are used clinically.

The diffusion tensor is a $3 \times 3$ symmetric matrix that describes diffusion displacement in the three major axis of diffusion:

\begin{equation}
D=
\begin{bmatrix}
    D_{xx} & D_{xy} & D_{xz} \\
    D_{xy} & D_{yy} & D_{yz} \\
    D_{xz} & D_{yz} & D_{zz}
\end{bmatrix}
\end{equation}
Where $D_{i,j}$ is related to:

\begin{equation}
\frac{S}{S_0} = e^{-\Big(\sum\limits_{i=x,y,z} \sum\limits_{j=x,y,z} b_{i,j} \cdot D_{i,j}\Big)}
\end{equation}

and

\begin{equation}
b_{i,j} = \gamma^2 G_i G_j \Big[ \delta^2 \Big( \Delta - \frac{\delta}{3} \Big) \Big]. 
\end{equation}

The diffusion tensor can then be diagonalized to find its eigen vectors, which are the principal axis of the diffusion ellipsoid, and from which we can obtain the eigen values (\ref{fig:tensor}):

\begin{equation}
\Lambda = 
\begin{bmatrix}
\lambda_1 	& 0 		& 0 		\\
0 			& \lambda_2	& 0 		\\
0 			& 0 		& \lambda_3 \\
\end{bmatrix}
= R \cdot D \cdot R^T.
\end{equation}

The eigen values are measured in the same unit as ADC, which is in $mm^2/s$. Their corresponding eigen vector is denoted as $\hat{\lambda_i}$ They are ordered from the greatest to the least diffusivity magnitude. $\hat{\lambda_1}$ points to the orientation of the greatest diffusion direction, and therefore $\lambda_1$ is also referred to as "longitudinal" or "axial" diffusivity. $\hat{\lambda_2}$ and $\hat{\lambda_3}$ are the lesser orthogonal diffusion directions, and their mean magnitude give rise to the cross-sectional, or "radial" diffusivity.

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.5in]{tensor.pdf}
\end{tabular}
\caption{Illustration of the eigen vectors of an ellipsoid tensor.}
\label{fig:tensor}
\end{center}
\end{figure}

\subsubsection{DTI Metrics}

The Gaussian tensor eigen values form the basis of diffusivity measures that are commonly used to quantify cerebral tissue. The most common diffusivity metrics include Trace, Axial(Parallel/Primary/Longitudinal) diffusivity (AD), Radial (Perpendicular) diffusivity (RD), Mean diffusivity (MD) and fractional anisotropy, these are listed as follows:

\begin{equation}
Trace = \sum_{i=1}^{3} \lambda_i
\end{equation}
\begin{equation}
AD = \lambda_1
\end{equation}
\begin{equation}
RD = \frac{(\lambda_2+\lambda_3)}{2}
\end{equation}
\begin{equation}
MD = \frac{\text{Trace}}{3}
\end{equation}
\begin{equation}
FA = \sqrt{\frac{3}{2}}\frac{\sqrt{(\lambda_1-MD)^2 + (\lambda_2-MD)^2 + (\lambda_3-MD)^2}}{\sqrt{\lambda_1^2+\lambda_2^2+\lambda_3^2}}
\end{equation}

FA has been found to correlate with a wide spectrum of neuromorphological changes, including but not limited to multiple sclerosis \cite{Cercignani2001,Ciccarelli2003d,Miron2012}, vestibular schwannomas \cite{Chen2011b,Taoka2006,Wai2009}, tumors \cite{Mori2002,Schonberg2006,Nimsky2005}, schizophrenia \cite{Rametti2009,Fitzsimmons2009}, traumatic brain injury \cite{Bendlin2008,Kinnunen2011c,Hellyer2012}.

RD is found to positively correlate with degree of demyelination in multiple sclerosis \cite{Song2005,Song2002,Klawiter2011,Janve2013}, and well as stages of white-matter development and aging \cite{Counsell2006,Sala2010}, including alzheimer's disease \cite{Acosta-Cabronero2012}.

AD is correlated positively with axonal integrity \cite{Song2003}, and is found to be a correlate of axonal injury \cite{Budde2009} such as those found in traumatic brain injury \cite{Kinnunen2011c} and Gulf War illness \cite{Rayhan2013},

MD is correlated with tissue inflammation and edema \cite{Senda2012g}, and is found to correlate with diffusivity changes as a result of gliomas \cite{Stadlbauer2007}, multiple sclerosis \cite{Cercignani2001,Senda2012g,DeGroot2013}, amyotrophic lateral sclerosis \cite{Sharma2012}, and aging \cite{Kantarci2011}.

\subsubsection{Limitations}

DTI is based on the Gaussian model of water diffusion, and therefore imposes an important limitation. The Gaussian tensor can derive only one diffusion direction at a given voxel, and therefore is incapable of modelling cross-fiber diffusion of two or more directions. By extension, its derived metrics (FA, AD, RD) can only characterize a single diffusion direction as well. Therefore the interpretation of these metrics in white matter that contains cross-fiber configurations is tenuous.

\subsection{High Angular Resolution Diffusion Imaging}

In order to characterize the additional diffusion components in voxels with multiple fibres, changes in DWI acquisition strategies is necessary. A natural extension is to acquire additional diffusion directions. This type of acquisition strategies, which enable higher order diffusion modelling algorithms, are generally referred to as high angular resolution diffusion imaging (HARDI). The most wide spread strategy is to equally distribute gradient directions as vertices of a triangle-tessellated geodesic sphere \cite{Frank2001}. Alternatively, it's also possible to equally spread gradient directions in a rectilinear grid, which forms the basis of diffusion spectral imaging (DSI) \cite{Wedeen2005}. Additional acquisition strategies such as distributing gradient directions across multiple concentric spherical shells is also actively explored \cite{Jeurissen2014a}.

Modifications to B-value needs to be considered for HARDI. The measured diffusion contrasts are dependent on the diffusion compartment variances. Therefore higher B-values will increase the measured signals variance contrasts, and allow better separation between crossing diffusion directions. Different HARDI algorithms have different requirements on B-values. For example Q-ball imaging needs B=3000 $s/mm^2$ \cite{Descoteaux2007a}, and multi-shell acquisitions will require stepped values of B values ranging from 500  $s/mm^2 $  to greater than 3500  $s/mm^2 $ \cite{Tournier2013}.

The determination of gradient sampling pattern and B-value directly affect acquisition time and image quality. More complex gradient sampling and higher B-values can improve tractography quality, however the acquisition time can become prohibitively long in a clinical setting. Other factors that must be considered include the nature of patient pathology, motion artifacts, and the inclusion of other imaging modalities during clinical scan sessions.

\subsection{Spherical Deconvolution}

An end-to-end spherical deconvolution diffusion model was first proposed by Tournier et al \cite{Tournier2004}. 
Consider a fiber population with distinct fiber components. The measured diffusion signal $S(\theta,\phi)$ can be written as:

\begin{equation}
S(\theta, \phi) = \sum_{i}^{} f_i \hat{A}_i R(\theta)
\end{equation}
where $\theta$ is the elevation angle,  $\phi$ the azumuth angle in spherical coordinates; $f_i$ is the volume fraction of the $i$-th fiber component, $\hat{A}$ is the rotation operator that orients fiber, and $R(\theta)$ is a axially symmetric response function.

$S(\theta,\phi) $ can be rewritten as a convolution over the unit sphere of $R(\theta)$ with a fiber orientation density function (FOD) $F(\theta, \phi)$, such that:
\begin{equation}
S(\theta, \phi) = F(\theta, \phi) \otimes R(\theta)
\end{equation}

With this approach, the FOD can be estimated using the reverse spherical deconvolution, using spherical and rotational harmonics \cite{Healy1998}. Spherical harmonics are orthonormal basis functions over a sphere, and therefore can approximate any symmetric diffusion geometry in the spherical q-space. The rotational harmonics similarly form the orthonormal basis over all possible rotations, and can represent any symmetric fiber configurations. The harmonics have two important parameters: the harmonic order $l$ and phase factor $m$, with $ -l \leq m \leq l $ for each harmonic order level $l$. The odd $l$ harmonics represent antisymmetric functions, while the even $l$ harmonics represent symmetric functions (\ref{fig:harmonics}).

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=100mm]{harmonics.pdf}
\end{tabular}
\caption{Visualization of the spherical harmonic functions up to l=3.}
\label{fig:harmonics}
\end{center}
\end{figure}

Spherical decovolution is considered to be the action of a set of rotations on a function over a sphere. Given $F^l $ as a vector of $l$th order spherical harmonic decomposition of  $F(\theta,\phi)$, with vector length $(2l + 1)$, and $R^l$ as the $l$th order rotational harmonic decomposite of $R(\theta)$, with vector length $(2l+1)^2$. The spherical harmonic representation of $S(\theta,\phi)$ is then:
\begin{equation}
S^l = R^l F^l
\end{equation}
To reduce the number of parameters to be computed, we can make the further assumption that the diffusion process is symmetric, and therefore the odd $l$ harmonic orders can be ignored, with only the even harmonics considered. 
The response function $R(\theta)$ can be directly estimated from the dataset by supplying DWI region with high anisotropy from a single fiber bundle, such as the body of corpus callosum.

The maximum haronic order $l_max$ that can be estimated is dependent on obtained number of directions. For example, $l_max$ = 8 requires 45 directions, $l_max$ = 10 requires 66 directions, and so on. Larger $l_max$ are more sensitive to noise, which is a weakness for the first spherical deconvolution algorithm \cite{Tournier2004}.

\subsubsection{Constrained Spherical Deconvolution}
Further advances to the estimation method of the spherical deconvolution model made it more robust to noise \cite{Tournier2007b}. The noise of SD estimation are represented as spurious negative lobes in FOD, which are physically impossible, and do not represent any real phenomenon. To mitigate this, a non-negativity constraint is added. This method is referred to as \textit{constrained spherical deconvolution} (CSD).

The CSD constraint is applied iteratively using Tikhonov regularization \cite{Hansen1998}. The procedure involves obtaining an initial FOD estimate, and a set of directional peaks are then identified. The Tikhonov regularization then constraints the negative peaks to zero, and a new set of FOD and peaks are obtained. This is applied iteratively until convergence, and the final constrained FOD is then supplied.

The Tikhonov regularization minimizes the weighted sum two terms:
\begin{equation}
\|{Af - b}^2\|^2 + \lambda^2 \| l(f - f')\|^2
\end{equation}
Where the first term is a lear least-square fit of solution $f$ to to data $b$ with problem matrix $A$. 
In terms of CSD, $f$ corresponds to the SH coefficients of the FOD, $b$ is the measured signal and $A$ is:
\begin{equation}
A = QR
\end{equation}
Where $Q$ is the SH coefficient to DW intensity mapping, and $R$ is the spherical deconvolution. 
The regularization term is the constrained matrix $L$ multiplying the norm of $f$ relative to an initial estimate $f'$. The parameter $\lambda$ controls the regularization amount, with $\lambda=0$ having no regularization, to $\lambda\rightarrow \infty $ making the solution tend to $f'$.

In CSD, the $f'=0$ and $L$ maps $f$ to orientations that have zero amplitude. During initiation, $f_0$ is estimated using spherical deconvolution with a low-pass filter \cite{Tournier2004}, and truncated to $l_max = 4$. The FOD amplitudes $u$ are then computed along N uniformly distributed directions (By default, MRTrix uses $N=300$) where
\begin{equation}
u = P f_i
\end{equation}
L is then formed as:
\begin{equation}
L_{m,n} = 
	\begin{cases}
		P_{m,n}, & u_m < \tau \\
		0 &	u_m \geq \tau 
	\end{cases}
\end{equation}
where $\tau$ is the amplitude threshold. In practice $\tau$ is set to 10\% of mean FOD amplitude. 
The improved iterative $f_i$ is then solved by incorporating the new FOD amplitude information by:
\begin{equation}
f_{i+1} = \text{argmin}\{ \|Af_i - b\|^2 + \lambda^2 \|L f_i\|^2  \}
\end{equation}
The FOD estimate becomes stable when matrix L converges, which take about 10 iterations.

CSD is robust to noise at relatively low B-values, and is capable of good quality FOD estimates at $B=1000 s/mm^2$. This means that existing DTI protocols can be extended for high order HARDI tractography by simply increasing the number of directions. With CSD, a 60 direction DWI acquisition can be used to confidently derive FOD with $l_{max}=8 $, and up to $l_{max}=12$ with super resolution enabled \cite{Tournier2007b} (\ref{fig:csd}).

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{csd-angles}
\end{tabular}
\caption{ Comparison of angle discrimination ability and CSD FOD. B-value=3000 \protect $s/mm^2$. Taken from \protect \cite{Tournier2007b} } 
\label{fig:csd}
\end{center}
\end{figure}

\subsection{Tractography}
Tractography is the construction, or propagation of streamline from an underlying field of orientation information, in order to visualize continuous structures within a volume.  The orientation field is generally provided as a 3D matrix of directional information obtained from DTI, CSD, or other methods. For example, in DTI the directions is the eigen vector with the greatest magnitude $\hat{\lambda_1}$; for CSD, the directional vectors would be peaks from the FOD.

Applying pathway tracking to DWI imaging was first explored by Mori \cite{Mori2002b} and Basser\cite{Basser2000}. The family of methods that propagates by deterministic algorithms based on local orientation data is collectively called deterministic tractography. The first tracking algorithms were applied to DTI using step-wise polyline propagators that followed local DT tensors \cite{Conturo1999}. Further refinement to the line propagator to include neighbouring DT tensors, and interpolation of the local DT directions in order to obtain a smooth curve \cite{Mori2002b}. Basser introduced Runge-Kutta integration to allow more robust path tracing when faced with noise in the local directions \cite{Basser2000}. Another similar path-integral method was also proposed by Tuch \cite{Tuch2000d}. The integrated tractography approaches are now standard practice for deterministic tractography \cite{Tournier2004,Pieper2004}.

Deterministic tractography lends itself naturally to anatomical interpretation and further geometric analysis. It was traditionally paired with DTI, and so was thought to be limited in the case of crossing fibers. Other methods attempted to circumvent the limitation with probabilistic approaches that include white matter tissue priors in the fiber propagation model. The most popular of the probabilistic approach is the FSL software's probabilistic tractography module (FDT) \cite{Behrens2007}, which also includes a model-based ball-and-stick diffusion model that can resolve 2-fiber crossings. The FDT software produces visitation images of its probabilistic streams that collates the probability of tract propagation at each voxel.  The visitation image however is hard to interpret and localize, and therefore does not rival deterministic methods in surgical visualization.

The availability of higher order HARDI methods, especially in CSD allows the reinterpretation of FOD as local fiber orientation probability distribution \cite{Tournier2010,Jeurissen2011b}. Development of a probabilistic streamline propagator based on CSD can now combine the advantage of probabilistic models with the geometric interpretability of deterministic tractography.

\subsubsection{Applications}
Tractography permits the mapping of white matter structures with computer graphic and statistic methods. It permits the study of brain structural connectivity at  individual level \cite{Boorman2007} and group level \cite{Meskaldji2013},  circuitry mapping \cite{Kirsch2015} and macro-structural insights between the brain hemispheres \cite{Kucyi2012a}. It is often paired with functional connectivity for studies in multiple sclerosis \cite{Zhong2016,Rocca2007a,Kucyi2015}, chronic pain \cite{Mansour2013,Seifert2011,Wiech2014}, and other areas of study such as empathy \cite{Bernhardt2014}, and aphasia \cite{Catani2013c}.

Tractography is also capable of fine-grained neuroanatomy visualizations \cite{Hodaie2010, Chen2015c}, and therefore is important in the advancement of surgical visualization \cite{Chen2011b,Rosen2015,Hodaie2012g,Taoka2006,Sherbondy2008,Muthusamy2007} and planning \cite{Golby2011}.

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{tractography1}
\end{tabular}
\caption{Example of cranial nerve reconstruction with deterministic tractography.} 
\label{fig:tract1}
\end{center}
\end{figure}

### 2017-11-24 thesis/contents/reviews/machine-learning.tex
Initial ShareLaTeX Import

\section{Machine Learning}
Machine learning (ML) is a branch of computer science that aims to learn and form predictions from data. In contrast with static algorithmic programming, where the structure of the algorithms are predefined, machine learning approaches construct models based on observed data, in order to extend its predictions to the unobserved. Strategies and algorithms for constructing various models overlap with the fields of artificial intelligence, computational statistics, pattern recognition, and neuroscience. Advanced in modern machine learning, especially in deep neural networks \cite{LeCun2015}, in recent years have lead to wide spread adaptation of machine learning based online services in translation \cite{bahdanau2014neural}, image search \cite{simonyan2014very}, and speech recognition \cite{graves2013speech}. There is also surging interest in deploy machine learning systems in radiology and disease diagnosis in medicine.

Machine learning training strategies can be categorized into Supervised Learning, Unsupervised Learning, and Reinforcement Learning. Supervised learning involves training data that are pre-labeled. For example the data supplied can be a list of patients' age and gender, and the label can be whether they or not they are in pain. Unsupervised learning involve data without any prior labels, and the algorithm attempts to discover patterns within the data itself, and is often used in data mining. Reinforcement learning involves continuously learning with a reward function. It is generally used to train agents that have a continuous stream of data and needs to interact with an external environment, and is widely used in robotics.

\subsection{Supervised Learning}
Supervised learning is the mostly widely used method of training strategy. Most of the existing machine learning system in production are based primarily on supervised training with human-supplied data labels. Different algorithms can be used to fit a function that forms a decision boundary based on data features and labels. Possible models ranges from Decision Trees, linear models such as Support Vector Machines (SVM), Deep Neural Networks (DNN), and bayesian-based models such as Naive Bayes and Gaussian Process (Figure \ref{fig:ml-bounds}).

\begin{figure}[h]
\includegraphics[width=\textwidth]{ml-decision-bounds.png}
\caption{Different of decision bounds as a result of different machine learning algorithms. Source: \protect\url{http://scikit-learn.org/stable/auto_examples/classification/plot_classifier_comparison.html} }\label{fig:ml-bounds}
 \end{figure}

\subsubsection{Gaussian Process}
The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} has sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}. However one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models. Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process \cite{gal2016dropout}. In fact it has been found that DNN with infinite number of hidden neurons approximates GP \cite{neal1996priors}.

Gaussian Process is defined \cite{rasmussen2006gaussian}) as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
\begin{equation}
	\begin{split}
		m(x) &= \mathbb{E}[f(x)], \\
		k(x,x') &= \mathbb{E}[(f(x)-m(x)(f(x')-m(x'))],
	\end{split}
\end{equation}
And the Gaussian Process is defined as
\begin{equation}
f(x) \sim GP(m(x), k(x, x')) 
\end{equation}

The covariance function is also termed called kernel function. The radial-basis function kernel or RBF kernel is widely used in machine learning tasks. The RBF kernel is also termed the square exponential, and is defined as

\begin{equation}
k(x, x') = exp\bigg( -\frac{1}{2}|x - x'|^2\bigg)
\end{equation}

Another commonly used co-variance function is the Mat\'{e}rn kernel, which is used as an approximation for the RBF kernel in machine learning. It is defined, given a distance $r$ between two points, as:
\begin{equation}
k_{matern}(r) = \frac{2^{1-\nu}}{\Gamma(\nu)}\bigg(\frac{\sqrt{2\nu}r}{l}\bigg)^{\nu}K_{\nu}\bigg(\frac{\sqrt{2\nu}r}{l}\bigg)
\end{equation}
Where $K_\nu$ is a modified Bessel function \cite{rasmussen2006gaussian}. The Mat\'{e}rn kernel is equivalent to the RBF kernel when $\nu\rightarrow\infty$. In machine learning applications, $\nu$ commonly takes the value of $1.5$ or $2.5$ such that:
\begin{equation}
\begin{split}
k_{\nu=1.5}(d) &= \bigg(1+\frac{\sqrt{3}r}{l}\bigg)exp\bigg(-\frac{\sqrt{3}r}{l}\bigg), \\
k_{\nu=2.5}(d) &= \bigg(1+\frac{\sqrt{5}r}{l}+\frac{5r^2}{3l^2}\bigg)exp\bigg(-\frac{\sqrt{5}r}{l}\bigg), \\
\end{split}
\end{equation}
Where the function becomes once and twice differentiable respectively, and thereby having different degree of smoothness of approximation.

\subsubsection{Training}
The process of fitting a ML model on labeled data to form predictions is called training. The objective of the training process is to obtain a model that can both correctly predict observed labels, and is able to generalize its predictions to new data that are not part of the training set. Depending on the model algorithm, there can be many training parameter that govern the model outcome. The tuning of these parameters is referred to as hyper-parameter tuning. It is possible for certain hyper-parameters that result in model over-fitting, which is when a model is over-specialized in achieving the most accurate training label prediction, but perform poorly on unseen data. To properly evaluate model performance, a dataset is split into training/testing and validation sets, usually in 4:1 or 9:1 ratios. Where model training is performed solely on the training set, and evaluation of model performance is on the validation set to assess the ability for the model to generalize.

There are cases when the number of available data is low, and each data point may have a substantial impact on model performance. In these cases cross-validation is used. The idea is to divide the data into all possible combinations of training and validation splits, usually in the form of 5 fold to 10 fold. Then separate models are trained and evaluated on each combination, and their performance is averaged to form the final evaluation. This gives a reasonable approximate on the performance of the model when trained using all available data. Leave-one-out cross-validation is an extreme form of this: given a dataset of N items, each split will consist of 1 and N-1 items, which result in N possible combinations. It is used when the available data is extremely low.

\subsection{Machine Learning, Neuroimaging, and Pain}

There has been a long history of applying statistical methods to the field of neuroimaging, from the very onset of the field. Neuroimaging studies often attempt to correlate physiological measures from neuro-activities with specific experimental variables or conditions. Neuro-activities may be measured from EGG sensors, or as voxels in functional or structural MR images. Traditionally data from each measured component was assumed to be independent. Correlative analysis were performed using general linear models, with post-hoc false-positive corrections. Therefore univariate models cannot take into account spatial or temporally correlative dynamics.

Increasing complexity in neuroimaging data have lead to widespread applications of multi-variable machine learning methods in neuroimaging. Application of ML in pain neuroimaging became controversial when Tor Wager \cite{Wager2013} demonstrated a Lasso-PCA model that classifies nociceptive pain with sensitivity and specificity of 90\% solely from fMRI imaging data. Additional studies showed consistently good performance of ML-based nociceptive pain discrimination. Discrimination of chronic pain from nociceptive pain proved to be more challenging, due to the high degree of similarity in pain-matrix activation patterns. The improvement in chronic pain detection lead to debates of the validity of presenting such methods as evidences in courtroom, where Davis et al argued against such actions \cite{Davis2012a}. While it is commonly accepted that pain activity in the brain involves multiple regions of the pain-matrix, evidences in some studies suggested that the posterior-insula is heavily involved in chronic pain. Segerdale et al claimed that the posterior insula is a primary pain region \cite{Segerdahl2015a}, thereby ignited renewed debates surrounding regional specific brain centers for chronic pain \cite{Davis2015}.

Despite the controversies, interest in machine learning in medical imaging prognosis have sparked both public and private efforts across the globe. Classification studies with using multi-modal data with features in functional connectivity and structural connectivity in pain syndromes such as trigeminal neuralgia continue to improve. It is clear that machine learning is here to stay in pain neuroimaging research.

### 2017-11-24 thesis/contents/reviews/mri.tex
Initial ShareLaTeX Import

%\graphicspath{{D:/project/thesis_preparation/PhDtemplateLATEX_V2/ch_Background_Work/figures/}}
\section{Magnetic Resonance Imaging}
Magnetic Resonance Imaging (MRI) is a medical imaging technology that allows non-invasive visualization of the body using magnetic fields. It is a key enabler of modern healthcare and medical research, and is used in the detection and diagnosis of neuropathology.

MRI is based on the principles of nuclear magnetic resonance (NMR), where atomic nuclei, when placed in a magnetic field, can absorb and emit electromagnetic radiation. A MRI machine applies a primary magnetic field (B0) to magnetize and align the hydrogen nuclei in the body. A radio frequency (RF) field is subsequently introduced perpendicular to B0, and reorients the spin axis. This causes the nuclei to produce NMR signals detectable by the MRI scanner. Based on the precession frequency decay, the signals are recorded to construct an image of the scanned area based on the contrasts in water and fat concentrations.

\subsection{Basic Principles}
\subsubsection{Spin and Magnetic Moment}
Spin is an intrinsic property of elementary particles, and can be either positive or negative. The spin quantum number  $s$ is defined as $ s = \frac{n}{2}$, where $n \in \{0,1,2,...\}$. Therefore $s$ is a quantized value in multiples of $\frac{1}{2}$. Protons, electrons and neutrons possess spin. Nuclei with odd number of protons or neutrons have non-zero nuclear spins; while those with even number of protons and neutrons have no spin. Particles with opposite spins can pair up to produce an overall spin of zero. Hydrogen nuclei, which makes up 63$\%$ the human body, possess spins of $\frac{1}{2}$. The spin of a particle has a magnetic moment vector that behaves like a magnet with north and south poles (Figure \ref{fig:spin}).

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=3.5in]{mri/spin.pdf}
\end{tabular}
\caption{Spin of a nucleus has a magnetic moment vector $\mu$, which behaves similar to a magnet with north (N) and south (S) poles.}
\label{fig:spin}
\end{center}
\end{figure}

The spinning of nuclear particles produces angular momentum  $\textbf{J}$, that is a vector with both magnitude and direction. The largest measurable component of $J$, expressed as $I$, represents the overall quantized spin value as the spin quantum number. The corresponding quantum number is known as the magnetic quantum number $m$, and can take a total of $2I + 1$ angular momentum states.  A non-zero spin is associated with a non-zero magnetic moment ($\vec{\mu}$) via the relation $\vec{\mu} =\gamma \textbf{J}$. The z-component of the angular momentum vector (\textbf{J}) is $J_z = m \hbar$. The z-component of the magnetic moment $\mu_z$ is thus:

$$\mu_z=\gamma J_z=\gamma m \hbar$$

Where $\gamma$ is the gyromagnetic ratio, $\hbar$ is the reduced Planck constant ($=1.0545726 \times 10^{-34} m^2kg/s$ ).

\subsubsection{Effect of Magnetic Field}

In the absence of an externally applied magnetic field, the magnetic dipoles are oriented randomly. When the proton is placed in an external magnetic field $B_0$, the spin vectors become oriented with the field, either parallel or anti-parallel with respect to $B_0$ (Figure \ref{fig:externalB0}). The nuclei precess about the magnetic field direction at some frequency called Larmor frequency:

$$\omega = \gamma B$$.

Where $\gamma$ is the gyromagnetic ratio, and $B$ the magnetic field magnitude.

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=4.5in]{mri/field.pdf}
\end{tabular}
\caption{Left panel-- magnetic moments with random orientations in the absence of external magnetic field; Right panel-- magnetic moments parallel or anti-parallel aligned with the external
field} \label{fig:externalB0}
\end{center}
\end{figure}

Protons aligned in the parallel alignment are more preferred as it has a lower energy state than in anti-parallel orientation. The energy E of a magnetic moment $\vec{\mu}$ when in a magnetic field $B_0$ is given by: $E=-\vec{\mu}\cdot \textbf{B}_0=-\mu_z B_0=-\gamma m \hbar B_0$. The lower energy state of protons corresponding to m = +1/2 in a static magnetic field $B_0$ is $E(\frac{1}{2})=-\frac{1}{2}\gamma \hbar B_0$. The higher energy state of protons corresponding to m =-1/2 is given by $E(-\frac{1}{2})=\frac{1}{2}\gamma \hbar B_0$.
Consequently, the energy differences between two quantum states can be calculated as $\bigtriangleup E = \gamma \hbar B_0$ (Figure \ref{fig:energy}).

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=3in]{mri/energy}
\end{tabular}
\caption{For protons with spin quantum number of 1/2, two energy
levels and the energy difference are shown.} \label{fig:energy}
\end{center}
\end{figure}

As protons aligned in the parallel orientation are more stable, the number of spins under this state is more than that under anti-parallel orientation. This number of difference is called spin excess and the magnetic moments of the excess spins add to form the net magnetization vector $M_z$. When the spin system is at equilibrium, net magnetization lies along the z axis, parallel to $B_0$.

\subsubsection{Resonance Condition}

The steady-state of the stable spin state under $B_0 $ can be disturbed by introducing additional energy $E$ into the system.

The proton would undergo a transition from the lower energy to higher energy state by the absorption of a photon. The energy difference between the nuclear spin levels is $\bigtriangleup E = \gamma \hbar B_0$. Hence, a magnetic resonance absorption will only occur when $\nu_0 = \gamma B_0/(2 \pi)$, which is the same as Larmor frequency. This is called the \textit{resonance condition}.

\subsubsection{Relaxation}
A transverse magnetization is introduced into the system that oscillates at radio frequency (RF) in order to modify the net magnetization, this is called the RF pulse. When exposing the nuclear spin system to energy of the RF pulse, the equilibrium of the spin system is disturbed. The RF pulse is designed to introduce energy E such that the net spin direction orients at $90^\circ$. The net magnetization now is denoted as $M_{xy}$ as it aligns with the xy-plane instead of the z-axis. Individual spins now precess in phase with the RF pulse, it gives rises to the MR signal in the receiver coil.

With the termination of the RF pulse, the nuclei in the excited state will return to the equilibrium condition, this process is called \textit{relaxation}. 
The energy transmitted decays rapidly by two relaxation processes:  spin-lattice and spin-spin relaxations. These are responsible for \textit{T1 relaxation} and \textit{T2 relaxation} respectively.

\paragraph{T1 Relaxation} 
As the transverse magnetization decays, the magnetic moment gradually realigns with the z-axis of $B_0$. 
The magnitude of the magnetic moment's xy-component also gradually diminishes, and the MR signal fades. 
The $M_z$ magnetic moment recovers in proportion, and is eventually restored. 
This processed is called \textit{longitudinal relaxation} or T1 recovery, where T1 refers to the mean time for an individual nucleus to
dissipate its excess energy (heat) to the surrounding environment ( the \textit{lattice}, and therefore the processed is also named \textit{spin-lattice relaxation})and return to its equilibrium state.

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/rf-pulse.pdf}
\end{tabular}
\caption{Illustration of the process of transverse relaxation. Initially, the spin precesses about the B0 field under steady-state. When a 90deg RF pulse is applied, the spin orients towards the xy-plane. When the RF pulse is switched off, the spin now relaxes towards the z-axis.  } \label{fig:energy}
\end{center}
\end{figure}

To describe how $M_z$ returns from 0 to its equilibrium during relaxation, the equation as a
function with respect to time $t$ is written as: 
$$M_z = M_0 (1 - e^{-t/T1})$$
where T1 is the time taken for approximately 63$\%$ of the longitudinal magnetization to be restored (Figure \ref{fig:T1T2}(A)). 
$M_0$ is the magnetic moment at equilibrium.

\paragraph{T2 Relaxation}

T2 relaxation is so named after the time constant of phase decoherance of the transverse magnetic moment.

Immediate after excitation, the phase precessions are synchronous, and therefore are in \textit{phase coherence}.  
However due to various reasons, the various spins begin to lose coherence, and their total sum magnetization $M_xy$ diminishes.

Phase coherence is lost in two ways:
\begin{enumerate}
\item Spins are associated with local magnetic fields, and interact randomly The energy transfer that occur in the local magnetic field interactions cause a mumulative loss of phase. It's a pure spin-spin interaction, and thereby not affected by $180^\circ$ refocusing pulse. It is also independent from the strength of $B_0$. 
The time constant which describes the return to equilibrium of the transverse magnetization, is called the \textit{spin-spin
relaxation time} or T2, and is defined in respect to time $t$ as:
$$M_{xy} =M_{xymax} e^{-t/T2}$$
where T2 is the time it takes for the transverse magnetization to decay to 37$\%$ of its original magnitude. 
T2 is always less than or equal to T1.

\item There are intrinsic inhomogeneities by the magnetic field generator, and the scanning subject. These intrinsic properties contribute to the dephasing, and has a shorter time decay than T2. This type of decay time constant is named T2*  (Figure \ref{fig:T1T2}(C)). The loss of the MR signal due to T2* is called\textit{ free induction decay} (FID). T2* is defined as follows: 
$$ \frac{1}{T_2^*} = \frac{1}{T_2} + \frac{1}{T_2inhomo}$$ 
\end{enumerate}

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=5.5in]{mri/T1T2}
\end{tabular}
\caption{Diagram depicting the relaxation processes of T1 (panel A), T2 (panel B) and $T_2^*$ (panel C). $T_2^*$ is shorter than T2.} \label{fig:T1T2}
\end{center}
\end{figure}

\subsubsection{MR Contrast}

MRI signal intensity depends primarily on three parameters: 
\begin{itemize}
\item Proton density: the number of excitable spins per unit volume, which determines the maximum amount of signal that can be measured from the tissue. It can be measured by minimizing T1 and T2 times. 
\item T1 time: the amount time for $M_xy$ to recover to $M_z$, to be ready for re-excitation.
\item T2 time: the amount of time for $M_xy$ signal to fade under decoherence. 
\end{itemize}
The contrast in brain tissues of the resulting MR image can be tuned by adjusting these three parameters in pulse sequence design.

\paragraph{Repetition Time}
T1 contrast is adjusted by tuning repetition time (TR), which is the length of relaxation interval between excitation pulses. When TR is long (> 1500 ms), $M_z$ recovers totally, and therefore subsequent excitation will again retain maximal signal strength; when the TR is short (< 600 ms), the $M_z$ has not totally recovered, and the next excitation pulse will not be able to gain strong MR signal. Different tissues have different intrinsic T1 differences, and therefore will show strong contrast under short TR. Therefore short TR are said to be T1 weighted.

\paragraph{Echo Time}
MR sequences have a gradient field which is used to control field  inhomogenieties and to encode spatial origin. The gradients also contribute to spin dephasing, therefore a refocusing pulse is necessary to adequately measure MR signals. The time delay between refocusing, and the measured signal under phase coherence (spin echo) is known as echo time (TE). When TE is short ($<$ 30 ms), there is little signal differences between tissue types; when TE is long ( $\geq$ 60 ms ), the intrinsic tissue differences will allow good contrast in the resulting image. Therefore long TE is said to be T2 weighted.

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/spin-echo.pdf}
\end{tabular}
\caption{Illustration of the process of spin-echo. Initially, the spin precesses about the B0 field under steady-state. When a 90deg RF pulse is applied, the spins orients towards the xy-plane. When the RF pulse is switched off, the spins now relaxes towards the z-axis, and the spin undergoes phase decoherance. A 180 degrees flip angle pulse is applied to prevent spin saturation, and then a refocusing pulse is applied to bring spins back into coherance.  } \label{fig:spin-echo}
\end{center}
\end{figure}

Typically, T1-weighted spin echo (SE)  sequence is acquired with TR/TE at 340/13 ms; T2-weighted fast spin echo (FSE) at TR/TE of 3500/120 ms. 
In the case of T1-weighted MRI, images are created with short TE and short TR times. They can differentiate fat from water - with fat brighter and water darker (\url{www.mr-tip.com}) by using a gradient echo (GRE) sequence. In the brain T1-weighted scans provide good gray matter/white matter contrast. In addition, the image contrast can be increased with the use of an inversion pulse as in an MP-RAGE sequence. Due to short TR, this scan can be run very quickly, allowing the collection of high resolution 3D structural image (Figure \ref{fig:T1T2PD}, Left). These structural images can provide detailed information of the brain's shape and size, and provide the base for cortical surface reconstruction and segmentation.

The T2-weighted sequence uses a long TR and long TE. The T2-weighted sequence can be employed as a dual echo sequence. Like the T1-weighted scan, fat is
differentiated from water, but with fat dark and water bright (Figure \ref{fig:T1T2PD}, Middle).

The first or shorter echo ($T_E<30msec$) is proton density (PD) weighted or a mixture of T1 and T2. This image can provide good contrast of tissue and CSF, thus very helpful for evaluating periventricular pathology (Figure \ref{fig:T1T2PD}, Right). $T_2^*$-weighted scans use a gradient echo (GRE) sequence, with long TE and long TR. It is prone to susceptibility losses at air/tissue boundaries, but can increase contrast for certain types of tissue, such as venous blood. It is used for functional MRI detection allowing people to visualize and map the parts of the brain used to perform tasks.

\subsubsection{3D encoding}

\paragraph{Slice Selection} Recall that the resonance condition can only be satisfied when the RF pulse is equal to the Larmor frequency of the nucleus precession under the magnetic field. 
Therefore, to allow slice discrimination, the magnetic field along the z-axis is made to be inhomogeneous by introducing a linear gradient field. It's an extra coil that adds or subtracts from the main magnetic field. With the linear gradient, only a particular portion of the z-axis would share a similar unique Larmor frequency, and can be selectively excited with an RF pulse. The steepness of the gradient determines the slice thickness, the steeper the gradient, the thinner the slices.

\paragraph{Spatial Encoding}
Once the spins are excited and precess in the xy-plane, two additional gradients are applied to allow discrimination of signal measures per slice.

\begin{itemize}
\item Phase encoding: The phase encoding gradient is applied in the y-direction. It alters the Larmor frequency of the spins based the location of the y-axis, and results in phase shifts of the spins relative to each other. The phase shift is determined by the duration and amplitude of the gradient, when the gradient is switched off, all spins return to their initial rate of precession, but maintains the relative phase shift. 
\item Frequency encoding: The frequency encoding gradient is applied on the x-direction. This applies a magnetic field that increase in strength from right to left. The change in Larmor frequency alters the precession speed of the spins such that the spins towards the right precess faster than that of the left. The MR signal collected represents the entire frequency spectrum of the spins. 
\end{itemize}

The measured MR signal with both encoding methods allows recovery of spatial information from the scanned area. Application of Fourier transform will decompose the signal into component frequencies along the frequency encoding direction (x). The phase distribution along each frequencies provide information on the phase encoding direction (y). Decomposing the phase distribution requires another Fourier transform on the multiple scans of different phase gradient strengths. This is called two-dimensional Fourier transform (2D-FT), the number of phase gradient measures is called the phase encoding step.

\paragraph{K-Space}

The collected MR data are encoded in a mathematical space termed the k-space. The k-space is a two dimensional feature space, where the x-axis $k_x$ represent the measured frequency domain, and $k_y$ the phase domain. Each step in $k_y$ represents a single phase encoding step. The x-axis origin $k_x=0$ is the gradient iso-center, and the frequency distribution is symmetric about the x-origin. The 2D-FT operation is applied to the k-space to transform the frequency domain information into image.

\begin{figure}[htb]
\begin{center}
%\begin{tabular}{c}
\includegraphics[width=5in]{mri/T1T2PD}
%\end{tabular}
\caption{Brain scan using T1 (left), T2 (middle) and proton density (right) measurements.}
\label{fig:T1T2PD}
\end{center}
%\vspace{-0.15in}

\end{figure}

### 2017-11-24 thesis/contents/reviews/multiplesclerosis.tex
Initial ShareLaTeX Import

\section{Multiple Sclerosis}

Multiple sclerosis (MS) is a chronic auto-immune demyelination disease of the central nervous system. It is a progressive disease characterized by wide-spread lesion, or plaques, within the brain, spinal cord, and optic nerve. Its pathogenesis is related to the focal inflammation and damage of the central myelin, where the disruptions of the myelin sheath negatively affect signal transduction of the central white-matter, and thereby result in severe disability. MS symptoms can include cognitive deficits, fatigue, depression, upper and lower motor disabilities, and neuropathic pain. Currently there is no cure for MS. 
An estimated of up to 2,300,000 people world wide are affected by MS, and 50\% of the patient will require motor assistance 15 years after disease onset \cite{Goldenberg2012}. The course of MS can be highly variable due to the distinct plaque distribution pattern in each individual, and therefor MS is difficult to diagnose and treat.

\subsection{Classification}

MS is divided into four major groups: 1) Relapsing-remitting (RRMS), 2) Secondary progressive (SPMS), 3) Primary-progressive (PPMS), and 4) Progressive-relapsing (PRMS). RRMS is the most common form of MS, affecting about 85\% of the people with MS. It is characterized by sudden relapse (major flare-up) of symptoms, followed by periods of remission where symptoms disappear. SPMS may develop in some RRMS patients, where the symptoms worsen or level off. PPMS occur in 10\% of the MS population, where there are no remission periods, and symptoms will steadily worsen, with occasional plateaus. PRMS is the rarest form of MS, with less than 5\% of the MS population affected. It’s progressive in the beginning, with periods of relapse, it has no remission periods. 
Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different episodes must occur in the disease course (time dissemination); and C) There is chronic inflammation of the CNS. One or more of these criteria result in a general diagnosis of MS. Both space and time dissemination criteria are confirmed with clinical MRI by radiologists. Detailed disability ratings are confirmed by the Expanded Disability Status Scale (EDSS)\cite{Kurtzke1983}, which tracks the progression of MS symptoms in an individual.

\subsection{Pathophysiology}

MS is defined by the existence of focal demyelination in the CNS that manifest in lesions or plaques. Ex-vivo dissections and ultra-structural examinations of this pathological tissue, obtained from autopsies of individuals who suffered from progressive MS or acute episodes. In acute tissue samples, the lesions showed signs of acute demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells. Partial remyelination by oligodendrocytes were also present, as well as signs of axonal swelling and injury. In chronic tissue samples, there is no presence of acute demyelination, and heavy loss of axons and myelin could be observed \cite{Filippi2012}.

\subsection{MR Imaging}

MR imaging is an important component of MS diagnosis, which permits more confident, and early detection of MS when combined with symptom criteria. The most important of these is the McDonald Criteria \cite{Polman2011}, and more recently the MAGNIMS guidelines \cite{Filippi2016}.

The MS lesions or plaques are characterized as focal hyper intensities in T2-weighted or proton-density images. Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium. Contrast agent usage however needs to be carefully tuned with the image acquisition parameters.
Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size. They are also found densely dispersed in paraventricular white matter. The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered when making a diagnosis. The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}. Paraventricular region require at least three or more lesions to be confirmed. Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.

Diffusion MRI (dMRI) is found to be sensitive to tissue changes such as edema, axonal injury and demyelination, and is becoming an important MS tissue quantification strategy. A number of diffusion metrics (Mean diffusivity (MD), radial diffusivity (RD), and axial diffusivity (AD)) have become important in MS tissue quantification and analysis.  They are particularly useful when study normal-appearing white matter. Normal-appearing white matter (NAWM) is microscopically abnormal white matter tissue that is visually indistinguishable from normal white matter in T2 MRI images. They are generally defined to be at least 1 cm away from a lesions. NAWM is found to include microscopic demyelination, gliosis, small round cell infiltration, macrophages and microglial activations. Only about 28 \% of NAWM is found to be normal, and there is a 12-42 \% reduction in axonal density in these tissues. DMRI study of NAWM have found decreased fractional anisotropy and increased mean diffusivity \cite{Ciccarelli2003d,DeGroot2013}.

\subsection{Pain and MS}
Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}. Common pain symptoms include optic neuritis, painful tonic seizure and trigeminal neuralgia. 91\% of the patients report pain at time of evaluation, and most experience daily pain. Both central and peripheral pain are reports in people of MS. Central pain in MS is reported to occur 27-58\% of the subjects \cite{O'Connor2008}. The most common central pain in MS are extremity pain, TN, and Lhermitte’s sign. TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population.

TN secondary to MS, or symptomatic TN (MS-TN), is indistinguishable from TN in terms of clinical presentation. It does however present with a demonstrable structural lesion other than vascular compression of CN V. MS-TN has a prevalence rate of about 2-4\% and in some studies, there are reports of prevalence rate as high as 6\%. Unilateral MS-TN is the most common pain type, however there is up to 30\% occurrences of bilateral pain.

### 2017-11-24 thesis/contents/reviews/pain.tex
Initial ShareLaTeX Import

\section{Pain}

Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species. In humans, the sensation of pain is generally described as having an unpleasant quality. Additionally, the experience of pain is complex, consisting of nociceptive, emotional and cognitive processes. The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.

Humanity have long sought to understand pain. In antiquity, thinkers in classical Greek traditions, including Plato and Aristotle considered pain as primarily an emotion instead of a physical senses\cite{Moayedi2012a}. Hippocrates and Galen believed that pain arose from imbalances in vital fluids of the body, and considered the heart to be the organ responsible for the experience of pain \cite{Melzack1999}.

The earliest theory that attribute pain to a dedicated physical pathway is commonly credited to Ren\'{e} Descartes (\ref{fig:descarte}). However, for Descartes, pain as an experience still could not be divorced from the idea of the soul, as he had stated in his letter to the French polymath Marin Mersenne: "I do not explain the feeling of pain without reference to the soul. For in my view exists only in the understanding. What I do explain is all the external movements which accompany this feeling in us." \cite[p. 148]{BookCottingham1991v3}. In his Treatise of Man, published in 1664, he stated: "Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch. In this way they pull the tiny fibre which you see attach to it, and simultaneously open the entrance to the pore, located opposite the point where this fibre terminates -- just as when you pull one of a string, you cause a bell hanging at the other end to ring at the same time. When the entrance to the pore or small tube is opened in this way, the animal spirits from cavity enter and are carried through it...the fibres cause a movement in the brain which gives occasion for the soul...to have the sensation of pain" \cite[p.101-103]{BookCottingham1991v1}.

\begin{figure}[ht]
\includegraphics[width=0.3\textwidth]{descartes-reflex.JPG}
\centering
\caption{Illustration of Descartes' pain pathway from the Treatise of Man}
\label{fig:descarte}
\end{figure}

Descartes' description was the first instance where the experience of pain was considered from a mechanistic perspective. The pain percept was for the first time framed as a physical system that could be understood and potentially treated.

Although the field of pain has changed significantly since the time of Descartes, a complete theory of pain still eludes modern neuroscience. Treatment of chronic pain is challenging still. However, investigations into unique syndromes such as trigeminal neuralgia offers an opportunity that further the understanding of pain and neuroscience at large.

\subsection{Theories of Pain}

The gate-control theory \cite{Melzack1965a} (\ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit primary afferent into the central transmission cells that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

 \begin{figure}[ht]
 \includegraphics[width=0.8\textwidth]{gate-control.jpg}
 \centering
 \caption{Melzack's gate control theory diagram.}
 \label{fig:gate-theory} 
 \end{figure}

The concept of a pain specific pathway was first put forward by Charles Bell where he proposed that the brain is a heterogeneous structure, and that nerves are a mix of heterogeneous neurons that became bundled to facilitate anatomical distribution \cite{Bell1868}. Evidence soon followed with the discovery of specific touch receptors and the demonstration that resection of posterior spinal tracts in animals by Schiff and Woroschiloff isolated innocuous touch from pain and temperature conductance (anterolateral tract) \cite{Dallenbach1939}. Von Frey demonstrated distinct receptive fields that respond to pressure that triggered tactile, warm, cold and pain. Sherrington further developed the specificity theory by coining the term “nocicipient” and unified attributes of intensity theory and specificity theory.

The classic pattern theory on the other hand, as first proposed by John Paul Nafe in the 1929 paper “A quantitative theory of feelings”\cite{Nafe1929}, postulates that the pain is not mediated by a separate nervous system or specialized receptors, but rather by units of undifferentiated afferent nervous impulses that only vary in their numbers, spatial and temporal relationships, including impulse frequency, duration, and density of nerve unit activations. The pattern theory postulates that all bodily sensations are a result of the brain's interpretation of nervous impulses. The discovery of high threshold mechanoreceptors and unmyelinated nociceptors however discredited the pattern theory and gave further credence to the theory of specificity.

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes all the necessary information for pain. That additional processing of information in the periphearla nd central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities is the pain sensory experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn ( laminae V ), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferent, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzacks, is the pain neuromatrix. Which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

\subsection{Pain Physiology}

\subsubsection{Peripheral Receptors}

\textit{Nociceptors}, the receptors in the peripheral nervous system that detect tissue damage can respond to a range of electrical or chemical signals released by noxious stimuli on the tissue, such as physical trauma or heat/cold temperatures exceeding the pain threshold. These receptors broadly separate into two categories, differentiated by the type of axonal innervation they receive. Acute pain receptors are innervated by $A\delta$ fibres, which permits fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick or acute trauma. Dull pain is triggered by receptors that are innervated by slow conducting C fibres ($<1$ m/s). $A\delta$ and C fibres can be co-activated in the presence of noxious stimuli.

\subsubsection{Nociception in the spinal cord}

The peripheral nociception signals are conducted from the axonal endings of the nociciption neurons, whose cell bodies are located in the spinal cord dorsal root gangions, to the dorsal horn of the spinal cord (\ref{fig:drg}). The $A\delta$ and C afferents project directly to lamina I of the dorsal horn. $A\delta$ afferents also directly project to lamina V. Lamina II, or the substantia gelatinosa, also receives inputs from C afferents, which indirectly synapse to lamina I neurons through interneurons. The dorsal horn also contains a dense layer of local interneurons in lamina III and IV, and receive input from innocuous somatosensory afferents at these layers.

\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{pain-drg.jpg}
\centering
\caption{The axonal projections of the fast conduction ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn.}
\label{fig:drg}
\end{figure}

The interneurons in the dorsal horn plays an important part in somatosensory integration as the sensory afferent pass from peripheral into the central nervous system. For example, the thermal grill illusion, discovered by Thunberg in 1896, involved the paradoxical feeling of heat or burning pain when adjacent skin regions are exposed to innocuous warm or cold stimuli \cite{Craig1994}. The mechanism is through the antagonistic interactions between innocuous cold A-fibers, cold heat pinch (CHP) C-fibers and warm C-fibers. Whereby the warm stimulus activates C-fibers that inhibit adjacent A-fiber activation, thereby unmasks the CHP C-fiber activations into the central nervous system.

Wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI. These neurons respond to both innocuous and painful stimuli, and are decidedly non-specific. They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing conclusive evidence that WDR neurons are sufficient for the percept of prolonged nociceptive pain.

In each spinal afferent level, the dorsal horn neurons project noceciptive signals to the brain via five distinct paths (\ref{fig:pain-pathways}). Lamina I, V-VII axons crosses the mid-line to form the \textit{spinothalamic} tract that ascend in the contralateral anterolateral column of the spinal cord, and project to thalamic nuclei. Laminae VII and VIII neurons from the \textit{spinorecticular} tract ascend via the anterolateral quadrant of the spinal cord, without crossing the midline, to reach the reticular formation and thalamus. The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, periaqueductual gray. The \textit{cervicothalamic} tract include only the white matters of the upper two cervical segments. It takes inputs from laminae III and IV, and projects to the thalamus via the medial lemniscus. The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions.

 \begin{figure}[ht]
 \includegraphics[width=\textwidth]{pain-pathways.jpg}
 \centering
 \caption{Afferent pain pathways} 
 \label{fig:pain-pathways}
 \end{figure}

\subsubsection{Thalamus}

Before reaching the neocortex, a majority of the second order pain afferents arrive at the thalamus. The spinothalamic tracts subserving face and body sensory information terminate in the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for face and body respectively. The VP complex sends third-order afferents to the primary sensory cortex (S1). Thus, the VP representation is somatotopic organized with the face region in the medial aspect, and lower limbs towards the lateral aspect. The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.

Another termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse. The MD nucleus projects third order afferents to the anterior cingulate cortex (ACC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.

An important nucleus has thought to be unique in humans is the ventromedial-posterior nucleus (VMpo)\cite{Willis2002,Craig2014}. It receives afferents primarily from dorsal horn lamina I, and can be considered to be contiguous with the basal ventral medial nucleus (VMb), that forms a relay for homeostatic afferents \cite{Craig2003}. The VMPo projects to the dorsal posterior insular cortex, which is thought to involve body interoception of pain.

The intralaminar nuclei centromedian (CL) and parafascicular (Pf) receive noxious heat afferents. CL receive afferents from laminae V and VII, while Pf from laminae I and V. They project to the basal ganglia, motor and parietal cortices, and are thought to be involved in general attention, arousal, and motor functions.

\subsubsection{Cortical projections of pain}\label{section:neuromatrix}

Pain percept is thought to involve distributed regions of the brain, coined as the “pain neuromatrix”\cite{Melzack1999a}. The regions are not functionally independent, but rather thought to be dynamically inter-related.

Neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a} have revealed consistent regions of BOLD co-activations when undergone painful stimulation. These classic fMRI pain regions include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). 
Regions such as ACC, PFC and insula are also shown to be active in social pain through social exclusion, and emotional stress through PTSD revealed changes in ACC and insula grey matter (Section \ref{fig:pain-matrix}).

 \begin{figure}[ht]
 \includegraphics[width=0.8\textwidth]{pain-matrix.png}
 \centering
 \caption{Cortical and sub-cortical regions of the pain matrix.} 
 \label{fig:pain-matrix}
 \end{figure}

Other studies suggest pain is correlated to the dynamically changing patterns of activities over time in the regions involved. Whereby the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}, and the dynamic rhythms of activities that attune to and away from saliency forms a complex pain connectome \cite{Kucyi2015}. Conditions such as chronic pain therefore show a differentiated tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable nociceptive stimuli \cite{Baliki2008,Legrain2009}.

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, which includes the location and intensity of pain include the primary sensory (S1) and secondary sensory (S2) cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. However, S1 has not been consistently reported in acute pain studies, possibily due to difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and is commonly found in pain studies. The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore it is believe to plays an important role in pain sensory-discrimination.

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain procecssing, which includes the unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions . The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}.

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Grecius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation. Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. DMN activation has been to correlate with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.

\paragraph{Insula}

The insula is an important neuroanatomical structure that has been implicated in both acute \cite{Lorenz2005,Starr2009} and chronic pain \cite{May2008,Jensen2016}. Anatomically it is subdivided into anterior and posterior components by the central insular sulcus. It can be further identified by its gyral features, consisting of 5 major gyri: anterior, middle and posterior short gyri in the anterior insula, and anterior long, posterior long gyri of the posterior insula. These gyri are not found consistently in the general population, but rather show individual variability \cite{Rosen2015}. Craig postulates that the insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}.

There is a recent debate pertaining the dorsal posterior insula as the "pain cortex" \cite{Segerdahl2015a}. Segerdahl et al. suggested that there is strong coupling between absolute cerebral blood flow and tonic pain intensity, and utilized arterial spin-labelling to quantify cortical activation in response to pain stimulation. The contralateral dorsal posterior insula emerged as the only cortical region with significant correlation. Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is strong supportive findings of a homologue region in humans. Critics of this finding raised issues with the low number of subjects (n=7) and long sampling duration (45 s), and the incompatibility of this interpretation with long standing bodies of neuroimaging literature \cite{Davis2015}.

\subsection{Chronic Pain}

Chronic pain, as defined by the International Association for the Study of Pain (IASP), is pain that persists or recurrent past the healing phase following an injury \cite{Apkarian2009}. It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months, other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}. There is a wide range of chronic pain conditions, including chronic back pain, migraine, arthritic pain, irritable bowel syndrome, phantom-limb pain, temporomandibular disorder, etc.

The new ICD-11 revision of chronic pain classification include a new category for chronic primary pain. It is defined as "pain in one or more anatomical regions that persists or recurs for longer than 3 months, and is associated with significant emotional stress or significant functional disability" \cite{Treede2015}. It includes chronic pain conditions where the etiology is unknown. Conditions such as fibromyalgia, and irritable bowel syndrome fall under this category.

Chronic pain conditions are considered to a type of maladaptive pain, whereby the pain response no longer serves its role in protecting the body from harm, but rather contributes negatively to the normal functions of life.

\subsubsection{Neuropathic Pain}

Neuropathic pain is a type of chronic pain, defined as "pain caused by a lesion or disease of the somatosensory nervous system" by the IASP. The pain experience may be constant or paroxysmal, related to a stimulus (evoked) or not-evoked. Animal models of neuropathic pain suggest that peripheral and spinal circuitry responsible for the transmission of nociception are altered under pathological conditions. The nervous system is believed to be altered by maladaptive plasticity following nervous system insult, and thereby lead to sustained neuropathic pain.

The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission and processing circuitry. 
Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex. The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation in the lower ganglions.

Peripheral neuropathic pain often occurs after lesion to the peripheral nervous system, and can be caused by physical trauma and infections. Central neuropathic pain is caused primarily by lesions in the central nervous system, including spinal cord injury, lesions relatting to multiple sclerosis or tumours.

\paragraph{Peripheral Sensitization}

Peripheral neuropathic pains can arise when there are disruptions in peripheral signal transmission. The disruption can be a result of injury to the first order peripheral afferent neurons, which result in increased sensitivity to stimuli and via ectopic action potentials. Changes in peripheral neurotropic factors as a result of an injury can cause increased membrane excitability or spurious sprouting of the synaptic terminals in the dorsal horn of the spinal cord. Additionally, maladaptive changes may occur in the secondary sensory neurons and interneurons in the dorsal horn laminae that amplifies or reroutes the noxious signals.

\paragraph{Central Sensitization}

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis, such that the synchronicity of action potential conduction between cortical and subcortical structures are disrupted, and thereby give rise to neuropathic pain. It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain (ref).

 The utilization of magnetic resonance imaging (MRI), coupled with novel imaging modalities such as functional MRI (fMRI) and diffusion MRI (dMRI), have been instrumental in our understanding of pain, permitting both \textit{in vivo} measures of brain activity during pain (functional studies), as well as assessment of changes to the brain microstructure.

 Studies have found that neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research.

 In general, brain regions that are involved in the somatosensation of pain process are involved in both acute and chronic pain. These regions include the posterior insula, lateral thalamus and dorsal striatum. In chronic pain conditions however, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Lorenz2005}. In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}. Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and meidal prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}. A shift in forebrain activities is observed in neuropathic versus acute pain, abnormal activities in anterior cingulate cortex and prefrontal cortex is commonly observed in these studies \cite{Davis2006}.

 Advances in modern machine learning methods led to their uses in attempting to improve neuropathic pain classification and diagnosis. Modern approaches combines cross-modal and additional analytical features, and can involve fMRI, diffusion imaging, as well as functional and/or structural connectivity. Machine learning in the form of multivariate pattern analysis to classify individuals with chronic back pain from normal controls, and achieve 92\% accuracy \cite{Callan2014}. Other studies in chronic pelvic pain using linear support vector machine was able to achieve 73\% accuracy \cite{Bagarinao2014}. Differentiating neuropathic and acute pain however is more difficult. Functional connectivity based classification on differentiating fibromyalgia, rheumatoid arthritis and controls was able to achieve a diagnositc accuracy of 69\%, and classification accuracy of 78\% \cite{Sundermann2014}.

### 2017-11-24 thesis/contents/reviews/trigeminal.tex
Initial ShareLaTeX Import

\section{Trigeminal Neuralgia}

Neuralgias by definition are pain without any detectable physical origin, and originate from the brain. Trigeminal neuralgia (TN), or Tic douloureux, is a type of facial neuropathic pain that involves the trigeminal nerve. TN symptoms involve paroxysmal, shock-like pain condition affecting one or more of the three facial sensory nerve branches. The pain episodes can be unpredictable, as the length of episode can range from 1 to 1462 days \cite{Katusic1990}. Patients often describe its symptom as the most painful experience a human can undergo. Its physical and emotional disturbances are often severely disabling.

The disorder has been documented as early as the first century AD, the clinical recognition of the disorder was by Nicolaus Andre in 1756, and by John Fothergill in 1776 \cite{Katusic1990}. In the three hundred years that followed, the aetiology of TN is still not completely understood. The modern understanding of the vascular compression of the trigeminal nerve root was first proposed by Janetta \cite{Jannetta1967}, where he elucidated that TN is a “nerve entrapment disorder” and that it may be a “consequence of aging … by arterialsclerotic elongation of the cerebrovasculature”.

Modern pain literature continuous to update the exactly diagnostic definition of TN.

An older classification scheme divides facial pain into 5 major subtypes \cite{Burchiel2003,Eller2005}: idiopathic (without apparent cause), trigeminal injury, symptomatic, postherpetic, and atypical facial pain. Idiopathic TN has 2 sub-types: 1) TN type 1, where the patient experiences greater episodic pain greater than 50\% of the time; 2) TN type 2, where constant pain occurs greater than 50\% of the time. Trigeminal injury can be unintentional: which can be due to facial trauma, surgery to the oral cavity; this type of pain would involve persistent and burning pain to the affected areas; and is termed trigeminal neuropathic pain. The injury can also be intentional: due to interventions such as rhizotomy or gangliolysis; these result in numbness and loss of sensation of the face; and is termed trigeminal deafferentation pain. Symptomatic TN, also termed MS-TN, is due to trigeminal nerve injury relating to multiple sclerosis related central white matter inflammation. Postherpetic TN is pain resulting from the infection of the herpes zoster virus, and can involving burning pain and deafferentation. Atypical facial pain is classified as somatoform pain disorder that can be of psychosomatic origin, although some literature can also include constant pain in the face into this classification.

The current The International Classification of Headache Disorders ICHD-3 Beta \cite{Society2013} classifies TN into classic TN, and painful trigeminal neuropathy. 
Classic TN's diagnostic criteria requires pain characteristics of: at least 3 attacks of unilateral facial pain; occuring in one or more trigeminal branches, with no radiation beyond the trigeminal distribution; severe intensity; electric shock-like, shooting, stabbing or sharp in quality; and precipatated by innocuous stimulation to the affected side of the face; and no clinically evident neurological deficit. Classic TN is further divided into purely paroxymal: recurrent attacks of unilateral facial pain with no persistent facial pain between attacks; and concomitant persistent: persistent facial pain of moderate intensity in the affected area. 
Painful trigeminal neuropathy in contrast are facial pain in one or more distributions of the trigeminal nerve caused by another disorder and indictive of neural damage. It can be further divided into categories relating to acute Herpes zoster, post-herpetic trigeminal neuropathy, multiple-sclerosis, space-occupying lesion, and other disorders.

There is also an definition \cite{Zakrzewska2016} that attempts to address the lack of symptomatic TN classification in ICHD-3 by classifying TN into idiopathic TN: without apparent cause; classic TN: caused by vascular compression of the trigeminal nerve root; and secondary TN: as the consequence of a major neurological disease such as MS.

\subsection{Anatomy}

The trigeminal nerve, or the fifth cranial nerve (CN V) is the primary afferent for both facial somatosensation and pain (\ref{fig:trig-branches}). It consists of, from superior to inferior, the ophthalmic (V1), maxillary (V2) and mandibular (V3) branches. The distribution of the branches consists of both motor and sensory components. The motor branch of CN V activates the muscles responsible for mastication, as well as tensor tympani, tensor velipalatini, mylohyoid and anterior belly of the digastric.

 \begin{figure}[ht]
 \includegraphics[width=120mm]{trig-branches.png}
 \centering
 \caption{Anatomy of the trigeminal nerve branches.} 
 \label{fig:trig-branches} 
 \end{figure}

The sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano and temperature pain, fine touch and vibration, such as the Messner's corpuscles, Pacinian and Ruffini corpuscles.

The V1 branch provides the sensory innervation to the superior aspects of the face and cranium, including forehead, upper eyelids, eyeball, frontal sinus, lacrimal gland, and dura of the anteior cranial fossa. It further divides into lacrimal, frontal, and nasociliary nerve branches. The lacrimal branch supplies the lacrimal; the frontal contains smaller branches that innervates the skin of the medial forehead, upper eyelid, and parts of the frontal sinus; the nasociliary branch innervate various parts of the eye, and mucous membrane of the anterior nasal septum, anterior portion of the lateral nasal wall, and the skin on the tip of the nose. V1 enters the skull base through the superior orbital fissure.

the V2 provides innervation to the temple, cheek, sides of the nose, and lower eyelids via the zygomaticofacial and zygomaticotemporal branches. It also innervates upper oral cavity. It enters through the foramen rotundum.

V3 contains sensory and motor branches. It innervates skins of the jaw, areas above the ear, lips, gum, and teeth. The V3 division carries the mastication efferent to the four muscles recruited for mastications: the masseter, temporal, medial and lateral pterygoids. The tensor tympani also receives some CN V innervation, which is thought to be important in dampening the sound of chewing. V3 enters through foramen ovale.

The branches then converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion), whose posterior extension forms the CN V. It is analogous to the spinal dorsal root ganglion, and contains all the trigeminal sensory cell bodies. It situates in the meckel's cave, on the petrous bone of the middle cranial fossa. It is somatotomically organized in respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.

The extension of the posterior aspect of the TRG forms the trigeminal nerve body, with both sensory and motor roots. The nerve body straddles the TRG and the trigeminal nucleus, by crossing the pontine cistern. The portion of the sensory nerve that enters the pons is termed the trigeminal root entry zone (REZ). The cistern trigeminal sensory nerve, similar to other peripheral nerves, is myelinated by Schwann cells. The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons. The CNS myelin can be several millimetres beyond the surface of the pons, before it give away to the transition area, termed Obersteiner-Redlich zone \cite{Peker2006}.

 \begin{figure}[ht]
 \includegraphics[width=100mm]{trig-pathway.png}
 \centering
 \caption{Overview of the trigeminal afferent pathways}
 \label{fig:trig-pathway}
 \end{figure}

The CN V passes into the pons (\ref{fig:trig-pathway}), through the pontine cerebellar peduncular bundles and enters the tegmentum of the pontine brainstem. CN V then curve towards the medial line before synapses onto the trigeminal nucleus in the basis of the pontine brainstem. The trigeminal nucleus is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei. The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.

The mesencephalic nucleus is responsible for proprioception, while the spinal trigeminal nucleus receives ipsilateral pain and temperature afferents not only from CN V, but also from CN VII, CN IX and CN X. Pars oralis transmits fine tactile sensations from the oralfacial region, pars interpolaris is also associated with dental pain, while pars caudalis is responsible for nociception and heat from the head.

The main sensory nucleus receives discriminative somatosensation from the oralfacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract. A separate branch responsible for oral cavity however project to the ipsilateral VC via the dorsal trigeminothalamic tract.

The VC thalamus projects to the primary sensory (S1) and secondary sensory (S2) cortical regions of the neocortex, thereby completing the afferent sensory projections of CN V.

\subsection{Pathophysiology}

Classic TN is believed to be caused by neurovascular compression (NVC) of the CN V at its root entry zone (REZ) by the surrounding vasculature, where the nerve interfaces with the pons. An aberrant loop of artery and less commonly a vein is seen in 80-90\% of the cases of TN to compressing against the nerve \cite{Love2001,McLaughlin1999}. This can be observed in conventional MR imaging \cite{Borges2010}. Generally, the superior cerebellar artery is responsible for most of the NVC cases (60-90\%), and then the anteroinferior cerebellar artery and basilar arteries (Lutz et al., 2011). There is also evidence that TN is preferentially lateralized to the right side in 61\% of the cases, and in very rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.  Other types of compressions may also cause TN, these can be in the forms of saccular aneurysm, arteriovenous malformation, vestibular schwannomas, meningiomas, epidermoid cysts, and other tumors.  However direct pathophysiological causal of TN pain is confounded by the fact observation of vascular compression does not always result in the expression of pain \cite{Desouza2013,Hodaie2013}. In some cases the TN pain is reported to be contralateral to the lesion \cite{Cheng2008,Haddad1990,Revuelta1995}. There are also 3.1-17\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006}.

Outside the brainstem, CN V is peripherally myelinated by Schwann cells, which are capable of remyelination. Inside the brainstem, the nerve is myelinated by oligodendrocytes. Study in 100 CN V samples from 50 cadavers \cite{Peker2006} showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm). The transition zone can be up to 25\% of the nerve length.

The NVC that occurs in TN is believed to disrupt the central myelin, and that the region of disruption would begin at the peripheral-central myelin transition zone, and extending into central myelin regions. Janetta \cite{Jannetta1967} proposed that NVC leads to focal demyelination of the trigeminal sensory root axons, and causes ephaptic nerve conductances that cause neuralgia. Ultrastructural examinations \cite{Love1998} of tissue samples obtained from CN V rhizotomy specimens showed oligodendrocyte demyelination at the site of root transition zone. In TN, adjacent to the site of compression-related focal demyelination, were cases of thinly myelinated sheaths that encapsulates a number of demyelinated axons, suggesting aberrant remyelination. The TRG is also postulated to be responsible for neuralgia \cite{Rappaport1994}. It's believed that ganglionic ectopic autonomic firing can be trigged by mechanical disturbances as a result of nerve compression. Additionally, ganglionic cross aftercharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding ganglionic neurons. Histology findings by Devor et al supported the ignition hypothesis \cite{Devor2002a}, where demyelinated axons were shown to be in close apposition without intervening glia.

\subsection{Treatment and outcomes}

For TN, anticonvulsant medication such as carbamazepine has 80\% response rate. However 50\% of patients will eventually need to undergo surgery as medication becomes ineffective due to tolerances. Nerve blocks such as peripheral neurectomy, neurolysis of the nerve or TRG can provide relief, albeit temporary, as pain always reoccurs \cite{Rappaport1994}. Long term treatment of TN involves the decompression of affected CN V \cite{Lovely1997}. This is achieved by either direct surgical intervention, in the form of microvascular decompression (MVD), where a Teflon\textcopyright padding is placed between the CN V root and the offending vessel. MVD showed 90\% pain relief immediately after surgery \cite{Zakrzewska2005}, and a long term success rate of 70\% at 10 years \cite{Barker1996}.
Partial sensory rhizotomy (PSR) can also be performed, where the offending CN V branch is severed, with a response rate of 88-70\% \cite{Young1993,Zakrzewska2005}.

Intervation can also be non-invasive, in the form of GammaKnife radiosurgery (GKRS) \cite{Hodaie2012g}. The GK system focuses gamma radiation from approximately 200 cobalt-60 ($^{60}$Co) sources onto specific targets along the trigeminal nerve. Earlier targets focused on the trigeminal ganglion on adjacent areas \cite{Leksell1983}, while more proximal targets towards the brainstem is now recently used\cite{Daugherty2015}. Proximal targets correlate with longer pain relief, but also higher chance of facial numbness\cite{Xu2014}. The procedure requires a stereotactic frame, and a dose of 70-90 Gy is deposited. The effect is delayed and pain relief can be experienced in 4-6 weeks post-radiosurgery. GKRS treatments result in 90\% initial pain relief \cite{Regis2015,Kondziolka2010}, about 50-80\% pain free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}. More recent studies suggested higher pain free probabilities at 3(71.8\%), 5(64.9\%), 7(59.7\%) and 10(45.3\%) years\cite{Regis2015}. More favorable treatment outcomes were related to no other accompanying symptoms, no prior surgery and shorter term pain history of less than three years.

### 2017-11-24 thesis/contents/sagit.tex
Initial ShareLaTeX Import

\graphicspath{{images/sagit/drafts/}}

\chapter{STUDY III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
\chaptermark{Study III}

This study was published in Frontiers In Neuroanatomy \textcopyright 2016 Chen et al. :

\bibentry{Chen2016}

\section{Abstract }
\textbf{Introduction:} Tractography analysis in group-based studies across large populations has been difficult to implement. We propose Selective Automated Group Integrated Tractography (SAGIT), an automated group tractography software platform that incorporates multiple dMRI practices which will allow great accessibility to group-wise dMRI. We use a merged tractography approach that permits evaluation of tractography datasets at the group level.  We also introduce an image normalized overlapping score (NOS) that measures the quality of the group tractography results. We deploy SAGIT to evaluate deterministic and probabilistic constrained spherical deconvolution (CST$_{det}$, CST$_{prob}$) tractography, extended streamline tractography (XST), and diffusion tensor tractography (DTT) in their ability to delineate different neuroanatomy, as well as validating NOS across these different brain regions.

\textbf{Methods:} MR sequences were acquired from 42 healthy adults. Anatomical and group registrations were performed using ANTs. Cortical segmentation was performed using FreeSurfer. Four tractography algorithms were used to delineate 6 sets of neuroanatomy: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation. The tracts were generated both with and without ROI filters. The generated visual reports were then evaluated by 5 neuroscientists.

\textbf{Results:} At a group level, merged tractography demonstrated that different methods have different fiber distribution characteristics. CST$_{prob}$ is prone to false-positives, and thereby suitable in anatomy with strong priors. CST$_{det}$ and XST are more conservative, but have greater difficulty resolving hemispherical decussation and distant crossing projections. DTT consistently shows the worst reproducibility across the anatomies. Linear regression of rater scores against NOS shows significant (p\textless0.05) correlation of the two sets of scores in filtered tractography. However, correlations are not significant (p\textgreater0.05) for unfiltered tractography.

\textbf{Conclusions:} The tractography results demonstrated reliable and consistent performance of SAGIT across multiple subjects and techniques. Through SAGIT, we quantifiably demonstrated that different algorithms showed different strength and weaknesses at a group level. While no single algorithm seems to be suitable for all anatomical tasks, it is useful to consider the use of a mix of algorithms for different anatomical segments. SAGIT appears to be a promising group-wise tractography analysis approach for this purpose.

\section{Introduction}

Diffusion magnetic resonance imaging (dMRI) tractography is an imaging analysis technique that permits non-invasive visualization of white matter anatomy $in$ $vivo$ \cite{Basser2000}.  It is based on the observation that the Brownian motion of water molecules within white matter fibers is constrained by the axonal bundles, and therefore such anisotropic water diffusion can be used to probe tissue microstructure \cite{Bihan2001a}. By scanning in multiple angular directions using diffusion weighted imaging (DWI) sequence, a model of diffusion can be estimated for each voxel of the scanned image volume. 
In the classic single-tensor dMRI (also known as diffusion tensor image; DTI) model \cite{Basser2002}, a single tensor is constructed at each voxel based on Gaussian model of diffusion, which describes the dominant diffusion direction. Using the tensor information, tractography algorithms \cite{Mori2002b} can be used to trace out structures within the DTI volume. The limitation of DTI is that there is insufficient information to resolve areas with crossing fibers with one tensor per voxel. Improvements over the limits of the DTI model, particularly in increasing angular resolution to improve crossing fibers resolution has been a subject of great interest \cite{Fritzsche2010d,Jeurissen2012d,Tuch2002f}.

To improve the crossing fiber information per voxel, significant changes in DWI acquisition strategies are required. These methods include diffusion spectrum imaging (DSI)\cite{Wedeen2005,Wedeen2008} which samples the Q-space in a Cartesian grid, and Q-ball imaging that is based on the Funk-Radon transform \cite{Cho2008,Descoteaux2007a} that samples the Q-space in a spherical shell. These approaches, as well as more complex multi-shell sampling strategies, aim to construct and sharpen an orientation distribution function (ODF) in order to provide better approximate of the underlying diffusion. These methods require complex and long DWI acquisitions that are often unsuitable for clinical applications. Alternatively, methods based on spherical deconvolution (SD) can be used on HARDI scans of lesser angular resolution. The approach is to model the HARDI signal as a convolution of the fiber orientations. The resulting deconvoluted fiber orientation distribution (FOD) shows better discrimination of fiber directions compared to Q-ball under similar scanning parameters, but is susceptible to image noise\cite{Anderson2005,Dell'Acqua2007,Tournier2004}. Alternatively, multi-tensor model-based approaches estimate the fiber configuration with the assumption that there are no more than two or three crossings in a given voxel. These include eXtended Streamline Tractography (XST), which can delineate lateral projections of the corticospinal tract in the motor cortex, in DWI scans with 50 gradient directions \cite{Qazi2009}. Other tractography algorithm include stochastically resolving tract propagation based on established white-matter prior probability (probabilistic tractography) \cite{Behrens2007}. Probabilistic approaches however are predominantly image-based, and result in a visitation volume image that needs to be visualized volumetrically.

There is increasing recognition in the field on the value of incorporating dMRI tractography into group-based studies in large populations. However, group tractography analysis has a number of challenges. The dMRI-derived geometries are difficult to register due to the large dimensionality of variables and low spatial resolution. Direct linear and non-linear deformations of either the tensor/ODF field can also confound diffusion metrics. Inter-subject anatomical variability also poses challenges for direct tractography geometric clustering and registration, making tract bundle identifications difficult. Selective anatomical selection also needs to be considered. TRACULA \cite{Yendiki2011} for example is able to perform group tractography analysis on pre-defined major white matter bundles based on ball-and-stick probabilistic tractography template, however there are no solutions available for fully customizable anatomy that incorporates new advances in dMRI tractography. Recently there are efforts to evaluate tractography algorithms with synthetic datasets such as Tractometer \cite{Cote2012,Cote2013}, as well as attempts to generate synthetic data from individuals \cite{Wilkins2015}. From these comparative studies, there is evidence that no single tractography algorithm is superior for the reconstruction of all white matter tracts. There are no detailed studies that have attempted to compare tractography algorithms across larger populations to determine their relative suitability for different white matter tract identification.

In the clinical setting, the anatomy of interest is often specific, and needs to be put in a context of high individual variability, therefore manual targeting of regions of interest (ROIs) are unavoidable. Across a population, manual ROI delineation is at the risk of high ROI placement variations due to operator bias. The complexity of tractography data across a population increases dramatically when iterative tuning of tractography parameters is considered. Therefore a combined and automated approach to group tractography in neuroimaging is highly desirable.

There is a lack of well-organized diffusion tractography software framework that bridges the gap between group image registration, diffusion image processing, large-scale tractography delineations and tractography evaluation. We propose an automated tractography software platform that incorporates existing and proven dMRI techniques, in order for group-wise dMRI to be more accessible to researchers. Selective Automated Group Integrated Tractography (SAGIT: https://github.com/sinkpoint/sagit) is a configurable and fully automated tractography generation and comparison pipeline. Its key contributions are: a) automated preprocessing, registration and tractography across an arbitrary number of subjects; b) flexibility in ROI definitions that allows high customizability for anatomical targeting; c) expressive ROI query that works with pre-defined segmentation masks such as FreeSurfer; d) compatibility with a number of popular tractography software that allows parameter iterations for consistent results.  SAGIT enables researchers to examine the result of different tractography algorithms at the group level.  We also introduce an image-based score (normalized overlap score, NOS) that can quantify the quality of the group tractography results across different tractography algorithms to further assist decision making for researchers.

Finally, we deploy our system to the evaluation of various tractography methods in their ability to delineate neuroanatomy that are highly specific but difficult to image. These include supra and infratentorial structures such as: small white matter bundles in heavy crossing fibers (cranial nerves), pontine decussation (rubrocerebellar pathway) and curved central pathways (fornix, visual and auditory radiations). We recruited 5 neuroanatomists to evaluate and rate the resulting visual reports generated by SAGIT. We then compared the results of NOS scores and the human raters to assess the automated NOS score as a tractography reproducibility metric.

\section{Methods}

\subsection{MRI Acquisitions}
Magnetic resonance images were acquired using GE Signa HDx 3 Tesla scanner with an 8-channel head-coil. MR sequences were acquired from 42 healthy adults (mean age $30.4\pm8.1$ years). Ethics approval was granted by the University Health Network Research Ethics Board (Toronto, Canada), MR images were acquired at the Toronto Western Hospital, and all subjects gave their informed written consent. Diffusion weighted images (DWI) were acquired with 1 B0 scan, 60 gradient directions, 3 mm slice thickness and in-plane resolution of 0.9375$\times$0.9375 mm, b=1000 s/mm$\textsuperscript{2}$, TE=86.4 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=240 mm, and matrix=256$\times$256. T$_1$ FSPGR anatomical scans were acquired with 1mm slice thickness and in-plane resolution of 0.9375$\times$0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256$\times$256.

\subsection{Preprocessing}
In order to automate the generation of tractography using multiple methods and minimize user-bias, the SAGIT framework is created (Figure 1). DWI sequences were corrected for eddy-current and motion distortions with appropriate rotational corrections to gradient vectors \cite{Leemans2009}. Fractional anisotropy (FA), axial diffusivity (AD) and radial diffusivity (RD) maps were generated from the DWI. A group specific average anatomical template was created with subject T$_1$ image, and intra-subject T$_1$ to DWI space was obtained using symmetric diffeomorphic registration (SyN) with Automated Normalization Tools (ANTs) \cite{Avants2008b}. Cortical and subcortical segmentations based on the T$_1$ images were obtained with FreeSurfer segmentation software \cite{Fischl2002b}.

\subsection{Tractography Delineation}
The tractography methods attempted to reconstruct the bilateral white-matter anatomy of: the fornix, facial/vestibular-cochlear cranial nerve complex (CN VII/VIII), vagus nerve (CN X), red nucleus pontine decussation (RN), lateral geniculate visual pathway (LGN) and medial geniculate auditory pathway (MGN). Seeding ROIs were defined on the group template, and projected to the individual DWI space.

Tractography filters were defined using a custom query expression to generate gray-matter/white-matter boundary inclusion and exclusion filter masks based on the FreeSurfer segmentations (See supplemental). No filtering rules were applied for the fornix. The fornix is used as a control group to judge rating bias. The unfiltered (filtering rules were not applied) versions of each anatomy were also generated for evaluation.

Single-tensor diffusion tensor tractography (DTT) using 3D Slicer version 3 \cite{Pieper2006,Tuch2000d}, two-tensor tractography using XST \cite{Qazi2009}, constrained spherical deconvolution based deterministic streamline tractography (CST$_{det}$) \cite{Tournier2012b} and constrained spherical deconvolution based probabilistic tractography (CST$_{prob}$) \cite{Tournier2010} using MRtrix version 3  were evaluated (See supplementary material).  
FA, AD and RD scalars were sampled and embedded into the tractography models using tri-linear interpolation from the corresponding image volumes in native DWI space. The affine and non-linear image registration transforms were applied to deform the native tracts from the DWI space to the template space; all corresponding tracts of the same anatomy in template space were then merged into one single tractography model, to obtain the merged group tractography geometry.

The individual tractography models were also converted to binary spatial images, and the resulting images were transformed to the template space. Multiple binary images were stacked together to form the conjunction percentage overlap image. The overlaps were then visualized using concentric isosurfaces created from step-wise (10\%) thresholds of the underlying overlap volume using the MayaVI data visualization library \cite{Ramachandran2011}. A color lookup table with even visual brightness falloff \cite{VanderWalt2015} was chosen to avoid visual judgment biases from the color presentation. Different viewpoints (axial, coronal, sagittal, and perspective) were created for each visualization, and then composed together to create the visual report.

\subsection{Tractography Evaluation}
The generated visual reports were then evaluated by 5 neuroscientists in a blinded study using a set of rating criteria (See supplemental). The conjunction images were at the same time used to generate NOS scores. The NOS score quantified the conjunction image generated by overlapping tractography masks to meaningfully determine the spatial agreement independent of anatomy. The assumption was that given a conjunction image, where its voxel value s denotes a range of overlap percentages between 0\% to 100\%, and value 1 that denotes the 100\% overlap value, we formalize it as $ s\in[0,1]$, and define NOS score as: 
\begin{equation}
n^{-1} \sum_{i=0}^{n-1} \frac{\ln(v_i)}{\ln(v_0)}
\end{equation}
where $n$ was the number of bins; $v_0$ was the number of voxels where $s \textgreater 0$;  $v_i$ was the number of voxels with $ 0 \textless s \ge \frac{i}{n}$; this study used $n=10$. Examples of NOS score behaviour are available in Figure 4. 
The rater scores were then normalized to each anatomy, and the normalized rater scores were linearly regressed against the NOS scores for correlations. Linear regression of rater scores against NOS was performed using the R-statistics library [R Development Core Team, 2012].

\begin{figure}[p]
\centering
\includegraphics[width=140mm]{figure1.png}
\caption{Flow diagram of the SAGIT group tractography framework. The SAGIT framework is designed to be fully configurable and extensible.}
\label{fig:SGfigure1}
\end{figure}

\afterpage{%
    \clearpage% Flush earlier floats (otherwise order might not be correct)
    \thispagestyle{empty}% empty page style (?)
    \begin{landscape}% Landscape page

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{figure4.png}
\caption{Examples of NOS scores of different conjunction images: The figure shows the coronal view of the left red nucleus tractography projections, as produced by different algorithms. The NOS score correlates closely with the visual colour scale. It can also resolve visually similar comparisons, such as the rating between XST and CST deterministic results.}
\label{fig:SGfigure4}
\end{figure}
	\end{landscape}
	\clearpage
}

\section{Results}

The fornix was shown to be consistent across the different algorithms. The visual report (Figure 2; the complete visual report can be found in supplemental) suggested that the subregions of the fornix were consistently delineated. The merged fornix tracts (Figure 3) showed different patterns of streamline distributions. CST$_{prob}$ and XST both showed wide spread streamline dispersions, while CST$_{det}$ and particularly DTT were limited to the region of the fornix. The fornix ratings (Figure 5) showed essentially no variability between raters.  The fornix NOS scores (Figure 6) similarly were consistent across the algorithms with little variability. Both filtered and unfiltered ratings were identical.

For the CN VII/VIII, CST$_{prob}$, CST$_{det}$ and XST were all able to delineate the cranial as well as brainstem segments of the nerve (Figure 2). CST$_{prob}$ however also delineated much of the cerebellum. DTT by Slicer was not able to delineate the brainstem portions of the fiber and therefore was scored lower than the other methods. These differences in delineation can be more clearly seen in the merged tracts (Figure 3), where the XST also showed further lateral delineations of CN VII/VIII. In unfiltered tractography, there was much more inter-rater variability for all except CST$_{prob}$. CST$_{det}$ and XST also showed lower average rating when unfiltered (Figure 5). The NOS showed similar trend where DTT resulted in the lowest score (Figure 6). The NOS also showed higher score for CST$_{det}$ when comparing to XST. There were no notable changes in NOS scores in unfiltered tractography.

For CN X, CST$_{prob}$ was rated the highest with no inter-rater variability. CST$_{det}$ and XST resulted in more rating variability and a lower average score (Figure 5). CST$_{prob}$ visually showed delineations that extended into the ipsilateral higher brain regions even with filters (Figures 2, 3). DTT showed the worst reproducibility, although its anatomy was recognizable in a small number of individuals.  In unfiltered tractography, CST$_{det}$ and XST showed reduced variability, and visually showed delineations of ascending and brainstem projects. DTT also showed improved ratings when unfiltered. NOS scores (Figure 6) for both filtered and unfiltered CN X tractography similarly showed an overall increase in score with unfiltered tractography.

For the red nucleus projections, visually CST$_{prob}$ showed the most consistent reproducibility of the decussation (Figure 2, Figure 3), with XST showing higher reproducibility over CST$_{det}$. DTT failed to delineate the pathway and instead delineated an erroneous path that decussated at the corpus callosum. CST$_{prob}$, CST$_{det}$ and XST resulted similar ratings, with unfiltered tracts showing lower rating and higher variability. DTT had the lowest rating, and very high variability (Figure 5). NOS score of the red nucleus projections highlighted the stronger performance of the CST$_{prob}$ over the other algorithms, and XST result was scored higher than CST$_{det}$ (Figure 6). The NOS scores of unfiltered tracts were notably higher for all algorithms, as the resulting tracts seemed to delineate much of the cortical spinal projections.

 The optic radiation showed the greatest reproducibility with CST$_{prob}$, where the Meyer’s loop could be seen in the result (Figure 2, Figure 3). CST$_{det}$ and XST showed low reproducibility of less than 20\%, and DTT failed to delineate any structure on the right side (see supplemental). The ratings reflected the observation (Figure 5), with CST$_{prob}$ having the full rating. CST$_{det}$ was rated higher in filtered tractography, while XST was rated higher when unfiltered. DTT also showed higher ratings when unfiltered. NOS score characterized the drop-off in reproducibility in filtered tractography for the deterministic methods (Figure 6), and the trend of rating under filtered tractography matched well with the ratings. NOS scores for unfiltered tractography however showed little match with ratings.

The auditory radiation was the least reproducible region. In filtered tractography, although CST$_{prob}$ showed high reproducibility in tractography (Figure 2, Figure 3), it was not clear if its delineation was correct. Visually there were wide spread area of false positives towards the occipital and anterior temporal lobe. The deterministic methods all showed poor ability to reach the Heschl’s gyrus. DTT was completely unable to produce any structures at all. For filtered ratings,  CST$_{prob}$ was the highest, and with XST rated higher than CST$_{det}$ (Figure 5). DTT scored zero due to the lack of delineated structures. When unfiltered, all algorithms showed widespread false positives in the hemispheres. Unfiltered ratings were over all higher than filtered ratings. NOS scores’ trends for filtered tractography matched well with ratings, but also highlighted the large differences in reproducibility (Figure 6). While NOS scores under unfiltered tractography did not correlate with ratings.

Linear regression of rater scores against NOS (Figure 7) showed significant (p\textless0.05) correlation of the two sets of scores in filtered tractography. However correlations were not significant (p\textgreater0.05) for unfiltered tractography. It can be observed that NOS score in unfiltered tractography agrees well with visual intuition, however rater variability and anatomical reports suggested greater false positives when presented with unfiltered results.

\afterpage{%
    \clearpage% Flush earlier floats (otherwise order might not be correct)
    \thispagestyle{empty}% empty page style (?)
    \begin{landscape}% Landscape page

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{figure2.png}
\caption{Example of the auto-generated visual panels. The panels that best represent the resulting anatomy are compiled. For anatomical reference, the best representative anatomy image slice is composed with the tract conjunction map. The color scale of the conjunction image represents the percentage reproducible for each region, where 1=100\%. For each algorithm (rows), different visual perspectives are created. Together 6 different anatomies were delineated (columns). To see the full visual reports generated, please see supplemental.}
\label{fig:SGfigure2}
\end{figure}

    \end{landscape}
    \clearpage% Flush page
}

\afterpage{%
    \clearpage% Flush earlier floats (otherwise order might not be correct)
    \thispagestyle{empty}% empty page style (?)
    \begin{landscape}% Landscape page
\begin{figure}[p]
\includegraphics[width=\linewidth]{figure3.png}
\caption{Example of merged tractography models for each of the techniques and anatomy. The fornix is colored to distinguish tracts from each of the 42 subjects, in order to demonstrate the effect of combining the group data. The other anatomies are coloured by FA intensity. The specific FA measures from each point were sampled from their native DWI space before tractography deformation. The grouping of the similar FA measures across the subjects show that the registrations were accurate. The Red Nuclei shows different metrics (FA and AD) depending on lateralization, in order to highlight the pontine decussation.}
\centering
\label{fig:SGfigure3}
\end{figure}
    \end{landscape}
    \clearpage% Flush page
}

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{figure5.png}
	\caption{Normalized anatomical scores as rated by experts. The error bars denote standard deviations of the ratings. The fornix serves as the control anatomy, therefore the represented filtered and unfiltered image are identical. It can be observed that unfiltered anatomy increases rater variability in some and decrease in others.}
	\label{fig:SGfigure5}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figure6.png}
\caption{NOS scores for each anatomy. It can be noted that having ROI filters severely reduces reproducibility in deterministic methods.}
\label{fig:SGfigure6}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figure7.png}
\caption{Correlation between rating scores and NOS. Filtered tractography NOS is significantly (p\textless 0.05) correlated with ratings. However unfiltered tractography show no significance.}
\label{fig:SGfigure7}
\end{figure}

\section{Discussion}

The present study performed automated group tractography generation and collation report on 6 sets of neuroanatomy in 42 subjects using 4 different types of tractography algorithms. The proposed automation pipeline performed with great reliablity. We have designed the automated pipeline to offer the ability to generate and collate tractography on a large scale. The key benefit of this approach is the vastly improved inter-subject tract reproducibility as a result of ROI filtering, and speed for tractography data generation. The method offers flexibility for researchers in ROI placements and parameter control for these tractography algorithms. The merged tractography permits additional analysis of the group-based anatomical reconstructions, and the auto-generated visual reports also reduce data confusion and improve research efficiency.

The most critical component of the pipeline is the reliability of image registrations. For this purpose, we picked ANTs since it is state-of-the-art and widely tested and deployed in numerous neuroimaging studies \cite{Avants2011,Klein2009}. For T$_1$ to DWI co-registration, we used mean diffusion weighted image (MDWI) as the registration intermediate. MDWI appears to be a reliable T$_1$-DWI co-registration intermediate in the absence of reverse-blip DWI acquisitions \cite{Chen2015b}. Future studies can also potentially improve T$_1$-DWI co-registration, for example with the possible use of anisotropic power image\cite{DellAcqua2014}.

We have deployed the pipeline to the task of delineating 6 sets of neuroanatomy from clinical DWI data. The anatomies were chosen to test the limits of the SAGIT pipeline as well as the tractography algorithms. These anatomies are often shown in an existing atlas, and contain curving projections that are difficult to delineate, but are nevertheless well defined and contain clear anatomical landmarks for ease of judgment. In total we included: two sets of cranial nerves (CN VII/VIII and CN X) for their small but precise anatomy; rubrocerebellar projections of the red nucleus, due to the difficulty of imaging the pontine decussation; optical radiation for its well defined landmarks and the Meyer's loop, which is difficult to image; auditory radiation, which is particularly difficult due to its course, that passes through a 3-way crossing in temporal lobe. The fornix serves as the control anatomical structure. We have previously studied fornix sub-region anatomies extensively \cite{Chen2015c}, and it is a popular anatomy to showcase the ability of new tractography algorithms due to the curvature of the forniceal crura \cite{Garyfallidis2014}.

It can be observed that results of averaged tractography image conjunction across a group are visually similar to probabilistic tractography in an individual, and that the group junction image seems to stabilize as the number of subject increase. It can be argued that the conjunction image captures the probable volume across a sampled population for a particular neuroanatomy in the template space. Their exact similarity and differences from probabilistic tractography at a group level should be explored in future studies. Additionally, the fidelity of the merged anatomy is often higher than what would be available from an individual. The result supports the validity of this technical approach. It is possible that the resulting average delineations can be used to generate population-specific anatomy atlases. It's also possible to use fiber density images rather than binary mask for conjunction. However the result of an average fiber density image is difficult to interpret, and therefore should be explored in future studies.

When attempting to judge tractography delineation quality, human experts often are able to assess a particular delineation by visually identifying known anatomical priors. Direct tractography visualizations suggest that wide streamline dispersions often obscure anatomical details in visualizations, making anatomical assessment difficult (figure 3). Therefore, image conjunctions offer better ability to assess reproducibility than direct tractography renderings alone. We aimed to develop an automated assessment of tractography reproducibility, since a) researchers are often prone to judgment bias when exposed to large number of tractography results, and that subtle differences in results are often hard to distinguish, and b) in an automated pipeline, the assessment score is necessary to allow iteratively fine tuning of tractography parameters. Since current neuroimaging techniques are not yet able to reliably classify neuroanatomy from tractography or image-based morphology, it is desirable for a computing pipeline to optimize and present the most reproducible tractography delineating across a population based on a particular set of ROIs as a first step. Subsequent anatomical assessment can then be made by a human expert, and the appropriate changes in ROI filter strategy can be made. This approach allows researchers with strong neuroanatomy backgrounds, but are less technically inclined, to make better judgments when performing group tractography.

The NOS and rater scores correlated significantly in filtered tractography. There was, however, little correlation in the case when tractography were unfiltered. The NOS scores are closely related to the change in visual scale of the averaged result; thus, suggesting that human ratings assessments of filtered tractography are well encapsulated by the NOS score, whereas the human decision factors for unfiltered tractography are more complex. Since the rating questionnaire is based on yes/no decisions on sub-anatomy identification, it is possible that the raters preferred to err more on false-positives than risking false-negatives, and this is what contributes to the differences in unfiltered tractography ratings. This means that human rating can become unpredictably biased with different filtering parameters and a more stable rating metric such as NOS may be more desirable.

For the specific tractography algorithms, CST$_{prob}$ produces the highest ratings, and performs particularly well in the optical radiation. It also results in more false-positives. This is most evident in CN VII/VIII and CN X results. In these cases the cranial nerve anatomies are very specific and local, and yet CST$_{prob}$ produces more false-positive projections to distant regions. In the fornix there are also erroneous tract extensions into the corpus callosum. Deterministic methods in comparison are more conservative and therefore result in lower scores. They perform well when the anatomy is regional and has finer features. Both CST$_{det}$ and XST performed well with cranial nerves. XST is able to image the rubro-cerebellar pontine decussation with more reliability than CST$_{det}$. While CST$_{prob}$ decussation delineation shows the highest reproducibility, it also produces wider projection coverage that’s harder to interpret. The low DTT score is not surprising, as the single-tensor's inability to resolve crossing-fibers is well known. Based on the result of the study, we recommend the use of CST$_{prob}$ for tasks in which the target anatomy is well defined and false positively can be easily recognized , or when interhemispheric and long projection distances are desired. For exploratory tractography, where filtering locations are not well defined, or when the anatomy features are close together, deterministic methods are recommended. When comparing CST$_{det}$ and XST, it appears that XST is more conservative and thereby results in less false-positives. It also has better ability in resolving hemispherical decussations, as evident by the results of the rubro-cerebellar projections. In practice the two methods are very close in performance. Note however that the comparison is not exhaustive of the possible parameter combinations on either method, and therefore automated tuning of the parameters can answer this question more definitively.

It is evident from this study that there is no single tractography algorithm that is superior in all aspects. The recent Tractometer challenge \cite{Neher2015}, which compared a wide range of tractography algorithms, including DTT and CST on a high resolution DWI phantom, has come to the same observation. It is clear that algorithm choice is highly task-dependent. Given that vast number of neuropathologies are not well understood, there is often no available ground truth of neuroanatomical measure in the clinical environment from which to form priors regarding algorithm performance.  Automatic tuning of tractography parameters in order to maximize anatomy and population-specific reproducibility, and thereby allowing task-specific tractography algorithm recommendations is a highly desirable future direction. The SAGIT platform and NOS score are a first step towards this goal.

\subsection{Limitations}
For this study the DWI images were acquired at 0.94$\times$0.94$\times$3 $mm\textsuperscript{3}$ voxel resolution. The dataset was acquired for the purpose of cranial nerve visualization on a 3T GE HDx MRI with 8-channels head coil, and is therefore incapable of less than 2.6 mm isovoxel DWI resolution at a clinically acceptable scanning time. For example, the average trigeminal nerve, one of the larger cranial nerves, has a diameter of about 2 mm, as of such we compromised on an anisotropic voxel resolution to gain in-plane resolution. We believe our findings is novel for the application of clinical tractography delineations in less than ideal conditions.

The limit of the NOS score is closed tied to the performance of its associated tractography algorithm. An algorithm and its associated parameters may consistently produce the wrong result in all subjects and produce a high NOS score, or it may produce no results, and result in a NOS score of 0. Both of these cases are regular occurrences in single-subject tractography analysis. With merged tractography, it is easier to recognize such faults. This is because in practice, incorrect tract path are often the result of imaging anomalies or algorithm limit, and are unstable across a population. This can be observed as a general trend in Figure 3, for example where DTI delineation of the rubrocerebellar tract is clearly incorrect, and results in low NOS score. Therefore, NOS score is capable of characterizing the uncertainty as a result of low algorithmic performance. In the case of false positives that are highly consistent, SAGIT can help researchers making more informed judgement.

\section{Conclusions}

In this study, the SAGIT platform was created as an automated group tractography software platform that incorporated existing and proven dMRI practices, in order for group-wise dMRI to be more accessible to researchers. The tractography results demonstrated reliable and consistent performance of SAGIT across multiple subjects and techniques. By deploying SAGIT on 42 subjects, we quantifiably demonstrated that merged tractography is able to demonstrate algorithmic differences at a group-level. CST$_{prob}$ is prone to false-positives, and thereby is suitable when the targeted anatomy is well known. CST$_{det}$ and XST are more conservative, but have more trouble resolving hemispherical decussation and distant crossing projections. The NOS score shows significant correlation with rater score for filtered tractography. Therefore it may be used for automated ratings of group tract results. As no single algorithm seems to be suitable for all anatomical tasks, it would be useful to consider using a mix of algorithms for different anatomical segments. Finally, we have demonstrated that merged tractography is a promising group-wise tractography analysis approach.

### 2017-11-24 thesis/contents/tn_sagit.tex
Initial ShareLaTeX Import

\graphicspath{{images/tn-gp-sagit/drafts/}}

\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
\chaptermark{Study IV}

\section{Abstract}
\textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial pain syndrome that involves the trigeminal sensory pathway. We demonstrate the use of an fully-automated group white-matter tractography, and Gaussian Process (GP) classification to pin-point key white-matter diffusivity changes that maximally differentiates patients and healthy controls.

\textbf{Methods:} We perform group merged tractography using SAGIT on 36 sex-matched TN patients (right-sided pain) and 36 controls. The trigeminal nerve (CNV), pontine decussation, and thalamocortico S1 pathways were reconstructed. GP classifiers were trained by scrolling a moving window over CNV and S1 tractography centroids. FA, GFA, RD, AD, and MD metrics were evaluated. Both TN vs. Control and Affect vs. Unaffected were analyzed. Classifiers that performed at greater-or-equal-to than 75\% accuracy were accepted.

\textbf{Results:} Both affected and unaffected CNV can be differentiated with up to 80\% accuracy from controls. Affected CNV differences were more focused in the cistern and root-transition-zone, whereas unaffected CNV show more focus towards the brainstem. The affected S1 showed focused MD differences towards the cortex with accuracy of 85\%.

\textbf{Conclusions:} We have demonstrated the first instance of group-wise merged tractography analysis of TN vs. control,  as well as the first white-matter machine learning classification with GP. The method reproduced previous findings with finer-grained regional specificity. The study also revealed specific diffusivity changes in deeper white matter.

\section{Introduction}
\subsubsection{Trigeminal Neuralgia}
Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxymal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CNV) branches. TN manifests classically in idiopathic fashion, or secondary to a range of neuropathology, including meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticerosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}. The pathophysiology of idiopathic TN is generally accepted to be vascular compression at the root of the CNV nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidences of vascular compression \cite{Lee2014}. We previously compared TN and MS-TN diffusivity differences in four segments of the CNV at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CNV diffusivity is disrupted in MS-TN. In all previous TN studies, placements of region-of-interests in the CNV were pre-determined and manually placed. In this study we further expand the study to A) analyze the CNV differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interests; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.

The trigeminal sensory system involves both the discriminative touch and pain pathways. The discriminative pathway synapsis onto the primary trigeminal sensory nucleus, and then project to the contralateral VC thalamus, and then to the S1 cortex as we have delineated. The pain pathway synapsis caudally onto the trigeminal spinal nucleus, and decussate at the level of the medulla and spinal cord; The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus; the pain pathway also projects to the medial nucleus of thalamus, where it then project medially to the cingulate cortex.

\subsubsection{Data Analysis from Streamlines}
Neuroimaging tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}. To obtain diffusivity metrics from tractography generally involves converting the streamline bundles into volumetric spatial mask, and then measure the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}. However this approach discards the orientation information in the streamlines. The naive form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-through. This becomes an issue when considering that an average 3 tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm. In white matter structures such as CNV, and other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.

Consider that a tractography streamline propagation algorithm can involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}. It is reasonable to consider these algorithms to be an interpolated subsampling scheme of the DWI volume that takes the diffusion-based orientation into account. Therefore, it is desirable to directly make use of the tractography streamlines for data measurement.

Along-the-tract analysis has been investigated by various authors \cite{Colby2012,ODonnell2009,Wang2015,Yeatman2012}. In these approaches, the bundles are selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and studies were limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and ucinate. The measures were pooled onto tract centroid in each individual subjects, and then aggregated at group level. The global tractography approach is unintuitive for certain anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is clear. A more focused ROI based delineation approach is therefore preferred in these circumstances.

There has been a number of centroid identification methods that were investigated. It’s important to distinguish the difference between streamline clustering and centroid identification. While a number of methods cluster the streamlines by identifying a polyline centroid, there are also methods that identify possible streamline clusters, but does not provide a best-representative centroid. Centroid identification involved resampling the streamlines into identical number of points, and then identifying point correspondences via different distance measures. For these approaches it’s important that the streamlines have similar length, and an assumption that the start and end of the streamlines across subjects are consistent. These studies focus on healthy subjects, but when consider that pathological brain tissue may negatively affect tractography delineation, then the geometry assumptions of these methods may not be satisfied when extending them to study patient populations. A number of these methods have running time and memory constraints that makes them infeasible to be deployed at a group level. One method for tract clustering and centroid identification that is shown to have linear running time in respect with streamline count is QuickBundles \cite{Garyfallidis2012}. We have adopted QuickBundles in order to obtain S1 streamline centroids.

The key advantage of a streamline-based approach is that we can retain the pathway orientation from its geometry. White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis. We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework (Chen et al., 2016) to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning. We then merge the tracts into a template space using non-linear deformations, and then perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.

\subsubsection{Gaussian Process}
The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} has sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}. However one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models. Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}. In fact it has been found that DNN with infinite number of hidden neurons approximates GP \cite{neal1996priors}.

Gaussian Process is defined \cite{rasmussen2006gaussian}) as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
\begin{equation}
	\begin{split}
		m(x) &= \mathbb{E}[f(x)], \\
		k(x,x') &= \mathbb{E}[(f(x)-m(x)(f(x')-m(x'))],
	\end{split}
\end{equation}
And the Gaussain processed is defined as
\begin{equation}
f(x) \sim GP(m(x), k(x, x')) 
\end{equation}

The covariance function is also termed called kernel function. The radial-basis function kernel or RBF kernel is widely used in machine learning tasks. GP have previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010}. It is particularly well suited for continuous data streams, such as tractography streamlines.

\section{Methods}
\subsubsection{Acquisitions}
The study recruited 36 idiopathic TN patients (28 females, 8 males; age 60.1$\pm$16 years; all right side pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).  All TN patients were scanned at 1 day pre-GK and 6 months post-GK. MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions. T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256. DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.

\subsubsection{Preprocessing}
DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}. Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains. The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}.

We deployed selective automatic group integrated tractography (SAGIT) framework to automate preprocessing, registration, and tractography. The processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations. The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}. Each individual DWI was registered to their T1 image with SDR, using mean DWI image for DWI-T1 co-registration. Freesurfer segmentation maps were first affine transformed from the Freesurfer space to the T1 space using FMRIB's Linear Image Registration Tool (FLIRT) \cite{Jenkinson2001,Jenkinson2002}, and then transformed using SDR transforms to the individual DWI space for selective ROI generation with SAGIT.

\subsubsection{Tractography}
ROIs were defined on the template image to bilaterally delineate: 1) CNV (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).

Constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} was used to delineate the three sets of anatomy. Deterministic CST was applied to delineate CNV and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy. Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration. Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = 45$\deg$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum lengh = 80 mm, initial cut-off = 0.15, cut-off = 0.15. Probabilistic tractography for TPT project was performed with step size = 1mm, and streamline count = 2000, and maximal length = 60 mm.

Tractography was first performed in the native DWI space for each individual.  FA, AD, RD, MD, GFA and AP metrics were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation. Additionally, subject ID and group ID scalar values were also inserted into each tractography files for identification.

\subsubsection{Merged Tractography}
All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis. A software library called “Fascicles” is created by the author to facilitate this process. The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and then all points were exported into a comma-delimited file. The points were then transformed by applying the subject DWI to T1, and T1 to template transforms using ANTs’  applyTransformToPoints command \cite{Avants2009}. The transformed points were then imported back into the SQLite database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved. Finally, all template-space tractography models across all the subjects are imported into a new SQLite database, and then exported as a single file for analysis.

\subsubsection{Along-the-tract Analysis}
The merged tract was first post-processed in 3D Slicer to remove erroneous and stray streamlines that will confound measurement. Streamlines removed include: remnant CNV streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels/noise; S1 streamlines that stray medially towards other regions such as the corpus callosum. The TPT streamlines were not examined due to inconsistent reproducibility across subjects.

Tractography centroids, which were the average curve best fit the tract bundle, was defined on the merged tracts in template space. CNV centroids were manually defined. The CNV centroids were defined by a list of 3 control points situated in the template world space. A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm. The spline is then resampled into 30 points that becomes the centroid definition with segment length of 1.3 mm. Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids. Diffusion metrics from the points were aggregated for each of the cross-sections.  S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threhold of 30, with centroid length of 62.1 mm and segment length of 2.07 mm.

\subsubsection{Gaussian Process Classification}

Diffusivity metrics of affected (right side) and unaffected CNV and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides. The affected and unaffected sides in TN patients were also compared with each other. 
Age was regressed out by using the residuals of the linear regression model from controls group. The diffusion values were subsequently normalized to zero mean with ranges $[-1, 1]$ for model training. 
Gaussian Process (GP) classification was exhaustively performed on CNV and S1 nerve segments in order to identify key segment differences. Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment. The entire machine learning process was performed using Scikit-learn machine learning software framework.
For each model, the subjects were split into train and validation sets using stratified 10-fold cross-validation. The GP model was initiated with Matern kernel (length scale = 1), and the model fitted using the training set. The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed segment. Segments accuracy greater than 75\% were accepted, and the segment length with the greatest accuracy was then determined.

\subsection{Results}
\subsubsection{Group Tractography}
Merged CNV delineations show anatomically consistency, and extended into the ipsilateral TGN. Changes in FA along the streamlines demonstrate clear differences between cistern and brainstem segments, suggesting good agreement in anatomical registration ( Figure \ref{fig:GPfigure1}). 
The TPT projections, when filtered strictly from the TGN to the ipsilateral VC thalamus region, showed distinct bundles of decussating pathways bewteen left and right. The resulting pathway are not symmetrical. Where the right TPT pathway diverts and envelopes the left (Figure \ref{fig:GPfigure2}). TPT streams lines were sparse, suggest low inter-subject reproducibility. Therefore no additional analysis were carried out for TPT.
S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}).

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{cnv-inferior-view.png}
\caption{Inferior view of the merged CNV delineations.}
\label{fig:GPfigure1}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{view-decussation.png}
\caption{Close up of the TPT decussation. Red: Right TPT; Blue: Left TPT. Note the Right TPT diverts around the left TPT bundle. }
\label{fig:GPfigure2}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{view-left-right.png}
\caption{The complete trigeminal discriminatory sensory pathway. }
\label{fig:GPfigure3}
\end{figure}

\subsubsection{CNV}
Both sides were distinguished from controls with classifiers that achieve greater than 75\% accuracy. 
Classifiers on the right side has performance ranging from 77\% (GFA) to 80\% (MD) accuracy. The left classifiers achieved accuracies ranging from 75\% (FA) to 80\%(GFA). The right side shows more consistent overlap of segment ranges between positions 10--17, in the region of the cistern/REZ; while the left side shows less consistent range overlaps, and AD segment range from positions 3 to 14, which include the brainstem regions. Overall, the right side features extend further into the distal space, into the cistern and even the ganglion, while the left side extend more into the brainstem. Comparing left/right, only the FA classifier achieved greater than 75\% accuracy, where its segments range from 11 to 19, which is the cisternal segment (Figure \ref{fig:GPfigure4}).

The best performing segments versus controls is GFA (segment 8--14) on the left (unaffected) side, with 80\% accuracy, 84\% precision, and AUC of 0.8; while MD (segment 10--17) performed the best on the right (affected) side, with 80\% accuracy, 84\% precision, and AUC of 0.79 (Table \ref{table:cnv}, Figure \ref{fig:GPfigure6}). The best performing left-versus-right classifier was FA (segment 11--19), with 75\% accuracy, 82\% precision, and AUC of 0.75.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{figure-GP-CNV.pdf}
\caption{CNV classifier performance}
\label{fig:GPfigure4}
\end{figure}

\begin{table}[h]
\centering
\csvautotabular{cnv_table.csv}
\caption{CNV GP classifier performance data}
\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
\label{table:cnv}
\end{table}

\subsubsection{S1}
Both sides of the S1 face region pathway were able to be distinguished by classifiers with greater than 75\% accuracy. 
The FA and RD classfiers at locations 617, representing the segment that emerge from the thalamus and merging into corona radiata, are common to both sides. The left side has higher classification accuracy of 80\% with FA, while the right side had a maximum accuracy of about 78\% with RD. 
The MD classifier mean while is unique to the left side, and achieved an accuracy of 85\% in locations 16--23, near the grey-white matter boundary, closer to the cortex (Figure \ref{fig:GPfigure5}).

The best performing segment versus controls was MD (segment 16--23) on the left (affected) side, with 85\% accuracy, 88\% precision, and AUC of 0.83; whereas the best on the right (unaffected) side was FA (segment 6--16), with 78\% accuracy, 81\% precision, and AUC of 0.76 (Table \ref{table:s1}, Figure \ref{fig:GPfigure6}).

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{figure-GP-S1.pdf}
\caption{S1 classifier performance}
\label{fig:GPfigure5}
\end{figure}

\begin{table}[h]
\centering
\csvautotabular{s1_table.csv}
\caption{S1 GP classifiers performance data}
\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
\label{table:s1}
\end{table}

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{figure-roc.pdf}
\caption{ROC curves of the top performing classifiers}
\label{fig:GPfigure6}
\end{figure}

\section{Discussion}
The study is the first instance of machine learning discrimination of TN vs. Controls with specific white matter analysis, using group-wise merged along-the-track tractography. In essence, we have demonstrated a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy solely from white-matter measures. We reproduced previous findings of TN diffusivity changes along the CNV with much higher regional specificity. We also examined TPT and S1 projections of the trigeminal sensory pathway.

The diffusivity measures of CNV clearly reproduced our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls. The consistent regional overlaps in the cistern/REZ region of the right (affected) CNV in MD, RD, AD shows the ability for the GP classifiers to auto-converge to focal white-matter abnormality. The left versus right GP classifier placed left-right differences at the cistern segment, in agreement with the pathophysiology of TN. The regional findings on left verus controls suggest that there is bilateral differences in CNV myeline in unilateral TN, despite the lack of symptoms on the unaffected side. The left regions also situate into the brainstem segments. It is possible that the left CNV is affected by global descending pain modulation, and warrant further investigation.

The S1 delineation was consistent, due to full automation using SAGIT and the incorporation of Freesurfer cortical segmentation maps. Reproducing streamlines into the face-region of S1 cortex in multiple subjects was difficult previously due to individual variations, as well as the multi-directional fiber crossings in the coronal radiata. Previous studies have suggested that S1 have key involvement in pain location discrimination \cite{bushnell1999pain}, and are found to have grey-matter abnormalities in TN pain \cite{Desouza2013c}. Our finding that the left (affected) thalamo-cortical S1 projection is uniquely different close to the cortex suggest that deep cortical white-matter/grey-matter changes may be related depending on the locality.

The TPT delineations showed that decussation pathway tractography is possibly asymmetric. It’s not clear if the asymmetry is a neuroanatomically factual, or an artifact from either algorithmic bias or partial volume bias of the underlying DWI. Resolving crossing fiber is still a difficult task in tractography, and while there are validations against MR phantoms, the anatomical validity of these methods can only be extrapolated, and the interpretation must involve anatomical priors from experts.

We included GFA in order to investigate its potential for measuring deep white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric. However S1 results demonstrated that it is not effective in pin-pointing deep white-matter abnormalities. GFA performed well in CNV, but not to any substantially degree than the Gaussian metrics.

Overall the GP classifiers resulted in higher precision when compare to accuracy. The overlapping regions revealed differences in localizations that may very well have been missed with singular diffusivity metric analysis. This suggest that cross-metric features should be considered in future studies. This would drastically increase the feature complexity, and thereby require subsequent demand in the number of subjects.

\subsubsection{Limitations}
The merged CNV delineations of all the subjects showed caudal projections towards the spinal cord. In multiple instances, diverging streamlines at the levels of the medulla strongly suggested the trigeminal spinal decussations. However the caudal projection was not available for majority of the TN group, and it was not a consistently producible pathway. This is likely due to that 1) 3T DWI scan has a limited 2mm isovoxel resolution; It is not enough to resolve CNV, medullary and spinal cord fiber details. We have modified our scan to 1x1x3mm voxel, to gain axial plane detail for this kind of small fiber tractography from clinical scans, at the sacrifice of coronal/sagittal resolution. 2) Large percent of the TN DWI scans has severe distortions at the pontine level. Similar distortions rarely show up in Control or MS-TN scans. The reason may be due to the changes in the vertebral and basilar blood vessels that gave rise to EPI distortions in these patients. These distortions narrow the brainstem in the DWI image, and when combined with limited resolution, making tractography delineations even more difficult. Therefore, it is highly recommended to perform reverse-blip DWI acquisitions in patients with cerebral vessel irregularities as part of MR DWI clinical practice.

The diffusivity metrics of the deeper white matter structures remains difficult to interpret, due to the heavy crossing fibers in these regions. Although metrics such as FA are popular in current neuroimaging literature, they nevertheless are derived from the Gaussian tensor model, which cannot adequately describe cross-fiber regional dynamics. For example, in a healthy crossing fiber region, where both bundles have equal myelin and axonal integrity, will have a low FA score due to a spherical tensor; its AD and RD will be roughly equal, and therefore MD will be higher. In a pathological cross region, where only one bundle is healthy, the regional FA actually increase due to the bias in one direction. Its RD will be lower, and its MD will be lower as well. Moreover, in diverging and converging structures, as well as in multi-crossing regions such as the corona radiata, the interpretations of Gaussian-based diffusivity metrics become even more unclear. Therefore, it is imperative that interpretable diffusivity measures for crossing regions be established and adopted by the neuroimaging tractography field for future studies.

\subsection{Conclusions}
We have demonstrated the first instance of group-wise merged along-the-track tractography analysis of TN vs. control. We have achieved examination of specific and targeted anatomy at a group level involving both healthy and pathological brain images. The method reproduced previous findings of TN diffusivity changes along the CNV with finer-grained regional specificity. We also examined TPT and S1 projections of the trigeminal sensory pathway. The study revealed that there are specific diffusivity changes on these deeper white matter structures, and demonstrated the power of along-the-tract analysis with machine-learning, which pin-points the exact location of the diffusivity disruption at millimetre level.

### 2017-12-02 thesis/contents/abstract.tex
checkpoint on edit

@@
 In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) software framework.
-Finally, in Study IV, we present an end-to-end machine-learning TN classification study that auto-discovers diffusivity disruptions in the cistern/REZ CNV and thalamo-cortico S1 pathway.
-In sum, this thesis presents a detailed roadmap of both the development and application of a novel end-to-end diffusion tractography machine learning classification.
+Finally, in Study IV, we present an end-to-end machine-learning TN classification study that auto-discovers diffusivity disruptions in the cistern/REZ CNV and thalamocortico S1 pathway.
+In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification.
 The application of this system to TN revealed specific diffusivity changes in deeper white matter structures, and pin-points the exact location of the diffusivity disruption at millimetre level.

### 2017-12-02 thesis/contents/discussions.tex
checkpoint on edit

@@
 \subsection{Delineating the trigeminal nerve into the brainstem}
-The primary motivation to visualize the trigeminal nerve via tractography, was due to the inability to discern both the brainstem trigeminal nerve body, nor the trigeminal nucleus in all known MRI modalities.
+The primary motivation to visualize the trigeminal nerve via tractography, was due to the inability to discern both the brainstem trigeminal nerve, nor the trigeminal nucleus in all known MRI modalities.
 Before tractography, the curvature of the trigeminal nerve could only be inferred from previous anatomical studies, or by postmortem dissection.
@@
 Discerning brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers.
-A number of methods has been devised, including probabilistic tractography made popular by the FSL software package, which generates image-based results that represents the propagation probability of a seed region.
+A number of methods have been devised, including probabilistic tractography made popular by the FSL software package, which generates image-based results that represent the propagation probability of a seed region.
 Neurosurgeons however often need visualizations that represent anatomy geometrically.
 Therefore streamline-based tractography visualizations are of broader appeal.
-In our experience, both XST and sphere harmic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions.
+In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions.
 Another often overlooked area is the seeding pattern.
@@
 \subsection{Tractography group registration}
-Tractography is a powerful methodology that reveals the morphology of brain white matter that were otherwise previously invisible.
-It's primary usage however has been limited to case studies and small scale anatomy comparative analysis.
+Tractography is a powerful methodology that reveals the morphology of brain white matters that were otherwise previously invisible.
+Its primary usage however has been limited to case studies and small-scale anatomy comparative analysis.
 The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of large groups.
 
-Group comparative studies requires the establishment of anatomical consensus, such that similar white matter structures between populations can be readily studied.
+Group comparative studies require the establishment of anatomical consensus, such that similar white matter structures between populations can be readily studied.
 Efforts in MRI medical image registration have paved the way to this end, especially in the fields of diffeomorphic registration.
-The difficulty in adapting registration to tractography, is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods.
-The best efforts to date applies affine transformation for between-tract registration \cite{Garyfallidis2015}, however true non-linear geometric registration is still a topic of future research.
+The difficulty in adapting registration to tractography is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods.
+The best efforts to date apply affine transformation for between-tract registrations \cite{Garyfallidis2015}.
+However non-linear geometric registration is still a topic of future research.
 Our approach is to recast tractography as a brain segmentation method that is spatially-aware.
-Tractography then can be considered a point-based sub-sampling method that are represented in the same coordinate frame as the native DWI space.
+Tractography then can be considered a point-based sub-sampling method that is represented in the same coordinate frame as the native DWI space.
 Any transforms applied to this DWI space is then applicable also to any tractography models that share its coordinate frame.
 Existing diffeomorphic registration solutions can then be used to non-linearly distort tractography, and circumvent the computational difficulty of geometry based non-linear registration.
-The advantage of our tractography registration strategy is that any new advances in the field of MR registration can be readily adopted to improve the output of our method.
+The advantage of our tractography registration strategy is that any new advances in the field of MR registration can be readily adapted to improve the output of our method.
 
@@
 Aggregating along-the-measure statistics is a form of clustering that requires projection of tract vertices onto a structural skeleton, or centroid.
-Common clustering methods calculates a polyline centroid based on different distancing functions between streamlines.
-These methods however require tractography streamlines to have similar length, and will underperform when tractography structures contain sharp turns.
+Common clustering methods calculate a polyline centroid based on different distancing functions between streamlines.
+These methods,  however, require tractography streamlines to have similar length, and will underperform when tractography structures contain sharp turns.
 We have explored a number of strategies, and considered adaptation of manifold-based methods, and point-cloud skeletal extraction method from computer graphics literature.
-These approaches however contain assumptions that do not suit our needs.
-For example, manifolds commonly consider that the latent structure of the data lie in a lower-dimensional surface, while tractography are common arranged in complex bundles; while point-cloud methods deals with point-cloud data obtained from 3D scanners, that are situation on a 3D shell, while tractography data are solid point clusters.
-White matter anatomy can have complex crossings and multi-way join and diverging structures that are difficult for existing 3D analytical methods that often need simpler assumptions to reduce the number of unknown parameters.
+These approaches, however, contain assumptions that do not suit our needs.
+For example, manifolds commonly consider that the latent structure of the data lies on a lower-dimensional surface, while tractography is commonly arranged in complex bundles.
+Moreover, point-cloud methods deal with point-cloud data obtained from 3D scanners, which situate on a 3D shell, while tractography points are distributed within a volume.
+White matter anatomy can have intricate crossings, multi-way join and diverging structures.
+These are difficult for existing 3D analytical methods, and require models with simplified assumptions to reduce the number of unknown parameters.
 A potentially promising future direction may be to automatically discover these more complex latent structures using deep learning methods.
@@
 
-The findings in Study I \ref{section:study1} revealed clearly localized CNV tissue differences in MS-TN and idiopathic TN.
-The implication is that disruption along the primary trigeminal afferent may lead to similar TN-like pain syndromes.
+The findings in Study I \ref{section:study1} revealed localized CN V tissue differences in MS-TN and idiopathic TN.
+The possible implication is that disruption anywhere along the primary trigeminal afferent may lead to similar TN-like pain syndromes.
 MS is a classic myelin inflammatory disease, and therefore these diseases may share similar disruptions in myelin ultrastructure.
-In Gamma-Knife radiosurgery treatment for TN, the target is often the Gessarian ganglion, which situates distally from the CNV nerve root, beyond the affect mid-cistern segment.
-The mechanism for GK treatment may function as a "volume control" for peripheral trigeminal afferents.
-Where some interaction between peripheral innocuous sensory conductance and the site of proximal CNV disruptions.
-These interaction may take the form of cross-talk between axons that carry discriminative sensory signals and pain fibers, or maladaptive convergences in the trigeminal nucleus.
-It's likely that these interaction are probabilistic in nature, due to the observation that TN pain is paroxysmal.
 
-Study IV further suggested that the unaffected CNV diffusivity is also different from that of controls.
+Study IV further suggested that the unaffected CN V diffusivity is also different from that of controls.
 The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain.
-In many studies it has also been used as a basis of comparision.
-This finding however is not as suprising as one may suspect.
+In many studies it has also been used as a basis of comparison.
+This finding however is not as surprising as one may suspect.
 It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases.
-Moreover, in third degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space.
-It is reasonable that these observable macroscopic changes can effect CNS tissue changes in the unaffected side.
-Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases a more centrally involved changes on the trigeminal nucleus and its associated pathways is possible.
+Moreover, in third-degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space.
+It is reasonable that these observable macroscopic changes can effect CNS tissue changes on the unaffected side.
+Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible.
+
+In Gamma-Knife radiosurgery treatment for TN, the target is on the cistern CN V segment, and situates proximally to the CN V REZ.
+Paradoxically, the GK treatment can be seen as a willful disruption of the CN V myelination.
+The question is, why would additional GK lesions on the CN V myelin effectively decrease TN pain.
+Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain.
+The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved.
+Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure.
+The force exerted onto the CN V from its neighbouring vasculature is therefore periodic.
+It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition".
+This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain.
+Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain.
+This model also explains that GK intervention disrupts the chance of resonance condition, thereby reducing TN pain.
+The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns.
+MVD, in this case can be explained as the physical disruption of the condition.
 
@@
 This was limited by the clinical 3-channel GE 3T HDx MR scanner, which places constraints both the acquisition resolution and time.
-Commonly, DWI voxel resolution on a 32-channel 3T research scanner can obtain up to 2mm iso-voxel resolution, on older 8-channel clinical scanners, maximum iso-voxel resolution is limitied at about 2.6mm iso-voxel.
-This is clearly insufficient for the study of cranial nerves such as CNV.
+Commonly, DWI voxel resolution on a 32-channel 3T research scanner can obtain up to 2mm isovoxel resolution, on older 8-channel clinical scanners, the maximum isovoxel resolution is limited at about 2.6mm isovoxel.
+This is insufficient for the study of cranial nerves such as CN V.
 Our resolution is limited to 1x1x3 mm voxel dimensions, in order to retain in-plane resolution at the sacrifice of slice thickness.
-Despite this, certain regions such as the TPT could not be reliably reconstructed in all of subjects.
-Anatomy featuring intricate but complex cross-sections, such as the thalamo-insula projections, which involves the claustrum, and medullary decussations of the spinal trigeminal fibers will remain out of reach of the current algorithms without substantial improvement in DWI resolution.
-Fortunately, SAGIT can adopt MR images of any resolution, so future studies in higher iso-voxel resolution can be immediately processed at soon as they become available.
+Despite this, certain regions such as the TPT could not be reliably reconstructed in all of the subjects.
+Anatomy featuring intricate but complex cross-sections, such as the thalamoinsular projections, which involves the claustrum, and medullary decussations of the spinal trigeminal fibers will remain out of reach of the current algorithms without substantial improvement in DWI resolution.
+Fortunately, SAGIT can adopt MR images of any resolution, so future studies in higher isovoxel resolution can be immediately processed as soon as they become available.
 
@@
 
-The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers are not limited to the application of TN.
+The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers,  are not limited to the application of TN.
 The methods can generalize to any anatomy where consistent tracts can be obtained at a group level.
@@
 For pain studies, detailed examination of the pain matrix, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest.
-Multi-modal and multi-variable study with function connectivity with this technique would also be possible.
+Multi-modal and multi-variable study with functional connectivity with this technique would also be possible.
 
-The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal important insights in the MS population.
+The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal additional insights in the MS population.
 Where focal lesion groups can be correlated with disease symptoms such as pain and motor disabilities.
 
-Gaussian Process can be used as an regressor to model tissue parameters along merged white-matter bundles.
+Gaussian Process can be used as a regressor to model tissue parameters along merged white-matter bundles.
 Any MR scalar image can be embedded into the tractography streamlines, so tissue quantifications such as G-ratio, myelin, proton-density, kurtosis, magnetization transer, PET images can be examined.
-Since GP is a bayesian method, the derived GP model can be used to provide predictive confidence on centroid measures from single subjects.
-When combined with the end-to-end tractography extraction ability of SAGIT, this provides a basis to build deployable framework for disease inference on new patients.
+Since GP is a Bayesian method, the derived GP model can be used to provide predictive confidence on centroid measures from single subjects.
+When combined with the end-to-end tractography extraction ability of SAGIT, this provides a basis to build a deployable framework for disease inference on new patients.
 
@@
 
-In sum, this thesis provides novel evidence on the diffusivity tissue disruptions of the CNV in TN, and that there is evidence of bilateral CNV tissue changes in TN patients when compare to controls.
-The examination was also extended to the thalamo-cortical S1 pathway and have demonstrated localized white matter changes near the white-grey matter boundary.
+In sum, this thesis provides novel evidence on the diffusivity tissue disruptions of the CN V in TN, and that there is evidence of bilateral CN V tissue changes in TN patients when compared to controls.
+The examination was also extended to the thalamocortical S1 pathway and have demonstrated localized white matter changes near the white-grey matter boundary.
 These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies as a whole.
-These result demonstrate that TN involve wide-spread changes at all levels of CNS white matter, and that the methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.
+These results demonstrate that TN involves wide-spread changes at all levels of CNS white matter, and that the methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.

### 2017-12-02 thesis/contents/paper1.tex
checkpoint on edit

@@
 
-This study was published in Multiple Sclerosis Journal \textcopyright 2015 Chen et al. :
+This study was published in Multiple Sclerosis Journal \textcopyright 2016 Chen et al. :
 
-\bibentry{Chen2015a}
+\bibentry{Chen2016a}
 
@@
 The major technical limitation of diffusion tractography to date has been the inability of resolving crossing fibres using a single tensor Gaussian diffusion model\cite{Farquharson2013}, which has proved unsuitable for the visualization of brainstem tracts.
-However, techniques that use high angular resolution imaging  with an increased number of gradient directions acquired during diffusion weighted imaging (DWI), and advances in multi-tensor tractography (MTT) have improved the ability to delineate fibres in brain regions that have dense fibre crossings, such as the brainstem \cite{Dell'Acqua2007,Descoteaux2009c,Fillard2011,Qazi2009}.
+However, techniques that use high angular resolution imaging  with an increased number of gradient directions acquired during diffusion weighted imaging (DWI), and advances in multi-tensor tractography (MTT) have improved the ability to delineate fibres in brain regions that have dense fibre crossings, such as the brainstem \cite{DellAcqua2007,Descoteaux2009c,Fillard2011,Qazi2009}.
 The microstructure of CN V in TN has recently been shown to be abnormal in terms of DTI metrics in the trigeminal REZ\cite{Desouza2013}.

### 2017-12-02 thesis/contents/reviews/dmri.tex
checkpoint on edit

@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 
-To properly characterize anisotropic diffusion, there must be more than 1 gradient directions available.
-In fact it is possible to describe voxel-wise diffusion using a Gaussian diffusion model, where the diffusion anisotropy can be visualized as an ellipsoidal tensor \cite{Jones2002}.
+To adequately characterize anisotropic diffusion, there must be more than one gradient directions available.
+In fact, it is possible to describe voxel-wise diffusion using a Gaussian diffusion model, where the diffusion anisotropy can be visualized as an ellipsoidal tensor \cite{Jones2002}.
 This type of imaging methodology is termed diffusion tensor imaging (DTI).
-At least 6 non-collinear gradient directions must be captured in order to compute the DT.
-In practice up to 32 gradients directions are used clinically.
+At least six non-collinear gradient directions must be captured to compute the DT.
+In practice, up to 32 gradients directions are used clinically.
 
-The diffusion tensor is a $3 \times 3$ symmetric matrix that describes diffusion displacement in the three major axis of diffusion:
+The diffusion tensor is a $3 \times 3$ symmetric matrix that describes diffusion displacement in the three major axes of diffusion:
 
@@
 $\hat{\lambda_1}$ points to the orientation of the greatest diffusion direction, and therefore $\lambda_1$ is also referred to as "longitudinal" or "axial" diffusivity.
-$\hat{\lambda_2}$ and $\hat{\lambda_3}$ are the lesser orthogonal diffusion directions, and their mean magnitude give rise to the cross-sectional, or "radial" diffusivity.
+$\hat{\lambda_2}$ and $\hat{\lambda_3}$ are the lesser orthogonal diffusion directions, and their mean magnitude gives rise to the cross-sectional, or "radial" diffusivity.
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 FA has been found to correlate with a wide spectrum of neuromorphological changes, including but not limited to multiple sclerosis \cite{Cercignani2001,Ciccarelli2003d,Miron2012}, vestibular schwannomas \cite{Chen2011b,Taoka2006,Wai2009}, tumors \cite{Mori2002,Schonberg2006,Nimsky2005}, schizophrenia \cite{Rametti2009,Fitzsimmons2009}, traumatic brain injury \cite{Bendlin2008,Kinnunen2011c,Hellyer2012}.
-
 RD is found to positively correlate with degree of demyelination in multiple sclerosis \cite{Song2005,Song2002,Klawiter2011,Janve2013}, and well as stages of white-matter development and aging \cite{Counsell2006,Sala2010}, including alzheimer's disease \cite{Acosta-Cabronero2012}.
-
-AD is correlated positively with axonal integrity \cite{Song2003}, and is found to be a correlate of axonal injury \cite{Budde2009} such as those found in traumatic brain injury \cite{Kinnunen2011c} and Gulf War illness \cite{Rayhan2013},
-
+AD is correlated positively with axonal integrity \cite{Song2003}, and is found to be a correlate of axonal injury \cite{Budde2009} such as those found in traumatic brain injury \cite{Kinnunen2011c} and Gulf War illness \cite{Rayhan2013}.
 MD is correlated with tissue inflammation and edema \cite{Senda2012g}, and is found to correlate with diffusivity changes as a result of gliomas \cite{Stadlbauer2007}, multiple sclerosis \cite{Cercignani2001,Senda2012g,DeGroot2013}, amyotrophic lateral sclerosis \cite{Sharma2012}, and aging \cite{Kantarci2011}.
@@
 This type of acquisition strategies, which enable higher order diffusion modelling algorithms, are generally referred to as high angular resolution diffusion imaging (HARDI).
-The most wide spread strategy is to equally distribute gradient directions as vertices of a triangle-tessellated geodesic sphere \cite{Frank2001}.
-Alternatively, it's also possible to equally spread gradient directions in a rectilinear grid, which forms the basis of diffusion spectral imaging (DSI) \cite{Wedeen2005}.
-Additional acquisition strategies such as distributing gradient directions across multiple concentric spherical shells is also actively explored \cite{Jeurissen2014a}.
+The most wide-spread strategy is to equally distribute gradient directions as vertices of a triangle-tessellated geodesic sphere \cite{Frank2001}.
+Alternatively, it is also possible to distribute gradient directions in a rectilinear grid, which forms the basis of diffusion spectral imaging (DSI) \cite{Wedeen2005}.
+Additional acquisition strategies such as distributing gradient directions across multiple concentric spherical shells are also actively explored \cite{Jeurissen2014a}.
 
@@
 The determination of gradient sampling pattern and B-value directly affect acquisition time and image quality.
-More complex gradient sampling and higher B-values can improve tractography quality, however the acquisition time can become prohibitively long in a clinical setting.
+More complex gradient sampling and higher B-values can improve tractography quality.
+However the acquisition time can become prohibitively long in a clinical setting.
 Other factors that must be considered include the nature of patient pathology, motion artifacts, and the inclusion of other imaging modalities during clinical scan sessions.
@@
 
-An end-to-end spherical deconvolution diffusion model was first proposed by Tournier et al \cite{Tournier2004}. 
+An end-to-end spherical deconvolution diffusion model was first proposed by Tournier et al. \cite{Tournier2004}. 
 Consider a fiber population with distinct fiber components. The measured diffusion signal $S(\theta,\phi)$ can be written as:
@@
 \end{equation}
-where $\theta$ is the elevation angle,  $\phi$ the azumuth angle in spherical coordinates; $f_i$ is the volume fraction of the $i$-th fiber component, $\hat{A}$ is the rotation operator that orients fiber, and $R(\theta)$ is a axially symmetric response function.
+where $\theta$ is the elevation angle,  $\phi$ the azimuth angle in spherical coordinates; $f_i$ is the volume fraction of the $i$-th fiber component, $\hat{A}$ is the rotation operator that orients fiber, and $R(\theta)$ is an axially symmetric response function.
 
@@
 The spherical harmonic representation of $S(\theta,\phi)$ is then:
+
 \begin{equation}
@@
 \end{equation}
+
 To reduce the number of parameters to be computed, we can make the further assumption that the diffusion process is symmetric, and therefore the odd $l$ harmonic orders can be ignored, with only the even harmonics considered.
-The response function $R(\theta)$ can be directly estimated from the dataset by supplying DWI region with high anisotropy from a single fiber bundle, such as the body of corpus callosum.
+The response function $R(\theta)$ can be directly estimated from the dataset by supplying DWI region with high anisotropy from a single fiber bundle, such as the body of the corpus callosum.
 
-The maximum haronic order $l_max$ that can be estimated is dependent on obtained number of directions.
+The maximum harmonic order $l_max$ that can be estimated is dependent on the obtained number of directions.
 For example, $l_max$ = 8 requires 45 directions, $l_max$ = 10 requires 66 directions, and so on.
@@
 Further advances to the estimation method of the spherical deconvolution model made it more robust to noise \cite{Tournier2007b}.
-The noise of SD estimation are represented as spurious negative lobes in FOD, which are physically impossible, and do not represent any real phenomenon.
+The noise of SD estimation is represented as spurious negative lobes in FOD, which are physically impossible, and do not represent any real phenomenon.
 To mitigate this, a non-negativity constraint is added.
@@
 The Tikhonov regularization then constraints the negative peaks to zero, and a new set of FOD and peaks are obtained.
-This is applied iteratively until convergence, and the final constrained FOD is then supplied.
+The process is applied iteratively until convergence, and the final constrained FOD is obtained.
 
-The Tikhonov regularization minimizes the weighted sum two terms:
+The Tikhonov regularization minimizes the weighted sum with two terms:
 \begin{equation}
@@
 \end{equation}
-Where the first term is a lear least-square fit of solution $f$ to to data $b$ with problem matrix $A$. 
+Where the first term is a linear least-square fit of solution $f$ to to data $b$ with problem matrix $A$. 
 In terms of CSD, $f$ corresponds to the SH coefficients of the FOD, $b$ is the measured signal and $A$ is:
@@
 This means that existing DTI protocols can be extended for high order HARDI tractography by simply increasing the number of directions.
-With CSD, a 60 direction DWI acquisition can be used to confidently derive FOD with $l_{max}=8 $, and up to $l_{max}=12$ with super resolution enabled \cite{Tournier2007b} (\ref{fig:csd}).
+With CSD, a 60 direction DWI acquisition can be used to confidently derive FOD with $l_{max}=8 $, and up to $l_{max}=12$ with super resolution enabled \cite{Tournier2007b} (Figure \ref{fig:csd}).
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 \subsection{Tractography}
-Tractography is the construction, or propagation of streamline from an underlying field of orientation information, in order to visualize continuous structures within a volume.
+Tractography is defined as the construction, or propagation of streamline from an underlying field of orientation information, in order to visualize continuous structures within a volume.
 The orientation field is generally provided as a 3D matrix of directional information obtained from DTI, CSD, or other methods.
-For example, in DTI the directions is the eigen vector with the greatest magnitude $\hat{\lambda_1}$; for CSD, the directional vectors would be peaks from the FOD.
+For example, in DTI the directions are the eigen vector with the greatest magnitude $\hat{\lambda_1}$; for CSD, the directional vectors would be peaks from the FOD.
 
@@
 The first tracking algorithms were applied to DTI using step-wise polyline propagators that followed local DT tensors \cite{Conturo1999}.
-Further refinement to the line propagator to include neighbouring DT tensors, and interpolation of the local DT directions in order to obtain a smooth curve \cite{Mori2002b}.
-Basser introduced Runge-Kutta integration to allow more robust path tracing when faced with noise in the local directions \cite{Basser2000}.
+Further refinement of the line propagator to include neighbouring DT tensors, and interpolation of the local DT directions in order to obtain a smooth curve \cite{Mori2002b}.
+Basser introduced Runge-Kutta integration to allow more robust path-tracing when faced with noise in the local directions \cite{Basser2000}.
 Another similar path-integral method was also proposed by Tuch \cite{Tuch2000d}.
@@
 The most popular of the probabilistic approach is the FSL software's probabilistic tractography module (FDT) \cite{Behrens2007}, which also includes a model-based ball-and-stick diffusion model that can resolve 2-fiber crossings.
-The FDT software produces visitation images of its probabilistic streams that collates the probability of tract propagation at each voxel.
-The visitation image however is hard to interpret and localize, and therefore does not rival deterministic methods in surgical visualization.
+The FDT software produces visitation images of its probabilistic streams that collate the probability of tract propagation at each voxel.
+The visitation image, however, is hard to interpret and localize, and therefore does not rival deterministic methods in surgical visualization.
 

### 2017-12-02 thesis/contents/reviews/machine-learning.tex
checkpoint on edit

@@
 Machine learning (ML) is a branch of computer science that aims to learn and form predictions from data.
-In contrast with static algorithmic programming, where the structure of the algorithms are predefined, machine learning approaches construct models based on observed data, in order to extend its predictions to the unobserved.
+In contrast with static algorithmic programming, where the structures of the algorithms are predefined, machine learning approaches construct models based on observed data, in order to extend its predictions to the unobserved.
 Strategies and algorithms for constructing various models overlap with the fields of artificial intelligence, computational statistics, pattern recognition, and neuroscience.
-Advanced in modern machine learning, especially in deep neural networks \cite{LeCun2015}, in recent years have lead to wide spread adaptation of machine learning based online services in translation \cite{bahdanau2014neural}, image search \cite{simonyan2014very}, and speech recognition \cite{graves2013speech}.
-There is also surging interest in deploy machine learning systems in radiology and disease diagnosis in medicine.
+Advanced in modern machine learning, especially in deep neural networks \cite{LeCun2015}, in recent years have led to wide spread adaptation of machine learning based online services in translation \cite{bahdanau2014neural}, image search \cite{simonyan2014very}, and speech recognition \cite{graves2013speech}.
+There is also surging interest in deploy machine learning systems in radiology and disease diagnosis in medicine\cite{Kamnitsas2016,Suk2014}.
 
 Machine learning training strategies can be categorized into Supervised Learning, Unsupervised Learning, and Reinforcement Learning.
-Supervised learning involves training data that are pre-labeled.
-For example the data supplied can be a list of patients' age and gender, and the label can be whether they or not they are in pain.
-Unsupervised learning involve data without any prior labels, and the algorithm attempts to discover patterns within the data itself, and is often used in data mining.
+Supervised learning involves training data that are pre-labelled.
+For example, the data supplied can be a list of patients' age and gender, and the label can be whether they or not they are in pain.
+Unsupervised learning involves data without any prior labels, and the algorithm attempts to discover patterns embedded within, and is often used in data mining.
 Reinforcement learning involves continuously learning with a reward function.
-It is generally used to train agents that have a continuous stream of data and needs to interact with an external environment, and is widely used in robotics.
+It is generally used to train agents that have a continuous stream of data and needs to interact with an external environment and is widely used in robotics.
 
 \subsection{Supervised Learning}
-Supervised learning is the mostly widely used method of training strategy.
-Most of the existing machine learning system in production are based primarily on supervised training with human-supplied data labels.
+Supervised learning is the most widely used method of training strategy.
+Most of the existing machine learning systems in production are based primarily on supervised training with human-supplied data labels.
 Different algorithms can be used to fit a function that forms a decision boundary based on data features and labels.
-Possible models ranges from Decision Trees, linear models such as Support Vector Machines (SVM), Deep Neural Networks (DNN), and bayesian-based models such as Naive Bayes and Gaussian Process (Figure \ref{fig:ml-bounds}).
+Possible models range from Decision Trees, linear models such as Support Vector Machines (SVM), Deep Neural Networks (DNN), and bayesian-based models such as Naive Bayes and Gaussian Process (Figure \ref{fig:ml-bounds}).
 
@@
 \subsubsection{Gaussian Process}
-The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} has sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}.
-However one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models.
+The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}.
+However, one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models.
 Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process \cite{gal2016dropout}.
-In fact it has been found that DNN with infinite number of hidden neurons approximates GP \cite{neal1996priors}.
+DNN with an infinite number of hidden unit approximates GP \cite{neal1996priors}.
 
-Gaussian Process is defined \cite{rasmussen2006gaussian}) as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
+Gaussian Process is defined as a collection of random variables, any finite number of which have a joint Gaussian distribution \cite{rasmussen2006gaussian}. 
 It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
@@
 \subsubsection{Training}
-The process of fitting a ML model on labeled data to form predictions is called training.
-The objective of the training process is to obtain a model that can both correctly predict observed labels, and is able to generalize its predictions to new data that are not part of the training set.
-Depending on the model algorithm, there can be many training parameter that govern the model outcome.
+The process of fitting an ML model on labelled data to form predictions is called training.
+The objective of the training process is to obtain a model that can both correctly predict observed labels, and can generalize its predictions to new data that are not part of the training set.
+Depending on the model algorithm, there can be many training parameters that govern the model outcome.
 The tuning of these parameters is referred to as hyper-parameter tuning.
-It is possible for certain hyper-parameters that result in model over-fitting, which is when a model is over-specialized in achieving the most accurate training label prediction, but perform poorly on unseen data.
-To properly evaluate model performance, a dataset is split into training/testing and validation sets, usually in 4:1 or 9:1 ratios.
-Where model training is performed solely on the training set, and evaluation of model performance is on the validation set to assess the ability for the model to generalize.
+It is possible for models trained with certain hyper-parameter combinations to result in model over-fitting, which is when a model is over-specialized in achieving the most accurate training label prediction, but perform poorly on unseen data.
+To properly evaluate model performance, a dataset is split into training/testing and validation sets, usually in 4:1 or 9:1 ratios, where model training is performed solely on the training set, and evaluation of model performance is on the validation set to assess the ability of the model to generalize.
 
-There are cases when the number of available data is low, and each data point may have a substantial impact on model performance.
-In these cases cross-validation is used.
-The idea is to divide the data into all possible combinations of training and validation splits, usually in the form of 5 fold to 10 fold.
+There are instances when the number of available data is low, and each data point may have a substantial impact on model performance.
+In these cases, cross-validation is applied.
+The purpose is to divide the data into all possible combinations of training and validation splits, usually in the form of 5 fold to 10 fold.
 Then separate models are trained and evaluated on each combination, and their performance is averaged to form the final evaluation.
 This gives a reasonable approximate on the performance of the model when trained using all available data.
-Leave-one-out cross-validation is an extreme form of this: given a dataset of N items, each split will consist of 1 and N-1 items, which result in N possible combinations.
-It is used when the available data is extremely low.
+Leave-one-out cross-validation is a special form where only one item is used for the validation split: given a dataset of N items, each split will consist of 1 and N-1 items, which result in N possible combinations.
+It is used when the available data are few.
 
@@
 
-There has been a long history of applying statistical methods to the field of neuroimaging, from the very onset of the field.
-Neuroimaging studies often attempt to correlate physiological measures from neuro-activities with specific experimental variables or conditions.
+There is an extensive history of applying statistical methods to the field of neuroimaging, from the very onset of the field.
+Neuroimaging studies often attempt to correlate physiological and image-based measures with neuro-activities with specific experimental variables or conditions.
 Neuro-activities may be measured from EGG sensors, or as voxels in functional or structural MR images.
-Traditionally data from each measured component was assumed to be independent.
+Traditionally data from each measured feature component was assumed to be independent.
+These components maybe individual sensor readings, or voxel-based measures.
 Correlative analysis were performed using general linear models, with post-hoc false-positive corrections.
+This type of univariate analysis do not have the ability to discover key relationships between feature components.
 Therefore univariate models cannot take into account spatial or temporally correlative dynamics.
 
-Increasing complexity in neuroimaging data have lead to widespread applications of multi-variable machine learning methods in neuroimaging.
+Increasing complexity in neuroimaging data has lead to widespread applications of multi-variable machine learning methods in neuroimaging.
 Application of ML in pain neuroimaging became controversial when Tor Wager \cite{Wager2013} demonstrated a Lasso-PCA model that classifies nociceptive pain with sensitivity and specificity of 90\% solely from fMRI imaging data.

### 2017-12-02 thesis/contents/reviews/mri.tex
checkpoint on edit

@@
 \section{Magnetic Resonance Imaging}
-Magnetic Resonance Imaging (MRI) is a medical imaging technology that allows non-invasive visualization of the body using magnetic fields.
-It is a key enabler of modern healthcare and medical research, and is used in the detection and diagnosis of neuropathology.
+Magnetic Resonance Imaging (MRI) is an imaging technology that allows non-invasive visualization of the body using magnetic fields.
+It is a crucial enabler of modern healthcare and medical research and is used for the detection and diagnosis of neuropathology.
 
 MRI is based on the principles of nuclear magnetic resonance (NMR), where atomic nuclei, when placed in a magnetic field, can absorb and emit electromagnetic radiation.
-A MRI machine applies a primary magnetic field (B0) to magnetize and align the hydrogen nuclei in the body.
+An MRI machine applies a primary magnetic field (B0) to magnetize and align the hydrogen nuclei in the body.
 A radio frequency (RF) field is subsequently introduced perpendicular to B0, and reorients the spin axis.
@@
 
-As protons aligned in the parallel orientation are more stable, the number of spins under this state is more than that under anti-parallel orientation.
+There are more protons with spins with parallel orientation than anti-parallel, due to the increased energy stability.
 This number of difference is called spin excess and the magnetic moments of the excess spins add to form the net magnetization vector $M_z$.
@@
 A transverse magnetization is introduced into the system that oscillates at radio frequency (RF) in order to modify the net magnetization, this is called the RF pulse.
-When exposing the nuclear spin system to energy of the RF pulse, the equilibrium of the spin system is disturbed.
+When exposing the nuclear spin system to the energy of the RF pulse, the equilibrium of the spin system is disturbed.
 The RF pulse is designed to introduce energy E such that the net spin direction orients at $90^\circ$.
@@
 When the RF pulse is switched off, the spin now relaxes towards the z-axis.
-} \label{fig:energy}
+} \label{fig:relaxation}
 \end{center}
@@
 
-T2 relaxation is so named after the time constant of phase decoherance of the transverse magnetic moment.
+T2 relaxation is so named after the time constant of phase decoherence of the transverse magnetic moment.
 
@@
 \begin{enumerate}
-\item Spins are associated with local magnetic fields, and interact randomly The energy transfer that occur in the local magnetic field interactions cause a mumulative loss of phase.
-It's a pure spin-spin interaction, and thereby not affected by $180^\circ$ refocusing pulse.
-It is also independent from the strength of $B_0$.
-The time constant which describes the return to equilibrium of the transverse magnetization, is called the \textit{spin-spin
+\item Spins are associated with local magnetic fields, and interact randomly.
+The energy transfers that occur in the local magnetic field interactions cause a cumulative loss of phase.
+These are pure spin-spin interactions, and thereby not affected by $180^\circ$ refocusing pulse.
+They are also independent of the strength of $B_0$.
+The time constant which describes the return to equilibrium of the transverse magnetization is called the \textit{spin-spin
 relaxation time} or T2, and is defined in respect to time $t$ as:
 $$M_{xy} =M_{xymax} e^{-t/T2}$$
-where T2 is the time it takes for the transverse magnetization to decay to 37$\%$ of its original magnitude. 
+where T2 is the duration of time for the transverse magnetization to decay to 37$\%$ of its original magnitude. 
 T2 is always less than or equal to T1.
@@
 T1 contrast is adjusted by tuning repetition time (TR), which is the length of relaxation interval between excitation pulses.
-When TR is long (> 1500 ms), $M_z$ recovers totally, and therefore subsequent excitation will again retain maximal signal strength; when the TR is short (< 600 ms), the $M_z$ has not totally recovered, and the next excitation pulse will not be able to gain strong MR signal.
-Different tissues have different intrinsic T1 differences, and therefore will show strong contrast under short TR.
+When TR is long (> 1500 ms), $M_z$ recovers totally, and therefore subsequent excitation will again retain maximal signal strength.
+When the TR is short (< 600 ms), the $M_z$ has not entirely recovered, and the next excitation pulse will not be able to regain signal strength.
+Different tissues have different intrinsic T1 differences, and will show strong contrast under short TR.
 Therefore short TR are said to be T1 weighted.
@@
 \paragraph{Echo Time}
-MR sequences have a gradient field which is used to control field  inhomogenieties and to encode spatial origin.
-The gradients also contribute to spin dephasing, therefore a refocusing pulse is necessary to adequately measure MR signals.
-The time delay between refocusing, and the measured signal under phase coherence (spin echo) is known as echo time (TE).
-When TE is short ($<$ 30 ms), there is little signal differences between tissue types; when TE is long ( $\geq$ 60 ms ), the intrinsic tissue differences will allow good contrast in the resulting image.
+MR sequences have a gradient field which is used to control field inhomogeneities and to encode spatial origin.
+The gradients also contribute to spin dephasing.
+Therefore a refocusing pulse is necessary to measure MR signals adequately.
+The time delay between refocusing and the measured signal under phase coherence (spin echo) is known as echo time (TE).
+When TE is short ($<$ 30 ms), there are little signal differences between tissue types; when TE is long ( $\geq$ 60 ms ), the intrinsic tissue differences will allow good contrast in the resulting image.
 Therefore long TE is said to be T2 weighted.
@@
 When a 90deg RF pulse is applied, the spins orients towards the xy-plane.
-When the RF pulse is switched off, the spins now relaxes towards the z-axis, and the spin undergoes phase decoherance.
-A 180 degrees flip angle pulse is applied to prevent spin saturation, and then a refocusing pulse is applied to bring spins back into coherance.
+When the RF pulse is switched off, the spins now relaxes towards the z-axis, and the spin undergoes phase decoherence.
+A 180 degrees flip angle pulse is applied to prevent spin saturation, and then a refocusing pulse is applied to bring spins back into coherence.
 } \label{fig:spin-echo}
@@
 Due to short TR, this scan can be run very quickly, allowing the collection of high resolution 3D structural image (Figure \ref{fig:T1T2PD}, Left).
-These structural images can provide detailed information of the brain's shape and size, and provide the base for cortical surface reconstruction and segmentation.
+These structural images can provide detailed information on the brain's shape and size, and provide the base for cortical surface reconstruction and segmentation.
 
@@
 Therefore, to allow slice discrimination, the magnetic field along the z-axis is made to be inhomogeneous by introducing a linear gradient field.
-It's an extra coil that adds or subtracts from the main magnetic field.
+It is an extra coil that adds or subtracts from the main magnetic field.
 With the linear gradient, only a particular portion of the z-axis would share a similar unique Larmor frequency, and can be selectively excited with an RF pulse.
@@
 \item Frequency encoding: The frequency encoding gradient is applied on the x-direction.
-This applies a magnetic field that increase in strength from right to left.
+This applies a magnetic field that increases in strength from right to left.
 The change in Larmor frequency alters the precession speed of the spins such that the spins towards the right precess faster than that of the left.
@@
 Application of Fourier transform will decompose the signal into component frequencies along the frequency encoding direction (x).
-The phase distribution along each frequencies provide information on the phase encoding direction (y).
+The phase distribution along each frequency provide information on the phase encoding direction (y).
 Decomposing the phase distribution requires another Fourier transform on the multiple scans of different phase gradient strengths.
@@
 The collected MR data are encoded in a mathematical space termed the k-space.
-The k-space is a two dimensional feature space, where the x-axis $k_x$ represent the measured frequency domain, and $k_y$ the phase domain.
+The k-space is a two-dimensional feature space, where the x-axis $k_x$ represent the measured frequency domain, and $k_y$ the phase domain.
 Each step in $k_y$ represents a single phase encoding step.
 The x-axis origin $k_x=0$ is the gradient iso-center, and the frequency distribution is symmetric about the x-origin.
-The 2D-FT operation is applied to the k-space to transform the frequency domain information into image.
+The 2D-FT operation is applied to the k-space to transform the frequency domain information into an image.
 

### 2017-12-02 thesis/contents/reviews/multiplesclerosis.tex
checkpoint on edit

@@
 
-The MS lesions or plaques are characterized as focal hyper intensities in T2-weighted or proton-density images.
+The MS lesions or plaques are characterized as focal hyperintensities in T2-weighted or proton-density images.
 Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium.
-Contrast agent usage however needs to be carefully tuned with the image acquisition parameters.
 Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size.
@@
 The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}.
-Paraventricular region require at least three or more lesions to be confirmed.
+Paraventricular region requires at least three or more lesions to be confirmed.
 Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.
@@
 Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}.
-Common pain symptoms include optic neuritis, painful tonic seizure and trigeminal neuralgia.
-91\% of the patients report pain at time of evaluation, and most experience daily pain.
+Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as trigeminal neuralgia.
+91\% of the patients report pain at the time of evaluation, and most experience daily pain.
 Both central and peripheral pain are reports in people of MS.
-Central pain in MS is reported to occur 27-58\% of the subjects \cite{O'Connor2008}.
-The most common central pain in MS are extremity pain, TN, and Lhermitte’s sign.
-TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population.
+Central pain in MS is reported to occur 27-58\% of the subjects
+\cite{OConnor2008}. The most common pain in MS include extremity pain and TN. TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population.
 
-TN secondary to MS, or symptomatic TN (MS-TN), is indistinguishable from TN in terms of clinical presentation.
+TN secondary to MS, or symptomatic TN (MS-TN), is frequently indistinguishable from TN in terms of clinical presentation.
 It does however present with a demonstrable structural lesion other than vascular compression of CN V.

### 2017-12-02 thesis/contents/reviews/pain.tex
checkpoint on edit

@@
 
+Trigeminal Neuralgia is a chronic facial pain syndrome.
+It is therefore imperative to review the evolving history and theories of pain.
 Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species.
@@
 
-The gate-control theory \cite{Melzack1965a} (\ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit primary afferent into the central transmission cells that relays touch or pain sensations.
+The gate-control theory \cite{Melzack1965a} (\ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations.
 The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission.
@@
 Von Frey demonstrated distinct receptive fields that respond to pressure that triggered tactile, warm, cold and pain.
-Sherrington further developed the specificity theory by coining the term “nocicipient” and unified attributes of intensity theory and specificity theory.
+Sherrington further developed the specificity theory by coining the term “nocicipient” and unified attributes of intensity theory and specificity theory \cite{Casey1982}.
 
@@
 The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes all the necessary information for pain.
-That additional processing of information in the periphearla nd central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities.
-This entire pattern of neural activities is the pain sensory experience.
-Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn ( laminae V ), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferent, and relay them to the thalamus, and subsequently the primary sensory cortex.
-A central concept, which was also introduced by Melzacks, is the pain neuromatrix.
-Which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}.
+That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities.
+This entire pattern of neural activities is the  sensory pain experience.
+Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn ( laminae V ), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex.
+A central concept, which was also introduced by Melzacks, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}.
 (See section \ref{section:neuromatrix})
@@
 \paragraph{Default mode network}
-The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Grecius2003,Raichle2015}.
+The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}.
 The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation.
@@
 Changes in peripheral neurotropic factors as a result of an injury can cause increased membrane excitability or spurious sprouting of the synaptic terminals in the dorsal horn of the spinal cord.
-Additionally, maladaptive changes may occur in the secondary sensory neurons and interneurons in the dorsal horn laminae that amplifies or reroutes the noxious signals.
+Additionally, maladaptive changes may occur in the secondary sensory neurons and interneurons in the dorsal-horn laminae that amplifies or reroutes the noxious signals \cite{Elmes2004}.
 
@@
 It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures.
-This may give rise or contribute to neuropathic pain (ref).
+This may give rise or contribute to neuropathic pain \cite{Geha2008}.
 
- The utilization of magnetic resonance imaging (MRI), coupled with novel imaging modalities such as functional MRI (fMRI) and diffusion MRI (dMRI), have been instrumental in our understanding of pain, permitting both \textit{in vivo} measures of brain activity during pain (functional studies), as well as assessment of changes to the brain microstructure.
+ The utilization of magnetic resonance imaging (MRI), coupled with novel imaging modalities such as functional MRI (fMRI) and diffusion MRI (dMRI), have been instrumental in our understanding of pain, permitting both \textit{in vivo} measures of brain activity during pain (functional studies) \cite{Davis2006}, as well as assessment of changes to the brain microstructure \cite{Hodaie2009a,Chen2015c}.
 

### 2017-12-02 thesis/contents/reviews/trigeminal.tex
checkpoint on edit

@@
 
-Neuralgias by definition are pain without any detectable physical origin, and originate from the brain.
-Trigeminal neuralgia (TN), or Tic douloureux, is a type of facial neuropathic pain that involves the trigeminal nerve.
+Trigeminal neuralgia (TN) is a chronic neuropathic facial pain disorder, that commonly affects one or more branches of the trigeminal nerve.
 TN symptoms involve paroxysmal, shock-like pain condition affecting one or more of the three facial sensory nerve branches.
-The pain episodes can be unpredictable, as the length of episode can range from 1 to 1462 days \cite{Katusic1990}.
+The pain episodes can be unpredictable, as the length of episodes can range from months to years \cite{Katusic1990}.
 Patients often describe its symptom as the most painful experience a human can undergo.
-Its physical and emotional disturbances are often severely disabling.
+Its physical and emotional disturbances can be severely disabling.
 
-The disorder has been documented as early as the first century AD, the clinical recognition of the disorder was by Nicolaus Andre in 1756, and by John Fothergill in 1776 \cite{Katusic1990}.
-In the three hundred years that followed, the aetiology of TN is still not completely understood.
-The modern understanding of the vascular compression of the trigeminal nerve root was first proposed by Janetta \cite{Jannetta1967}, where he elucidated that TN is a “nerve entrapment disorder” and that it may be a “consequence of aging … by arterialsclerotic elongation of the cerebrovasculature”.
+This disorder has been documented as early as the first century AD, the clinical recognition of the disorder was by Nicolaus Andre in 1756, and by John Fothergill in 1776 \cite{Katusic1990}.
+In the three hundred years that followed, the etiology of TN is still not understood.
+The modern understanding of the vascular compression of the trigeminal nerve root was first proposed by Janetta \cite{Jannetta1967}, where he elucidated that TN is a “nerve entrapment disorder” and that it may be a “consequence of ageing … by arterialsclerotic elongation of the cerebrovasculature”.
 
-Modern pain literature continuous to update the exactly diagnostic definition of TN.
-
-An older classification scheme divides facial pain into 5 major subtypes \cite{Burchiel2003,Eller2005}: idiopathic (without apparent cause), trigeminal injury, symptomatic, postherpetic, and atypical facial pain.
-Idiopathic TN has 2 sub-types: 1) TN type 1, where the patient experiences greater episodic pain greater than 50\% of the time; 2) TN type 2, where constant pain occurs greater than 50\% of the time.
+Modern pain literature continues to update the exactly diagnostic definition of TN.
+The Burchiel classification scheme divides facial pain into 5 major subtypes \cite{Burchiel2003,Eller2005}: idiopathic (without apparent cause), trigeminal injury, symptomatic, postherpetic, and atypical facial pain.
+Idiopathic TN has 2 sub-types: 1) TN type 1, also referred to as classic TN, where the patient experiences greater episodic pain greater than 50\% of the time; 2) TN type 2, where constant pain occurs greater than 50\% of the time.
 Trigeminal injury can be unintentional: which can be due to facial trauma, surgery to the oral cavity; this type of pain would involve persistent and burning pain to the affected areas; and is termed trigeminal neuropathic pain.
@@
 Symptomatic TN, also termed MS-TN, is due to trigeminal nerve injury relating to multiple sclerosis related central white matter inflammation.
-Postherpetic TN is pain resulting from the infection of the herpes zoster virus, and can involving burning pain and deafferentation.
-Atypical facial pain is classified as somatoform pain disorder that can be of psychosomatic origin, although some literature can also include constant pain in the face into this classification.
+Postherpetic TN is pain resulting from the infection of the herpes zoster virus, and can involve burning pain and deafferentation.
+In this classification, the definition of atypical facial pain is reserved for somatoform pain disorder that can be of psychosomatic origin, although some literature can also include constant pain in the face into this classification.
+This is in contrast with the common clinical definition, where atypical facial pain is used for the definition of trigeminal pain whose symptomatology deviates from classic TN.
 
 The current The International Classification of Headache Disorders ICHD-3 Beta \cite{Society2013} classifies TN into classic TN, and painful trigeminal neuropathy. 
-Classic TN's diagnostic criteria requires pain characteristics of: at least 3 attacks of unilateral facial pain; occuring in one or more trigeminal branches, with no radiation beyond the trigeminal distribution; severe intensity; electric shock-like, shooting, stabbing or sharp in quality; and precipatated by innocuous stimulation to the affected side of the face; and no clinically evident neurological deficit.
-Classic TN is further divided into purely paroxymal: recurrent attacks of unilateral facial pain with no persistent facial pain between attacks; and concomitant persistent: persistent facial pain of moderate intensity in the affected area.
-Painful trigeminal neuropathy in contrast are facial pain in one or more distributions of the trigeminal nerve caused by another disorder and indictive of neural damage.
+Classic TN's diagnostic criteria requires pain characteristics of: at least 3 attacks of unilateral facial pain; occurring in one or more trigeminal branches, with no radiation beyond the trigeminal distribution; severe intensity; electric shock-like, shooting, stabbing or sharp in quality; and precipitated by innocuous stimulation to the affected side of the face; and no clinically evident neurological deficit.
+Classic TN is further divided into purely paroxysmal: recurrent attacks of unilateral facial pain with no persistent facial pain between attacks; and concomitant persistent: persistent facial pain of moderate intensity in the affected area.
+Painful trigeminal neuropathy, in contrast, is facial pain in one or more distributions of the trigeminal nerve caused by another disorder and indicative of neural damage.
 It can be further divided into categories relating to acute Herpes zoster, post-herpetic trigeminal neuropathy, multiple-sclerosis, space-occupying lesion, and other disorders.
 
-There is also an definition \cite{Zakrzewska2016} that attempts to address the lack of symptomatic TN classification in ICHD-3 by classifying TN into idiopathic TN: without apparent cause; classic TN: caused by vascular compression of the trigeminal nerve root; and secondary TN: as the consequence of a major neurological disease such as MS.
+There is also a definition \cite{Zakrzewska2016} that attempts to address the lack of symptomatic TN classification in ICHD-3 by classifying TN into idiopathic TN: without apparent cause; classic TN: caused by vascular compression of the trigeminal nerve root; and secondary TN: as the consequence of a major neurological disease such as MS.
 
-\subsection{Anatomy}
+\subsection{Trigeminal Anatomy}
 
@@
 
-The V1 branch provides the sensory innervation to the superior aspects of the face and cranium, including forehead, upper eyelids, eyeball, frontal sinus, lacrimal gland, and dura of the anteior cranial fossa.
+The V1 branch provides the sensory innervation to the superior aspects of the face and cranium, including forehead, upper eyelids, eyeball, frontal sinus, lacrimal gland, and dura of the anterior cranial fossa.
 It further divides into lacrimal, frontal, and nasociliary nerve branches.
-The lacrimal branch supplies the lacrimal; the frontal contains smaller branches that innervates the skin of the medial forehead, upper eyelid, and parts of the frontal sinus; the nasociliary branch innervate various parts of the eye, and mucous membrane of the anterior nasal septum, anterior portion of the lateral nasal wall, and the skin on the tip of the nose.
+The lacrimal branch supplies the lacrimal; the frontal contains smaller branches that innervate the skin of the medial forehead, upper eyelid, and parts of the frontal sinus; the nasociliary branch innervate various parts of the eye, and mucous membrane of the anterior nasal septum, anterior portion of the lateral nasal wall, and the skin on the tip of the nose.
 V1 enters the skull base through the superior orbital fissure.
 
-the V2 provides innervation to the temple, cheek, sides of the nose, and lower eyelids via the zygomaticofacial and zygomaticotemporal branches.
+The V2 provides innervation to the temple, cheek, lateral aspects of the nose, and lower eyelids via the zygomaticofacial and zygomaticotemporal branches.
 It also innervates upper oral cavity.
-It enters through the foramen rotundum.
+It exists the skull through the foramen rotundum.
 
@@
 The tensor tympani also receives some CN V innervation, which is thought to be important in dampening the sound of chewing.
-V3 enters through foramen ovale.
+V3 exists the skull through foramen ovale.
 
@@
 It is analogous to the spinal dorsal root ganglion, and contains all the trigeminal sensory cell bodies.
-It situates in the meckel's cave, on the petrous bone of the middle cranial fossa.
-It is somatotomically organized in respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.
+It situates in the Meckel's cave, on the petrous bone of the middle cranial fossa.
+It is somatotopically organized in respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.
 
@@
 The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons.
-The CNS myelin can be several millimetres beyond the surface of the pons, before it give away to the transition area, termed Obersteiner-Redlich zone \cite{Peker2006}.
+The CNS myelin can be several millimetres beyond the surface of the pons, before it gives away to the transition area, termed root entry zone, or less commonly, Obersteiner-Redlich zone \cite{Peker2006}.
+Study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm).
+The transition zone can be up to 25\% of the nerve length \cite{Peker2006}.
 
@@
 
-The CN V passes into the pons (\ref{fig:trig-pathway}), through the pontine cerebellar peduncular bundles and enters the tegmentum of the pontine brainstem.
-CN V then curve towards the medial line before synapses onto the trigeminal nucleus in the basis of the pontine brainstem.
+CN V passes into the pons (\ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem.
+It then curve towards the medial line before synapses onto the trigeminal nucleus in the basis of the pontine brainstem.
 The trigeminal nucleus is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei.
@@
 The mesencephalic nucleus is responsible for proprioception, while the spinal trigeminal nucleus receives ipsilateral pain and temperature afferents not only from CN V, but also from CN VII, CN IX and CN X.
-Pars oralis transmits fine tactile sensations from the oralfacial region, pars interpolaris is also associated with dental pain, while pars caudalis is responsible for nociception and heat from the head.
+Pars oralis transmits fine tactile sensations from the oralfacial region, pars interpolaris is associated with dental pain, while pars caudalis is responsible for nociception and heat from the head.
 
 The main sensory nucleus receives discriminative somatosensation from the oralfacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract.
-A separate branch responsible for oral cavity however project to the ipsilateral VC via the dorsal trigeminothalamic tract.
+A separate branch responsible for oral cavity, however, projects to the ipsilateral VC via the dorsal trigeminothalamic tract.
 
@@
 Classic TN is believed to be caused by neurovascular compression (NVC) of the CN V at its root entry zone (REZ) by the surrounding vasculature, where the nerve interfaces with the pons.
-An aberrant loop of artery and less commonly a vein is seen in 80-90\% of the cases of TN to compressing against the nerve \cite{Love2001,McLaughlin1999}.
-This can be observed in conventional MR imaging \cite{Borges2010}.
-Generally, the superior cerebellar artery is responsible for most of the NVC cases (60-90\%), and then the anteroinferior cerebellar artery and basilar arteries (Lutz et al., 2011).
-There is also evidence that TN is preferentially lateralized to the right side in 61\% of the cases, and in very rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.
-Other types of compressions may also cause TN, these can be in the forms of saccular aneurysm, arteriovenous malformation, vestibular schwannomas, meningiomas, epidermoid cysts, and other tumors.
-However direct pathophysiological causal of TN pain is confounded by the fact observation of vascular compression does not always result in the expression of pain \cite{Desouza2013,Hodaie2013}.
-In some cases the TN pain is reported to be contralateral to the lesion \cite{Cheng2008,Haddad1990,Revuelta1995}.
-There are also 3.1-17\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006}.
-
-Outside the brainstem, CN V is peripherally myelinated by Schwann cells, which are capable of remyelination.
-Inside the brainstem, the nerve is myelinated by oligodendrocytes.
-Study in 100 CN V samples from 50 cadavers \cite{Peker2006} showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm).
-The transition zone can be up to 25\% of the nerve length.
+An aberrant loop of artery and less commonly a vein is seen in 80-90\% of the cases of TN compressing the nerve \cite{Love2001,McLaughlin1999}.
+This can also be observed in conventional MR imaging \cite{Borges2010}.
+The superior cerebellar artery is responsible for most of the NVC (60-90\%), followed by the anteroinferior cerebellar artery and basilar arteries (Lutz et al., 2011).
+There is a slight lateral prevalence, with right-sided TN occurring in  in 60\% of the cases, and in rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.
+Other types of compressions may also cause TN; these can be in the forms of saccular aneurysm, arteriovenous malformation, vestibular schwannomas, meningiomas, epidermoid cysts, and other tumours.
+However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the expression of pain \cite{Desouza2013,Hodaie2013}.
+In some cases the TN pain is reported to be contralateral to the NVC \cite{Cheng2008,Haddad1990,Revuelta1995}.
+There are also 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.
 
@@
 Ultrastructural examinations \cite{Love1998} of tissue samples obtained from CN V rhizotomy specimens showed oligodendrocyte demyelination at the site of root transition zone.
-In TN, adjacent to the site of compression-related focal demyelination, were cases of thinly myelinated sheaths that encapsulates a number of demyelinated axons, suggesting aberrant remyelination.
+Adjacent to the site of compression-related focal demyelination, thinly myelinated sheaths that encapsulate a number of demyelinated axons could be observed, suggesting aberrant remyelination.
 The TRG is also postulated to be responsible for neuralgia \cite{Rappaport1994}.
-It's believed that ganglionic ectopic autonomic firing can be trigged by mechanical disturbances as a result of nerve compression.
+It's believed that ganglionic ectopic autonomic firing can be triggered by mechanical disturbances as a result of nerve compression.
 Additionally, ganglionic cross aftercharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding ganglionic neurons.
-Histology findings by Devor et al supported the ignition hypothesis \cite{Devor2002a}, where demyelinated axons were shown to be in close apposition without intervening glia.
+Histology findings by Devor et al.
+supported the ignition hypothesis \cite{Devor2002a}, where demyelinated axons were shown to be in close apposition without intervening glia.
 
@@
 
-For TN, anticonvulsant medication such as carbamazepine has 80\% response rate.
-However 50\% of patients will eventually need to undergo surgery as medication becomes ineffective due to tolerances.
-Nerve blocks such as peripheral neurectomy, neurolysis of the nerve or TRG can provide relief, albeit temporary, as pain always reoccurs \cite{Rappaport1994}.
-Long term treatment of TN involves the decompression of affected CN V \cite{Lovely1997}.
-This is achieved by either direct surgical intervention, in the form of microvascular decompression (MVD), where a Teflon\textcopyright padding is placed between the CN V root and the offending vessel.
-MVD showed 90\% pain relief immediately after surgery \cite{Zakrzewska2005}, and a long term success rate of 70\% at 10 years \cite{Barker1996}.
+The initial mode of treatment of TN consists in pharmacotherapy, primarily using anticonvulsant medication such as carbamazepine.
+While there is an initial, very good response to medical treatment, eventually nearly 50\% of patients will fail medical therapy, either due to side effects, or requirements of a progressive increase in dose \cite{Rappaport1994}.
+Multiple surgical treatments are available, from more central procedures addressing NVC to peripheral procedures such as neurectomy.
+The procedure that provides the greatest likelihood of pain relief is microvascular decompression (MVD) \cite{Lovely1997}, where the nerve is decompressed from the offending vascular structure through the placement of a piece of Teflon\textcopyright gauze.
+This is thought to dampen the pulsations of the artery on the nerve.
+MVD showed 90\% pain relief immediately after surgery \cite{Zakrzewska2005}, and a long-term success rate of 70\% at 10 years \cite{Barker1996}.
 Partial sensory rhizotomy (PSR) can also be performed, where the offending CN V branch is severed, with a response rate of 88-70\% \cite{Young1993,Zakrzewska2005}.
 
-Intervation can also be non-invasive, in the form of GammaKnife radiosurgery (GKRS) \cite{Hodaie2012g}.
-The GK system focuses gamma radiation from approximately 200 cobalt-60 ($^{60}$Co) sources onto specific targets along the trigeminal nerve.
-Earlier targets focused on the trigeminal ganglion on adjacent areas \cite{Leksell1983}, while more proximal targets towards the brainstem is now recently used\cite{Daugherty2015}.
+Intervention can also be non-invasive, in the form of Gamma Knife radiosurgery (GKRS) \cite{Hodaie2012g}.
+The Perfexion GK system focuses gamma radiation from 192 cobalt-60 ($^{60}$Co) sources onto specific targets along the trigeminal nerve.
+The target is either on the cisternal segment of CN V or just outside of the REZ \cite{Daugherty2015}.
 Proximal targets correlate with longer pain relief, but also higher chance of facial numbness\cite{Xu2014}.
@@
 The effect is delayed and pain relief can be experienced in 4-6 weeks post-radiosurgery.
-GKRS treatments result in 90\% initial pain relief \cite{Regis2015,Kondziolka2010}, about 50-80\% pain free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}.
-More recent studies suggested higher pain free probabilities at 3(71.8\%), 5(64.9\%), 7(59.7\%) and 10(45.3\%) years\cite{Regis2015}.
-More favorable treatment outcomes were related to no other accompanying symptoms, no prior surgery and shorter term pain history of less than three years.
+GKRS treatments result in 90\% initial pain relief \cite{Tuleasca2015,Kondziolka2010}, about 50-80\% pain-free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}.
+More recent studies suggested higher pain-free probabilities at 3(71.8\%), 5(64.9\%), 7(59.7\%) and 10(45.3\%) years\cite{Tuleasca2015}.
+More favourable treatment outcomes were related to no other accompanying symptoms, no prior surgery and shorter term pain history of fewer than three years.

### 2017-12-02 thesis/contents/sagit.tex
checkpoint on edit

@@
 The approach is to model the HARDI signal as a convolution of the fiber orientations.
-The resulting deconvoluted fiber orientation distribution (FOD) shows better discrimination of fiber directions compared to Q-ball under similar scanning parameters, but is susceptible to image noise\cite{Anderson2005,Dell'Acqua2007,Tournier2004}.
+The resulting deconvoluted fiber orientation distribution (FOD) shows better discrimination of fiber directions compared to Q-ball under similar scanning parameters, but is susceptible to image noise\cite{Anderson2005,DellAcqua2007,Tournier2004}.
 Alternatively, multi-tensor model-based approaches estimate the fiber configuration with the assumption that there are no more than two or three crossings in a given voxel.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\centering
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering

### 2017-12-02 thesis/contents/tn_sagit.tex
checkpoint on edit

@@
 \textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial pain syndrome that involves the trigeminal sensory pathway.
-We demonstrate the use of an fully-automated group white-matter tractography, and Gaussian Process (GP) classification to pin-point key white-matter diffusivity changes that maximally differentiates patients and healthy controls.
+We demonstrate the use of a fully-automated group white-matter tractography, and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate patients and healthy controls.
 
 \textbf{Methods:} We perform group merged tractography using SAGIT on 36 sex-matched TN patients (right-sided pain) and 36 controls.
-The trigeminal nerve (CNV), pontine decussation, and thalamocortico S1 pathways were reconstructed.
-GP classifiers were trained by scrolling a moving window over CNV and S1 tractography centroids.
+The trigeminal nerve (CN V), pontine decussation, and thalamocortical S1 pathways were reconstructed.
+GP classifiers were trained by scrolling a moving window over CN V and S1 tractography centroids.
 FA, GFA, RD, AD, and MD metrics were evaluated.
@@
 
-\textbf{Results:} Both affected and unaffected CNV can be differentiated with up to 80\% accuracy from controls.
-Affected CNV differences were more focused in the cistern and root-transition-zone, whereas unaffected CNV show more focus towards the brainstem.
-The affected S1 showed focused MD differences towards the cortex with accuracy of 85\%.
+\textbf{Results:} Both affected and unaffected CN V can be differentiated with up to 80\% accuracy from controls.
+Affected CN V differences were more focused in the cistern and root-transition-zone, whereas unaffected CN V show more focus towards the brainstem.
+The affected S1 showed focused MD differences towards the cortex with an accuracy of 85\%.
 
@@
 \subsubsection{Trigeminal Neuralgia}
-Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxymal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CNV) branches.
-TN manifests classically in idiopathic fashion, or secondary to a range of neuropathology, including meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticerosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
-The pathophysiology of idiopathic TN is generally accepted to be vascular compression at the root of the CNV nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidences of vascular compression \cite{Lee2014}.
-We previously compared TN and MS-TN diffusivity differences in four segments of the CNV at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CNV diffusivity is disrupted in MS-TN.
-In all previous TN studies, placements of region-of-interests in the CNV were pre-determined and manually placed.
-In this study we further expand the study to A) analyze the CNV differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interests; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.
+Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CN V) branches.
+TN manifests classically in idiopathic fashion, or secondary to a range of neuropathology, including meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
+The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}.
+We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is disrupted in MS-TN.
+In all previous TN studies, placements of region-of-interests in the CN V were pre-determined and manually placed.
+In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.
 
@@
 The discriminative pathway synapsis onto the primary trigeminal sensory nucleus, and then project to the contralateral VC thalamus, and then to the S1 cortex as we have delineated.
-The pain pathway synapsis caudally onto the trigeminal spinal nucleus, and decussate at the level of the medulla and spinal cord; The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus; the pain pathway also projects to the medial nucleus of thalamus, where it then project medially to the cingulate cortex.
+The pain pathway synapsis caudally onto the trigeminal spinal nucleus, and decussate at the level of the medulla and spinal cord.
+The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus.
+The pain pathway also projects to the medial nucleus of the thalamus, where it then project medially to the cingulate cortex.
 
@@
 Neuroimaging tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}.
-To obtain diffusivity metrics from tractography generally involves converting the streamline bundles into volumetric spatial mask, and then measure the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}.
+To obtain diffusivity metrics from tractography involves converting the streamline bundles into a volumetric spatial mask, and then measure the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}.
 However this approach discards the orientation information in the streamlines.
 The naive form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-through.
-This becomes an issue when considering that an average 3 tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm.
-In white matter structures such as CNV, and other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
+This becomes an issue when considering that an average 3-tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm.
+In white matter structures such as CN V, and other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
 
-Consider that a tractography streamline propagation algorithm can involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
+Consider that a tractography streamline propagation algorithm involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
 It is reasonable to consider these algorithms to be an interpolated subsampling scheme of the DWI volume that takes the diffusion-based orientation into account.
@@
 Along-the-tract analysis has been investigated by various authors \cite{Colby2012,ODonnell2009,Wang2015,Yeatman2012}.
-In these approaches, the bundles are selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and studies were limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and ucinate.
-The measures were pooled onto tract centroid in each individual subjects, and then aggregated at group level.
-The global tractography approach is unintuitive for certain anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is clear.
+In these approaches, the bundles are selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and studies were limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and uncinate.
+The measures were pooled onto tract centroid in each individual subjects, and then aggregated at the group level.
+The global tractography approach is unintuitive for anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is clear.
 A more focused ROI based delineation approach is therefore preferred in these circumstances.
 
-There has been a number of centroid identification methods that were investigated.
-It’s important to distinguish the difference between streamline clustering and centroid identification.
-While a number of methods cluster the streamlines by identifying a polyline centroid, there are also methods that identify possible streamline clusters, but does not provide a best-representative centroid.
-Centroid identification involved resampling the streamlines into identical number of points, and then identifying point correspondences via different distance measures.
-For these approaches it’s important that the streamlines have similar length, and an assumption that the start and end of the streamlines across subjects are consistent.
-These studies focus on healthy subjects, but when consider that pathological brain tissue may negatively affect tractography delineation, then the geometry assumptions of these methods may not be satisfied when extending them to study patient populations.
-A number of these methods have running time and memory constraints that makes them infeasible to be deployed at a group level.
-One method for tract clustering and centroid identification that is shown to have linear running time in respect with streamline count is QuickBundles \cite{Garyfallidis2012}.
-We have adopted QuickBundles in order to obtain S1 streamline centroids.
+It is important to note that streamline clustering and centroid identification can be considered as independent tasks.
+Some clustering methods do not provide a best-representative streamline centroid.
+Centroid identification involved resampling the streamlines into an identical number of points, and then identifying point correspondences via different distance measures.
+For these approaches, it is vital that the streamlines have similar length, and an assumption that the start and end of the streamlines across subjects are consistent.
+However, pathological brain tissue may negatively affect tractography delineation.
+Therefore the geometry assumptions of such approaches may not be satisfied when extending them to study patient populations.
+Running time and memory constraints also need to be taken into consideration, when applying such algorithms at a group level.
+One method for tract clustering and centroid identification that is shown to have linear running time in respect to streamline count is QuickBundles \cite{Garyfallidis2012}.
+We have adopted QuickBundles to obtain S1 streamline centroids.
 
-The key advantage of a streamline-based approach is that we can retain the pathway orientation from its geometry.
+The principal advantage of a streamline-based approach is that we can retain the pathway orientation from its geometry.
 White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis.
@@
 \subsubsection{Gaussian Process}
-The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} has sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}.
-However one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models.
+The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}.
+However, one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models.
 Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}.
-In fact it has been found that DNN with infinite number of hidden neurons approximates GP \cite{neal1996priors}.
+In fact, it has been found that DNN with an infinite number of hidden units approximates GP \cite{neal1996priors}.
 
@@
 The radial-basis function kernel or RBF kernel is widely used in machine learning tasks.
-GP have previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010}.
+GP has previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010}.
 It is particularly well suited for continuous data streams, such as tractography streamlines.
@@
 T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256.
-DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.
+DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.
 
@@
 \subsubsection{Tractography}
-ROIs were defined on the template image to bilaterally delineate: 1) CNV (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).
+ROIs were defined on the template image to bilaterally delineate: 1) CN V (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).
 
 Constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} was used to delineate the three sets of anatomy.
-Deterministic CST was applied to delineate CNV and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy.
+Deterministic CST was applied to delineate CN V and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy.
 Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration.
@@
 The merged tract was first post-processed in 3D Slicer to remove erroneous and stray streamlines that will confound measurement.
-Streamlines removed include: remnant CNV streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels/noise; S1 streamlines that stray medially towards other regions such as the corpus callosum.
+Streamlines removed include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise.
+S1 streamlines that stray medially towards other regions such as the corpus callosum were also removed.
 The TPT streamlines were not examined due to inconsistent reproducibility across subjects.
@@
 Tractography centroids, which were the average curve best fit the tract bundle, was defined on the merged tracts in template space.
-CNV centroids were manually defined.
-The CNV centroids were defined by a list of 3 control points situated in the template world space.
+CN V centroids were manually defined.
+The CN V centroids were defined by a list of 3 control points situated in the template world space.
 A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm.
-The spline is then resampled into 30 points that becomes the centroid definition with segment length of 1.3 mm.
+The spline is then resampled into 30 points that become the centroid definition with segment length of 1.3 mm.
 Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids.
 Diffusion metrics from the points were aggregated for each of the cross-sections.
-S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threhold of 30, with centroid length of 62.1 mm and segment length of 2.07 mm.
+S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, with centroid length of 62.1 mm and segment length of 2.07 mm.
 
@@
 
-Diffusivity metrics of affected (right side) and unaffected CNV and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides.
+Diffusivity metrics of affected (right side) and unaffected CN V and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides.
 The affected and unaffected sides in TN patients were also compared with each other.
@@
 The diffusion values were subsequently normalized to zero mean with ranges $[-1, 1]$ for model training.
-Gaussian Process (GP) classification was exhaustively performed on CNV and S1 nerve segments in order to identify key segment differences.
+Gaussian Process (GP) classification was exhaustively performed on CN V and S1 nerve segments to identify key segment differences.
 Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment.
@@
 \subsubsection{Group Tractography}
-Merged CNV delineations show anatomically consistency, and extended into the ipsilateral TGN.
+Merged CN V delineations show anatomical consistency, and extended into the ipsilateral TGN.
 Changes in FA along the streamlines demonstrate clear differences between cistern and brainstem segments, suggesting good agreement in anatomical registration ( Figure \ref{fig:GPfigure1}).
-The TPT projections, when filtered strictly from the TGN to the ipsilateral VC thalamus region, showed distinct bundles of decussating pathways bewteen left and right.
-The resulting pathway are not symmetrical.
-Where the right TPT pathway diverts and envelopes the left (Figure \ref{fig:GPfigure2}).
+The TPT projections, when filtered strictly from the TGN to the ipsilateral VC thalamus region, showed distinct bundles of decussating pathways between left and right.
+The resulting pathway was not symmetrical, where the right TPT pathway diverts and envelopes the left (Figure \ref{fig:GPfigure2}).
 TPT streams lines were sparse, suggest low inter-subject reproducibility.
-Therefore no additional analysis were carried out for TPT.
+Therefore no additional analysis was carried out for TPT.
 S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}).
@@
 \centering
-\includegraphics[width=\linewidth]{cnv-inferior-view.png}
-\caption{Inferior view of the merged CNV delineations.}
+\includegraphics[width=\linewidth]{CN V-inferior-view.png}
+\caption{Inferior view of the merged CN V delineations.}
 \label{fig:GPfigure1}
@@
 
-\subsubsection{CNV}
+\subsubsection{CN V}
 Both sides were distinguished from controls with classifiers that achieve greater than 75\% accuracy. 
-Classifiers on the right side has performance ranging from 77\% (GFA) to 80\% (MD) accuracy.
-The left classifiers achieved accuracies ranging from 75\% (FA) to 80\%(GFA).
+Classifiers on the right side have performance ranging from 77\% (GFA) to 80\% (MD) accuracy.
+The left classifiers achieved accuracy ranging from 75\% (FA) to 80\%(GFA).
 The right side shows more consistent overlap of segment ranges between positions 10--17, in the region of the cistern/REZ; while the left side shows less consistent range overlaps, and AD segment range from positions 3 to 14, which include the brainstem regions.
-Overall, the right side features extend further into the distal space, into the cistern and even the ganglion, while the left side extend more into the brainstem.
+Overall, the right side features extend further into the distal space, into the cistern and even the ganglion, while the left side extends more into the brainstem.
 Comparing left/right, only the FA classifier achieved greater than 75\% accuracy, where its segments range from 11 to 19, which is the cisternal segment (Figure \ref{fig:GPfigure4}).
 
-The best performing segments versus controls is GFA (segment 8--14) on the left (unaffected) side, with 80\% accuracy, 84\% precision, and AUC of 0.8; while MD (segment 10--17) performed the best on the right (affected) side, with 80\% accuracy, 84\% precision, and AUC of 0.79 (Table \ref{table:cnv}, Figure \ref{fig:GPfigure6}).
+The best performing segments versus controls is GFA (segment 8--14) on the left (unaffected) side, with 80\% accuracy, 84\% precision, and AUC of 0.8; while MD (segment 10--17) performed the best on the right (affected) side, with 80\% accuracy, 84\% precision, and AUC of 0.79 (Table \ref{table:CN V}, Figure \ref{fig:GPfigure6}).
 The best performing left-versus-right classifier was FA (segment 11--19), with 75\% accuracy, 82\% precision, and AUC of 0.75.
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
-\includegraphics[width=\linewidth]{figure-GP-CNV.pdf}
-\caption{CNV classifier performance}
+\includegraphics[width=\linewidth]{figure-GP-CN V.pdf}
+\caption{CN V classifier performance}
 \label{fig:GPfigure4}
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \centering
-\csvautotabular{cnv_table.csv}
-\caption{CNV GP classifier performance data}
-\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
-\label{table:cnv}
+\csvautotabular{CN V_table.csv}
+\caption{CN V GP classifier performance data}
+\caption*{List of the best accuracy for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
+\label{table:CN V}
 \end{table}
@@
 Both sides of the S1 face region pathway were able to be distinguished by classifiers with greater than 75\% accuracy. 
-The FA and RD classfiers at locations 617, representing the segment that emerge from the thalamus and merging into corona radiata, are common to both sides.
+The FA and RD classifiers at locations 617, representing the segment that emerge from the thalamus and merging into corona radiata, are common to both sides.
 The left side has higher classification accuracy of 80\% with FA, while the right side had a maximum accuracy of about 78\% with RD.
-The MD classifier mean while is unique to the left side, and achieved an accuracy of 85\% in locations 16--23, near the grey-white matter boundary, closer to the cortex (Figure \ref{fig:GPfigure5}).
+The MD classifier meanwhile  is unique to the left side, and achieved an accuracy of 85\% in locations 16--23, near the grey-white matter boundary, closer to the cortex (Figure \ref{fig:GPfigure5}).
 
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
 In essence, we have demonstrated a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy solely from white-matter measures.
-We reproduced previous findings of TN diffusivity changes along the CNV with much higher regional specificity.
+We reproduced previous findings of TN diffusivity changes along the CN V with much higher regional specificity.
 We also examined TPT and S1 projections of the trigeminal sensory pathway.
 
-The diffusivity measures of CNV clearly reproduced our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls.
-The consistent regional overlaps in the cistern/REZ region of the right (affected) CNV in MD, RD, AD shows the ability for the GP classifiers to auto-converge to focal white-matter abnormality.
+The diffusivity measures of CN V reproduced our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls.
+The consistent regional overlaps in the cistern/REZ region of the right (affected) CN V in MD, RD, AD shows the ability for the GP classifiers to auto-converge to focal white-matter abnormality.
 The left versus right GP classifier placed left-right differences at the cistern segment, in agreement with the pathophysiology of TN.
-The regional findings on left verus controls suggest that there is bilateral differences in CNV myeline in unilateral TN, despite the lack of symptoms on the unaffected side.
+The regional findings on left versus controls suggest that there is bilateral differences in CN V myeline in unilateral TN, despite the lack of symptoms on the unaffected side.
 The left regions also situate into the brainstem segments.
-It is possible that the left CNV is affected by global descending pain modulation, and warrant further investigation.
+It is possible that the left CN V is affected by global descending pain modulation, and warrant further investigation.
 
@@
 Previous studies have suggested that S1 have key involvement in pain location discrimination \cite{bushnell1999pain}, and are found to have grey-matter abnormalities in TN pain \cite{Desouza2013c}.
-Our finding that the left (affected) thalamo-cortical S1 projection is uniquely different close to the cortex suggest that deep cortical white-matter/grey-matter changes may be related depending on the locality.
+Our finding that the left (affected) thalamocortical S1 projection is uniquely different close to the cortex suggest that deep cortical white-matter/grey-matter changes may be related depending on the locality.
 
 The TPT delineations showed that decussation pathway tractography is possibly asymmetric.
-It’s not clear if the asymmetry is a neuroanatomically factual, or an artifact from either algorithmic bias or partial volume bias of the underlying DWI.
+It is not clear if the asymmetry is a neuroanatomically factual, or an artifact from either algorithmic bias or partial volume bias of the underlying DWI.
 Resolving crossing fiber is still a difficult task in tractography, and while there are validations against MR phantoms, the anatomical validity of these methods can only be extrapolated, and the interpretation must involve anatomical priors from experts.
 
-We included GFA in order to investigate its potential for measuring deep white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric.
-However S1 results demonstrated that it is not effective in pin-pointing deep white-matter abnormalities.
-GFA performed well in CNV, but not to any substantially degree than the Gaussian metrics.
+We included GFA to investigate its potential for measuring deep white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric.
+However S1 results demonstrated that it is not effective in pinpoint deep white-matter abnormalities.
+GFA performed well in CN V, but not to any substantially degree than the Gaussian-base diffusivity metrics.
 
-Overall the GP classifiers resulted in higher precision when compare to accuracy.
+Overall the GP classifiers resulted in higher precision when compared to accuracy.
 The overlapping regions revealed differences in localizations that may very well have been missed with singular diffusivity metric analysis.
-This suggest that cross-metric features should be considered in future studies.
+This suggests that cross-metric features should be considered in future studies.
 This would drastically increase the feature complexity, and thereby require subsequent demand in the number of subjects.
@@
 \subsubsection{Limitations}
-The merged CNV delineations of all the subjects showed caudal projections towards the spinal cord.
+The merged CN V delineations of all the subjects showed caudal projections towards the spinal cord.
 In multiple instances, diverging streamlines at the levels of the medulla strongly suggested the trigeminal spinal decussations.
-However the caudal projection was not available for majority of the TN group, and it was not a consistently producible pathway.
-This is likely due to that 1) 3T DWI scan has a limited 2mm isovoxel resolution; It is not enough to resolve CNV, medullary and spinal cord fiber details.
+However the caudal projection was not available for the majority of the TN group, and it was not a consistently producible pathway.
+This is likely due to that 1) 3T DWI scan has a limited 2mm isovoxel resolution; It is not enough to resolve CN V, medullary and spinal cord fiber details.
 We have modified our scan to 1x1x3mm voxel, to gain axial plane detail for this kind of small fiber tractography from clinical scans, at the sacrifice of coronal/sagittal resolution.
-2) Large percent of the TN DWI scans has severe distortions at the pontine level.
+2) A large percent of the TN DWI scans has severe distortions at the pontine level.
 Similar distortions rarely show up in Control or MS-TN scans.
@@
 
-The diffusivity metrics of the deeper white matter structures remains difficult to interpret, due to the heavy crossing fibers in these regions.
+The diffusivity metrics of the deeper white matter structures remains difficult to interpret, due to the complex crossing fibers in these regions.
 Although metrics such as FA are popular in current neuroimaging literature, they nevertheless are derived from the Gaussian tensor model, which cannot adequately describe cross-fiber regional dynamics.
-For example, in a healthy crossing fiber region, where both bundles have equal myelin and axonal integrity, will have a low FA score due to a spherical tensor; its AD and RD will be roughly equal, and therefore MD will be higher.
-In a pathological cross region, where only one bundle is healthy, the regional FA actually increase due to the bias in one direction.
-Its RD will be lower, and its MD will be lower as well.
+For example, in the crossing-fiber region of healthy tissue, where both bundles have equal myelin and axonal integrity, will have a low FA score due to a spherical tensor; its AD and RD will be roughly equal, and therefore MD will be higher.
+In a pathological cross region, where only one bundle is healthy, the FA increases due to the diffusivity bias in one direction; the RD will be lower, and MD will be lower as well.
 Moreover, in diverging and converging structures, as well as in multi-crossing regions such as the corona radiata, the interpretations of Gaussian-based diffusivity metrics become even more unclear.
@@
 We have achieved examination of specific and targeted anatomy at a group level involving both healthy and pathological brain images.
-The method reproduced previous findings of TN diffusivity changes along the CNV with finer-grained regional specificity.
+The method reproduced previous findings of TN diffusivity changes along the CN V with finer-grained regional specificity.
 We also examined TPT and S1 projections of the trigeminal sensory pathway.
-The study revealed that there are specific diffusivity changes on these deeper white matter structures, and demonstrated the power of along-the-tract analysis with machine-learning, which pin-points the exact location of the diffusivity disruption at millimetre level.
+The study revealed that there are specific diffusivity changes on these deeper white matter structures, and demonstrated the power of along-the-tract analysis with machine-learning, which pinpoints the exact location of the diffusivity disruption at millimetre level.

### 2018-01-09 thesis/contents/reviews/multiplesclerosis.tex
changes with moji

@@
 \section{Multiple Sclerosis}
-
-Multiple sclerosis (MS) is a chronic auto-immune demyelination disease of the central nervous system.
+This chapter will review multiple sclerosis (MS) as a disorder, given the importance of MS-related TN as an independent TN subtype.
+MS is a chronic autoimmune demyelination disease of the central nervous system.
 It is a progressive disease characterized by wide-spread lesion, or plaques, within the brain, spinal cord, and optic nerve.
 Its pathogenesis is related to the focal inflammation and damage of the central myelin, where the disruptions of the myelin sheath negatively affect signal transduction of the central white-matter, and thereby result in severe disability.
-MS symptoms can include cognitive deficits, fatigue, depression, upper and lower motor disabilities, and neuropathic pain.
+MS symptoms can include cognitive deficits, fatigue, depression, motor weakness affecting the upper and lower extremities and neuropathic pain.
 Currently there is no cure for MS.
-An estimated of up to 2,300,000 people world wide are affected by MS, and 50\% of the patient will require motor assistance 15 years after disease onset \cite{Goldenberg2012}.
+An estimated 2,300,000 people world wide are affected by MS, and 50\% of the patient will require motor assistance 15 years after disease onset \cite{Goldenberg2012}.
 The course of MS can be highly variable due to the distinct plaque distribution pattern in each individual, and therefor MS is difficult to diagnose and treat.
 
-\subsection{Classification}
-
-MS is divided into four major groups: 1) Relapsing-remitting (RRMS), 2) Secondary progressive (SPMS), 3) Primary-progressive (PPMS), and 4) Progressive-relapsing (PRMS).
+MS is divided into four major clinical groups: 1) Relapsing-remitting (RRMS), 2) Secondary progressive (SPMS), 3) Primary-progressive (PPMS), and 4) Progressive-relapsing (PRMS).
 RRMS is the most common form of MS, affecting about 85\% of the people with MS.
@@
 PRMS is the rarest form of MS, with less than 5\% of the MS population affected.
-It’s progressive in the beginning, with periods of relapse, it has no remission periods.
-Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different episodes must occur in the disease course (time dissemination); and C) There is chronic inflammation of the CNS.
+It is progressive in the beginning, with periods of relapse, it has no remission periods.
+
+Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different episodes must occur in the disease course (time dissemination); and C) Chronic inflammation of the CNS.
 One or more of these criteria result in a general diagnosis of MS.
-Both space and time dissemination criteria are confirmed with clinical MRI by radiologists.
+Both space and time dissemination criteria are confirmed with clinical MRI.
 Detailed disability ratings are confirmed by the Expanded Disability Status Scale (EDSS)\cite{Kurtzke1983}, which tracks the progression of MS symptoms in an individual.
 
-\subsection{Pathophysiology}
-
-MS is defined by the existence of focal demyelination in the CNS that manifest in lesions or plaques.
-Ex-vivo dissections and ultra-structural examinations of this pathological tissue, obtained from autopsies of individuals who suffered from progressive MS or acute episodes.
-In acute tissue samples, the lesions showed signs of acute demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells.
-Partial remyelination by oligodendrocytes were also present, as well as signs of axonal swelling and injury.
-In chronic tissue samples, there is no presence of acute demyelination, and heavy loss of axons and myelin could be observed \cite{Filippi2012}.
+MS is defined by the existence of focal demyelination in the CNS that manifests as lesions or plaques.
+Ultrastructural lesion analysis shows demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells.
+Partial remyelination by oligodendrocytes can be present, as well as signs of axonal swelling and injury.
+In tissue samples of longstanding MS there is no presence of acute demyelination, and heavy loss of axons and myelin could be observed \cite{Filippi2012}.
 
@@
 They are also found densely dispersed in paraventricular white matter.
-The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered when making a diagnosis.
+The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered for the diagnosis.
 The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}.

### 2018-01-09 thesis/contents/reviews/pain.tex
changes with moji

@@
 
-Trigeminal Neuralgia is a chronic facial pain syndrome.
-It is therefore imperative to review the evolving history and theories of pain.
+Trigeminal Neuralgia is a chronic neuropathic facial pain syndrome.
+The quest for the understanding of the pathophysiology of this syndrome requires a review of the history and theories of pain.
 Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species.
-In humans, the sensation of pain is generally described as having an unpleasant quality.
-Additionally, the experience of pain is complex, consisting of nociceptive, emotional and cognitive processes.
+In humans, the sensation of pain is typically described as an unpleasant quality.
+The experience of pain is complex, consisting of interaction between nociceptive, emotional and cognitive processes.
 The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.
@@
 Humanity have long sought to understand pain.
-In antiquity, thinkers in classical Greek traditions, including Plato and Aristotle considered pain as primarily an emotion instead of a physical senses\cite{Moayedi2012a}.
+In antiquity, thinkers in classical Greek traditions including Plato and Aristotle considered pain as primarily an emotion \cite{Moayedi2012a}.
 Hippocrates and Galen believed that pain arose from imbalances in vital fluids of the body, and considered the heart to be the organ responsible for the experience of pain \cite{Melzack1999}.
+These broad historical interpretations contrast more modern physiological approaches to the study of pain \cite{Melzack1999,Moayedi2012a}.
 
-The earliest theory that attribute pain to a dedicated physical pathway is commonly credited to Ren\'{e} Descartes (\ref{fig:descarte}).
-However, for Descartes, pain as an experience still could not be divorced from the idea of the soul, as he had stated in his letter to the French polymath Marin Mersenne: "I do not explain the feeling of pain without reference to the soul.
+The earliest theory that attributes pain to a dedicated physical pathway is commonly credited to Ren\'{e} Descartes (\ref{fig:descarte}).
+However, for Descartes, pain as an experience  could not be divorced from the idea of the soul, as he had stated in his letter to the French polymath Marin Mersenne: "I do not explain the feeling of pain without reference to the soul.
 For in my view exists only in the understanding.
@@
 148]{BookCottingham1991v3}.
-In his Treatise of Man, published in 1664, he stated: "Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch.
+In his \textit{Treatise of Man}, published in 1664, he stated: "Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch.
 In this way they pull the tiny fibre which you see attach to it, and simultaneously open the entrance to the pore, located opposite the point where this fibre terminates -- just as when you pull one of a string, you cause a bell hanging at the other end to ring at the same time.
@@
 
-Although the field of pain has changed significantly since the time of Descartes, a complete theory of pain still eludes modern neuroscience.
-Treatment of chronic pain is challenging still.
-However, investigations into unique syndromes such as trigeminal neuralgia offers an opportunity that further the understanding of pain and neuroscience at large.
+\subsection{Theories of Pain}
 
-\subsection{Theories of Pain}
+The concept of a pain specific pathway was first put forward by Charles Bell where he proposed that the brain is a heterogeneous structure, and that nerves are a mix of heterogeneous neurons that became bundled to facilitate anatomical distribution \cite{Bell1868}.
+Evidence soon followed with the discovery of specific touch receptors and the demonstration that resection of posterior spinal tracts in animals by Schiff and Woroschiloff isolated innocuous touch from pain and temperature conductance (anterolateral tract) \cite{Dallenbach1939}.
+Von Frey demonstrated distinct receptive fields that respond to pressure that triggered tactile, warm, cold and pain.
+Sherrington further developed the specificity theory by coining the term “nocicipient” and unified attributes of intensity theory and specificity theory \cite{Casey1982}.
+
+The classic pattern theory on the other hand, as first proposed by John Paul Nafe in the 1929 paper “A quantitative theory of feelings”\cite{Nafe1929}, postulates that the pain is not mediated by a separate nervous system or specialized receptors, but rather by units of undifferentiated afferent nervous impulses that only vary in their numbers, spatial and temporal relationships, including impulse frequency, duration, and density of nerve unit activations.
+The pattern theory postulates that all bodily sensations are a result of the brain's interpretation of nervous impulses.
+The discovery of high threshold mechanoreceptors and unmyelinated nociceptors however discredited the pattern theory and gave further credence to the theory of specificity.
 
@@
 
-The concept of a pain specific pathway was first put forward by Charles Bell where he proposed that the brain is a heterogeneous structure, and that nerves are a mix of heterogeneous neurons that became bundled to facilitate anatomical distribution \cite{Bell1868}.
-Evidence soon followed with the discovery of specific touch receptors and the demonstration that resection of posterior spinal tracts in animals by Schiff and Woroschiloff isolated innocuous touch from pain and temperature conductance (anterolateral tract) \cite{Dallenbach1939}.
-Von Frey demonstrated distinct receptive fields that respond to pressure that triggered tactile, warm, cold and pain.
-Sherrington further developed the specificity theory by coining the term “nocicipient” and unified attributes of intensity theory and specificity theory \cite{Casey1982}.
-
-The classic pattern theory on the other hand, as first proposed by John Paul Nafe in the 1929 paper “A quantitative theory of feelings”\cite{Nafe1929}, postulates that the pain is not mediated by a separate nervous system or specialized receptors, but rather by units of undifferentiated afferent nervous impulses that only vary in their numbers, spatial and temporal relationships, including impulse frequency, duration, and density of nerve unit activations.
-The pattern theory postulates that all bodily sensations are a result of the brain's interpretation of nervous impulses.
-The discovery of high threshold mechanoreceptors and unmyelinated nociceptors however discredited the pattern theory and gave further credence to the theory of specificity.
-
 The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory.
-The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes all the necessary information for pain.
+The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes the necessary information for pain.
 That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities.
-This entire pattern of neural activities is the  sensory pain experience.
-Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn ( laminae V ), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex.
+This entire pattern of neural activities constitute the sensory component of the pain experience.
+Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn (laminae V), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex.
 A central concept, which was also introduced by Melzacks, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}.
@@
 These receptors broadly separate into two categories, differentiated by the type of axonal innervation they receive.
-Acute pain receptors are innervated by $A\delta$ fibres, which permits fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick or acute trauma.
+Acute pain receptors are innervated by $A\delta$ fibres, which permits fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma.
 Dull pain is triggered by receptors that are innervated by slow conducting C fibres ($<1$ m/s).
@@
 
-Before reaching the neocortex, a majority of the second order pain afferents arrive at the thalamus.
+Before reaching the neocortex, several important second order pain afferents arrive at the thalamus.
 The spinothalamic tracts subserving face and body sensory information terminate in the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for face and body respectively.
 The VP complex sends third-order afferents to the primary sensory cortex (S1).
-Thus, the VP representation is somatotopic organized with the face region in the medial aspect, and lower limbs towards the lateral aspect.
+Thus, the VP representation is somatotopically organized with the face region in the medial aspect, and lower limbs towards the lateral aspect.
 The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.
@@
 
-The intralaminar nuclei centromedian (CL) and parafascicular (Pf) receive noxious heat afferents.
-CL receive afferents from laminae V and VII, while Pf from laminae I and V.
-They project to the basal ganglia, motor and parietal cortices, and are thought to be involved in general attention, arousal, and motor functions.
-
 \subsubsection{Cortical projections of pain}\label{section:neuromatrix}
 
-Pain percept is thought to involve distributed regions of the brain, coined as the “pain neuromatrix”\cite{Melzack1999a}.
+Pain percept, defined as the totality of the experience of pain, is thought to involve distributed regions of the brain, coined as the “pain neuromatrix”\cite{Melzack1999a}.
 The regions are not functionally independent, but rather thought to be dynamically inter-related.
 
-Neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a} have revealed consistent regions of BOLD co-activations when undergone painful stimulation.
-These classic fMRI pain regions include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1).
-Regions such as ACC, PFC and insula are also shown to be active in social pain through social exclusion, and emotional stress through PTSD revealed changes in ACC and insula grey matter (Section \ref{fig:pain-matrix}).
+ Important cortical regions involved in pain include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1).
+The role of these cortical regions in pain were discovered through neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during painful stimulation.
+Regions such as ACC, PFC and insula are also shown to be active in social pain during social exclusion and emotional stress associated with PTSD revealed changes in ACC and insula grey matter (Section \ref{fig:pain-matrix}).
 
@@
 
-Other studies suggest pain is correlated to the dynamically changing patterns of activities over time in the regions involved.
-Whereby the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}, and the dynamic rhythms of activities that attune to and away from saliency forms a complex pain connectome \cite{Kucyi2015}.
+A recent study suggests that pain is correlated to the dynamically changing patterns of BOLD activities over time in the regions involved.
+Thus, the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}.
+The dynamic rhythms of activities that attune to and away from saliency, form a complex pain connectome \cite{Kucyi2015}.
 Conditions such as chronic pain therefore show a differentiated tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable nociceptive stimuli \cite{Baliki2008,Legrain2009}.
@@
 \paragraph{Sensory-Discrimination}
-Cortical regions associated with sensory-discriminative aspect of pain, which includes the location and intensity of pain include the primary sensory (S1) and secondary sensory (S2) cortices.
+Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the primary sensory (S1) and secondary sensory (S2) cortices.
 The S1 cortex is responsible for the discrimination of fine-touch sensations.
 S1 receives inputs directly from VP thalamus.
-However, S1 has not been consistently reported in acute pain studies, possibily due to difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within \cite{bushnell1999pain}.
-S2, however, receives direct nociceptive input from VPI thalamus, and is commonly found in pain studies.
+Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}.
+S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies.
 The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}.
@@
 \paragraph{Affective-Motivational}
-The cortical regions that are involved in affective-motivational pain procecssing, which includes the unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions .
+The cortical regions that are involved in affective-motivational pain procecssing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions.
 The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus.
@@
 Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}.
-DMN activation has been to correlate with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.
+DMN activation has been correlated with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.
 
@@
 The contralateral dorsal posterior insula emerged as the only cortical region with significant correlation.
-Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is strong supportive findings of a homologue region in humans.
-Critics of this finding raised issues with the low number of subjects (n=7) and long sampling duration (45 s), and the incompatibility of this interpretation with long standing bodies of neuroimaging literature \cite{Davis2015}.
+Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is a strong supportive finding of a homologue region in humans.
+Critics of this finding raised issues with the low number of subjects, long sampling duration and the incompatibility of this interpretation with long standing bodies of neuroimaging literature \cite{Davis2015}.
 
@@
 
-Chronic pain, as defined by the International Association for the Study of Pain (IASP), is pain that persists or recurrent past the healing phase following an injury \cite{Apkarian2009}.
+Chronic pain is pain that persists or recurrent past the healing phase following an injury \cite{Apkarian2009,Merskey2012}.
 It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months, other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}.
-There is a wide range of chronic pain conditions, including chronic back pain, migraine, arthritic pain, irritable bowel syndrome, phantom-limb pain, temporomandibular disorder, etc.
-
 The new ICD-11 revision of chronic pain classification include a new category for chronic primary pain.
@@
 Conditions such as fibromyalgia, and irritable bowel syndrome fall under this category.
-
-Chronic pain conditions are considered to a type of maladaptive pain, whereby the pain response no longer serves its role in protecting the body from harm, but rather contributes negatively to the normal functions of life.
+Chronic pain conditions are considered to a type of maladaptive pain, whereby the pain response no longer serves its role in protecting the body from harm, but rather contributes negatively to the normal functions of life \cite{May2008}.
 
@@
 
-Neuropathic pain is a type of chronic pain, defined as "pain caused by a lesion or disease of the somatosensory nervous system" by the IASP.
-The pain experience may be constant or paroxysmal, related to a stimulus (evoked) or not-evoked.
+Neuropathic pain is a type of chronic pain, defined as "pain caused by a lesion or disease of the somatosensory nervous system"\cite{Merskey2012}.
+The pain experience may be constant or paroxysmal, related to a stimulus (evoked) or spontaneous.
 Animal models of neuropathic pain suggest that peripheral and spinal circuitry responsible for the transmission of nociception are altered under pathological conditions.
-The nervous system is believed to be altered by maladaptive plasticity following nervous system insult, and thereby lead to sustained neuropathic pain.
+The nervous system is believed to be altered by maladaptive plasticity following a nervous system insult thereby resulting in sustained neuropathic pain.
 
@@
 Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex.
-The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation in the lower ganglions.
+The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation.
 
@@
 
- Studies have found that neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research.
-
- In general, brain regions that are involved in the somatosensation of pain process are involved in both acute and chronic pain.
+ Neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research.
+Brain regions that are involved in the somatosensation of pain process are involved in both acute and chronic pain.
 These regions include the posterior insula, lateral thalamus and dorsal striatum.
@@
 In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}.
-Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and meidal prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}.
+Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and medial prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}.
 A shift in forebrain activities is observed in neuropathic versus acute pain, abnormal activities in anterior cingulate cortex and prefrontal cortex is commonly observed in these studies \cite{Davis2006}.

### 2018-01-09 thesis/contents/reviews/trigeminal.tex
changes with moji

@@
 
-Modern pain literature continues to update the exactly diagnostic definition of TN.
+Modern pain literature continues to update the exact diagnostic definition of TN.
 The Burchiel classification scheme divides facial pain into 5 major subtypes \cite{Burchiel2003,Eller2005}: idiopathic (without apparent cause), trigeminal injury, symptomatic, postherpetic, and atypical facial pain.
@@
 Trigeminal injury can be unintentional: which can be due to facial trauma, surgery to the oral cavity; this type of pain would involve persistent and burning pain to the affected areas; and is termed trigeminal neuropathic pain.
-The injury can also be intentional: due to interventions such as rhizotomy or gangliolysis; these result in numbness and loss of sensation of the face; and is termed trigeminal deafferentation pain.
+The injury can also be intentional, due to interventions such as rhizotomy or gangliolysis; these result in numbness and loss of sensation of the face and are termed trigeminal deafferentation pain.
 Symptomatic TN, also termed MS-TN, is due to trigeminal nerve injury relating to multiple sclerosis related central white matter inflammation.
@@
 
-There is also a definition \cite{Zakrzewska2016} that attempts to address the lack of symptomatic TN classification in ICHD-3 by classifying TN into idiopathic TN: without apparent cause; classic TN: caused by vascular compression of the trigeminal nerve root; and secondary TN: as the consequence of a major neurological disease such as MS.
+There is also a definition \cite{Zakrzewska2016} that attempts to address the lack of symptomatic TN classification in ICHD-3 by classifying TN into idiopathic, classic TN, caused by vascular compression of the trigeminal nerve root and secondary TN, as the consequence of a major neurological disease such as MS.
 
@@
 
-The sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano and temperature pain, fine touch and vibration, such as the Messner's corpuscles, Pacinian and Ruffini corpuscles.
+The sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messner's, Pacinian and Ruffini corpuscles.
 
@@
 It further divides into lacrimal, frontal, and nasociliary nerve branches.
-The lacrimal branch supplies the lacrimal; the frontal contains smaller branches that innervate the skin of the medial forehead, upper eyelid, and parts of the frontal sinus; the nasociliary branch innervate various parts of the eye, and mucous membrane of the anterior nasal septum, anterior portion of the lateral nasal wall, and the skin on the tip of the nose.
-V1 enters the skull base through the superior orbital fissure.
+The lacrimal branch supplies the lacrimal gland; the frontal contains smaller branches that innervate the skin of the medial forehead, upper eyelid, and parts of the frontal sinus; the nasociliary branch innervate various parts of the eye, and mucous membrane of the anterior nasal septum, anterior portion of the lateral nasal wall, and the skin on the tip of the nose.
+V1 exits the skullbase through the superior orbital fissure.
 
@@
 It also innervates upper oral cavity.
-It exists the skull through the foramen rotundum.
+It exits the skull through the foramen rotundum.
 
@@
 The V3 division carries the mastication efferent to the four muscles recruited for mastications: the masseter, temporal, medial and lateral pterygoids.
-The tensor tympani also receives some CN V innervation, which is thought to be important in dampening the sound of chewing.
-V3 exists the skull through foramen ovale.
+The tensor tympani also receives some CN V innervation, but it is primarily innervated by the facial nerve.
+It is thought that CNV innervation is important in dampening the sound of chewing.
+V3 exits the skull through foramen ovale.
 
-The branches then converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion), whose posterior extension forms the CN V.
-It is analogous to the spinal dorsal root ganglion, and contains all the trigeminal sensory cell bodies.
-It situates in the Meckel's cave, on the petrous bone of the middle cranial fossa.
-It is somatotopically organized in respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.
+The afferent branches converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion), which then forms the trunk or cisternal component of CN V.
+The TRG contains all the trigeminal sensory cell bodies.
+It situates in the Meckel's cave, in the middle cranial fossa.
+It is anatomically organized with respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.
 
@@
 The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons.
-The CNS myelin can be several millimetres beyond the surface of the pons, before it gives away to the transition area, termed root entry zone, or less commonly, Obersteiner-Redlich zone \cite{Peker2006}.
+The CNS myelin can be several millimetres beyond the surface of the pons, before it gives away to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}.
 Study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm).
@@
 This can also be observed in conventional MR imaging \cite{Borges2010}.
-The superior cerebellar artery is responsible for most of the NVC (60-90\%), followed by the anteroinferior cerebellar artery and basilar arteries (Lutz et al., 2011).
+The superior cerebellar artery is responsible for most of the NVC (60-90\%), followed by the anterior inferior cerebellar artery (AICA) and basilar arteries \cite{Lutz2011}.
 There is a slight lateral prevalence, with right-sided TN occurring in  in 60\% of the cases, and in rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.
-Other types of compressions may also cause TN; these can be in the forms of saccular aneurysm, arteriovenous malformation, vestibular schwannomas, meningiomas, epidermoid cysts, and other tumours.
-However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the expression of pain \cite{Desouza2013,Hodaie2013}.
-In some cases the TN pain is reported to be contralateral to the NVC \cite{Cheng2008,Haddad1990,Revuelta1995}.
-There are also 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.
+Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms .
+However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the clinical expression of pain \cite{Desouza2013,Hodaie2013}.
+In approximately 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.
 
-The NVC that occurs in TN is believed to disrupt the central myelin, and that the region of disruption would begin at the peripheral-central myelin transition zone, and extending into central myelin regions.
 Janetta \cite{Jannetta1967} proposed that NVC leads to focal demyelination of the trigeminal sensory root axons, and causes ephaptic nerve conductances that cause neuralgia.
@@
 Adjacent to the site of compression-related focal demyelination, thinly myelinated sheaths that encapsulate a number of demyelinated axons could be observed, suggesting aberrant remyelination.
-The TRG is also postulated to be responsible for neuralgia \cite{Rappaport1994}.
-It's believed that ganglionic ectopic autonomic firing can be triggered by mechanical disturbances as a result of nerve compression.
-Additionally, ganglionic cross aftercharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding ganglionic neurons.
-Histology findings by Devor et al.
-supported the ignition hypothesis \cite{Devor2002a}, where demyelinated axons were shown to be in close apposition without intervening glia.
+It is therefore plausible that NCV disrupts the central myelin.
+The region of disruption may include the peripheral-central myelin transition zone, and extending into central myelin regions.The TRG is also postulated to be responsible for neuralgia \cite{Rappaport1994}.
+It is believed that ganglionic ectopic autonomic firing can be triggered by mechanical disturbances as a result of nerve compression.
+Additionally, ganglionic cross aftercharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding neurons.
+Histological findings by Devor et al.
+support the ignition hypothesis \cite{Devor2002a}, where demyelinated axons are shown to be in close apposition without intervening glia.
 
@@
 Multiple surgical treatments are available, from more central procedures addressing NVC to peripheral procedures such as neurectomy.
-The procedure that provides the greatest likelihood of pain relief is microvascular decompression (MVD) \cite{Lovely1997}, where the nerve is decompressed from the offending vascular structure through the placement of a piece of Teflon\textcopyright gauze.
+The procedure that provides the greatest likelihood of pain relief is microvascular decompression (MVD) \cite{Lovely1997,Hodaie2013}, where the nerve is decompressed from the offending vascular structure through the placement of a piece of Teflon\textcopyright gauze.
 This is thought to dampen the pulsations of the artery on the nerve.
-MVD showed 90\% pain relief immediately after surgery \cite{Zakrzewska2005}, and a long-term success rate of 70\% at 10 years \cite{Barker1996}.
+MVD can result in 90\% pain relief immediately after surgery \cite{Zakrzewska2005}, and a long-term success rate of 70\% at 10 years \cite{Barker1996}.
 Partial sensory rhizotomy (PSR) can also be performed, where the offending CN V branch is severed, with a response rate of 88-70\% \cite{Young1993,Zakrzewska2005}.
@@
 Proximal targets correlate with longer pain relief, but also higher chance of facial numbness\cite{Xu2014}.
-The procedure requires a stereotactic frame, and a dose of 70-90 Gy is deposited.
+The procedure requires a stereotactic frame, and a dose of 70-90 Gy delivered to the nerve.
 The effect is delayed and pain relief can be experienced in 4-6 weeks post-radiosurgery.
-GKRS treatments result in 90\% initial pain relief \cite{Tuleasca2015,Kondziolka2010}, about 50-80\% pain-free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}.
-More recent studies suggested higher pain-free probabilities at 3(71.8\%), 5(64.9\%), 7(59.7\%) and 10(45.3\%) years\cite{Tuleasca2015}.
+GKRS treatments result in 75\% initial pain relief \cite{Tuleasca2015,Kondziolka2010}, about 50-80\% remain pain-free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}.
 More favourable treatment outcomes were related to no other accompanying symptoms, no prior surgery and shorter term pain history of fewer than three years.
+
+\subsection{Neuroimaging of Trigeminal Neuralgia}
+
+Neuroimaging of the TN has been challenging, due to the small diameter of the CN V and since the contrast of the CNS trigeminal pathways is difficult to distinguish from the surrounding on conventional MRI \cite{Borges2010}.
+Manual segmentation of T1 and T2 MRI requires great effort, with mixed results \cite{Miller2008a}.
+Diffusion MRI provided fiber orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}.
+
+Application of tractography permitted visual reconstructions of human cranial nerves in vivo \cite{Hodaie2010}.
+Cisternal CN V could be reliably delineated with Gaussian tensors with just 25 MR gradient directions.
+This permitted high fidelity visualization and targeted characterization of CN V diffusivity in TN, demonstrating significant reduction in fractional anisotropy (FA) post-surgery at the site of the GK target \cite{Hodaie2009a}.
+Study of CN V diffusivity in conjunction with deep white matter using TBSS, again revealed reduced FA on the affected CN V, with also bilateral increased in radial, axial, and mean diffusivity.
+Disruptions were also found in the cognitive-affective, attention, and motor regions of the cortex, consistent with cortical patterns in chronic pain \cite{Desouza2013}.
+Lowered FA was reported in multiple TN studies \cite{Herweh2007, Lutz2011}.
+Study in temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}, thereby demonstrating that CN V FA is related to nociception.
+The TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.

### 2018-01-09 thesis/contents/tn_sagit.tex
changes with moji

@@
 \section{Abstract}
-\textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial pain syndrome that involves the trigeminal sensory pathway.
+\textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial neuropathic pain syndrome involving the trigeminal sensory pathway.
 We demonstrate the use of a fully-automated group white-matter tractography, and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate patients and healthy controls.
 
-\textbf{Methods:} We perform group merged tractography using SAGIT on 36 sex-matched TN patients (right-sided pain) and 36 controls.
+\textbf{Methods:} We performed group merged tractography using SAGIT on 36 sex-matched TN patients (right-sided pain) and 36 controls.
 The trigeminal nerve (CN V), pontine decussation, and thalamocortical S1 pathways were reconstructed.
@@
 Unaffected were analyzed.
-Classifiers that performed at greater-or-equal-to than 75\% accuracy were accepted.
+Classifiers that performed at greater-or-equal-to 75\% accuracy were accepted.
 
-\textbf{Results:} Both affected and unaffected CN V can be differentiated with up to 80\% accuracy from controls.
+\textbf{Results:} Both affected and unaffected trigeminal nerves in TN patients can be differentiated with up to 80\% accuracy from controls.
 Affected CN V differences were more focused in the cistern and root-transition-zone, whereas unaffected CN V show more focus towards the brainstem.
@@
 Tractography was first performed in the native DWI space for each individual.
-FA, AD, RD, MD, GFA and AP metrics were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.
+FA, AD, RD, MD, and GFA metrics were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.
 Additionally, subject ID and group ID scalar values were also inserted into each tractography files for identification.
@@
 The affected and unaffected sides in TN patients were also compared with each other.
-Age was regressed out by using the residuals of the linear regression model from controls group.
+Age was regressed out by using the residuals of the linear regression model from the controls group.
 The diffusion values were subsequently normalized to zero mean with ranges $[-1, 1]$ for model training.
@@
 \centering
-\includegraphics[width=\linewidth]{CN V-inferior-view.png}
+\includegraphics[width=\linewidth]{cnv-inferior-view.png}
 \caption{Inferior view of the merged CN V delineations.}
@@
 \centering
-\includegraphics[width=\linewidth]{figure-GP-CN V.pdf}
+\includegraphics[width=\linewidth]{figure-GP-CNV.pdf}
 \caption{CN V classifier performance}
@@
 \centering
-\csvautotabular{CN V_table.csv}
+\csvautotabular{cnv_table.csv}
 \caption{CN V GP classifier performance data}

### 2018-01-09 thesis/contents/hypotheses.tex
- changes to hypothesis started

@@
 
-The general aim of this thesis is quantify the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways in order to identify specific diffusivity pattern that differentiates between TN and non-TN CN V tissue.
+The general aim of this thesis is quantify the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways in order to identify specific diffusivity pattern that differentiates between TN and non-TN.
 In achieve this aim, a generalizable group diffusion imaging tractography software framework must be developed to overcome the technical barrier of large scale group tractography in arbitrary populations.
 A number of confounds on data registration and measure must be resolved in order to gain confidence in the result of this methodology.
-These include a) establish the feasibility of diffusion tractography of brainstem CN V originating from the trigeminal nucleus; b) establish the best way to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects; c) Make possible non-linear registration of tractography geometries, and to quantify the geometry registration performance at the group level; d) apply previously established methodology to the reconstruction and quantification of the trigeminal sensory pathway.
+The roadmap and rational for the specific studies are described below.
+
+\section{Study I}
+Symptomatic TN, or MS-TN is thought to relate to the brainstem MS lesions.
+However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI, and the variability of MS lesion across individuals.
+Due to these factors, image-registration-based techniques such as voxel-based mophometry becomes unreliable.
+Previous single-tensor tractography studies of CN V has been limited to the cistern nerve segment.
+In this study, we have the opportunity to study sub-segments of CN V in TN and MS-TN, with the assistance multi-tensor tractography delineation of the brainstem CN V.
+Thereby establishing the feasibility of multi-tensor tractography in improving anatoimcal delineation of CN V, as well as its usefulness in the study of TN.
+
+\paragraph{Main Aim} Establish the feasibility of diffusion tractography in delineating the brainstem CN V, and contribute towards the study of pathological CN V tissue.
+
+\paragraph{Specific Aim}
+\begin{itemize}
+    \item Use multi-tensor tractography to delineate brainstem CN V, in order to quantify tissue diffusivity measurements in targeted sub-regions.
+    \item Differentiate TN and MS-TN based solely on the obtained CN V diffusivity measurements.
+\end{itemize}
+\paragraph{Specific Hypothesis}
+
+\section{Study II}
+
+\paragraph{Main Aim}
+establish the best way to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects
+\paragraph{Specific Aim}
+\paragraph{Specific Hypothesis}
+
+\section{Study III}
+
+\paragraph{Main Aim}
+Make possible non-linear registration of tractography geometries, and to quantify the geometry registration performance at the group level
+\paragraph{Specific Aim}
+\paragraph{Specific Hypothesis}
+
+\section{Study IV}
+\paragraph{Main Aim}
+apply previously established methodology to the reconstruction and quantification of the trigeminal sensory pathway.
+
+\paragraph{Specific Aim}
+\paragraph{Specific Hypothesis}

### 2018-01-10 thesis/contents/hypotheses.tex
changes to hypothesis

--- 2017-11-24 thesis/contents/hypotheses.tex
@@
 
-The general aim of this thesis is quantify the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways in order to identify specific diffusivity pattern that differentiates between TN and non-TN CN V tissue.
-In achieve this aim, a generalizable group diffusion imaging tractography software framework must be developed to overcome the technical barrier of large scale group tractography in arbitrary populations.
-A number of confounds on data registration and measure must be resolved in order to gain confidence in the result of this methodology.
-These include a) establish the feasibility of diffusion tractography of brainstem CN V originating from the trigeminal nucleus; b) establish the best way to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects; c) Make possible non-linear registration of tractography geometries, and to quantify the geometry registration performance at the group level; d) apply previously established methodology to the reconstruction and quantification of the trigeminal sensory pathway.
+The general aim of this thesis is to quantify the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, in order to identify specific diffusivity pattern that differentiates between TN and non-TN.
+Towards this aim, a generalizable group diffusion imaging tractography software framework must be developed to overcome the technical barrier of large-scale group tractography.
+The roadmap and rational for the specific studies are described below.
+
+\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
+MS-TN, also known as Symptomatic TN, is thought to relate to the brainstem MS lesions.
+However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI, and the variability of MS lesion across individuals.
+Due to these factors, image-registration-based techniques such as voxel-based morphometry becomes unreliable.
+Previous single-tensor tractography studies of CN V has been limited to the cistern nerve segment.
+In this study, we have the opportunity to study sub-segments of CN V in TN and MS-TN by using multi-tensor tractography to delineate brainstem CN V.
+Thereby establishing the feasibility of multi-tensor tractography in improving anatomical delineation of CN V, as well as its usefulness in the study of TN.
+
+\subsection{Main Aim} 
+Establish the feasibility of diffusion tractography in delineating the brainstem CN V, and contribute towards the study of pathological CN V tissue.
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Use multi-tensor tractography to delineate brainstem CN V, in order to quantify tissue diffusivity measurements in targeted sub-regions.
+    \item Differentiate TN and MS-TN based solely on the obtained CN V diffusivity measurements.
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item TN is known to relate to vascular compression of the cistern REZ, therefore its diffusivity signature will occur in the cistern region.
+    \item MS-TN should be related to brainstem MS lesions.
+Therefore CN V segments close to the lesions will show more significant changes in tissue diffusivity, whereas segments far away from the lesions will not.
+    \item MS is a demeylination disease, therefore MS-TN diffusivity signature must be related to demyelination, such as increase in RD, and decrease in FA. 
+\end{itemize}
+
+\section{Study II: Diffusion Weighted Image to T1 Co-registration Without Reverse-blip: Investigation of Best Practices}
+The task of establishing inter-subject anatomical correspondences is imperative for extending tractography to multiple subjects.
+Population studies with inter-subject T1 registration are well-established, where a normalized T1 template is generated from numerous subjects to facilitate data analysis.
+Similarly, DWI of each subject can be co-registered to their corresponding T1, to extend the T1 template space to tractography space.
+Therefore the DWI-T1 co-registration accuracy is crucial to the reliability of inter-subject tractography alignment.
+DWI-T1 co-registration is challenging due to the DWI eddy-current and EPI field distortions.
+While new DWI sequences with reverse-blips have been shown to reduce non-linear distortions, they have yet to be commonly adopted in clinical practice.
+While T1 inter-subject registration methodology is well documented, the feasibility, and strategy for DWI-T1 co-registration on sequence without reverse-blips have not been documented, and forms a potential bottleneck on the performance of any DWI-T1 population registration pipeline.
+
+\subsection{Main Aim}
+Establish the best strategy to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects.
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Quantify the degree of improvement in registration between affine-only versus symmetric-diffeomorphic registrations.
+    \item Determine which commonly-used DWI scalar image types can provide the best registration result when used as the registration-intermediate.
+    \item Determine if anisotropic-power image, an experimental DWI scalar map that closely resemble T1 contrast, can provide even more registration improvements. 
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item Symmetric-diffeomorphic registration should result in the best accuracy, but with more diminishing asymptotic improvement over affine registration alone.
+    \item Averaged DWI should be the best scalar image, as it provides a balanced contrast between white and grey matter.
+    \item Anisotropic-power Image will show more substantial improvement over the existing scalar images.
+\end{itemize}
+
+\section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
+Once the feasibility of multi-tensor tractography to the brainstem trigeminal pathways is established, and DWI-T1 co-registration optimization identified, the software framework for group tractography is constructed.
+The software aims to automate the tractography process at large-scale, and also aims to reduce the cognitive load of researchers by providing region-of-interest, and tractography parameter management.
+It also provides the ability to report the tractography output at a group level both visually, as well as quantifying the output by an original scalar score called Normalized Overlap Score (NOS) score.
+The study demonstrates the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population.
+At a scale that would be time-prohibitive with manual methods.
+The comparisons determine the best tractography methods to use for different types of neuroanatomy.
+
+\subsection{Main Aim}
+Make possible non-linear registration of tractography geometries, and to quantify the geometry registration performance at the group level
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD in 42 subjects. 
+    \item Delineate neuropathways that are known to be difficult to delineate: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
+    \item Determine if NOS score perform as expected on real-world data, by comparing with manual rating by neuroanatomy experts.
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item At a group level, the delineations should show reasonable convergence, and clearly recognizable. 
+    \item The group tractography should reveal known limits of DTI, as well as improvements on XST and CSD methods.
+    \item Deterministic tractography should preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
+    \item NOS score should correlate with expert ratings, and show improved consistency.
+\end{itemize}
+
+\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
+With the software framework thoroughly tested, and the best-suited tractography methods determined, the developed methods are applied to the trigeminal sensory pathway.
+The study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group.
+The pathway is divided into three major regions: CN V, trigeminopontothalamal (TPT) segment, and the thalamocortical S1 segment.
+With the increased volume of data, more detailed along-the-tract analysis can be carried out on each segment, and machine learning is used to auto-determine relevant segments that maximally differentiates TN from controls to minimize human bias.
+S1 and thalamus activation were consistently found in chronic pain fMRI, and therefore their related pathway should also show similar findings.
+
+\subsection{Main Aim}
+Apply previously established methodology to the reconstruction and quantification of the trigeminal sensory pathway.
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Delineate and visualize the three segments of the trigeminal sensory pathway: CN V, trigeminopontothalamal segment, and the thalamocortical S1 segment. 
+    \item Determine diffusivity signatures that maximally differentiate the affected and unaffected sides of the TN patients.
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item CN V diffusivity signature should match findings from Study I.
+    \item At the level of CN V, the affected cistern segment should show differentiation from the unaffected side.
+    \item The affected S1 segment should differ from the unaffected side, due to more frequent nociception.
+    \item The affected TPT segment might show more differentiation near the end points, due to possible differences in trigeminal nucleus and thalamus diffusivity.
+However, its mid-segment diffusivity might be inconclusive, as the diffusivity metrics such as FA and RD might not be adequate in describing the complex crossings at the pontine decussation.
+\end{itemize}

### 2018-02-02 thesis/contents/discussions.tex
Updates from ShareLaTeX

@@
 \chapter{General Discussions}
-\section{Tractography}
-\subsection{Delineating the trigeminal nerve into the brainstem}
-The primary motivation to visualize the trigeminal nerve via tractography, was due to the inability to discern both the brainstem trigeminal nerve, nor the trigeminal nucleus in all known MRI modalities.
-Before tractography, the curvature of the trigeminal nerve could only be inferred from previous anatomical studies, or by postmortem dissection.
-For successful tractography of the CN V, both DWI sequence and the algorithm choice is paramount.
-Given a DWI sequence, at 25 gradient directions, cisternal CN V could be readily reconstructed with Gaussian tensors.
-Resolution of the DWI sequence would affect the output.
-Unlike major white matter bundles, cranial nerves diameters are in the milimeter range, and therefore voxel sizes greater than 2 mm will result in severe partial volume of cranial nerves.
-Achieving 2mm isovoxel on a 3T scanner requires 32 channel head coil.
-Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable.
-For this reason, the acquisition in-plane resolution can be tune to 1x1 mm, and slice thickness at 3mm.
-While the acquisition is not ideal, it sacrifices resolution in only the z-axis, the tractography visual fidelity is improved.
-Discerning brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers.
-A number of methods have been devised, including probabilistic tractography made popular by the FSL software package, which generates image-based results that represent the propagation probability of a seed region.
-Neurosurgeons however often need visualizations that represent anatomy geometrically.
-Therefore streamline-based tractography visualizations are of broader appeal.
-In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions.
-Another often overlooked area is the seeding pattern.
-When tracing the streamlines, an algorithm will need to generate a number of starting points in a defined volume.
-Common patterns in prepackaged software include in a regularly spaced grid, where seed spacing can make decisive differences in visuals, and random sampling.
-The seeding pattern is important in both the final visual style of the tracts, as well as if key anatomical landmarks can be properly tracked.
-Another consideration is the reproducibility of the desired anatomy, as randomly seeded bundles cannot be reproduced exactly.
-Grid seeding also can produce sparse spacing of the tract bundles when viewed from a certain angle, while random seeding can reduce this kind of visual artifact.
-A good balance might be to have a reproducible sampling scheme that also is robust to different viewing directions, for example a phyllotaxis pattern.
 
-\subsection{Tractography group registration}
-Tractography is a powerful methodology that reveals the morphology of brain white matters that were otherwise previously invisible.
-Its primary usage however has been limited to case studies and small-scale anatomy comparative analysis.
-The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of large groups.
+\section{Summary Of Findings}
 
-Group comparative studies require the establishment of anatomical consensus, such that similar white matter structures between populations can be readily studied.
-Efforts in MRI medical image registration have paved the way to this end, especially in the fields of diffeomorphic registration.
-The difficulty in adapting registration to tractography is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods.
-The best efforts to date apply affine transformation for between-tract registrations \cite{Garyfallidis2015}.
-However non-linear geometric registration is still a topic of future research.
-Our approach is to recast tractography as a brain segmentation method that is spatially-aware.
-Tractography then can be considered a point-based sub-sampling method that is represented in the same coordinate frame as the native DWI space.
-Any transforms applied to this DWI space is then applicable also to any tractography models that share its coordinate frame.
-Existing diffeomorphic registration solutions can then be used to non-linearly distort tractography, and circumvent the computational difficulty of geometry based non-linear registration.
-The advantage of our tractography registration strategy is that any new advances in the field of MR registration can be readily adapted to improve the output of our method.
+The contribution of this thesis are as follows:
 
-Group registration of tractography towards a common template space requires two separate steps: T1---DWI co-registration, and T1---Template registration.
-Any registration errors in these two steps will become multiplicative in the end result.
-T1 registration is well studied, however T1---DWI registration has few literature.
-Therefore it was imperative that we investigated the best method to minimize registration errors that were compatible with our image data format.
+\begin{itemize}
 
-\subsection{Measurements from tracts}
-Aggregating along-the-measure statistics is a form of clustering that requires projection of tract vertices onto a structural skeleton, or centroid.
-Common clustering methods calculate a polyline centroid based on different distancing functions between streamlines.
-These methods,  however, require tractography streamlines to have similar length, and will underperform when tractography structures contain sharp turns.
-We have explored a number of strategies, and considered adaptation of manifold-based methods, and point-cloud skeletal extraction method from computer graphics literature.
-These approaches, however, contain assumptions that do not suit our needs.
-For example, manifolds commonly consider that the latent structure of the data lies on a lower-dimensional surface, while tractography is commonly arranged in complex bundles.
-Moreover, point-cloud methods deal with point-cloud data obtained from 3D scanners, which situate on a 3D shell, while tractography points are distributed within a volume.
-White matter anatomy can have intricate crossings, multi-way join and diverging structures.
-These are difficult for existing 3D analytical methods, and require models with simplified assumptions to reduce the number of unknown parameters.
-A potentially promising future direction may be to automatically discover these more complex latent structures using deep learning methods.
+\item We are the first to image the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography.
+Prior to this work, CN V tractography was limited to single-tensor DTI models and cannot be delineated into the pons.
+Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increase the range of trigeminal-related pathology, for example MS-TN, that can be studied with neuroimaging.
+
+\item The study, for the first time, clearly differentiated TN and MS-TN by sub-segment diffusivity differences, between cistern (TN) and brainstem peri-lesional (MS-TN) segments.
+This distinction in the trigeminal REZ between these two conditions have been suspected, but never confirmed.
+The study is the first to suggest that disruptions along the CN V in different locations may trigger similar pain symptoms.
+
+\item We overcame the major technical barrier to detailed diffusion tractography study of the trigeminal pathway in a large population group, by developing and validate the Selective Automated Group Integrated Tractography (SAGIT) software platform.
+In the process, we determined the optimal T1-DWI co-registration methodology with clinical DWI sequences.
+Prior to this project, the co-registration method has never been exhaustively determined and reported in literature.
+
+\item We are the first to perform merged group tractography study of the trigeminal nerve, and its primary sensory pathway, and the first to conduct such study in TN, and the first to perform Gaussian Process machine learning to the study of TN.
+We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in its cross-sectional detail.
+We discovered that the unaffected CN V diffusivity differs from controls as well as the affected side and that the affected TPT pathway and S1 pathway diffusivity differs from controls.
+
+\end{itemize}
 
@@
 
-The findings in Study I \ref{section:study1} revealed localized CN V tissue differences in MS-TN and idiopathic TN.
-The possible implication is that disruption anywhere along the primary trigeminal afferent may lead to similar TN-like pain syndromes.
-MS is a classic myelin inflammatory disease, and therefore these diseases may share similar disruptions in myelin ultrastructure.
+The findings in Study I (Chapter \ref{section:study1}), which aimed to identify trigeminal nerve diffusivity signatures in TN and MS-TN, identified for the first time a method that can distinguish between microstructural changes in these conditions, and revealed localized CN V diffusivity differences in MS-TN and idiopathic TN.
+The possible implication is that disruption anywhere along both the cisternal and pontine segments of the trigeminal nerve may lead to similar TN-like pain syndromes.
+Central demyelination in the trigeminal root transition zone is associated with vascular compression in TN.
+Moreover MS is a classic myelin inflammatory disease.
+Therefore both diseases may share pathophysiology due to disruptions in myelin ultrastructure.
+The study opened the possibility that multi-tensor tractograpy aided analysis added a measure of objective assessment in examining pain conditions, and may facilitate similar studies that provide insight into the fundamental changes that occur at the level of white matter.
 
-Study IV further suggested that the unaffected CN V diffusivity is also different from that of controls.
+Study IV (Chapter \ref{section:study4}), which applied group tractography to the study of the trigeminal sensory pathway, further suggested that the unaffected CN V diffusivity is also different from that of controls.
+DeSouza et al.
+\cite{Desouza2013} also presented similar findings, where the unaffected RD, MD, and AD were found to differ from controls.
+Desouza performed the study with manual ROI placement at a single measurement position in the REZ, Whereas we performed GP classification along the nerve automatically.
 The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain.
 In many studies it has also been used as a basis of comparison.
-This finding however is not as surprising as one may suspect.
+This finding, however, is not surprising.
 It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases.
@@
 Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible.
+With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between affected and healthy controls.
+Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what degree of difference does it result in pain.
+Investigations into bilateral TN pain, and vascular compression in the absence of pain, may reveal additional insight into the relationship between white matter changes and TN pain in general.
 
@@
 MVD, in this case can be explained as the physical disruption of the condition.
+
+\section{Tractography of the trigeminal nerve and brainstem fibers}
+
+The primary motivation for the study of the trigeminal nerve with tractography was the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
+Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissection.
+
+Resolving cross fibers of the cerebral white matter was a fundamental advancement in diffusion tractography.
+The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging.
+A widely adopted technique was probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map.
+This kind of image-based probabilistic tractography was useful for whole-brain white matter studies.
+However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways.
+Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves.
+Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition.
+The acquisition time of these sequences was impractical for the clinical environment.
+Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
+The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development.
+SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions.
+This meant that the methods are more robust to noise, and therefore require less number of gradient acquisitions.
+CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions.
+Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.
+
+These key algorithmic developments decidedly brought cross-fiber tractography in clinical neuroimaging to the realm of possibility.
+This is significant as the vast majority of imaging of neuropathologies in patients only exist as clinical acquisitions.
+We took advantage of this opportunity and explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST.
+The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also that the trigeminal nerve has a number of unique properties as a tractography target: 1) delineation quality is easy to judge due to its simple morphology.
+2) it situates in both single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.
+
+The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging.
+It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid.
+Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks.
+Such manual methods were highly subjective and became problematic with patient images, where distortions to the nerve could not be assessed.
+
+The ability to segment the entire trigeminal nerve also opens the avenue to measure its tissue diffusivity from tractography directly.
+Moreover, specific tissue-metrics such as myelin image, or PET images, can also be measured with confidence in reproducibility.
+This extends these methods to facilitate disease studies well beyond TN.
+We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transits from pontine tissue to the cistern.
+Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable, due to the noise in the cistern CSF.
+Turns out that the merged tractography strategy simplified this task by allowed a single manual curve to be defined manually in the template tractography space.
+This strategy can similarly be extended for obtaining along-the-track measure for other highly variable streamline bundles.
+
+We applied machine learning to the analysis of TN in Study IV in order to avoid our own confirmation bias.
+The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method.
+We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret without inducing biases.
+The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency.
+2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive.
+
+\section{Neuroimaging of deeper white matter for the study of pain}
+
+The pain neuromatrix, as studied by various functional and structural connectivity studies, distributes over the full range of neocortical areas.
+These areas are associated with complex white-matter pathways that extends to all regions of the brain.
+These include sharp curvatures of the anterior pain pathways of the ACC, the distant projections of the temporal cortices.
+Some pathways are just beyond the feasibility of current methods, these include the projections of the insula and claustrum, small decending pathways to PAG, and the spinal trigeminal pathways in the brainstem.
+
+We have shown in Study III (\ref{section:study3}), for the first time, that different tractography algorithms are suited to different neuropathways when comparing merged reconstructions in 42 subjects.
+Maier-Hein et al came to similar conclusions after examining 96 reconstructions from a single set of simulated data \cite{Maier-Hein2017}.
+Therefore care must be taken to ensure tractography parameters are targeted to the anatomy of study, as more consistent delineation lead to more robust tissue measurements.
+
+We observed that merged tractography can delineate more complete pathway delineations than from any single subject.
+This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high quality templates.
+Therefore merged tractography may serve to produce high quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals.
+We have explored the use of SAGIT in generating the first neonatal porcine tractography template \cite{Zhong2016a}.
+Such template does not exist for humans to the best of our knowledge, and can be readily created from Human Connectome Project data.
+In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low resolution clinical acquisitions, these template tracts may be used as proxy to enable analysis.
+
+\section{Neurosurgery and Tractography}
+
+The biggest value of tractography to neurosugery is to assit in pre-operative planning.
+Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D.
+We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}.
+Tractography and 3D anatomical model ultimately are standard 3D geometries in computer graphics.
+Therefore advances in 3D printing can be easily adopted in neurosurgery to allow surgeons to physically examine tractography reconstructions.
+On the otherhand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making.
+Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.
+
+\section{Method Discussions}
+
+\subsection{Delineating the brainstem fibers of the trigeminal nerve}
+
+For successful tractography of the CN V, both DWI sequence and the algorithm choice are paramount.
+Given a DWI sequence, cisternal CN V could be readily reconstructed with Gaussian tensors.
+Unlike major white matter bundles, cranial nerves diameters are in the millimetre range, and therefore voxel sizes greater than 2 mm may result in partial volume averaging.
+Achieving 2mm isovoxel on a 3T scanner requires 32 channel head coil.
+Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable.
+For this reason, the acquisition in-plane resolution can be tuned to 1x1 mm, and slice thickness at 3mm.
+While the acquisition is not ideal, it sacrifices resolution only in the z-axis, however the tractography visual fidelity is improved.
+Imaging the brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers.
+A number of methods have been devised, including probabilistic tractography (FSL)\cite{Behrens2007}, which generates image-based results that represent the propagation probability of a seed region.
+Neurosurgical visualization however requires three dimensional representation of the fibers, and therefore streamline-based tractography visualizations are of broader appeal.
+In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions.
+
+An important technical point, but rarely mentioned in literally, is the tractography seeding pattern.
+Therefore I will address it here: When tracing the streamlines, an algorithm will need to generate a number of starting points in a defined volume.
+Common patterns in prepackaged software include in a regularly spaced grid and random sampling.
+The seeding pattern is important in both the final visual representation of the tracts, and to determine whether key anatomical landmarks can be properly tracked.
+Another consideration is the reproducibility of the desired anatomy, as randomly seeded bundles cannot be reproduced deterministically.
+Grid seeding may result in sparse spacing of the tract bundles when viewed from a certain angle, while random seeding can reduce this kind of visual artefact.
+A good balance might be to have a reproducible sampling scheme that also is robust to different viewing directions, for example a phyllotaxis pattern.
+
+\subsection{Tractography group registration}
+Tractography is a powerful methodology that permits the visualization of white matter fibers in three dimensions.
+Its primary use however has been limited to case studies and small-scale anatomy comparative analysis.
+The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of larger groups.
+
+Group comparative studies require the establishment of anatomical consensus, such that similar white matter structures between populations can be readily studied.
+Efforts in MRI medical image registration have paved the way to this end, especially in the fields of diffeomorphic registration.
+The difficulty in adapting registration to tractography is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods, as they are computationally untenable.
+To date there is no direct non-linear 3D deformation methods for tractography.
+The best efforts to date apply affine transformation for between-tract registrations \cite{Garyfallidis2015}.
+Our approach is to recast tractography as a brain segmentation method while also make use of its orientation information.
+Tractography then can be considered a point-based sub-sampling method that is represented in the same coordinate frame as the native DWI space.
+Any transforms applied to this DWI space is then applicable also to any tractography models that share its coordinate frame.
+Existing diffeomorphic registration solutions can then be used to non-linearly distort tractography, and circumvent the computational difficulty of geometry based non-linear registration.
+The advantage of our tractography registration strategy is that any new advances in the field of MR registration can be readily adapted to improve the output of our method.
+
+Group registration of tractography towards a common template space requires two separate steps: T1---DWI co-registration, and T1---Template registration.
+Any registration errors in these two steps will become multiplicative in the end result.
+T1 registration is well studied, however T1---DWI registration has few literature.
+This demonstrates the need to investigate the best method to minimize registration errors that were compatible with our image data format.
+
+\subsection{Measurements from tracts}
+Aggregating along-the-measure statistics is a form of clustering that requires projection of tract vertices onto a structural skeleton, or centroid.
+Common clustering methods calculate a polyline centroid based on different distancing functions between streamlines.
+These methods require tractography streamlines to have similar length, and will underperform when tractography structures contain sharp turns.
+We have explored a number of strategies, and considered adaptation of manifold-based methods, and point-cloud skeletal extraction method from computer graphics literature.
+These approaches, however, contain assumptions that do not suit our needs.
+For example, manifolds commonly consider that the latent structure of the data lies on a lower-dimensional surface, while tractography is commonly arranged in complex bundles.
+Moreover, point-cloud methods deal with point-cloud data obtained from 3D scanners, which situate on a 3D shell, while tractography points are distributed within a volume.
+White matter anatomy can have intricate crossings, multi-way join and diverging structures.
+These are difficult for existing 3D analytical methods, and require models with simplified assumptions to reduce the number of unknown parameters.
+A potentially promising future direction may be to automatically discover these more complex latent structures using deep learning methods.
 

### 2018-02-02 thesis/contents/hypotheses.tex
Updates from ShareLaTeX

--- 2017-11-24 thesis/contents/hypotheses.tex
@@
 
-The general aim of this thesis is quantify the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways in order to identify specific diffusivity pattern that differentiates between TN and non-TN CN V tissue.
-In achieve this aim, a generalizable group diffusion imaging tractography software framework must be developed to overcome the technical barrier of large scale group tractography in arbitrary populations.
-A number of confounds on data registration and measure must be resolved in order to gain confidence in the result of this methodology.
-These include a) establish the feasibility of diffusion tractography of brainstem CN V originating from the trigeminal nucleus; b) establish the best way to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects; c) Make possible non-linear registration of tractography geometries, and to quantify the geometry registration performance at the group level; d) apply previously established methodology to the reconstruction and quantification of the trigeminal sensory pathway.
+The general aim of this thesis is to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease.
+This necessitates detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls.
+With an increasing number of subjects, single subject analysis becomes cumbersome.
+Critical methodological advances are necessary for the feasibility of the general aim.
+These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy.
+Lastly, these methodological advances will be applied to the large-scale study of TN versus healthy control.
+The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, to identify specific diffusivity pattern that differentiates between TN and non-TN.
+The roadmap and rationale for the specific studies are described below.
+
+\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
+MS-TN is thought to relate pathophysiologically to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons.
+However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals.
+Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment.
+In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.
+
+\subsection{Main Aim} 
+Establish the feasibility of multi-tensor diffusion tractography in delineating the cisternal, root entry zone and brainstem segments of CN V and how these changes correlate with pain in MS-TN.
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Use multi-tensor tractography to delineate CN V into the the pontine brainstem, in order to quantify tissue diffusivity measurements in targeted sub-regions.
+    \item Differentiate TN and MS-TN based on the obtained CN V diffusivity measurements.
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item TN is known to relate to vascular compression of the cistern REZ, therefore we expect its diffusivity signature to occur in the cisternal region.
+    \item MS-TN should be related to brainstem MS lesions.
+Therefore CN V segments close to the lesions are expected to show more significant changes in tissue diffusivity, whereas segments far away from the lesions will not.
+    \item As a demyelinating disease, we expect to find a pattern of change in diffusivity signatures that reflects demyelination, including increase in RD, and decrease in FA. 
+\end{itemize}
+
+\section{Study II: Diffusion Weighted Image to T1 Co-registration Without Reversed-blip: Investigation of Best Practices}
+The task of establishing inter-subject anatomical correspondences is imperative for extending tractography to multiple subjects.
+Population studies with inter-subject T1 registration are well-established, where a normalized T1 template is generated from numerous subjects to facilitate data analysis.
+Similarly, DWI of each subject can be co-registered to their corresponding T1, to extend the T1 template space to tractography space.
+Therefore the DWI-T1 co-registration accuracy is crucial to the reliability of inter-subject tractography alignment.
+DWI-T1 co-registration is challenging due to the DWI eddy-current and EPI field distortions.
+While new DWI sequences with reversed-blips have been shown to reduce non-linear distortions, they have yet to be commonly adopted in clinical practice.
+While T1 inter-subject registration methodology is well documented, the feasibility, and strategy for DWI-T1 co-registration on sequence without reversed-blips have not been documented, and forms a potential bottleneck on the performance of any DWI-T1 population registration pipeline.
+
+\subsection{Main Aim}
+Establish the best strategy to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects.
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Quantify the degree of improvement in registration between affine-only versus symmetric-diffeomorphic registrations.
+    \item Determine which commonly-used DWI scalar image types can provide the best registration result when used as the registration-intermediate.
+    \item Determine if anisotropic-power image, an experimental DWI scalar map that closely resemble T1 contrast, can provide even more registration improvements. 
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item Symmetric-diffeomorphic registration will result in the best accuracy, but with limited asymptotic improvement over affine registration alone.
+    \item Averaged DWI should be the best scalar image, as it provides a balanced contrast between white and grey matter.
+    \item Anisotropic-power (AP) image will show more substantial improvement over the existing scalar images.
+\end{itemize}
+
+\section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
+The construction of an automated group tractography software framework depends on the establishment of the feasiblity of CN V brainstem delineation using multi-tensor tractography, and the optimal DWI-T1 co-registration strategy.
+The software aims to automate the tractography process on a large-scale, and also aims to streamline research workflow by providing region-of-interest, and tractography parameter management.
+It also provides the ability to report the tractography output at a group level both visually, as well as quantifying the output by an original scalar score called Normalized Overlap Score (NOS) score.
+
+\subsection{Main Aim}
+The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population.
+The tasks are performed  at a scale that would be prohibitive with manual methods.
+The comparisons determine the best tractography methods to use for different types of neuroanatomy.
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD in 42 subjects. 
+    \item Use the above techniques to study a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
+    \item Validate NOS scores by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item DTI is known to not able to delineate cross-fibers regions.
+If the group tractography works consistently, then its output should reflect such known algorithmic limitations, and DTI group tracts should not be able to perform well on more complex anatomies.
+The grouped tracts should however show consistent regional anatomical overlap across individuals.
+    \item By merging the tracts of many individuals, the tractography delinations can complement each other, and this may improve the anatomical fidelity over individual cases.
+    \item Specific tractography algorithms may be preferentially suited for the visualization of specific areas in the brain.
+For examples, deterministic tractography should preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
+    \item NOS score should correlate with expert ratings, and show improved rating consistency over human rater variability.
+\end{itemize}
+
+\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
+The developed methodology is applied to the study of the trigeminal sensory pathway.
+The method used is now thoroughly tested, and the best-suited tractography methods determined from the previous study.
+The study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group.
+The pathway is divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment.
+With the increased datapoints, more detailed along-the-tract analysis can be carried out on each segment.
+Following this, we use machine learning algorithms to auto-determine relevant segments that maximally differentiates TN from controls.
+
+\subsection{Main Aim}
+The aim of this project is the reconstruction of the trigeminal pathways from the cisternal segment to the level of S1 sensory cortex, and examination using Gaussian Process classification.
+
+\subsection{Specific Aim}
+\begin{itemize}
+    \item Delineate and visualize the three segments of the trigeminal sensory pathway: CN V, trigeminopontothalamic and  thalamocortical S1 segment. 
+    \item At each segment, we will look for microstructural features to distinguish the affected vs. unaffected sides.
+\end{itemize}
+
+\subsection{Specific Hypothesis}
+\begin{itemize}
+    \item To validate the study, CN V diffusivity signature should match findings from Study I.
+    \item At the level of CN V, the affected cistern segment should show differentiation from the unaffected side.
+    \item The affected S1 segment should differ from the unaffected side.
+    \item The affected TPT segment might show more differentiation near the end points, due to possible differences in trigeminal nucleus and thalamus diffusivity.
+However, its mid-segment diffusivity might be inconclusive, as the diffusivity metrics such as FA and RD might not be adequate in describing the complex crossings at the pontine decussation.
+\end{itemize}

### 2018-02-02 thesis/contents/reviews/mri.tex
Updates from ShareLaTeX

@@
 Protons, electrons and neutrons possess spin.
-Nuclei with odd number of protons or neutrons have non-zero nuclear spins; while those with even number of protons and neutrons have no spin.
+Nuclei with an odd number of protons or neutrons have non-zero nuclear spins; while those with an even number of protons and neutrons have no spin.
 Particles with opposite spins can pair up to produce an overall spin of zero.
@@
 This number of difference is called spin excess and the magnetic moments of the excess spins add to form the net magnetization vector $M_z$.
-When the spin system is at equilibrium, net magnetization lies along the z axis, parallel to $B_0$.
+When the spin system is at equilibrium, net magnetization lies along the z-axis, parallel to $B_0$.
 
@@
 \subsubsection{Relaxation}
-A transverse magnetization is introduced into the system that oscillates at radio frequency (RF) in order to modify the net magnetization, this is called the RF pulse.
+Transverse magnetization is introduced into the system that oscillates at radio frequency (RF) to modify the net magnetization, and this is called the RF pulse.
 When exposing the nuclear spin system to the energy of the RF pulse, the equilibrium of the spin system is disturbed.
@@
 The net magnetization now is denoted as $M_{xy}$ as it aligns with the xy-plane instead of the z-axis.
-Individual spins now precess in phase with the RF pulse, it gives rises to the MR signal in the receiver coil.
+Individual spins now precess in phase with the RF pulse, and it gives rises to the MR signal in the receiver coil.
 
-With the termination of the RF pulse, the nuclei in the excited state will return to the equilibrium condition, this process is called \textit{relaxation}. 
+With the termination of the RF pulse, the nuclei in the excited state will return to the equilibrium condition. This process is called \textit{relaxation}. 
 The energy transmitted decays rapidly by two relaxation processes:  spin-lattice and spin-spin relaxations. These are responsible for \textit{T1 relaxation} and \textit{T2 relaxation} respectively.

### 2018-02-02 thesis/contents/reviews/pain.tex
Updates from ShareLaTeX

@@
 The experience of pain is complex, consisting of interaction between nociceptive, emotional and cognitive processes.
-The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.
+The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\cite{Merskey2012}.
 
@@
  \centering
- \caption{Afferent pain pathways} 
+ \caption{Afferent pain pathways \protect\cite{kandel2000principles}}
  \label{fig:pain-pathways}
@@
 Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}.
-S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies.
+S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies \cite{Wager2013,Brodersen2012}.
 The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}.
@@
 \paragraph{Affective-Motivational}
-The cortical regions that are involved in affective-motivational pain procecssing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions.
+The cortical regions that are involved in affective-motivational pain processing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions.
 The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus.
@@
 These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}.
+The unpleasantness of pain has been thought to be a motivational force \cite{Auvray2010}, that biases actions towards avoidance.
+There is evidence that the motor cortex, which is involved in action planning, and the ACC are activated during pain \cite{Erpelding2013}.
+The cortical excitability of the motor cortex is also shown to be altered in chronic pain \cite{Parker2016}.
+The motivational system is also involved in pain-reward and involves the ACC, lateral orbitofrontal cortex (OFC), and the prefrontal cortex (PFC, both ventromedial and anterior aspects) \cite{Navratilova2015}.
+The nucleus accumbens (NAc) BOLD activity correlated with anticipation of pain relief \cite{Baliki2010}.
+The reward/motivational circuitry is thought to be disrupted in chronic pain, and such disruption is related to the emotional and cognitive deficits in chronic pain sufferers.
 

### 2018-02-02 thesis/contents/reviews/trigeminal.tex
Updates from ShareLaTeX

@@
  \centering
- \caption{Anatomy of the trigeminal nerve branches.} 
+ \caption{Anatomy of the trigeminal nerve branches. \protect\cite{gray1918anatomy}} 
  \label{fig:trig-branches} 
@@
 There is a slight lateral prevalence, with right-sided TN occurring in  in 60\% of the cases, and in rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.
-Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms .
+Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms \cite{Haller2016,Love2001}.
 However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the clinical expression of pain \cite{Desouza2013,Hodaie2013}.
@@
 Neuroimaging of the TN has been challenging, due to the small diameter of the CN V and since the contrast of the CNS trigeminal pathways is difficult to distinguish from the surrounding on conventional MRI \cite{Borges2010}.
+In conventional MRI, the brainstem nerve segment can only be deduced from its surrounding structures \cite{Casselman2008}.
+Sequences that can visualize myelinated structures are chosen, these include T2-weighted MRI  sequences such as FIESTA and CISS.
 Manual segmentation of T1 and T2 MRI requires great effort, with mixed results \cite{Miller2008a}.
 Diffusion MRI provided fiber orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}.
+Additionally, DTI showed strong ability to visualize spatial relationships in white matter anatomy, for example in delineating fornix sub-regions \cite{Chen2015c}, and in visualizing the distortion of CN VII/VIII by vestibular schwannomas \cite{Chen2011b}.
 

### 2018-02-02 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 
-\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
+\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia using Merged Tractography}
 \chaptermark{Study IV}
+\label{section:study4}
 
@@
 \textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial neuropathic pain syndrome involving the trigeminal sensory pathway.
-We demonstrate the use of a fully-automated group white-matter tractography, and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate patients and healthy controls.
+We demonstrate the use of a fully-automated group white-matter tractography and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate patients and healthy controls.
 
-\textbf{Methods:} We performed group merged tractography using SAGIT on 36 sex-matched TN patients (right-sided pain) and 36 controls.
-The trigeminal nerve (CN V), pontine decussation, and thalamocortical S1 pathways were reconstructed.
+\textbf{Methods:} We used SAGIT based group merged tractography for 36 sex-matched TN patients (right-sided pain) and 36 controls, examining the following white matter pathways: trigeminal nerve (CN V), pontine decussation, and thalamocortical fibers.
 GP classifiers were trained by scrolling a moving window over CN V and S1 tractography centroids.
-FA, GFA, RD, AD, and MD metrics were evaluated.
-Both TN vs.
-Control and Affect vs.
-Unaffected were analyzed.
+FA, GFA, RD, AD, and MD metrics were evaluated for both groups, analysing TN vs.
+Control and affected vs.
+unaffected groups.
 Classifiers that performed at greater-or-equal-to 75\% accuracy were accepted.
@@
 
-\textbf{Conclusions:} We have demonstrated the first instance of group-wise merged tractography analysis of TN vs.
-control,  as well as the first white-matter machine learning classification with GP.
+\textbf{Conclusions:} We  demonstrate the first instance of group-wise merged tractography analysis of TN vs.
+control and TN white-matter machine learning classification using GP.
 The method reproduced previous findings with finer-grained regional specificity.
@@
 Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CN V) branches.
-TN manifests classically in idiopathic fashion, or secondary to a range of neuropathology, including meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
+TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
 The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}.
@@
 The trigeminal sensory system involves both the discriminative touch and pain pathways.
-The discriminative pathway synapsis onto the primary trigeminal sensory nucleus, and then project to the contralateral VC thalamus, and then to the S1 cortex as we have delineated.
-The pain pathway synapsis caudally onto the trigeminal spinal nucleus, and decussate at the level of the medulla and spinal cord.
+Anatomicall, the discriminative pathway synapses onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex.
+The pain pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord.
 The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus.
@@
 \subsubsection{Data Analysis from Streamlines}
-Neuroimaging tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}.
-To obtain diffusivity metrics from tractography involves converting the streamline bundles into a volumetric spatial mask, and then measure the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}.
+White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}.
+Obtaining diffusivity metrics from tractography involves converting streamline bundles into a volumetric spatial mask, followed by measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}.
 However this approach discards the orientation information in the streamlines.
 The naive form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-through.
-This becomes an issue when considering that an average 3-tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm.
-In white matter structures such as CN V, and other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
+This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm.
+In white matter structures such as CN V as well as other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
 
-Consider that a tractography streamline propagation algorithm involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
+In this light, it must be considered that a tractography streamline propagation algorithm involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
 It is reasonable to consider these algorithms to be an interpolated subsampling scheme of the DWI volume that takes the diffusion-based orientation into account.
-Therefore, it is desirable to directly make use of the tractography streamlines for data measurement.
+Therefore, it would be desirable to directly make use of the tractography streamlines for data measurement.
 
 Along-the-tract analysis has been investigated by various authors \cite{Colby2012,ODonnell2009,Wang2015,Yeatman2012}.
-In these approaches, the bundles are selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and studies were limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and uncinate.
+In these approaches, the bundles were selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and uncinate.
 The measures were pooled onto tract centroid in each individual subjects, and then aggregated at the group level.
@@
 Some clustering methods do not provide a best-representative streamline centroid.
-Centroid identification involved resampling the streamlines into an identical number of points, and then identifying point correspondences via different distance measures.
-For these approaches, it is vital that the streamlines have similar length, and an assumption that the start and end of the streamlines across subjects are consistent.
+Centroid identification involves resampling the streamlines into an identical number of points, and then identifying point correspondences via different distance measures.
+For these approaches, streamlines require similar length, and an assumption that the start and end of the streamlines across subjects is consistent.
 However, pathological brain tissue may negatively affect tractography delineation.
-Therefore the geometry assumptions of such approaches may not be satisfied when extending them to study patient populations.
-Running time and memory constraints also need to be taken into consideration, when applying such algorithms at a group level.
-One method for tract clustering and centroid identification that is shown to have linear running time in respect to streamline count is QuickBundles \cite{Garyfallidis2012}.
-We have adopted QuickBundles to obtain S1 streamline centroids.
+Therefore the geometry assumptions of such approaches may not be satisfied when studying patient populations.
+Running time and memory constraints also need to be taken into consideration when applying such algorithms at a group level.
+We have adopted QuickBundles since it is a practical method of tract clustering and centroid identification with linear running time to obtain S1 streamline centroids.
+\cite{Garyfallidis2012}
 
@@
 We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework (Chen et al., 2016) to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning.
-We then merge the tracts into a template space using non-linear deformations, and then perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.
+We then merge the tracts into a template space using non-linear deformations, and  perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.
 
@@
 The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}.
-However, one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models.
+However, one disadvantage of these networks is in its inability to quantify uncertainty and provide interpretable models.
 Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}.
@@
 
-Gaussian Process is defined \cite{rasmussen2006gaussian}) as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
+Gaussian Process is defined \cite{rasmussen2006gaussian} as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
 It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
@@
 
-The covariance function is also termed called kernel function.
-The radial-basis function kernel or RBF kernel is widely used in machine learning tasks.
-GP has previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010}.
-It is particularly well suited for continuous data streams, such as tractography streamlines.
+The covariance function is termed called kernel function.
+The radial-basis function kernel (RBF kernel) is widely used in machine learning tasks.
+GP has previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010} and it is particularly well suited for continuous data streams, such as tractography streamlines.
 
@@
 \subsubsection{Acquisitions}
-The study recruited 36 idiopathic TN patients (28 females, 8 males; age 60.1$\pm$16 years; all right side pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).
+We studied 36 patients with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).
 All TN patients were scanned at 1 day pre-GK and 6 months post-GK.
@@
 The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}.
-
-We deployed selective automatic group integrated tractography (SAGIT) framework to automate preprocessing, registration, and tractography.
+We deployed selective automatic group integrated tractography (SAGIT) framework to automate preprocessing, registration, and tract generation.
 The processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations.
@@
 \subsubsection{Tractography}
-ROIs were defined on the template image to bilaterally delineate: 1) CN V (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).
+ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).
 
@@
 
-Tractography was first performed in the native DWI space for each individual.
-FA, AD, RD, MD, and GFA metrics were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.
+Tractography was first performed in the native DWI space for each subject.
+Diffusivity metrics of FA, AD, RD, MD, and GFA  were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.
 Additionally, subject ID and group ID scalar values were also inserted into each tractography files for identification.
@@
 A software library called “Fascicles” is created by the author to facilitate this process.
-The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and then all points were exported into a comma-delimited file.
+The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported into a comma-delimited file.
 The points were then transformed by applying the subject DWI to T1, and T1 to template transforms using ANTs’  applyTransformToPoints command \cite{Avants2009}.
@@
 Tractography centroids, which were the average curve best fit the tract bundle, was defined on the merged tracts in template space.
-CN V centroids were manually defined.
-The CN V centroids were defined by a list of 3 control points situated in the template world space.
+CN V centroids were manually defined, using 3 control points situated in the template world space.
 A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm.
@@
 The left versus right GP classifier placed left-right differences at the cistern segment, in agreement with the pathophysiology of TN.
-The regional findings on left versus controls suggest that there is bilateral differences in CN V myeline in unilateral TN, despite the lack of symptoms on the unaffected side.
+The regional findings on left versus controls suggest that there is bilateral differences in CN V myelin in unilateral TN, despite the lack of symptoms on the unaffected side.
 The left regions also situate into the brainstem segments.

### 2018-02-03 thesis/contents/tn_sagit.tex

@@
 
-\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia using Merged Tractography}
+\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia using Merged Group Tractography}
 \chaptermark{Study IV}
@@
 \section{Abstract}
-\textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial neuropathic pain syndrome involving the trigeminal sensory pathway.
-We demonstrate the use of a fully-automated group white-matter tractography and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate patients and healthy controls.
+\textbf{Introduction:}  Imaging of trigmeinal neuralgia (TN) has demonstrated key diffusivity alterations in the trigeminal nerve but the difficulties of group tractography analysis have previously prevented us from assessing the entire trigeminal pathway.
+We apply fully-automated group white-matter tractography to now image the trigeminal sensory pathway, and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate TN in patients and healthy controls.
 
@@
 unaffected groups.
-Classifiers that performed at greater-or-equal-to 75\% accuracy were accepted.
+Classifiers that performed at greater-or-equal-to 70\% accuracy were accepted.
 

### 2018-02-03 thesis/contents/tn_sagit.tex

@@
 
-\textbf{Results:} Both affected and unaffected trigeminal nerves in TN patients can be differentiated with up to 80\% accuracy from controls.
-Affected CN V differences were more focused in the cistern and root-transition-zone, whereas unaffected CN V show more focus towards the brainstem.
-The affected S1 showed focused MD differences towards the cortex with an accuracy of 85\%.
+\textbf{Results:} Both affected and unaffected trigeminal nerves in TN patients can be differentiated with 80\% accuracy from controls.
+Affected CN V FA segments that are differentiated from trons have more overlap towards the distal segments in the cistern and gangion, whereas unaffected CN V segment overlaps extend more towards the brainstem.
+The bilateral middle segment of S1 can be differentiated from controls.
+Where the affected S1 RD classifier achieved higher accuracy of 87\%.
 
@@
 The method reproduced previous findings with finer-grained regional specificity.
-The study also revealed specific diffusivity changes in deeper white matter.
+The study also revealed specific diffusivity changes in TPT and S1 white matter pathways.
 
@@
 The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}.
-We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is disrupted in MS-TN.
-In all previous TN studies, placements of region-of-interests in the CN V were pre-determined and manually placed.
-In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.
+We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.
 
 The trigeminal sensory system involves both the discriminative touch and pain pathways.
-Anatomicall, the discriminative pathway synapses onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex.
-The pain pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord.
+Anatomically, the discriminative pathway synapses onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex.
+The nociceptive pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord.
 The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus.
@@
 
-\subsubsection{Data Analysis from Streamlines}
+\subsubsection{Along-the-tract analysis}
 White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}.
@@
 However this approach discards the orientation information in the streamlines.
-The naive form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-through.
+The simplest form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-throughs.
 This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm.
@@
 
-The principal advantage of a streamline-based approach is that we can retain the pathway orientation from its geometry.
+The principal advantage of a streamline-based approach is that we can preserve the fiber orientations.
 White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis.
@@
 The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}.
-However, one disadvantage of these networks is in its inability to quantify uncertainty and provide interpretable models.
+However, one disadvantage of deep neural networks is in its inability to quantify uncertainty and provide interpretable models.
 Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}.
-In fact, it has been found that DNN with an infinite number of hidden units approximates GP \cite{neal1996priors}.
 
@@
 
+In all previous TN studies, placements of measurement region-of-interests in the CN V were pre-planned and manually placed by researchers.
+Manual tractography generation with manual ROI placements was a roadblock in performing more indepth group-wise tractography examination of the trigeminal pathways.
+We had developed and validated the SAGIT tractography framework to allow automated group tractography \cite{Chen2016}.
+In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.
+
 \section{Methods}
@@
 We studied 36 patients with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).
-All TN patients were scanned at 1 day pre-GK and 6 months post-GK.
+All patients underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment.
 MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions.
@@
 DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}.
-Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains.
+Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains \cite{Chen2016}.
 The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}.
-We deployed selective automatic group integrated tractography (SAGIT) framework to automate preprocessing, registration, and tract generation.
-The processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations.
+We deployed selective automatic group integrated tractography (SAGIT) framework \cite{Chen2016} to automate preprocessing, registration, and tract generation.
+The SAGIT processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations.
 The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}.
@@
 \subsubsection{Tractography}
-ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).
+ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibers must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary; it must exclude the brainstem).
 
-Constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} was used to delineate the three sets of anatomy.
+We used constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} to delineate the three defined anatomical regions.
+Deterministic and probabilistic streamline algorithms were used to delineate different regions, based on findings in the SAGIT study \cite{Chen2016}.
 Deterministic CST was applied to delineate CN V and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy.
-Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration.
+The parameters for each were as follows: Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration.
 Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = 45$\deg$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum lengh = 80 mm, initial cut-off = 0.15, cut-off = 0.15.
 Probabilistic tractography for TPT project was performed with step size = 1mm, and streamline count = 2000, and maximal length = 60 mm.
-
 Tractography was first performed in the native DWI space for each subject.
-Diffusivity metrics of FA, AD, RD, MD, and GFA  were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.
-Additionally, subject ID and group ID scalar values were also inserted into each tractography files for identification.
+FA, AD, RD, MD, and GFA were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.
 
@@
 All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis.
-A software library called “Fascicles” is created by the author to facilitate this process.
-The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported into a comma-delimited file.
-The points were then transformed by applying the subject DWI to T1, and T1 to template transforms using ANTs’  applyTransformToPoints command \cite{Avants2009}.
-The transformed points were then imported back into the SQLite database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved.
-Finally, all template-space tractography models across all the subjects are imported into a new SQLite database, and then exported as a single file for analysis.
+A software library\footnote{https://github.com/sinkpoint/fascicle} is created by the author to facilitate this process, the steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}.
+The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved.
+Finally, all template-space tractography models across all the subjects are merged into a single file.
 
 \subsubsection{Along-the-tract Analysis}
-The merged tract was first post-processed in 3D Slicer to remove erroneous and stray streamlines that will confound measurement.
+The merged tract was post-processed in 3D Slicer to remove erroneous streamlines that confounded measurement.
 Streamlines removed include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise.
+TPT streamlines that did not decussate at the level of the pons, as well as fibers that stray into the contralateral TGN were removed.
 S1 streamlines that stray medially towards other regions such as the corpus callosum were also removed.
-The TPT streamlines were not examined due to inconsistent reproducibility across subjects.
 
-Tractography centroids, which were the average curve best fit the tract bundle, was defined on the merged tracts in template space.
-CN V centroids were manually defined, using 3 control points situated in the template world space.
+Tractography centroids, which were the average curve best fit the tract bundle, were identified on the merged tracts in template space.
+CN V centroids were manually defined using 3 control points.
 A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm.
@@
 Diffusion metrics from the points were aggregated for each of the cross-sections.
-S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, with centroid length of 62.1 mm and segment length of 2.07 mm.
+S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, resulting in centroid length of 62.1 mm and segment length of 2.07 mm.
 
@@
 The diffusion values were subsequently normalized to zero mean with ranges $[-1, 1]$ for model training.
-Gaussian Process (GP) classification was exhaustively performed on CN V and S1 nerve segments to identify key segment differences.
+Gaussian Process (GP) classification was performed on CN V and S1 nerve segments to identify key segment differences.
 Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment.
@@
 \subsection{Results}
-\subsubsection{Group Tractography}
+\subsubsection{Tractography}
 Merged CN V delineations show anatomical consistency, and extended into the ipsilateral TGN.
@@
 The resulting pathway was not symmetrical, where the right TPT pathway diverts and envelopes the left (Figure \ref{fig:GPfigure2}).
-TPT streams lines were sparse, suggest low inter-subject reproducibility.
-Therefore no additional analysis was carried out for TPT.
 S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}).
@@
 
-\subsubsection{CN V}
-Both sides were distinguished from controls with classifiers that achieve greater than 75\% accuracy. 
-Classifiers on the right side have performance ranging from 77\% (GFA) to 80\% (MD) accuracy.
-The left classifiers achieved accuracy ranging from 75\% (FA) to 80\%(GFA).
-The right side shows more consistent overlap of segment ranges between positions 10--17, in the region of the cistern/REZ; while the left side shows less consistent range overlaps, and AD segment range from positions 3 to 14, which include the brainstem regions.
+\subsubsection{Diffusivity Differences}
+\paragraph{CN V}
+Both sides were distinguished from controls with classifiers that achieve around 80\% accuracy (Figure \ref{fig:GPfigure4}). 
+Classifiers on the right side have performance ranging from 76\% (GFA) to 86\% (FA) accuracy.
+The left classifiers achieved accuracy ranging from 79\% (MD) to 81\%(FA).
+The right side shows more overlap of segment ranges between positions 14--24, in the region of the cistern/gangion; while the left side shows segments that covers the entire the nerve, and extends to cover position 0--4 in the brainstem.
 Overall, the right side features extend further into the distal space, into the cistern and even the ganglion, while the left side extends more into the brainstem.
-Comparing left/right, only the FA classifier achieved greater than 75\% accuracy, where its segments range from 11 to 19, which is the cisternal segment (Figure \ref{fig:GPfigure4}).
-
-The best performing segments versus controls is GFA (segment 8--14) on the left (unaffected) side, with 80\% accuracy, 84\% precision, and AUC of 0.8; while MD (segment 10--17) performed the best on the right (affected) side, with 80\% accuracy, 84\% precision, and AUC of 0.79 (Table \ref{table:CN V}, Figure \ref{fig:GPfigure6}).
-The best performing left-versus-right classifier was FA (segment 11--19), with 75\% accuracy, 82\% precision, and AUC of 0.75.
+FA consistently achieved the highest accuracy in both sides, with right FA at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cister).
+Comparing left/right, the FA classifier achieved 75\% accuracy, where its segments range from 11 to 19, which is the cisternal segment.
+The GFA classifier achieved 70\% accuracy in region 9--12.
+Demonstrating that affected-unaffected diffusivity differences are found in the cistern/REZ (Table \ref{table:CN V}).
 
@@
 
-\subsubsection{S1}
-Both sides of the S1 face region pathway were able to be distinguished by classifiers with greater than 75\% accuracy. 
-The FA and RD classifiers at locations 617, representing the segment that emerge from the thalamus and merging into corona radiata, are common to both sides.
-The left side has higher classification accuracy of 80\% with FA, while the right side had a maximum accuracy of about 78\% with RD.
-The MD classifier meanwhile  is unique to the left side, and achieved an accuracy of 85\% in locations 16--23, near the grey-white matter boundary, closer to the cortex (Figure \ref{fig:GPfigure5}).
+\paragraph{Trigeminopontothalamic tract}
+Both left and right TPT pathways can be distinguished from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}).
+The right maximum accuracy was 85\% (FA), and minimum accuracy 83\% (RD), while the left maximum accuracy was 89\% (FA) and minimum 71\% (RD).
+The regions covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage.
+The left-right classifiers achieved maximum accuracy of 98\% (FA) and minimum of 88\% (MD).
+The left-right FA covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:tpt}).
 
-The best performing segment versus controls was MD (segment 16--23) on the left (affected) side, with 85\% accuracy, 88\% precision, and AUC of 0.83; whereas the best on the right (unaffected) side was FA (segment 6--16), with 78\% accuracy, 81\% precision, and AUC of 0.76 (Table \ref{table:s1}, Figure \ref{fig:GPfigure6}).
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\linewidth]{figure-GP-TPT.pdf}
+\caption{TPT classifier performance}
+\label{fig:GPfigureTPT}
+\end{figure}
+
+\begin{table}[ht]
+\centering
+\csvautotabular{tpt_table.csv}
+\caption{TPT GP classifiers performance data}
+\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
+\label{table:tpt}
+\end{table}
+
+\paragraph{S1}
+Both sides of the S1 face region pathway were able to be distinguished by classifiers with about 85\% accuracy.
+The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at psotion 8--26 (Figure \ref{fig:GPfigure5}).
+Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifieres ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:s1}).
+Both sides differ from controls in the branching region between the thalamus and cortex.
+Left-right classifiers showed maximum accuracy of 78\% (RD), and minimum of 76\% (FA).
+The left-right classifiers overlap in the regions 5--10, at the segment where the pathway emerges from the thalamus.
 
@@
 \end{table}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\linewidth]{figure-roc.pdf}
-\caption{ROC curves of the top performing classifiers}
-\label{fig:GPfigure6}
-\end{figure}
 

### 2018-02-05 thesis/contents/discussions.tex
Updates from ShareLaTeX

@@
 
-The biggest value of tractography to neurosugery is to assit in pre-operative planning.
+The biggest value of tractography to neurosugery is to assist in pre-operative planning.
 Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D.
 We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}.
+Another contribution is to visualize tissue changes from surgical interventions that can otherwise be invisible.
+There are cases where post-operative GammaKnife ablation target on the trigeminal nerve did not result in dicernable gadonenium enhancement in standard MR, but the nerve FA changes can be seen in the delineated tractography.
+Therefore tractography can provide both spatial, as well as tissue-based insights to the surgeons to aid in assessment.
 Tractography and 3D anatomical model ultimately are standard 3D geometries in computer graphics.

### 2018-02-05 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 A software library\footnote{https://github.com/sinkpoint/fascicle} is created by the author to facilitate this process, the steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}.
-The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved.
+The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group associations preserved.
 Finally, all template-space tractography models across all the subjects are merged into a single file.
@@
 The diffusivity measures of CN V reproduced our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls.
-The consistent regional overlaps in the cistern/REZ region of the right (affected) CN V in MD, RD, AD shows the ability for the GP classifiers to auto-converge to focal white-matter abnormality.
-The left versus right GP classifier placed left-right differences at the cistern segment, in agreement with the pathophysiology of TN.
+The consistent regional overlaps in the cistern/ganglion region of the right (affected) CN V showed the ability for the GP classifiers to auto-converge to focal white-matter abnormality.
+The left versus right GP classifier placed left-right differences at the cistern/REZ segment, in agreement with the pathophysiology of TN.
 The regional findings on left versus controls suggest that there is bilateral differences in CN V myelin in unilateral TN, despite the lack of symptoms on the unaffected side.
@@
 It is possible that the left CN V is affected by global descending pain modulation, and warrant further investigation.
+
+The TPT delineations showed that decussation pathway tractography is possibly asymmetric.
+It demonstrated very strong differences from the classifier results, with overlapping differences in the ascending segment.
+There also seemed to be diffusivity differences in the decussation segments between affected and unaffected sides.
+The decussaing fibers of the TPT emerges from the TGN, therefore the difference seemed reasonable.
+It is not clear however, whether the tractography asymmetry of the TPT is neuroanatomically factual, or an artifact from either algorithmic bias or partial volume bias of the underlying DWI.
+Resolving crossing fiber is still a difficult task in tractography, and while there are validations against MR phantoms, the anatomical validity of these methods can only be extrapolated, and the interpretation must involve anatomical priors from experts.
 
@@
 
-The TPT delineations showed that decussation pathway tractography is possibly asymmetric.
-It is not clear if the asymmetry is a neuroanatomically factual, or an artifact from either algorithmic bias or partial volume bias of the underlying DWI.
-Resolving crossing fiber is still a difficult task in tractography, and while there are validations against MR phantoms, the anatomical validity of these methods can only be extrapolated, and the interpretation must involve anatomical priors from experts.
-
 We included GFA to investigate its potential for measuring deep white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric.
-However S1 results demonstrated that it is not effective in pinpoint deep white-matter abnormalities.
-GFA performed well in CN V, but not to any substantially degree than the Gaussian-base diffusivity metrics.
+However results demonstrated that it is not effective in pinpoint deep white-matter abnormalities.
+GFA did not perform to any substantially degree than the Gaussian-base diffusivity metrics.
 

### 2018-02-05 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 \textbf{Results:} Both affected and unaffected trigeminal nerves in TN patients can be differentiated with 80\% accuracy from controls.
-Affected CN V FA segments that are differentiated from trons have more overlap towards the distal segments in the cistern and gangion, whereas unaffected CN V segment overlaps extend more towards the brainstem.
+Affected CN V FA segments that are differentiated from controls have more overlap towards the distal segments in the cistern and gangion, whereas unaffected CN V segment overlaps extend more towards the brainstem.
+Bilateral TPT differ from controls with at least 85\%.
+TPT left-right classification achieved 98\%.
 The bilateral middle segment of S1 can be differentiated from controls.

### 2018-02-21 thesis/contents/discussions.tex
Updates from ShareLaTeX

@@
 Therefore both diseases may share pathophysiology due to disruptions in myelin ultrastructure.
+Prior to the study, identification of the CN V nerve inside the brainstem from MR images were highly inconsistent, and the precision of multiple ROI placements could not be maintained.
+Multi-tensor tractography was instrumental in the feasibility of the method.
 The study opened the possibility that multi-tensor tractograpy aided analysis added a measure of objective assessment in examining pain conditions, and may facilitate similar studies that provide insight into the fundamental changes that occur at the level of white matter.
@@
 \cite{Desouza2013} also presented similar findings, where the unaffected RD, MD, and AD were found to differ from controls.
-Desouza performed the study with manual ROI placement at a single measurement position in the REZ, Whereas we performed GP classification along the nerve automatically.
-The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain.
-In many studies it has also been used as a basis of comparison.
+Desouza performed the study with manual ROI placement at a single measurement position in the REZ, whereas we performed GP classification along the nerve automatically.
+The GP classifier produced results by exhaustively examine moving windows at all position and sizes, and the results are auto-discovered by the algorithm, in contrast to the prior hand-placement of the measurement ROI in previous studies.
+The consistency in CN V findings in the location and diffusivity across both manual and automatic methods, suggest that the evidences for diffusivity changes in the unaffected CN V are convincing.
 This finding, however, is not surprising.
@@
 Moreover, in third-degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space.
-It is reasonable that these observable macroscopic changes can effect CNS tissue changes on the unaffected side.
+It is reasonable that these observable macroscopic disturbances can effect CNS tissue changes on the unaffected side.
 Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible.
-With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between affected and healthy controls.
+The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain.
+In many studies it has also been used as a basis of comparison.
+With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between pathological and healthy tissue.
 Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what degree of difference does it result in pain.
@@
 
-In Gamma-Knife radiosurgery treatment for TN, the target is on the cistern CN V segment, and situates proximally to the CN V REZ.
-Paradoxically, the GK treatment can be seen as a willful disruption of the CN V myelination.
-The question is, why would additional GK lesions on the CN V myelin effectively decrease TN pain.
-Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain.
-The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved.
-Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure.
-The force exerted onto the CN V from its neighbouring vasculature is therefore periodic.
-It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition".
-This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain.
-Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain.
-This model also explains that GK intervention disrupts the chance of resonance condition, thereby reducing TN pain.
-The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns.
-MVD, in this case can be explained as the physical disruption of the condition.
-
 \section{Tractography of the trigeminal nerve and brainstem fibers}
 
-The primary motivation for the study of the trigeminal nerve with tractography was the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
-Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissection.
+The primary motivation for the study of the trigeminal nerve with tractography was to overcome the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
+Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissections.
 
-Resolving cross fibers of the cerebral white matter was a fundamental advancement in diffusion tractography.
+The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging.
+It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid.
+Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks.
+Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be assessed.
+
+The brainstem CN V nerve in some subjects could be seen to bend caudally into the medulla.
+Althought not reported in Study IV, sometime we could delineate what appeared to be the spinal trigeminal decussation at the level of the medulla.
+The fibers were sparse, and not consistently reproducible across our subjects.
+Moreover, the mesencephalic trigeminal nucleus could be delineated in some subjects as well.
+Merge tractography demonstrated that these anatomical features have low reproducibility, and therefore they were not included as part of the study.
+In theory, with enough resolution, these smaller structures can be imaged.
+It would also be interesting to attempt to delineate the motor branch of the trigeminal nerve.
+
+Resolving cross-fibers of the cerebral white matter was a fundamental advancement in diffusion tractography.
 The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging.
@@
 
-The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging.
-It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid.
-Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks.
-Such manual methods were highly subjective and became problematic with patient images, where distortions to the nerve could not be assessed.
+The ability to segment the entire trigeminal nerve also opened the avenue to measure its tissue diffusivity from tractography directly.
+Moreover, this method can extend to any scalar images that can be registered to the DWI space.
+These include specific tissue-metrics such as myelin, G-ratio, and PET images.
+This extended these methods to facilitate disease studies well beyond TN.
+We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transited from pontine tissue to the cistern.
+The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure.
+In theory the number of divisions could be infinitely large, given that a centroid curve was identified.
+In practice the cross-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered.
+Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF.
+White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chaism.
+A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research.
+Turns out that the merged tractography strategy simplified this task by allowed a single curve to be defined manually in the template tractography space.
+This greatly reduced the amount of labour for researchers, as centroids no longer need to be defined repeatedly for each subject.
+Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.
 
-The ability to segment the entire trigeminal nerve also opens the avenue to measure its tissue diffusivity from tractography directly.
-Moreover, specific tissue-metrics such as myelin image, or PET images, can also be measured with confidence in reproducibility.
-This extends these methods to facilitate disease studies well beyond TN.
-We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transits from pontine tissue to the cistern.
-Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable, due to the noise in the cistern CSF.
-Turns out that the merged tractography strategy simplified this task by allowed a single manual curve to be defined manually in the template tractography space.
-This strategy can similarly be extended for obtaining along-the-track measure for other highly variable streamline bundles.
+\section{Machine Learning}
 
@@
 The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method.
-We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret without inducing biases.
+We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret.
+This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics.
+Eventually it was decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which contradicts frequentist assumptions.
 The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency.
 2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive.
+More specifically, the streamline measurements resemble time-series data, in that it exhibits positional dependency between measurements.
+Unlike a time-series however, the streamline measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency, as suppose to uni-directional in time.
+Posing the study as a supervised machine learning problem, the feature vector would be a N dimension vector representing N measures at each division, and a classification label.
+SVM would assume the features to be points in a N-dimensional space, and each measurement would assume independence.
+This makes the result of the SVM output difficult to interpret.
+The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output.
+Therefore the GP model is able to take the positional dependency into account, and the result output is also much more intuitive.
+Moreover, changing the GP classification model into a regression one is trivial.
+This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a patient's CN V, what is the probability that is resembles a pathological population, what about a healthy population?
+It is possible to answer these questions trivially with the GP model using regression.
+The full implication is outside of the scope of this thesis, but its implementation is has already been seeded within.
 
@@
 
-The pain neuromatrix, as studied by various functional and structural connectivity studies, distributes over the full range of neocortical areas.
-These areas are associated with complex white-matter pathways that extends to all regions of the brain.
-These include sharp curvatures of the anterior pain pathways of the ACC, the distant projections of the temporal cortices.
-Some pathways are just beyond the feasibility of current methods, these include the projections of the insula and claustrum, small decending pathways to PAG, and the spinal trigeminal pathways in the brainstem.
+The findings in S1 in Study IV (\ref{section:study4}) showed TN diffusivity differences in regions surrounding the thalamus, as well as in the mid-segments of the S1 pathways.
+These findings were in agreement with pain literature, where thalamus and S1/S2 activations are often reported as part of the pain neuromatrix.
+The TPT pathway resulted in differences near the TGN and the thalamus, areas that are expected to be involved in TN pain.
+It would be of great interest to re-examine the MS-TN population in these pathways, in order to determine any relationship between deeper white-matter MS-lesion, MS-TN, and non-TN pain in the people with MS.
 
-We have shown in Study III (\ref{section:study3}), for the first time, that different tractography algorithms are suited to different neuropathways when comparing merged reconstructions in 42 subjects.
+We have shown in Study III (\ref{section:study3}), for the first time, that different tractography algorithms are differentially suited to delineate different neuropathways when comparing merged reconstructions in 42 subjects.
 Maier-Hein et al came to similar conclusions after examining 96 reconstructions from a single set of simulated data \cite{Maier-Hein2017}.
@@
 
-We observed that merged tractography can delineate more complete pathway delineations than from any single subject.
-This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high quality templates.
-Therefore merged tractography may serve to produce high quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals.
-We have explored the use of SAGIT in generating the first neonatal porcine tractography template \cite{Zhong2016a}.
-Such template does not exist for humans to the best of our knowledge, and can be readily created from Human Connectome Project data.
-In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low resolution clinical acquisitions, these template tracts may be used as proxy to enable analysis.
+Quantification of tissue measures from tractography, as demonstrated in Study IV, implies the use of points in tract models as a sampling method for the scalar values in DWI space.
+We demonstrated in Study III, that different tratography algorithms exhibit different geometric point distributions.
+Eventhough we average the point-measures during along-the-tract analysis, which may be considered a form of smoothing/down-sampling.
+It is not clear whether consistency in comparing measures drawn from the same sampling distribution should be considered as a factor.
+In other words, given two tracts generatef from identical ROI definitions, with similar cross-sectional area, but of different algorithm, are comparisons between them still valid.
+Intuitively the algorithm choices and parameters should be identical, and that's what we did in Study III.
+However, there maybe cases where different parameters are better suited for certain pathologies, and this is may be a concern when deploying tractogrpahy analysis towards clinical diagnosis in the future, and therefore merit further investigation.
 
@@
 
-\section{Future development}
+\section{Future Directions}
 
 The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers,  are not limited to the application of TN.
-The methods can generalize to any anatomy where consistent tracts can be obtained at a group level.
+The methods can generalize to any white matter anatomy where consistent tracts can be obtained at a group level.
 The limitation of its examination power is often in the data, both in data volume, resolution, and quality.
 
-An immediate application is to re-examine TPT tracts, both to establish its diffusivity baseline, and to confirm its asymmetry.
-This can be immediately deployed on the Human Connectome Project (HCP) dataset.
-Additional examine in a diseased population requires the availability of clinical MR images that at minimum matches the parameters of HCP.
+Extensions to the current studies can follow three lines of investigations:
 
-For pain studies, detailed examination of the pain matrix, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest.
-Multi-modal and multi-variable study with functional connectivity with this technique would also be possible.
+I) Keep pushing the limit of anatomical delineations towards other areas of the trigeminal system.
+This include the spinal trigeminal pathways, which include the spinal trigeminal ganglion and its sub-gangions, and its medullary decussations.
+Re-examine TPT tracts, both to establish its diffusivity baseline, and to confirm its asymmetry.
+Image the medial pain pathways, and descending projections to the PAG.
+These tasks are currently limited by the resolution of availabile DWI images.
+Immediate gains may be obtained by using SAGIT to delineate these structure on the Human Connectome Project (HCP) dataset.
+Additionally, investigation into 7T DWI images will be necessary.
 
+II) Re-examine TN and other pain in the MS population with S1/S2 and TPT pathways.
 The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal additional insights in the MS population.
 Where focal lesion groups can be correlated with disease symptoms such as pain and motor disabilities.
+Comparitive studies with other trigeminal pain syndromes such as Herpetic neuralgia and trigeminal deafferentation pain would also be a valuable direction.
 
+III) Continue to advance machine learning to enable disease diagnosis and prognosis from white-matter tractography.
+Investigate how to diagnose and provide treatment targets from MR images with minimal turn around time.
+This can involve apply GP regression to the prediction of patient disease, and multi-modal learning, possibly with deep neural networks.
 Gaussian Process can be used as a regressor to model tissue parameters along merged white-matter bundles.
@@
 When combined with the end-to-end tractography extraction ability of SAGIT, this provides a basis to build a deployable framework for disease inference on new patients.
+
+We observed that merged tractography can delineate more complete pathway delineations than from any single subject.
+This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high quality templates.
+Therefore merged tractography may serve to produce high quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals.
+We have explored the use of SAGIT in generating the first neonatal porcine tractography template \cite{Zhong2016a}.
+Such template does not exist for humans to the best of our knowledge, and can be readily created from Human Connectome Project data.
+In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low resolution clinical acquisitions, these template tracts may provide a standard population distribution for GP regression, and can used as proxy to enable advanced analysis.
+
+The pain neuromatrix, as studied by various functional and structural connectivity studies, distributes over the full range of neocortical areas.
+These areas are associated with complex white-matter pathways that extends to all regions of the brain.
+These include sharp curvatures of the anterior pain pathways of the ACC, the distant projections of the temporal cortices.
+Some pathways are just beyond the feasibility of current methods, these include the projections of the insula and claustrum, small decending pathways to PAG, and the spinal trigeminal pathways in the brainstem.
+Detailed examination of the pain matrix, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest.
+Multi-modal and multi-variable study with functional connectivity with this technique would also be possible.
+
+From my observations across the studies in TN, I postulate a resonance hypothesis of TN.
+I am aware that there lacks substantial physiological measurements for these claims, nevertheless I feel it is important to state them here as a possible direction of investigation for the future: In Gamma-Knife radiosurgery treatment for TN, the target is on the cistern CN V segment, and situates proximally to the CN V REZ.
+Paradoxically, the GK treatment can be seen as a willful disruption of the CN V myelination.
+The question is, why would additional GK lesions on the CN V myelin effectively decrease TN pain.
+Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain.
+The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved.
+Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure.
+The force exerted onto the CN V from its neighbouring vasculature is therefore periodic.
+It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition".
+This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain.
+Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain.
+This model also explains that GK intervention disrupts the chance of resonance condition, thereby reducing TN pain.
+The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns.
+MVD, in this case can be explained as the physical disruption of the condition.
 

### 2018-02-21 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

--- 2018-02-05 thesis/contents/tn_sagit.tex
@@
 \section{Abstract}
-\textbf{Introduction:}  Imaging of trigmeinal neuralgia (TN) has demonstrated key diffusivity alterations in the trigeminal nerve but the difficulties of group tractography analysis have previously prevented us from assessing the entire trigeminal pathway.
-We apply fully-automated group white-matter tractography to now image the trigeminal sensory pathway, and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate TN in patients and healthy controls.
+\textbf{Introduction:}  Imaging of trigeminal neuralgia (TN) has demonstrated key diffusivity alterations in the trigeminal nerve but technical limitations in group tractography analysis have  prevented us from assessing the trigeminal nerve and its connections in the central nervous system.
+We apply fully-automated group white-matter tractography to image the trigeminal sensory pathway as well as machine learning classification using Gaussian Process (GP) classifier to pinpoint key white-matter diffusivity changes that maximally differentiate TN patients from healthy controls.
 
-\textbf{Methods:} We used SAGIT based group merged tractography for 36 sex-matched TN patients (right-sided pain) and 36 controls, examining the following white matter pathways: trigeminal nerve (CN V), pontine decussation, and thalamocortical fibers.
+\textbf{Methods:} We used SAGIT based group merged tractography for 36 sex-matched TN patients (right-sided pain) and 36 controls, examining the following trigeminal related white matter pathways: trigeminal nerve (CN V), pontine decussation, and thalamocortical fibers.
 GP classifiers were trained by scrolling a moving window over CN V and S1 tractography centroids.
-FA, GFA, RD, AD, and MD metrics were evaluated for both groups, analysing TN vs.
-Control and affected vs.
-unaffected groups.
-Classifiers that performed at greater-or-equal-to 70\% accuracy were accepted.
+FA, GFA, RD, AD, and MD metrics were evaluated for both groups, analyzing TN vs.
+control groups and affected vs.
+unaffected sides.
+Classifiers that performed at greater-or-equal-to 70\% accuracy were included.
 
-\textbf{Results:} Both affected and unaffected trigeminal nerves in TN patients can be differentiated with 80\% accuracy from controls.
-Affected CN V FA segments that are differentiated from trons have more overlap towards the distal segments in the cistern and gangion, whereas unaffected CN V segment overlaps extend more towards the brainstem.
-The bilateral middle segment of S1 can be differentiated from controls.
-Where the affected S1 RD classifier achieved higher accuracy of 87\%.
+\textbf{Results:} GP classifier consistently demonstrated bilateral trigeminal changes, differentiating them from controls with an accuracy of 80\%, even though the clinical expression of pain was strictly unilateral.
+Affected and unaffected sides could also be differentiated from each other with an accuracy of 75\%.
+Assessment of the trigeminal subregions demonstrated that the segments in affected CN V differ from controls from the region of the REZ and distally.
+Bilateral TPT could be distinguished from controls with at least 85\% accuracy while TPT left-right classification achieved 98\% accuracy.
+Bilateral S1 could be differentiated from controls, where the affected S1 RD classifier achieved an accuracy of 87\%.
 
-\textbf{Conclusions:} We  demonstrate the first instance of group-wise merged tractography analysis of TN vs.
-control and TN white-matter machine learning classification using GP.
-The method reproduced previous findings with finer-grained regional specificity.
-The study also revealed specific diffusivity changes in TPT and S1 white matter pathways.
+\textbf{Conclusions:} We  demonstrate the first instance of group-wise merged tractography analysis of TN white-matter machine learning classification using GP, and analysis of the complete trigeminal pathways from the pons to S1 cortex.
+The previous reported findings of bilateral changes in the REZ of TN patients is observed.
+Our study suggests additional, bilateral changes extend beyond the REZ and pinpoints these changes with greater detail in regional specificity.
 
@@
 TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
-The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}.
+The pathophysiology of classic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although classic TN can also occur without evidence of vascular compression \cite{Lee2014}.
 We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.
@@
 The simplest form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-throughs.
-This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm.
+This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm \cite{Neher2015,Wilkins2015}.
 In white matter structures such as CN V as well as other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
 
-In this light, it must be considered that a tractography streamline propagation algorithm involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
+In this light, a tractography streamline propagation algorithm must involve all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
 It is reasonable to consider these algorithms to be an interpolated subsampling scheme of the DWI volume that takes the diffusion-based orientation into account.
@@
 The measures were pooled onto tract centroid in each individual subjects, and then aggregated at the group level.
-The global tractography approach is unintuitive for anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is clear.
+The global tractography approach is unintuitive for anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is predetermined.
 A more focused ROI based delineation approach is therefore preferred in these circumstances.
@@
 Centroid identification involves resampling the streamlines into an identical number of points, and then identifying point correspondences via different distance measures.
-For these approaches, streamlines require similar length, and an assumption that the start and end of the streamlines across subjects is consistent.
-However, pathological brain tissue may negatively affect tractography delineation.
-Therefore the geometry assumptions of such approaches may not be satisfied when studying patient populations.
-Running time and memory constraints also need to be taken into consideration when applying such algorithms at a group level.
-We have adopted QuickBundles since it is a practical method of tract clustering and centroid identification with linear running time to obtain S1 streamline centroids.
+For these approaches, streamlines require similar lengths, and an assumption that the start and end of the streamlines across subjects is consistent.
+However, pathological brain tissue may  affect tractography delineation.
+Therefore the geometry assumptions of these clustering approaches may not be satisfied when studying patient populations.
+Computation time and memory constraints also need to be taken into consideration when applying such algorithms at a group level.
+We have adopted QuickBundles since it is a practical method of tract clustering and centroid identification with linear running time to obtain TPT and S1 streamline centroids.
 \cite{Garyfallidis2012}
 
-The principal advantage of a streamline-based approach is that we can preserve the fiber orientations.
+The principal advantage of a streamline-based approach is that we can preserve the geometric orientations of the pathways.
 White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis.
-We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework (Chen et al., 2016) to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning.
+We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework \cite{Chen2016} to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning.
 We then merge the tracts into a template space using non-linear deformations, and perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.
@@
 Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}.
+
+The streamline measurements resembles time-series data, in that it exhibits positional dependency between measurements.
+Unlike a time-series however, the measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency.
+Posing the study as a supervised machine learning problem, the feature vector would be a N dimensional vector representing N measures at each division.
+The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output.
+Therefore the GP model is able to take the positional dependency of the measurements into account, and is well suited for streamline-based classification.
 
@@
 \section{Methods}
+
 \subsubsection{Acquisitions}
@@
 DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-method.pdf}
+\caption{Diagram of the processing steps}
+\label{fig:GPMethods}
+\end{figure}
 
@@
 All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis.
-A software library\footnote{https://github.com/sinkpoint/fascicle} is created by the author to facilitate this process, the steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}.
+We have previously created a software library\footnote{https://github.com/sinkpoint/fascicle} to facilitate this process, the steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}.
 The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group associations preserved.
 Finally, all template-space tractography models across all the subjects are merged into a single file.
-
-\subsubsection{Along-the-tract Analysis}
-The merged tract was post-processed in 3D Slicer to remove erroneous streamlines that confounded measurement.
-Streamlines removed include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise.
+The merged tract was post-processed in 3D Slicer to remove anatomically erroneous streamlines.
+These include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise.
 TPT streamlines that did not decussate at the level of the pons, as well as fibers that stray into the contralateral TGN were removed.
@@
 
+\subsubsection{Along-the-tract Analysis}
 Tractography centroids, which were the average curve best fit the tract bundle, were identified on the merged tracts in template space.
+The spline was then resampled into 30 points and became the centroid definition.
+Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids.
+Diffusion metrics from the tractography points were averaged for each of the cross-sections.
 CN V centroids were manually defined using 3 control points.
-A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm.
-The spline is then resampled into 30 points that become the centroid definition with segment length of 1.3 mm.
-Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids.
-Diffusion metrics from the points were aggregated for each of the cross-sections.
-S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, resulting in centroid length of 62.1 mm and segment length of 2.07 mm.
+A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}.
+TPT and S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30.
 
@@
 
-Diffusivity metrics of affected (right side) and unaffected CN V and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides.
+Diffusivity metrics of affected (right side) and unaffected CN V, TPT, and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides.
 The affected and unaffected sides in TN patients were also compared with each other.
 Age was regressed out by using the residuals of the linear regression model from the controls group.
-The diffusion values were subsequently normalized to zero mean with ranges $[-1, 1]$ for model training.
-Gaussian Process (GP) classification was performed on CN V and S1 nerve segments to identify key segment differences.
+The diffusion values were subsequently normalized to zero-mean with ranges $[-1, 1]$ for model training.
+Gaussian Process (GP) classification was performed to identify key segment differences.
 Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment.
@@
 The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed segment.
-Segments accuracy greater than 75\% were accepted, and the segment length with the greatest accuracy was then determined.
+Segments accuracy greater than 70\% were accepted, and the segment length with the greatest accuracy was then determined.
 
 \subsection{Results}
-\subsubsection{Tractography}
+\subsubsection{Anatomical Delineations}
 Merged CN V delineations show anatomical consistency, and extended into the ipsilateral TGN.
-Changes in FA along the streamlines demonstrate clear differences between cistern and brainstem segments, suggesting good agreement in anatomical registration ( Figure \ref{fig:GPfigure1}).
-The TPT projections, when filtered strictly from the TGN to the ipsilateral VC thalamus region, showed distinct bundles of decussating pathways between left and right.
-The resulting pathway was not symmetrical, where the right TPT pathway diverts and envelopes the left (Figure \ref{fig:GPfigure2}).
-S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}).
+Changes in FA along the streamlines demonstrate clear differences between cistern and brainstem segments, suggesting good agreement in anatomical registration (Figure \ref{fig:GPfigure1}).
+The CN V centroid curve length was determined as 38.91 mm, with the length of each subdivision being 1.3 mm.
+The TPT projections from the TGN to the ipsilateral VC thalamus region showed distinct bundles of decussating pathways in both left and right projections.
+The resulting pathways were not symmetrical, where the right TPT pathway visibly diverts and envelopes the left (Figure \ref{fig:GPfigure2}).
+S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}). The S1 centroid curve length was calculated to 62.1 mm and sub-division length was 2.07 mm.
 
@@
 \includegraphics[width=\linewidth]{cnv-inferior-view.png}
-\caption{Inferior view of the merged CN V delineations.}
+\caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines.
+The expect increase in FA as the fibers enter the brainstem can be visualized, and denotes good registration between subjects.
+}
 \label{fig:GPfigure1}
@@
 \includegraphics[width=\linewidth]{view-decussation.png}
-\caption{Close up of the TPT decussation. Red: Right TPT; Blue: Left TPT. Note the Right TPT diverts around the left TPT bundle. }
+\caption{Focused view of the TPT decussation. Red: Right TPT; Blue: Left TPT. Note the Right TPT diverts around the left TPT bundle. }
 \label{fig:GPfigure2}
@@
 \includegraphics[width=\linewidth]{view-left-right.png}
-\caption{The complete trigeminal discriminatory sensory pathway. }
+\caption{The complete trigeminal discriminatory sensory pathway, FA is depicted by the colour scales (Red: right side; Blue: left side).
+The S1 pathways projects to the lateral face region of the S1 cortex.
+}
 \label{fig:GPfigure3}
@@
 \paragraph{CN V}
-Both sides were distinguished from controls with classifiers that achieve around 80\% accuracy (Figure \ref{fig:GPfigure4}). 
-Classifiers on the right side have performance ranging from 76\% (GFA) to 86\% (FA) accuracy.
-The left classifiers achieved accuracy ranging from 79\% (MD) to 81\%(FA).
-The right side shows more overlap of segment ranges between positions 14--24, in the region of the cistern/gangion; while the left side shows segments that covers the entire the nerve, and extends to cover position 0--4 in the brainstem.
-Overall, the right side features extend further into the distal space, into the cistern and even the ganglion, while the left side extends more into the brainstem.
-FA consistently achieved the highest accuracy in both sides, with right FA at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cister).
-Comparing left/right, the FA classifier achieved 75\% accuracy, where its segments range from 11 to 19, which is the cisternal segment.
+The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves.Both sides were distinguished from controls with classifiers that achieved around 80\% accuracy (Figure \ref{fig:GPfigure4}).
+Classifiers on the right side had accuracie ranging from 76\% (GFA) to 86\% (FA).
+The left classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA).
+The right side showed more overlap of segment ranges between positions 14--24, in the region of the cistern/gangion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4 in the brainstem.
+Overall, the right side classifiers extend further into the distal space, into the cistern and even the ganglion, while the left side extended more into the brainstem.
+FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cister).
+The FA classifier achieved 75\% accuracy in differentiating affected versus unaffected nerves, where its segments ranged from 11 to 19, which was the cisternal segment.
 The GFA classifier achieved 70\% accuracy in region 9--12.
-Demonstrating that affected-unaffected diffusivity differences are found in the cistern/REZ (Table \ref{table:CN V}).
+Demonstrating that affected-unaffected diffusivity differences were found in the cistern/REZ (Table \ref{table:CN V}).
 
@@
 \paragraph{Trigeminopontothalamic tract}
-Both left and right TPT pathways can be distinguished from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}).
+Both left and right TPT pathways could be distinguished from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}).
 The right maximum accuracy was 85\% (FA), and minimum accuracy 83\% (RD), while the left maximum accuracy was 89\% (FA) and minimum 71\% (RD).
@@
 The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at psotion 8--26 (Figure \ref{fig:GPfigure5}).
-Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifieres ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:s1}).
+Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifiers ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:s1}).
 Both sides differ from controls in the branching region between the thalamus and cortex.
@@
 This suggests that cross-metric features should be considered in future studies.
-This would drastically increase the feature complexity, and thereby require subsequent demand in the number of subjects.
+This would drastically increase the feature complexity, and thereby demand substantial increase in the number of subjects.
 

### 2018-02-22 thesis/contents/discussions.tex
Updates from ShareLaTeX

@@
 
-\item We are the first to image the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography.
-Prior to this work, CN V tractography was limited to single-tensor DTI models and cannot be delineated into the pons.
-Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increase the range of trigeminal-related pathology, for example MS-TN, that can be studied with neuroimaging.
+\item Imaging of the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography.
+Prior to this work, CN V tractography was limited to single-tensor DTI models and trigeminal fibers could not be visualized as they enter into the pons.
+Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increasing the range of study of trigeminal-related pathology that can be studied with neuroimaging, such as MS-TN.
 
-\item The study, for the first time, clearly differentiated TN and MS-TN by sub-segment diffusivity differences, between cistern (TN) and brainstem peri-lesional (MS-TN) segments.
-This distinction in the trigeminal REZ between these two conditions have been suspected, but never confirmed.
-The study is the first to suggest that disruptions along the CN V in different locations may trigger similar pain symptoms.
+\item Differentiation between TN and MS-TN by examining sub-segment diffusivities, and distinguishing between cistern (TN) and brainstem peri-lesional (MS-TN) segments.
+This pathophysiological distinction between these two entities had not been previously demonstrated using neuroimaging tools, nor had it been correlated with the clinical expression of pain.
 
-\item We overcame the major technical barrier to detailed diffusion tractography study of the trigeminal pathway in a large population group, by developing and validate the Selective Automated Group Integrated Tractography (SAGIT) software platform.
-In the process, we determined the optimal T1-DWI co-registration methodology with clinical DWI sequences.
-Prior to this project, the co-registration method has never been exhaustively determined and reported in literature.
+\item Overcoming the major technical barrier to diffusion tractography study of the trigeminal pathway in a large population group, by developing and validating the Selective Automated Group Integrated Tractography (SAGIT) software platform.
+For the process, we determined the optimal T1-DWI co-registration methodology with clinical DWI sequences.
+Prior to this project, the optimal co-registration method had not been exhaustively studied.
 
-\item We are the first to perform merged group tractography study of the trigeminal nerve, and its primary sensory pathway, and the first to conduct such study in TN, and the first to perform Gaussian Process machine learning to the study of TN.
-We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in its cross-sectional detail.
-We discovered that the unaffected CN V diffusivity differs from controls as well as the affected side and that the affected TPT pathway and S1 pathway diffusivity differs from controls.
+\item Merged group tractography study of the trigeminal nerve and its primary sensory pathway as a novel use of Gaussian Process machine learning to the study of TN.
+We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in cross-sectional detail.
+Our report defines distinct bilateral changes in trigeminal neuralgia subjects, as well as changes between affected and unaffected sides.
 
@@
 
-The primary motivation for the study of the trigeminal nerve with tractography was to overcome the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
-Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissections.
+The study of trigeminal neuralgia has revolutionarized the approach to the investigation of this unique pain disorder.
+Additionally, an important motivation for the use of tractography to study of the trigeminal nerve was the ability to overcome the significant limitations of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
+Prior to the use of tractography, the fibers of the trigeminal nerve and connected pathways could only be inferred from anatomical studies or postmortem dissections.
 
-The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging.
-It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid.
-Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks.
-Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be assessed.
+The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus signified that for the first time we could use software methods to segment the pontine trigeminal anatomy in a reliable manner.
+Prior to this, the ROI identification was done visually, and sometimes with simple heuristics methods based on distance from surrounding landmarks.
+Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be always assessed.
 
@@
 
-Resolving cross-fibers of the cerebral white matter was a fundamental advancement in diffusion tractography.
+Diffusion tensor imaging revolutionarized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter is a limitation of diffusion tractography.
 The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging.
-A widely adopted technique was probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map.
+Techniques to improve visualization of white matter fibers consist in probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map.
 This kind of image-based probabilistic tractography was useful for whole-brain white matter studies.
@@
 SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions.
-This meant that the methods are more robust to noise, and therefore require less number of gradient acquisitions.
+This meant that these methods are more robust to noise, and therefore require less number of gradient acquisitions.
 CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions.
@@
 
-These key algorithmic developments decidedly brought cross-fiber tractography in clinical neuroimaging to the realm of possibility.
-This is significant as the vast majority of imaging of neuropathologies in patients only exist as clinical acquisitions.
-We took advantage of this opportunity and explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST.
-The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also that the trigeminal nerve has a number of unique properties as a tractography target: 1) delineation quality is easy to judge due to its simple morphology.
-2) it situates in both single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.
+These algorithmic developments brought imaging of crossing-fibers using tractography to the realm of possibility.
+We explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST.
+The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images.
+2) it courses through single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.
 
-The ability to segment the entire trigeminal nerve also opened the avenue to measure its tissue diffusivity from tractography directly.
+The segmentation of the subregions of the trigeminal pathway allowed for the possibility of measuring the respective diffusivities.
 Moreover, this method can extend to any scalar images that can be registered to the DWI space.
@@
 This extended these methods to facilitate disease studies well beyond TN.
-We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transited from pontine tissue to the cistern.
+We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
 The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure.

### 2018-02-22 thesis/contents/tn_sagit.tex

@@
 The spline was then resampled into 30 points and became the centroid definition.
-Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids.
-Diffusion metrics from the tractography points were averaged for each of the cross-sections.
+Nearest-neighbor point assignment was used to divide the merged tract into 30 sections along the centroid.
+Diffusion metrics from the tractography points were averaged for each of the sections.
 CN V centroids were manually defined using 3 control points.
@@
 TPT and S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-path-analysis.pdf}
+\caption{Diagram for the along-the-tract data processing.}
+\label{fig:GPfigure-method-centroid}
+\end{figure}
 

### 2018-02-23 thesis/contents/discussions.tex

@@
 
-The segmentation of the subregions of the trigeminal pathway allowed for the possibility of measuring the respective diffusivities.
+The segmentation of the subregions of the trigeminal pathway has allowed for the possibility of measuring the respective diffusivities.
 Moreover, this method can extend to any scalar images that can be registered to the DWI space.
 These include specific tissue-metrics such as myelin, G-ratio, and PET images.
-This extended these methods to facilitate disease studies well beyond TN.
-We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
+This methodology is therefore highly applicable and not specific to the study of TN.
+We divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
 The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure.
@@
 I) Keep pushing the limit of anatomical delineations towards other areas of the trigeminal system.
-This include the spinal trigeminal pathways, which include the spinal trigeminal ganglion and its sub-gangions, and its medullary decussations.
+This include the spinal trigeminal pathways, which include the spinal trigeminal ganglion and its sub-ganglions, and its medullary decussations.
 Re-examine TPT tracts, both to establish its diffusivity baseline, and to confirm its asymmetry.

### 2018-02-23 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 Gaussian Process (GP) classification was performed to identify key segment differences.
-Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment.
+Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment (Figure \ref{fig:GPfigure-gp-train}).
 The entire machine learning process was performed using Scikit-learn machine learning software framework.
-For each model, the subjects were split into train and validation sets using stratified 10-fold cross-validation.
-The GP model was initiated with Matern kernel (length scale = 1), and the model fitted using the training set.
+For each window, the subjects were split into train and validation sets using stratified 10-fold cross-validation, and a separate GP classifier was trained for each fold.
+GP models were initiated with Matern kernel (length scale = 1).
 The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed segment.
-Segments accuracy greater than 70\% were accepted, and the segment length with the greatest accuracy was then determined.
+Segments accuracy greater than 70\% were accepted, and the window segment with the greatest accuracy was then determined.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-GP-training.pdf}
+\caption{Diagram for generating the best performing classifier for a particular window size.
+The data for each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position.
+Finally the window position with the best accuracy was selected.
+The same process is repeated on every diffusivity metric, and for all window sizes ranging from 5 to 30.
+}
+\label{fig:GPfigure-gp-train}
+\end{figure}
 

### 2018-02-23 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 The trigeminal sensory system involves both the discriminative touch and pain pathways.
-Anatomically, the discriminative pathway synapses onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex.
+Pain and discriminitive touch stimuli trigger innervations of the three trigeminal nerve branches (V1, V2, and V3) in the head and face, these primary affrents converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion) in the middle cranial fossa.
+The discriminitve touch afferents then pass through the cisternal segment of CN V, which then enters the pons, and synapse onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex.
 The nociceptive pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord.
@@
 \subsubsection{Along-the-tract Analysis}
-Tractography centroids, which were the average curve best fit the tract bundle, were identified on the merged tracts in template space.
-The spline was then resampled into 30 points and became the centroid definition.
-Nearest-neighbor point assignment was used to divide the merged tract into 30 sections along the centroid.
-Diffusion metrics from the tractography points were averaged for each of the sections.
+For the purpose of quantifying the trigeminal pathways, a centroid was defined for the merged tractography of each pathway (Figure \ref{fig:GPfigure-method-centroid}).
+The centroids were defined as the average curve that best fit the tract bundles, and were in template space.
 CN V centroids were manually defined using 3 control points.
 A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}.
+The spline was then resampled into 30 points.
 TPT and S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30.
+Nearest-neighbor point assignment was used to cluster  the points of the merged tract model to the centroid control points, to obtain the corresponding 30 sections along each centroid.
+Diffusion metrics from the tractography points were averaged for each of the sections for each subject.
 
@@
 The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves.Both sides were distinguished from controls with classifiers that achieved around 80\% accuracy (Figure \ref{fig:GPfigure4}).
-Classifiers on the right side had accuracie ranging from 76\% (GFA) to 86\% (FA).
+Classifiers on the right side had accuracies ranging from 76\% (GFA) to 86\% (FA).
 The left classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA).
-The right side showed more overlap of segment ranges between positions 14--24, in the region of the cistern/gangion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4 in the brainstem.
+The right side showed more overlap of segment ranges between positions 14--24, in the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4 in the brainstem.
 Overall, the right side classifiers extend further into the distal space, into the cistern and even the ganglion, while the left side extended more into the brainstem.
-FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cister).
+FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cistern).
 The FA classifier achieved 75\% accuracy in differentiating affected versus unaffected nerves, where its segments ranged from 11 to 19, which was the cisternal segment.
@@
 \includegraphics[width=\linewidth]{figure-GP-CNV.pdf}
-\caption{CN V classifier performance}
+\caption{CN V classifier performances.
+The left panel shows the coverage (y-axis) and accuracies (x-axis) of the GP classifiers with greater than 70\% accuracy.
+The middle panel is a heatmap of the FA values of the corresponding centroid positions.
+The right panel shows the centroid and point clusterings for anatomical reference.
+The colours of the points correspond to their respective cluster index on a rainbow spectrum.
+}
 \label{fig:GPfigure4}
@@
 \paragraph{Trigeminopontothalamic tract}
-Both left and right TPT pathways could be distinguished from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}).
-The right maximum accuracy was 85\% (FA), and minimum accuracy 83\% (RD), while the left maximum accuracy was 89\% (FA) and minimum 71\% (RD).
-The regions covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage.
-The left-right classifiers achieved maximum accuracy of 98\% (FA) and minimum of 88\% (MD).
-The left-right FA covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:tpt}).
+Classifiers could distinguish bilateral TPT pathways (affected and unaffected) from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}).
+The right FA classifier achieved the greatest accuracy at 85\%, and right RD classifier was least accuracy at 83\%.
+The left FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuray at 71\%.
+The classifiers covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage.
+The left-versus-right classifiers achieved maximum accuracy of 98\% (FA) and minimum of 88\% (MD).
+The FA classifier covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:tpt}).
 
@@
 \includegraphics[width=\linewidth]{figure-GP-TPT.pdf}
-\caption{TPT classifier performance}
+\caption{TPT classifier performance, see Figure \ref{fig:GPfigure4} for details on annotations.}
 \label{fig:GPfigureTPT}
@@
 \csvautotabular{tpt_table.csv}
-\caption{TPT GP classifiers performance data}
 \caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
@@
 \paragraph{S1}
-Both sides of the S1 face region pathway were able to be distinguished by classifiers with about 85\% accuracy.
-The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at psotion 8--26 (Figure \ref{fig:GPfigure5}).
+Classifiers distinguished bilateral S1 face region pathways from controls with about 85\% accuracy.
+The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at position 8--26 (Figure \ref{fig:GPfigure5}).
 Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifiers ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:s1}).
-Both sides differ from controls in the branching region between the thalamus and cortex.
-Left-right classifiers showed maximum accuracy of 78\% (RD), and minimum of 76\% (FA).
-The left-right classifiers overlap in the regions 5--10, at the segment where the pathway emerges from the thalamus.
+Classifiers of both sides covered the the branching region between the thalamus and cortex.
+Left-versus-right classifiers showed maximum accuracy of 78\% (RD), and minimum of 76\% (FA).
+The left-versus-right classifiers overlap in the regions 5--10, at the segment where the pathway emerges from the thalamus.
 
@@
 \includegraphics[width=\linewidth]{figure-GP-S1.pdf}
-\caption{S1 classifier performance}
+\caption{S1 classifier performance, see Figure \ref{fig:GPfigure4} for details on annotations.}
 \label{fig:GPfigure5}

### 2018-02-23 thesis/contents/discussions.tex
Updates from ShareLaTeX

@@
 \item Merged group tractography study of the trigeminal nerve and its primary sensory pathway as a novel use of Gaussian Process machine learning to the study of TN.
-We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in cross-sectional detail.
+We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (thalamocortical, S1) white patter pathways in sub-regional detail.
 Our report defines distinct bilateral changes in trigeminal neuralgia subjects, as well as changes between affected and unaffected sides.
@@
 Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be always assessed.
+Improvements in tractography of the brainstem trigeminal nerve allowed the shift towards more algorithmic driven and automated segmentation and ROI placement.
 
-The brainstem CN V nerve in some subjects could be seen to bend caudally into the medulla.
-Althought not reported in Study IV, sometime we could delineate what appeared to be the spinal trigeminal decussation at the level of the medulla.
-The fibers were sparse, and not consistently reproducible across our subjects.
-Moreover, the mesencephalic trigeminal nucleus could be delineated in some subjects as well.
-Merge tractography demonstrated that these anatomical features have low reproducibility, and therefore they were not included as part of the study.
-In theory, with enough resolution, these smaller structures can be imaged.
-It would also be interesting to attempt to delineate the motor branch of the trigeminal nerve.
-
-Diffusion tensor imaging revolutionarized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter is a limitation of diffusion tractography.
-The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging.
+Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter is a limitation of diffusion tractography.
+The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure therefore posed a fundamental challenge for neuroimaging.
 Techniques to improve visualization of white matter fibers consist in probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map.
@@
 The acquisition time of these sequences was impractical for the clinical environment.
-Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
+Reduction in acquisition time for these algorithms is currently an active area of research.
 The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development.
@@
 The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images.
-2) it courses through single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.
+2) it courses through simple (cistern) and complex (pons) fiber environments; 3) it is a prominent cranial nerve, and therefore a good benchmark for imaging small anatomy.
 
-The segmentation of the subregions of the trigeminal pathway has allowed for the possibility of measuring the respective diffusivities.
-Moreover, this method can extend to any scalar images that can be registered to the DWI space.
-These include specific tissue-metrics such as myelin, G-ratio, and PET images.
-This methodology is therefore highly applicable and not specific to the study of TN.
 We divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
@@
 In theory the number of divisions could be infinitely large, given that a centroid curve was identified.
-In practice the cross-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered.
+In practice the sub-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered.
 Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF.
-White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chaism.
+White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chiasm.
 A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research.
@@
 Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.
+
+The segmentation of the subregions of the trigeminal pathway has allowed for the possibility of measuring the respective diffusivities.
+Moreover, this method can extend to any scalar images that can be registered to the DWI space.
+These may include specific tissue-metrics such as myelin, G-ratio, and PET images.
+This methodology is therefore highly applicable and not specific to the study of TN.
 
@@
 This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics.
-Eventually it was decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which contradicts frequentist assumptions.
+Eventually it was decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which may not fit with frequentist assumptions.
 The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency.
@@
 
-The biggest value of tractography to neurosugery is to assist in pre-operative planning.
 Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D.
@@
 From my observations across the studies in TN, I postulate a resonance hypothesis of TN.
-I am aware that there lacks substantial physiological measurements for these claims, nevertheless I feel it is important to state them here as a possible direction of investigation for the future: In Gamma-Knife radiosurgery treatment for TN, the target is on the cistern CN V segment, and situates proximally to the CN V REZ.
-Paradoxically, the GK treatment can be seen as a willful disruption of the CN V myelination.
-The question is, why would additional GK lesions on the CN V myelin effectively decrease TN pain.
+I am aware that there lacks substantial physiological measurements for these claims, nevertheless I feel it is important to state them here as a possible direction of investigation for the future: In surgical interventions for TN, the target is on the cistern CN V segment, and situates close to the CN V REZ.
+Paradoxically, the intervention can be seen as a willful disruption of the CN V myelination.
+The question is, why would additional disruptions on the CN V myelin effectively decrease TN pain.
 Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain.
@@
 Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain.
-This model also explains that GK intervention disrupts the chance of resonance condition, thereby reducing TN pain.
+This model also explains that surgical intervention disrupts the chance of resonance condition, thereby reducing TN pain.
 The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns.

### 2018-02-23 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}.
-The spline was then resampled into 30 points.
+The spline was then resampled into 30 subpoints.
 TPT and S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30.
-Nearest-neighbor point assignment was used to cluster  the points of the merged tract model to the centroid control points, to obtain the corresponding 30 sections along each centroid.
-Diffusion metrics from the tractography points were averaged for each of the sections for each subject.
+Nearest-neighbor point assignment was used to cluster the points of the merged tract model to the centroid subpoints, to obtain the corresponding 30 divisions along each centroid.
+Diffusion metrics from the tractography points were averaged for each of the divisions for each subject.
 
@@
 \includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-path-analysis.pdf}
-\caption{Diagram for the along-the-tract data processing.}
+\caption{Diagram for the along-the-tract data processing.
+Process starts with 2 specific goals: 1) Derive the spatial structure that best represent the merged streamlines, 2) Deriving sub-regional tissue information from these.
+A) First the best-fit curve is obtained to represent the collective streamlines, which is defined as the centroid.
+B) In areas where this can be automated, Quickbundles was used.
+For the CN V, the centroid was defined manually given the unpredictability of the lengths of streamlines.
+C) Representative drawing of the centroid onto the streamline points, representing the collective spatial information.
+The centroid is resampled to 30 subpoints.
+D) To derive tissue information, a nearest neighbor algorithm was used to assign streamline points to form 30 divisions.
+E) The average diffusivity metrics was pooled for each of the 36 subjects at each division.
+The colours represent the spatial position of each of the 30 divisions, and the streamline points that were assigned to each particular division.
+The data was then used for machine learning pipeline (see Figure \ref{fig:GPfigure-gp-train}).
+}
 \label{fig:GPfigure-method-centroid}
@@
 \caption{Diagram for generating the best performing classifier for a particular window size.
-The data for each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position.
+The data was derived from Figure \ref{fig:GPfigure-method-centroid}, each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position.
 Finally the window position with the best accuracy was selected.
@@
 The GFA classifier achieved 70\% accuracy in region 9--12.
-Demonstrating that affected-unaffected diffusivity differences were found in the cistern/REZ (Table \ref{table:CN V}).
+Demonstrating that affected-unaffected diffusivity differences were found in the cistern/REZ (Table \ref{table:GP}).
 
@@
 
-\begin{table}[ht]
-\centering
-\csvautotabular{cnv_table.csv}
-\caption{CN V GP classifier performance data}
-\caption*{List of the best accuracy for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
-\label{table:CN V}
-\end{table}
-
 \paragraph{Trigeminopontothalamic tract}
@@
 The left-versus-right classifiers achieved maximum accuracy of 98\% (FA) and minimum of 88\% (MD).
-The FA classifier covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:tpt}).
+The FA classifier covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:GP}).
 
@@
 
-\begin{table}[ht]
-\centering
-\csvautotabular{tpt_table.csv}
-\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
-\label{table:tpt}
-\end{table}
-
 \paragraph{S1}
@@
 The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at position 8--26 (Figure \ref{fig:GPfigure5}).
-Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifiers ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:s1}).
+Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifiers ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:GP}).
 Classifiers of both sides covered the the branching region between the thalamus and cortex.
@@
 \centering
-\csvautotabular{s1_table.csv}
-\caption{S1 GP classifiers performance data}
-\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
-\label{table:s1}
+\csvautotabular{tn_sagit_table.csv}
+\caption{List of the best GP models for different diffusivity meatures and window segments.
+Only models with at least 70\% accuracy were accepted.
+Precision and f1 scores are also provided for reference}
+\label{table:GP}
 \end{table}

### 2018-02-23 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 \caption{Diagram for generating the best performing classifier for a particular window size.
-The data was derived from Figure \ref{fig:GPfigure-method-centroid}, each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position.
+The data was derived from Figure \ref{fig:GPfigure-method-centroid}, where the top bar represents data from each centroid division, where each position represents the anatomical region of the trigeminal pathways within it.
+Each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position.
 Finally the window position with the best accuracy was selected.
@@
 \includegraphics[width=\linewidth]{cnv-inferior-view.png}
-\caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines.
-The expect increase in FA as the fibers enter the brainstem can be visualized, and denotes good registration between subjects.
+\caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
+The expected increase in FA as the fibers enter the brainstem can be visualized, and denotes good registration between subjects.
 }
@@
 \includegraphics[width=\linewidth]{view-decussation.png}
-\caption{Focused view of the TPT decussation. Red: Right TPT; Blue: Left TPT. Note the Right TPT diverts around the left TPT bundle. }
+\caption{Focused view of the TPT and its decussation, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
+Colours were chosen arbitrarily to denote left and right pathways, however the colour intensity represents FA values as depicted by FA scale, top right corner.
+Red: Right TPT; Blue: Left TPT.
+Note the Right TPT appears to divert around the left TPT bundle.
+}
 \label{fig:GPfigure2}
@@
 \includegraphics[width=\linewidth]{view-left-right.png}
-\caption{The complete trigeminal discriminatory sensory pathway, FA is depicted by the colour scales (Red: right side; Blue: left side).
+\caption{Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.
+Images obtained prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
+FA is depicted by the colour scales (Red: right side; Blue: left side).
 The S1 pathways projects to the lateral face region of the S1 cortex.

### 2018-02-23 thesis/contents/tn_sagit.tex

@@
 \caption{CN V classifier performances.
-The left panel shows the coverage (y-axis) and accuracies (x-axis) of the GP classifiers with greater than 70\% accuracy.
-The middle panel is a heatmap of the FA values of the corresponding centroid positions.
-The right panel shows the centroid and point clusterings for anatomical reference.
-The colours of the points correspond to their respective cluster index on a rainbow spectrum.
-}
+This figure consists of 3 panels.
+Panel A, y axis represents the anatomical subregions of the trigeminal nerve as defined by the centroid subdivisions (see Figure \ref{fig:GPfigure-method-centroid}).
+Panel A, x-axis denotes the accuracy of the classifiers (scale of 0-1, with 1 representing 100\%)).
+Each diffusivity is denoted with respect to the window size (see Figure \ref{fig:GPfigure-method-centroid}) and position along the centroid where the accuracy is derived from.
+3 instances are studied: comparison of left (unaffected) vs.
+control (average of both trigeminal nerves), comparison of right (affected) vs.
+control and comparison of left (unaffected) vs.
+right (affected).
+The colours for the diffusivity in each panel in A is arbitrary.
+Panel B is a heatmap (raw visualization) of FA values of the corresponding centroid positions.
+Panel C shows the centroid and point assignments for anatomical reference.
+The colours of the points correspond to their respective subdivision position as described in figure \ref{fig:GPfigure-method-centroid} and \ref{fig:GPfigure-method-centroid}} Throughout the figure, the position of the centroid subdivisions are aligned and expressed with respect to the anatomical position as per panel B.
 \label{fig:GPfigure4}
@@
 The right FA classifier achieved the greatest accuracy at 85\%, and right RD classifier was least accuracy at 83\%.
-The left FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuray at 71\%.
+The left FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuracy at 71\%.
 The classifiers covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage.
@@
 \csvautotabular{tn_sagit_table.csv}
-\caption{List of the best GP models for different diffusivity meatures and window segments.
+\caption{List of the best GP models for different diffusivity measures and window segments.
 Only models with at least 70\% accuracy were accepted.

### 2018-02-23 thesis/contents/tn_sagit.tex

@@
 \csvautotabular{tn_sagit_table.csv}
-\caption{List of the best GP models for different diffusivity measures and window segments.
+\caption{List of the best GP models for different diffusivity metrics and centroid windows.
 Only models with at least 70\% accuracy were accepted.
-Precision and f1 scores are also provided for reference}
+Table is sorted by anatomical region, classifier type and accuracy.
+Start and end denote the beginning and end of the window subdivision positions along the centroid.
+Length is the number of subdivisions covered by the window.
+Precision and f1 scores are provided for reference}
 \label{table:GP}
@@
 \section{Discussion}
+We have demonstrated the use of a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy strictly from white-matter measures.
+Our work is in agreement with previous findings of TN diffusivity changes along the CN V with much higher regional specificity.
+We also examined TPT and S1 projections of the trigeminal sensory pathway.
 The study is the first instance of machine learning discrimination of TN vs.
 Controls with specific white matter analysis, using group-wise merged along-the-track tractography.
-In essence, we have demonstrated a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy solely from white-matter measures.
-We reproduced previous findings of TN diffusivity changes along the CN V with much higher regional specificity.
-We also examined TPT and S1 projections of the trigeminal sensory pathway.
 
-The diffusivity measures of CN V reproduced our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls.
+The diffusivity measures of CN V agreed with our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls.
 The consistent regional overlaps in the cistern/ganglion region of the right (affected) CN V showed the ability for the GP classifiers to auto-converge to focal white-matter abnormality.
-The left versus right GP classifier placed left-right differences at the cistern/REZ segment, in agreement with the pathophysiology of TN.
-The regional findings on left versus controls suggest that there is bilateral differences in CN V myelin in unilateral TN, despite the lack of symptoms on the unaffected side.
+The GP classifier placed affected vs unaffected differences at the cistern/REZ segment, in agreement with the pathophysiology of TN.
+The findings on unaffected versus controls suggest that there is bilateral differences in CN V myelin in unilateral TN, despite the lack of symptoms on the unaffected side.
 The left regions also situate into the brainstem segments.
@@
 There also seemed to be diffusivity differences in the decussation segments between affected and unaffected sides.
-The decussaing fibers of the TPT emerges from the TGN, therefore the difference seemed reasonable.
+The decussating fibers of the TPT emerges from the TGN, therefore the difference seemed reasonable.
 It is not clear however, whether the tractography asymmetry of the TPT is neuroanatomically factual, or an artifact from either algorithmic bias or partial volume bias of the underlying DWI.

### 2018-02-26 thesis/contents/discussions.tex
Updates from ShareLaTeX

@@
 
-\item Overcoming the major technical barrier to diffusion tractography study of the trigeminal pathway in a large population group, by developing and validating the Selective Automated Group Integrated Tractography (SAGIT) software platform.
+\item Overcoming the major technical barriers to diffusion tractography study of the trigeminal pathway in a large population group, by developing and validating the Selective Automated Group Integrated Tractography (SAGIT) software platform.
 For the process, we determined the optimal T1-DWI co-registration methodology with clinical DWI sequences.
-Prior to this project, the optimal co-registration method had not been exhaustively studied.
 
 \item Merged group tractography study of the trigeminal nerve and its primary sensory pathway as a novel use of Gaussian Process machine learning to the study of TN.
-We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (thalamocortical, S1) white patter pathways in sub-regional detail.
-Our report defines distinct bilateral changes in trigeminal neuralgia subjects, as well as changes between affected and unaffected sides.
+This study involved spatial defintion of groups tracts in space, as well as measurement of diffusivity metrics from the subregions of trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (thalamocortical, S1).
+We demonstrate distinct bilateral changes in trigeminal neuralgia subjects, as well as differences between affected and unaffected sides.
 
@@
 
-\section{Trigeminal Neuralgia}
+\section{A new approach to the study of Trigeminal Neuralgia}
 
-The findings in Study I (Chapter \ref{section:study1}), which aimed to identify trigeminal nerve diffusivity signatures in TN and MS-TN, identified for the first time a method that can distinguish between microstructural changes in these conditions, and revealed localized CN V diffusivity differences in MS-TN and idiopathic TN.
-The possible implication is that disruption anywhere along both the cisternal and pontine segments of the trigeminal nerve may lead to similar TN-like pain syndromes.
-Central demyelination in the trigeminal root transition zone is associated with vascular compression in TN.
-Moreover MS is a classic myelin inflammatory disease.
-Therefore both diseases may share pathophysiology due to disruptions in myelin ultrastructure.
-Prior to the study, identification of the CN V nerve inside the brainstem from MR images were highly inconsistent, and the precision of multiple ROI placements could not be maintained.
+The introduction of DTI in the study of TN has revolutionized our understanding of this disorder.
+Moving away from clinical studies alone, DTI is able to derive metrics that provide important information on the microstructure of the nerve which can then be related to specific processes such as changes in myelination or the microstructure of the axons.
+Thus DTI provides an important objective measure in the study of pain, which was unavailable up unto this point.
+The earliest work on the feasibility of accurate reconstruction of cranial nerve with DTI tractography \cite{Hodaie2010} in 2010.
+This study was able to image CN II, III, and importantly CN V.
+Similarly, we were able to establish the methods to image the TRG and the cisternal CN V with manual delineation, followed by visualization of the CN V postganglionic GK target, prior and after surgery \cite{Hodaie2012g}.
+We found alternations in both the CN V reconstruction, as well as FA in the surgical target location, that improved our understanding beyond what is visible with post-surgical gadolinium enhancements.
+This has now led our team to study whether CN V GK target diffusivity at 6 months post-surgery can predict long-term pain relief in TN \cite{Tohyama2018}.
+Similarly whether the diffusivity changes are specific enough that they can serve as predictors of outcome \cite{Hung2017}.
+The tractography study of TN has been directly linked with the study of gray and white matter abnormalities of the TN brain \cite{Desouza2013c}.
+The value of DTI studies in TN is therefore unique, and creates a strong context for the work done in this thesis.
+
+The findings in Study I (Chapter \ref{section:study1}), point to the clear role of DTI in distinguishing the localized, focal changes in diffusivity that characterize TN and MS-TN.
+While TN is associated with compression of the REZ and possible associated focal central demyelination \cite{Devor2002a,Peker2006}, MS is associated with inflammation and myelin changes resulting in CNS plaques.
+While both diseases involve the same clinical expression of pain, the pathophysiology of pain differs, but this had not been shown directly using DTI based imaging techniques.
+From a technical perspective, the identification of the CN V nerve inside the brainstem from MR images had been previously highly inconsistent, and the precision of multiple ROI placements could not be maintained.
 Multi-tensor tractography was instrumental in the feasibility of the method.
@@
 
-Study IV (Chapter \ref{section:study4}), which applied group tractography to the study of the trigeminal sensory pathway, further suggested that the unaffected CN V diffusivity is also different from that of controls.
-DeSouza et al.
-\cite{Desouza2013} also presented similar findings, where the unaffected RD, MD, and AD were found to differ from controls.
-Desouza performed the study with manual ROI placement at a single measurement position in the REZ, whereas we performed GP classification along the nerve automatically.
-The GP classifier produced results by exhaustively examine moving windows at all position and sizes, and the results are auto-discovered by the algorithm, in contrast to the prior hand-placement of the measurement ROI in previous studies.
-The consistency in CN V findings in the location and diffusivity across both manual and automatic methods, suggest that the evidences for diffusivity changes in the unaffected CN V are convincing.
-This finding, however, is not surprising.
-It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases.
-Moreover, in third-degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space.
-It is reasonable that these observable macroscopic disturbances can effect CNS tissue changes on the unaffected side.
-Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible.
+As a result of the methodology developed in Study I, a recent publication details the investigation of whether presurgical CN V segment diffusivity values can predict response to TN surgery \cite{Hung2017}.
+The important study found that lowered cistern MD and AD pre-surgery was correlated with positive surgical outcome, whereas lowered REZ FA and increased pontine AD was correlated with the lack of response from surgery.
+This provided further evidence of the role of cisternal CN V diffusivities in the study of TN.
+The fact that the novel finding of pontine diffusivity in classic TN favours a non-responder outcome points to the role of CNS alterations in classic TN.
+
+Study IV (Chapter \ref{section:study4}), applied group tractography to the study of the trigeminal sensory pathway, and highlighted the observation that the unaffected CN V diffusivity in TN differs from that of controls.
+There have been other studies suggesting bilateral abnormalities in TN.
+A group level diffusivity study of TN was conducted by DeSouza et al.
+in 2013 \cite{Desouza2013} presented similar findings.
+DeSouza manually defined ROI in the trigeminal REZ, as well as performing full-brain Tract-Based Spatial Statistics.
+Bilateral differences in the REZ between TN patients and controls were found in MD, RD, and AD.
+Wide-spread white matter differences were also found, especially in the corpus callosum, cingulum, and longitudinal fasciculus.
+DeSouza had also found increased cortical and grey-matter volume changes in TN patients \cite{Desouza2013c}, including increase in contralateral S1 and thalamus volumes.
+Using SAGIT in study IV, the ROIs were projected to each individual subject using an automated pipeline.
+The GP classifier produced results by exhaustively examine moving windows at all position and sizes, and the results are auto-discovered by the algorithm, in contrast to manual-placement of the measurement ROI in previous studies.
+There is therefore consistency in CN V findings in the location and diffusivity across both manual and automated methods.
+
+The observation of bilateral differences has important implications in TN pain.
+From previous studies, it is clear that altered diffusivities in the cisternal and pontine segments of CN V are correlated with TN pain in the affected side.
+However there are now a number of studies that point to bilateral REZ/pontine CN V diffusivities being altered in unilateral TN, although each point to significant differences between affected and unaffected sides as well.
+The exact nature of bilateral diffusivity alterations is not yet clear.
+Perhaps the earliest indication that TN can affect the nerves bilaterally is a clinical paper by the Burchiel group, where they point to a surprisingly large number of patients with unilateral TN that have evidence of bilateral neurovascular compression, in contrast with healthy controls \cite{Miller2009}.
+Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, result in central changes in the trigeminal nucleus and its associated pathways.
+
 The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain.
@@
 With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between pathological and healthy tissue.
-Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what degree of difference does it result in pain.
+Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what point does the clinical expression of pain start.
 Investigations into bilateral TN pain, and vascular compression in the absence of pain, may reveal additional insight into the relationship between white matter changes and TN pain in general.
+The newly developed SAGIT and machine learning methods can additionally be leveraged to conduct large cohort studies, that allow investigation into the brain regions identified by DeSouza et al, as well as examining important areas in the modulation of pain, such as the insula and its relationship to TN.
+
+Finally, with the methods developed throughout this thesis, future studies are no longer restricted by manual ROI placements for each subject, and tract subregions can be studied in detail.
+This opens the avenue to the study of associated pathways of TN pain, and allow the development of more sophisticated prediction models to further the understanding, and treatment of TN.
 
@@
 
-The study of trigeminal neuralgia has revolutionarized the approach to the investigation of this unique pain disorder.
-Additionally, an important motivation for the use of tractography to study of the trigeminal nerve was the ability to overcome the significant limitations of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
+An important motivation for the use of tractography to study the trigeminal nerve was the ability to overcome the significant limitations of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
 Prior to the use of tractography, the fibers of the trigeminal nerve and connected pathways could only be inferred from anatomical studies or postmortem dissections.
@@
 
-Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter is a limitation of diffusion tractography.
-The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure therefore posed a fundamental challenge for neuroimaging.
-Techniques to improve visualization of white matter fibers consist in probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map.
-This kind of image-based probabilistic tractography was useful for whole-brain white matter studies.
-However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways.
-Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves.
-Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition.
-The acquisition time of these sequences was impractical for the clinical environment.
+Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter remains a limitation of diffusion tractography.
+The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure therefore poses a fundamental challenge for neuroimaging.
+Techniques to improve visualization of white matter fibers consist of probabilistic tractography, which uses tissue priors to sample a large number of directions at each step stochastically, and combines the final result as a volumetric probability map.
+This kind of image-based probabilistic tractography has proved useful for whole-brain white matter studies.
+However, it is constrained to the resolution of the diffusion acquisition, which can be insufficient for localized and small diameter pathways.
+Moreover, the algorithm depends on pre-generated tissue-priors, which excludes cranial nerves.
+Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempt to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions where required) in the diffusion acquisition.
+The acquisition time of these sequences can be impractical for the clinical environment.
 Reduction in acquisition time for these algorithms is currently an active area of research.
-The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development.
-SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions.
-This meant that these methods are more robust to noise, and therefore require less number of gradient acquisitions.
+The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), are a significant development.
+SH-based methods model the fiber orientation distribution as a decomposition of SH basis functions.
+This means that these methods are more robust to noise, and therefore require less number of gradient acquisitions.
 CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions.
@@
 
-We divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
+As part of our focusing on ???
+and for the purposes of imaging the trigeminal pathways, where a significant number of crossing fibers exist, we divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
 The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure.
 In theory the number of divisions could be infinitely large, given that a centroid curve was identified.
-In practice the sub-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered.
-Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF.
-White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chiasm.
-A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research.
-Turns out that the merged tractography strategy simplified this task by allowed a single curve to be defined manually in the template tractography space.
-This greatly reduced the amount of labour for researchers, as centroids no longer need to be defined repeatedly for each subject.
+Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths are variable.
+White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales.
+A generalized algorithm that can reliably calculate a centroid for these structure is of great value and yet still a topic of research.
+Merged tractography  simplified this task by allowing a single curve to be defined manually in the template tractography space.
+This greatly simplied the task, since centroids no longer need to be defined repeatedly for each subject.
 Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.
@@
 Moreover, this method can extend to any scalar images that can be registered to the DWI space.
-These may include specific tissue-metrics such as myelin, G-ratio, and PET images.
+These may include specific tissue-metrics such as myelin and G-ratio images.
 This methodology is therefore highly applicable and not specific to the study of TN.
 
-\section{Machine Learning}
+\section{Neuroimaging of deep white matter}
 
-We applied machine learning to the analysis of TN in Study IV in order to avoid our own confirmation bias.
-The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method.
-We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret.
-This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics.
-Eventually it was decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which may not fit with frequentist assumptions.
-The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency.
-2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive.
-More specifically, the streamline measurements resemble time-series data, in that it exhibits positional dependency between measurements.
-Unlike a time-series however, the streamline measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency, as suppose to uni-directional in time.
-Posing the study as a supervised machine learning problem, the feature vector would be a N dimension vector representing N measures at each division, and a classification label.
-SVM would assume the features to be points in a N-dimensional space, and each measurement would assume independence.
-This makes the result of the SVM output difficult to interpret.
-The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output.
-Therefore the GP model is able to take the positional dependency into account, and the result output is also much more intuitive.
-Moreover, changing the GP classification model into a regression one is trivial.
-This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a patient's CN V, what is the probability that is resembles a pathological population, what about a healthy population?
-It is possible to answer these questions trivially with the GP model using regression.
-The full implication is outside of the scope of this thesis, but its implementation is has already been seeded within.
-
-\section{Neuroimaging of deeper white matter for the study of pain}
-
-The findings in S1 in Study IV (\ref{section:study4}) showed TN diffusivity differences in regions surrounding the thalamus, as well as in the mid-segments of the S1 pathways.
+The findings in S1 in Study IV (Chapter \ref{section:study4}) showed TN diffusivity differences in regions surrounding the thalamus, as well as in the mid-segments of the S1 pathways.
 These findings were in agreement with pain literature, where thalamus and S1/S2 activations are often reported as part of the pain neuromatrix.
@@
 
-We have shown in Study III (\ref{section:study3}), for the first time, that different tractography algorithms are differentially suited to delineate different neuropathways when comparing merged reconstructions in 42 subjects.
+We have shown in Study III (Chapter \ref{section:study3}) that different tractography algorithms are suited to delineate different neuropathways when comparing merged reconstructions in 42 subjects.
 Maier-Hein et al came to similar conclusions after examining 96 reconstructions from a single set of simulated data \cite{Maier-Hein2017}.
-Therefore care must be taken to ensure tractography parameters are targeted to the anatomy of study, as more consistent delineation lead to more robust tissue measurements.
+As tractography is applied to the investigation of white matter elsewhere in the brain, care must be taken to ensure  parameters are targeted to the anatomy of study, as more consistent delineation leads to more robust tissue measurements.
 
 Quantification of tissue measures from tractography, as demonstrated in Study IV, implies the use of points in tract models as a sampling method for the scalar values in DWI space.
-We demonstrated in Study III, that different tratography algorithms exhibit different geometric point distributions.
-Eventhough we average the point-measures during along-the-tract analysis, which may be considered a form of smoothing/down-sampling.
-It is not clear whether consistency in comparing measures drawn from the same sampling distribution should be considered as a factor.
-In other words, given two tracts generatef from identical ROI definitions, with similar cross-sectional area, but of different algorithm, are comparisons between them still valid.
-Intuitively the algorithm choices and parameters should be identical, and that's what we did in Study III.
-However, there maybe cases where different parameters are better suited for certain pathologies, and this is may be a concern when deploying tractogrpahy analysis towards clinical diagnosis in the future, and therefore merit further investigation.
+We demonstrated in Study III, that different tractography algorithms exhibit different geometric point distributions.
+In other words, given two tracts generated from identical ROI definitions, with similar cross-sectional area, but of different algorithm, the underlying spacial distribution of the points are different.
+It can be argued that we average the point-metrics during along-the-tract analysis, which may be considered a form of smoothing/down-sampling, and therefore the spatial distribution of the tractography algorithm might not be important.
+However, until the exact influence of spatial distribution have on the measured distribution of diffusivity metrics can be understood, comparing between similar tracts from two different tractography algorithms should be avoided.
 
-\section{Neurosurgery and Tractography}
+\section{Machine Learning}
 
-Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D.
-We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}.
-Another contribution is to visualize tissue changes from surgical interventions that can otherwise be invisible.
-There are cases where post-operative GammaKnife ablation target on the trigeminal nerve did not result in dicernable gadonenium enhancement in standard MR, but the nerve FA changes can be seen in the delineated tractography.
-Therefore tractography can provide both spatial, as well as tissue-based insights to the surgeons to aid in assessment.
-Tractography and 3D anatomical model ultimately are standard 3D geometries in computer graphics.
-Therefore advances in 3D printing can be easily adopted in neurosurgery to allow surgeons to physically examine tractography reconstructions.
-On the otherhand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making.
-Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.
+The availability of the SAGIT, full automation of tractography processing enabled a dramatic increase in the volume of data available for statistical analysis.
+This lead to some challenges in the scalability of the analytical process, but also opportunities in leveraging more expressive machine learning approaches.
+The need for data to improve diagnosis and treat of disease, particularly in the emergence of artificial intelligence in medical imaging is increasingly been recognized.
+Data is now the foundation from which new techniques and discoveries are possible.
+Initiatives such as the UK Biobank, Allen Brain Institute, and Ontario Brain Institute all collect and provide their neuroimaging data to researchers.
+With this deluge of imaging data, the complexity of analysis that demand of researchers proportionally increases.
+The need to include large amount of variables, patient attributes, study measures, time sequences, can now provide valuable insight that is beyond the mean of traditional statistics.
+In this thesis we developed an machine learning approach that is scalable, and one that can quantify uncertainty as demanded by the medical need.
+The data processing and machine learning is end-to-end, meaning that the system can be practically deployed in a clinical setting.
 
-\section{Method Discussions}
+We applied machine learning to the analysis of TN in Study IV.
+The analysis of the trigeminal nerve in TN versus controls agreed with our earlier findings, and therefore validates the study's GP classification method.
+We performed traditional multiple comparison statistical analysis for our model, but found the result hard to interpret.
+This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics.
+We therefore decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which may not fit with frequentist assumptions.
+The GP classifier was chosen over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency.
+2) SVM results are hard to interpret, whereas GP is a bayesian method, and therefore it is much more intuitive.
+More specifically, the streamline measurements resemble time-series data, in that it exhibits positional dependency between measurements.
+Unlike a time-series however, the streamline measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency, as opposed to uni-directionally dependent in time.
+Posing the study as a supervised machine learning problem, the feature vector would be a N dimension vector representing N measures at each division, and a classification label.
+SVM would assume the features to be points in a N-dimensional space, and each measurement would assume independence.
+This makes the result of the SVM output difficult to interpret.
+The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output \cite{rasmussen2006gaussian}.
+Therefore the GP model is able to take the positional dependency into account.
+Changing the GP classification model into a regression one is very simple.
+This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a single patient's CN V, what is the probability that is resembles a pathological population as opposed to a healthy population?
+It is possible to answer these questions trivially with the GP model using regression.
+The full implication is outside of the scope of this thesis, but this work will serve as a strong baseline for the applications in this direction.
+
+\section{Novel Technical developments}
 
@@
 
-From my observations across the studies in TN, I postulate a resonance hypothesis of TN.
-I am aware that there lacks substantial physiological measurements for these claims, nevertheless I feel it is important to state them here as a possible direction of investigation for the future: In surgical interventions for TN, the target is on the cistern CN V segment, and situates close to the CN V REZ.
-Paradoxically, the intervention can be seen as a willful disruption of the CN V myelination.
-The question is, why would additional disruptions on the CN V myelin effectively decrease TN pain.
-Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain.
-The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved.
-Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure.
-The force exerted onto the CN V from its neighbouring vasculature is therefore periodic.
-It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition".
-This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain.
-Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain.
-This model also explains that surgical intervention disrupts the chance of resonance condition, thereby reducing TN pain.
-The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns.
-MVD, in this case can be explained as the physical disruption of the condition.
+Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D.
+We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}.
+Another contribution is to visualize tissue changes from surgical interventions that can otherwise be invisible.
+There are cases where post-operative GammaKnife ablation target on the trigeminal nerve did not result in dicernable gadonenium enhancement in standard MR, but the nerve FA changes can be seen in the delineated tractography.
+Therefore tractography can provide both spatial, as well as tissue-based insights to the surgeons to aid in assessment.
+Tractography and 3D anatomical model ultimately are standard 3D geometries in computer graphics.
+Therefore advances in 3D printing can be easily adopted in neurosurgery to allow surgeons to physically examine tractography reconstructions.
+On the otherhand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making.
+Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.
 

### 2018-02-26 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 \graphicspath{{images/tn-gp-sagit/drafts/}}
-
-\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia using Merged Group Tractography}
-\chaptermark{Study IV}
-\label{section:study4}
-
-\section{Abstract}
-\textbf{Introduction:}  Imaging of trigeminal neuralgia (TN) has demonstrated key diffusivity alterations in the trigeminal nerve but technical limitations in group tractography analysis have  prevented us from assessing the trigeminal nerve and its connections in the central nervous system.
-We apply fully-automated group white-matter tractography to image the trigeminal sensory pathway as well as machine learning classification using Gaussian Process (GP) classifier to pinpoint key white-matter diffusivity changes that maximally differentiate TN patients from healthy controls.
-
-\textbf{Methods:} We used SAGIT based group merged tractography for 36 sex-matched TN patients (right-sided pain) and 36 controls, examining the following trigeminal related white matter pathways: trigeminal nerve (CN V), pontine decussation, and thalamocortical fibers.
-GP classifiers were trained by scrolling a moving window over CN V and S1 tractography centroids.
-FA, GFA, RD, AD, and MD metrics were evaluated for both groups, analyzing TN vs.
-control groups and affected vs.
-unaffected sides.
-Classifiers that performed at greater-or-equal-to 70\% accuracy were included.
-
-\textbf{Results:} GP classifier consistently demonstrated bilateral trigeminal changes, differentiating them from controls with an accuracy of 80\%, even though the clinical expression of pain was strictly unilateral.
-Affected and unaffected sides could also be differentiated from each other with an accuracy of 75\%.
-Assessment of the trigeminal subregions demonstrated that the segments in affected CN V differ from controls from the region of the REZ and distally.
-Bilateral TPT could be distinguished from controls with at least 85\% accuracy while TPT left-right classification achieved 98\% accuracy.
-Bilateral S1 could be differentiated from controls, where the affected S1 RD classifier achieved an accuracy of 87\%.
-
-\textbf{Conclusions:} We  demonstrate the first instance of group-wise merged tractography analysis of TN white-matter machine learning classification using GP, and analysis of the complete trigeminal pathways from the pons to S1 cortex.
-The previous reported findings of bilateral changes in the REZ of TN patients is observed.
-Our study suggests additional, bilateral changes extend beyond the REZ and pinpoints these changes with greater detail in regional specificity.
 
@@
 \subsubsection{Trigeminal Neuralgia}
-Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CN V) branches.
+Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more divisions of the three trigeminal nerve (The fifth cranial nerve: CN V) branches.
 TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
 The pathophysiology of classic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although classic TN can also occur without evidence of vascular compression \cite{Lee2014}.
-We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.
+While the brainstem fibers of the trigeminal nerve have not been analyzed in detail, We recently compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.
+Similarly, we demonstrated, using multi-tensor tractography, that altered diffusivity at the level of the pons prognosticates treatment non-response \cite{Hung2017}.
+The tractography analysis of the trigeminal projects beyond the brainstem have never been studied.
 
-The trigeminal sensory system involves both the discriminative touch and pain pathways.
+The trigeminal sensory system involves both the discriminative touch and pain pathways \cite{Henssen2016}.
 Pain and discriminitive touch stimuli trigger innervations of the three trigeminal nerve branches (V1, V2, and V3) in the head and face, these primary affrents converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion) in the middle cranial fossa.
-The discriminitve touch afferents then pass through the cisternal segment of CN V, which then enters the pons, and synapse onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex.
+The discriminitve touch afferents then course through the cisternal segment of CN V, enter the pons, and synapse onto the primary trigeminal sensory nucleus, after which they project to the contralateral VC thalamus and the S1 cortex.
 The nociceptive pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord.
@@
 White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}.
-Obtaining diffusivity metrics from tractography involves converting streamline bundles into a volumetric spatial mask, followed by measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}.
+This analysis invovles obtaining diffusivity metrics from tractography after converting streamline bundles into a volumetric spatial mask and measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}.
 However this approach discards the orientation information in the streamlines.
@@
 This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm \cite{Neher2015,Wilkins2015}.
-In white matter structures such as CN V as well as other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
+In white matter structures such as CN V as well as other small pathways, the structure diameter may be fairly small (less than 3mm) and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
 
-In this light, a tractography streamline propagation algorithm must involve all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
+For a tractography streamline propagation algorithm to address these limitations, it must involve all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}.
 It is reasonable to consider these algorithms to be an interpolated subsampling scheme of the DWI volume that takes the diffusion-based orientation into account.
@@
 Along-the-tract analysis has been investigated by various authors \cite{Colby2012,ODonnell2009,Wang2015,Yeatman2012}.
-In these approaches, the bundles were selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and uncinate.
-The measures were pooled onto tract centroid in each individual subjects, and then aggregated at the group level.
-The global tractography approach is unintuitive for anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is predetermined.
+Commonly, the bundles were selectively filtered from global tractography \cite{Wang2015,Yeatman2012}.
+An spatial representative curve, termed the tract centroid, was determined for the streamlines that consist of the bundle.
+The centroid was calculated by a variety of distance measures between the points of the streamlines \cite{Garyfallidis2012}, and resampled to a certain number of subpoints.
+The diffusivity metrics from each streamline points were pooled onto the tract centroid in each individual subjects, and then aggregated at the group level.
+In this way, diffusivity metrics within a tractography bundle in 3D space would be projected to a 1D centroid for detailed analysis.
+
+The global tractography approach is limited to major white matter bundles, and unintuitive for anatomy-specific tasks in a clinical environment, where the anatomy of interest is predetermined.
 A more focused ROI based delineation approach is therefore preferred in these circumstances.
-
-It is important to note that streamline clustering and centroid identification can be considered as independent tasks.
-Some clustering methods do not provide a best-representative streamline centroid.
-Centroid identification involves resampling the streamlines into an identical number of points, and then identifying point correspondences via different distance measures.
-For these approaches, streamlines require similar lengths, and an assumption that the start and end of the streamlines across subjects is consistent.
-However, pathological brain tissue may  affect tractography delineation.
-Therefore the geometry assumptions of these clustering approaches may not be satisfied when studying patient populations.
-Computation time and memory constraints also need to be taken into consideration when applying such algorithms at a group level.
-We have adopted QuickBundles since it is a practical method of tract clustering and centroid identification with linear running time to obtain TPT and S1 streamline centroids.
-\cite{Garyfallidis2012}
-
-The principal advantage of a streamline-based approach is that we can preserve the geometric orientations of the pathways.
-White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis.
 We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework \cite{Chen2016} to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning.
-We then merge the tracts into a template space using non-linear deformations, and perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.
+This technique permits that tracts to be merged into a template space using non-linear deformations.
+A centroid can be directly defined on the merged tractography bundle, and the diffusivity metrics from multiple subjects can be pooled onto the centroid directly.
+The extract diffusivity data can then be analyzed by machine learning.
 
@@
 The streamline measurements resembles time-series data, in that it exhibits positional dependency between measurements.
-Unlike a time-series however, the measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency.
+Unlike a time-series however, the measurements represent physical locations along a pathway, thus exhibiting bi-directional dependency.
 Posing the study as a supervised machine learning problem, the feature vector would be a N dimensional vector representing N measures at each division.
@@
 In all previous TN studies, placements of measurement region-of-interests in the CN V were pre-planned and manually placed by researchers.
-Manual tractography generation with manual ROI placements was a roadblock in performing more indepth group-wise tractography examination of the trigeminal pathways.
+Manual tractography generation with manual ROI placements was a roadblock in performing more in-depth group-wise tractography examination of the trigeminal pathways.
 We had developed and validated the SAGIT tractography framework to allow automated group tractography \cite{Chen2016}.
@@
 
+The overall processing steps are a) generate merged tractography of each trigeminal pathway; b) define the centroids that spatially represent each pathway, and pool the diffusivity data from streamlines onto the centroid; and c) perform GP classification.
+The details are as follows.
+
 \subsubsection{Acquisitions}
-We studied 36 patients with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).
-All patients underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment.
+We studied 36 TN subjects with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).
+All TN subjects underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment.
 MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions.
@@
 For the purpose of quantifying the trigeminal pathways, a centroid was defined for the merged tractography of each pathway (Figure \ref{fig:GPfigure-method-centroid}).
-The centroids were defined as the average curve that best fit the tract bundles, and were in template space.
+The centroids were defined as the average curve that best fit the tract bundles and were in template space.
 CN V centroids were manually defined using 3 control points.
@@
 Diffusivity metrics of affected (right side) and unaffected CN V, TPT, and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides.
-The affected and unaffected sides in TN patients were also compared with each other.
+The affected and unaffected sides in TN subjects were also compared with each other.
 Age was regressed out by using the residuals of the linear regression model from the controls group.
 The diffusion values were subsequently normalized to zero-mean with ranges $[-1, 1]$ for model training.
-Gaussian Process (GP) classification was performed to identify key segment differences.
-Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment (Figure \ref{fig:GPfigure-gp-train}).
+Gaussian Process (GP) classification was performed to identify windows that resulted in the highest classification accuracies.
+Multiple moving windows of length between 5 to 30 were used to scroll through the centroid positions, and a Gaussian Process model was trained for each window (Figure \ref{fig:GPfigure-gp-train}).
 The entire machine learning process was performed using Scikit-learn machine learning software framework.
@@
 GP models were initiated with Matern kernel (length scale = 1).
-The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed segment.
-Segments accuracy greater than 70\% were accepted, and the window segment with the greatest accuracy was then determined.
+The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed.
+Classification accuracy greater than 70\% were accepted, and the window with the greatest accuracy was then determined.
 
@@
 \caption{Diagram for generating the best performing classifier for a particular window size.
-The data was derived from Figure \ref{fig:GPfigure-method-centroid}, where the top bar represents data from each centroid division, where each position represents the anatomical region of the trigeminal pathways within it.
+The data was derived from Figure \ref{fig:GPfigure-method-centroid}, where the top bar represents data from each centroid division, and each position represents the anatomical region of the trigeminal pathways within it.
 Each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position.
@@
 \subsubsection{Anatomical Delineations}
-Merged CN V delineations show anatomical consistency, and extended into the ipsilateral TGN.
-Changes in FA along the streamlines demonstrate clear differences between cistern and brainstem segments, suggesting good agreement in anatomical registration (Figure \ref{fig:GPfigure1}).
+The findings in each of the three trigeminal areas are as follows: Merged CN V delineations show anatomical consistency from their position in the cisternal segment of the nerve and extending into the ipsilateral TGN.
+Changes in FA along the streamlines demonstrate clear step-wise increase at the interface between cistern and pontine segments, which is consistent with observations in single subjects.
+The FA metric of each point was sampled from the native DWI space of each subject, prior to the template space deformations.
+Therefore the sharp transition in FA in this anatomical landmark suggesting good agreement in the SAGIT anatomical registration (Figure \ref{fig:GPfigure1}).
 The CN V centroid curve length was determined as 38.91 mm, with the length of each subdivision being 1.3 mm.
 The TPT projections from the TGN to the ipsilateral VC thalamus region showed distinct bundles of decussating pathways in both left and right projections.
-The resulting pathways were not symmetrical, where the right TPT pathway visibly diverts and envelopes the left (Figure \ref{fig:GPfigure2}).
+The resulting pathways were not symmetrical, where the right TPT pathway appears to divert around the left (Figure \ref{fig:GPfigure2}).
 S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}). The S1 centroid curve length was calculated to 62.1 mm and sub-division length was 2.07 mm.
@@
 \centering
-\includegraphics[width=\linewidth]{cnv-inferior-view.png}
+\includegraphics[width=\linewidth]{cnv-inferior-view-annotated.png}
 \caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
@@
 Images obtained prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
-FA is depicted by the colour scales (Red: right side; Blue: left side).
-The S1 pathways projects to the lateral face region of the S1 cortex.
-}
+FA is depicted by the colour intensities (Red: right side; Blue: left side).
+The trigeminal nerves, including the cisternal segments are seen in a coronal perspective (CN V), following which two pathways can be seen.
+The first pathway appears to head into the spine, while the second pathway consists of the TPT.
+Following the decussation, the fibers proceed to the area of the sensory thalamus and from there to the cortical sensory area S1.}
 \label{fig:GPfigure3}
@@
 \paragraph{CN V}
-The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves.Both sides were distinguished from controls with classifiers that achieved around 80\% accuracy (Figure \ref{fig:GPfigure4}).
-Classifiers on the right side had accuracies ranging from 76\% (GFA) to 86\% (FA).
-The left classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA).
-The right side showed more overlap of segment ranges between positions 14--24, in the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4 in the brainstem.
-Overall, the right side classifiers extend further into the distal space, into the cistern and even the ganglion, while the left side extended more into the brainstem.
+The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves.
+Both sides were distinguished from controls with classifiers that achieved near 80\% accuracy (Figure \ref{fig:GPfigure4}).
+Classifiers on the right (affected) side had accuracies ranging from 76\% (GFA) to 86\% (FA).
+The left (unaffected) classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA).
+The right side showed more overlap of segment ranges between positions 14--24, which corresponds to the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4, within the substrance of the trigeminal fibers in the brainstem.
+Overall, the right side classifiers extend further into the distal subdivisions, into the cistern and the area of the ganglion, while the left side extended more into the brainstem.
 FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cistern).
@@
 The GFA classifier achieved 70\% accuracy in region 9--12.
-Demonstrating that affected-unaffected diffusivity differences were found in the cistern/REZ (Table \ref{table:GP}).
+Demonstrating that affected-unaffected diffusivity differences were found primarily in the cistern/REZ (Table \ref{table:GP}).
 
@@
 Classifiers could distinguish bilateral TPT pathways (affected and unaffected) from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}).
-The right FA classifier achieved the greatest accuracy at 85\%, and right RD classifier was least accuracy at 83\%.
-The left FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuracy at 71\%.
+The right (affected) FA classifier achieved the greatest accuracy at 85\%, and right RD classifier was least accuracy at 83\%.
+The left (unaffected) FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuracy at 71\%.
 The classifiers covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage.
@@
 \paragraph{S1}
-Classifiers distinguished bilateral S1 face region pathways from controls with about 85\% accuracy.
-The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at position 8--26 (Figure \ref{fig:GPfigure5}).
+Classifiers distinguished bilateral S1 face region pathways from controls with 85\% accuracy.
+The left (affected) RD classifier achieved maximal accuracy of 87\% at position 13--20, while the right (unaffected) RD classifier achieved maximal accuracy of 83\% at position 8--26 (Figure \ref{fig:GPfigure5}).
 Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifiers ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:GP}).
@@
 The GP classifier placed affected vs unaffected differences at the cistern/REZ segment, in agreement with the pathophysiology of TN.
-The findings on unaffected versus controls suggest that there is bilateral differences in CN V myelin in unilateral TN, despite the lack of symptoms on the unaffected side.
-The left regions also situate into the brainstem segments.
-It is possible that the left CN V is affected by global descending pain modulation, and warrant further investigation.
 
-The TPT delineations showed that decussation pathway tractography is possibly asymmetric.
-It demonstrated very strong differences from the classifier results, with overlapping differences in the ascending segment.
-There also seemed to be diffusivity differences in the decussation segments between affected and unaffected sides.
-The decussating fibers of the TPT emerges from the TGN, therefore the difference seemed reasonable.
-It is not clear however, whether the tractography asymmetry of the TPT is neuroanatomically factual, or an artifact from either algorithmic bias or partial volume bias of the underlying DWI.
-Resolving crossing fiber is still a difficult task in tractography, and while there are validations against MR phantoms, the anatomical validity of these methods can only be extrapolated, and the interpretation must involve anatomical priors from experts.
+The findings on unaffected (left) versus controls suggests bilateral differences in CN V diffusivities in unilateral TN, despite the lack of symptoms on the unaffected side.
+RD and AD have previously shown to relate to myelination \cite{Song2002} and axonal integrity \cite{Budde2009}.
+Therefore this finding is significant, in which it suggest that the REZ and brainstem CN V ultrastructure might be bilaterally affected in unilateral TN pain.
+A process that may play important role in TN pain chronification, and responsiveness to surgery.
+The classifier windows of unaffected side also situate into the brainstem subdivisions.
+It is possible that the unaffected CN V is modulated by descending pathways, and warrant further investigation.
 
-The S1 delineation was consistent, due to full automation using SAGIT and the incorporation of Freesurfer cortical segmentation maps.
-Reproducing streamlines into the face-region of S1 cortex in multiple subjects was difficult previously due to individual variations, as well as the multi-directional fiber crossings in the coronal radiata.
-Previous studies have suggested that S1 have key involvement in pain location discrimination \cite{bushnell1999pain}, and are found to have grey-matter abnormalities in TN pain \cite{Desouza2013c}.
-Our finding that the left (affected) thalamocortical S1 projection is uniquely different close to the cortex suggest that deep cortical white-matter/grey-matter changes may be related depending on the locality.
+The TPT left-verus-right FA/RD/GFA classifiers demonstrated very high accuracy (>90\%), with overlapping windows in the ascending segment.
+The TPT merged tractography delineations showed that decussation pathway tractography is possibly asymmetric.
+The asymmetry of this decussation has not been demonstrated before.
+It is not clear whether this represents the anatomical arrangement of these fibers, or whether decussations are always purely symmetrical (akin to a perfectly shuffled deck of cards).
+Further assessment of other decussation points will be helpful to understand this process.
 
-We included GFA to investigate its potential for measuring deep white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric.
-However results demonstrated that it is not effective in pinpoint deep white-matter abnormalities.
-GFA did not perform to any substantially degree than the Gaussian-base diffusivity metrics.
+We were able to successfully delineate the course of the fibers from the thalamus to S1 given the use of SAGIT and incorporation of Freesurfer cortical segmentation maps.
+Previous studies have suggested that S1 is involvement in pain location discrimination \cite{bushnell1999pain}, and have been found to exhibit grey-matter abnormalities in TN pain \cite{Desouza2013c}.
+Our finding that while the left (affected) S1 RD classifier had higher accuracy (87\%) compare to the right S1 RD classifier (83\%), the left-verus-right classifier windows showed more overlap in subdivisions close the thalamus.
+Suggesting that while there is bilateral difference in the S1 pathways from controls, there is little differentiation between the segments closer to the S1 cortex.
 
-Overall the GP classifiers resulted in higher precision when compared to accuracy.
-The overlapping regions revealed differences in localizations that may very well have been missed with singular diffusivity metric analysis.
-This suggests that cross-metric features should be considered in future studies.
-This would drastically increase the feature complexity, and thereby demand substantial increase in the number of subjects.
+The classifier windows showed a large variation of size.
+For example, amongst CN V classifiers (Figure \ref{fig:GPfigure3}), right MD consisted of 18 subdivisions, while right FA consisted of 7; The left classifiers showed a variety of sizes, yet the accuracy derived from each classifier were about 80\%.
+In the same manner, we observed variabilities in TPT and S1 classifiers window sizes.
+This suggests that classification according to a single diffusivity metric may be difficult to interpret, and thus multiple metric classifiers should be considered together.
 
-\subsubsection{Limitations}
-The merged CN V delineations of all the subjects showed caudal projections towards the spinal cord.
-In multiple instances, diverging streamlines at the levels of the medulla strongly suggested the trigeminal spinal decussations.
-However the caudal projection was not available for the majority of the TN group, and it was not a consistently producible pathway.
-This is likely due to that 1) 3T DWI scan has a limited 2mm isovoxel resolution; It is not enough to resolve CN V, medullary and spinal cord fiber details.
-We have modified our scan to 1x1x3mm voxel, to gain axial plane detail for this kind of small fiber tractography from clinical scans, at the sacrifice of coronal/sagittal resolution.
-2) A large percent of the TN DWI scans has severe distortions at the pontine level.
-Similar distortions rarely show up in Control or MS-TN scans.
-The reason may be due to the changes in the vertebral and basilar blood vessels that gave rise to EPI distortions in these patients.
-These distortions narrow the brainstem in the DWI image, and when combined with limited resolution, making tractography delineations even more difficult.
-Therefore, it is highly recommended to perform reverse-blip DWI acquisitions in patients with cerebral vessel irregularities as part of MR DWI clinical practice.
+FA classifiers consistently achieved the highest accuracy in the different comparisons, with the exception of S1.
+This might be due to the fact that it is known to have trouble in regions of heavy crossing-fibers.
+The TPT pathway also situations in regions of complex crossing fibers.
+One explanation is that because the TPT courses through the narrow passage of pontine/hypothalamic regions, the fiber structures are highly contrained.
+Therefore FA still remains as a sensitive region for the TPT.
 
-The diffusivity metrics of the deeper white matter structures remains difficult to interpret, due to the complex crossing fibers in these regions.
+We included GFA to investigate its potential for measuring cross-fiber white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric.
 Although metrics such as FA are popular in current neuroimaging literature, they nevertheless are derived from the Gaussian tensor model, which cannot adequately describe cross-fiber regional dynamics.
@@
 Moreover, in diverging and converging structures, as well as in multi-crossing regions such as the corona radiata, the interpretations of Gaussian-based diffusivity metrics become even more unclear.
+However results demonstrated that GFA did not perform to any substantially degree than the Gaussian-based diffusivity metrics such as FA.
+
+Overall the GP classifiers windows revealed areas in the TN pathways that may very well have been missed with singular diffusivity metric analysis.
+This suggests that cross-metric features may have a greater value and should be considered in future studies.
+
+\subsubsection{Limitations}
+The merged CN V delineations of all the subjects showed caudal projections towards the spinal cord.
+The consistent observation of diverging streamlines at the levels of the medulla strongly suggests that these fibers correspond to trigeminal spinal decussations.
+However the caudal projection were not consistently seen.
+This is likely due to the limited size of voxel resolution that in turn does not permit adequate resolution of medullary and spinal cord fiber details.
+
+The diffusivity metrics of the deeper white matter structures remains difficult to interpret, due to the complex crossing fibers in these regions.
 Therefore, it is imperative that interpretable diffusivity measures for crossing regions be established and adopted by the neuroimaging tractography field for future studies.
@@
 \subsection{Conclusions}
-We have demonstrated the first instance of group-wise merged along-the-track tractography analysis of TN vs.
+We have demonstrated group-wise along-the-track tractography machine learning analysis of TN vs.
 control.
-We have achieved examination of specific and targeted anatomy at a group level involving both healthy and pathological brain images.
-The method reproduced previous findings of TN diffusivity changes along the CN V with finer-grained regional specificity.
+We have achieved the examination of specific and targeted anatomy at a group level involving both healthy and pathological brain images.
+The method agreed with previous findings of TN diffusivity changes along the CN V with finer-grained regional specificity.
 We also examined TPT and S1 projections of the trigeminal sensory pathway.

### 2018-02-26 thesis/contents/tn_sagit_abstract.tex
Updates from ShareLaTeX

\textbf{Introduction:}  Imaging of trigeminal neuralgia (TN) has demonstrated key DTI based diffusivity alterations in the trigeminal nerve but imaging has primarily focused on the peripheral nerve segment, since technical limitations have  prevented us from assessing the trigeminal nerve and its connections in the central nervous system. We apply fully-automated group white-matter tractography to image the trigeminal sensory pathway and machine learning classification using a Gaussian Process (GP) classifier to pinpoint key white-matter diffusivity changes that maximally differentiate TN subjects from healthy controls.

\textbf{Methods:} We used SAGIT based group merged tractography to study 36 sex-matched TN subjects (right-sided pain) and 36 controls, examining the following trigeminal related white matter pathways: trigeminal nerve (CN V), pontine decussation (TPT), and thalamocortical fibers (S1). GP classifiers were trained by scrolling a moving window over CN V, TPT, and S1 tractography centroids. FA, GFA, RD, AD, and MD metrics were evaluated for both groups, analyzing TN vs. control groups and affected vs. unaffected sides. Classifiers that performed at greater-or-equal-to 70\% accuracy were included.

\textbf{Results:} GP classifier consistently demonstrated bilateral trigeminal changes, differentiating them from controls with an accuracy of 80\%, even though the clinical expression of pain was strictly unilateral. Affected and unaffected sides could also be differentiated from each other with an accuracy of 75\%. Assessment of the trigeminal subregions demonstrated that the segments in affected CN V differ from controls from the region of the REZ and distally. Bilateral TPT could be distinguished from controls with at least 85\% accuracy while TPT left-right classification achieved 98\% accuracy. Bilateral S1 could be differentiated from controls, where the affected S1 RD classifier achieved an accuracy of 87\%.

\textbf{Conclusions:} This is the first study in TN that combines group-wise merged tractography, machine learning classification using GP, and analysis of the complete trigeminal pathways from the the trigeminal peripheral fibers to S1 cortex. This analysis demonstrates that TN is characterized by bilateral changes throughout the trigeminal pathway compared with healthy controls, as well as changes between clinically affected and unaffected sides. The combination of group tractography and machine learning have proven to be a powerful approach for the study of white matter diffusivities.

### 2018-02-27 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX
- tn-sagit paper changes

@@
-\graphicspath{{images/tn-gp-sagit/drafts/}}
-
 \section{Introduction}
 \subsubsection{Trigeminal Neuralgia}
-Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more divisions of the three trigeminal nerve (The fifth cranial nerve: CN V) branches.
-TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
+Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more divisions of the trigeminal nerve (The fifth cranial nerve, CN V) branches.
+TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
 The pathophysiology of classic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although classic TN can also occur without evidence of vascular compression \cite{Lee2014}.
-While the brainstem fibers of the trigeminal nerve have not been analyzed in detail, We recently compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.
+While the focus of the study of TN has been primarily the peripheral segments of the trigeminal nerve, we recently compared TN and MS-TN diffusivity differences in four segments of the CN V, including at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.
 Similarly, we demonstrated, using multi-tensor tractography, that altered diffusivity at the level of the pons prognosticates treatment non-response \cite{Hung2017}.
-The tractography analysis of the trigeminal projects beyond the brainstem have never been studied.
+It appears therefore that TN is associated with unique peripheral abnormalities and brainstem related central abnormalities.
 
-The trigeminal sensory system involves both the discriminative touch and pain pathways \cite{Henssen2016}.
+Investigation of potential gray and white matter findings in TN has demonstrated that there are key abnormalities in areas important for discrimination and modulation of pain, including the thalamus, basal ganglia, primary somatosensory cortex, primary motor cortex, orbitofrontal cortex, cingulate cortex, and insula \cite{Desouza2013c, Desouza2013}.
+
+The trigeminal sensory system involves both discriminative touch and pain pathways \cite{Henssen2016}.
 Pain and discriminitive touch stimuli trigger innervations of the three trigeminal nerve branches (V1, V2, and V3) in the head and face, these primary affrents converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion) in the middle cranial fossa.
-The discriminitve touch afferents then course through the cisternal segment of CN V, enter the pons, and synapse onto the primary trigeminal sensory nucleus, after which they project to the contralateral VC thalamus and the S1 cortex.
+The discriminative touch afferents then course through the cisternal segment of CN V, enter the pons, and synapse onto the primary trigeminal sensory nucleus, after which they project to the contralateral VC thalamus and the S1 cortex.
 The nociceptive pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord.
 The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus.
-The pain pathway also projects to the medial nucleus of the thalamus, where it then project medially to the cingulate cortex.
+The pain pathway also projects to the medial nucleus of the thalamus, after which it courses towards the fibers of the cingulate cortex.
+
+The tractography and diffusivity parameters of the trigeminal pathways more central to the brainstem have not been studied, primarily due to methodological limitations.
+However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain.
+It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibers.
 
@@
 We had developed and validated the SAGIT tractography framework to allow automated group tractography \cite{Chen2016}.
-In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.
+We further expand, in the current study to A) analyze the CN V differences between TN and healthy controls across the entire peripheral segment of the nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.
 
@@
 
-The overall processing steps are a) generate merged tractography of each trigeminal pathway; b) define the centroids that spatially represent each pathway, and pool the diffusivity data from streamlines onto the centroid; and c) perform GP classification.
+The overall processing steps are a) generation of merged tractography bundle of each trigeminal pathway; b) definition of centroids that spatially represent each pathway, and pooling the diffusivity data from streamlines onto the centroid; and c) scroll a moving window along the centroid positions, and train a GP classifier at each step, the windows with the best accuracy is then determined.
 The details are as follows.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\linewidth]{figure-method.pdf}
+\caption{Diagram of the processing steps}
+\label{fig:GPMethods}
+\end{figure}
 
@@
 DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-method.pdf}
-\caption{Diagram of the processing steps}
-\label{fig:GPMethods}
-\end{figure}
 
@@
 \subsubsection{Tractography}
-ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibers must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary; it must exclude the brainstem).
+ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (specifications: ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibers must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary and exclude the brainstem).
 
@@
 All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis.
-We have previously created a software library\footnote{https://github.com/sinkpoint/fascicle} to facilitate this process, the steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}.
+We have previously created a software library\footnote{https://github.com/sinkpoint/fascicle} to facilitate this process.
+Detailed steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}.
 The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group associations preserved.
@@
 \centering
-\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-path-analysis.pdf}
+\includegraphics[width=\linewidth]{figure-path-analysis.pdf}
 \caption{Diagram for the along-the-tract data processing.
@@
 Multiple moving windows of length between 5 to 30 were used to scroll through the centroid positions, and a Gaussian Process model was trained for each window (Figure \ref{fig:GPfigure-gp-train}).
-The entire machine learning process was performed using Scikit-learn machine learning software framework.
+The machine learning process was performed using Scikit-learn machine learning software framework.
 For each window, the subjects were split into train and validation sets using stratified 10-fold cross-validation, and a separate GP classifier was trained for each fold.
@@
 \centering
-\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-GP-training.pdf}
+\includegraphics[width=\linewidth]{figure-GP-training.pdf}
 \caption{Diagram for generating the best performing classifier for a particular window size.
@@
 The left (unaffected) classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA).
-The right side showed more overlap of segment ranges between positions 14--24, which corresponds to the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4, within the substrance of the trigeminal fibers in the brainstem.
-Overall, the right side classifiers extend further into the distal subdivisions, into the cistern and the area of the ganglion, while the left side extended more into the brainstem.
+The right side showed greater overlap of segment ranges between positions 14--24, which corresponds to the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4, within the substance of the trigeminal fibers in the brainstem.
+Overall, the windows of the right side classifiers extend further into the distal subdivisions, into the cistern and the area of the ganglion, while the left side extended more into the brainstem.
 FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cistern).
@@
 
-\begin{figure}[ht]
+\begin{figure}[p]
 \centering
-\includegraphics[width=\linewidth]{figure-GP-CNV.pdf}
+\includegraphics[width=1.2\textwidth,center]{figure-GP-CNV.pdf}
 \caption{CN V classifier performances.
@@
 Panel C shows the centroid and point assignments for anatomical reference.
-The colours of the points correspond to their respective subdivision position as described in figure \ref{fig:GPfigure-method-centroid} and \ref{fig:GPfigure-method-centroid}} Throughout the figure, the position of the centroid subdivisions are aligned and expressed with respect to the anatomical position as per panel B.
+The colours of the points correspond to their respective subdivision position as described in figure \ref{fig:GPfigure-method-centroid} and figure \ref{fig:GPfigure-method-centroid}} throughout the figure, the position of the centroid subdivisions are aligned and expressed with respect to the anatomical position as per panel B.
 \label{fig:GPfigure4}
@@
 Classifiers could distinguish bilateral TPT pathways (affected and unaffected) from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}).
-The right (affected) FA classifier achieved the greatest accuracy at 85\%, and right RD classifier was least accuracy at 83\%.
+The right (affected) FA classifier achieved the greatest accuracy at 85\% versus controls, and right RD classifier was least accuracy at 83\%.
 The left (unaffected) FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuracy at 71\%.
@@
 \centering
-\includegraphics[width=\linewidth]{figure-GP-TPT.pdf}
+\includegraphics[width=1.2\textwidth,center]{figure-GP-TPT.pdf}
 \caption{TPT classifier performance, see Figure \ref{fig:GPfigure4} for details on annotations.}
@@
 \centering
-\includegraphics[width=\linewidth]{figure-GP-S1.pdf}
+\includegraphics[width=1.2\textwidth,center]{figure-GP-S1.pdf}
 \caption{S1 classifier performance, see Figure \ref{fig:GPfigure4} for details on annotations.}
@@
 
-\begin{table}[ht]
+\begin{table}[p]
 \centering
+\small
 \csvautotabular{tn_sagit_table.csv}
@@
 We have demonstrated the use of a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy strictly from white-matter measures.
-Our work is in agreement with previous findings of TN diffusivity changes along the CN V with much higher regional specificity.
+Our work is in agreement with previous findings of TN diffusivity changes along the CN V, yet the present technique offers much higher regional specificity.
 We also examined TPT and S1 projections of the trigeminal sensory pathway.
 The study is the first instance of machine learning discrimination of TN vs.
-Controls with specific white matter analysis, using group-wise merged along-the-track tractography.
+controls with specific white matter analysis, using group-wise merged along-the-track tractography.
 
-The diffusivity measures of CN V agreed with our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls.
+The diffusivity measures of CN V agreed with our previous findings \cite{Chen2016a}, that the cistern, REZ differentiated TN from controls.
 The consistent regional overlaps in the cistern/ganglion region of the right (affected) CN V showed the ability for the GP classifiers to auto-converge to focal white-matter abnormality.
-The GP classifier placed affected vs unaffected differences at the cistern/REZ segment, in agreement with the pathophysiology of TN.
+The GP classifier placed affected vs unaffected differences at the cistern/REZ segment, in agreement with the pathophysiology of focal neurovascular compression in TN.
 
@@
 RD and AD have previously shown to relate to myelination \cite{Song2002} and axonal integrity \cite{Budde2009}.
-Therefore this finding is significant, in which it suggest that the REZ and brainstem CN V ultrastructure might be bilaterally affected in unilateral TN pain.
-A process that may play important role in TN pain chronification, and responsiveness to surgery.
+Therefore this finding is significant, since it suggests that the REZ and brainstem CN V ultrastructure might be bilaterally affected in unilateral TN pain, a process that may play important role in TN pain chronification, and responsiveness to surgery.
 The classifier windows of unaffected side also situate into the brainstem subdivisions.
-It is possible that the unaffected CN V is modulated by descending pathways, and warrant further investigation.
+It is possible that the unaffected CN V is modulated by descending pathways \cite{Davis2013,Bushnell2013}.
+This point warrants further investigation.
 
@@
 Previous studies have suggested that S1 is involvement in pain location discrimination \cite{bushnell1999pain}, and have been found to exhibit grey-matter abnormalities in TN pain \cite{Desouza2013c}.
-Our finding that while the left (affected) S1 RD classifier had higher accuracy (87\%) compare to the right S1 RD classifier (83\%), the left-verus-right classifier windows showed more overlap in subdivisions close the thalamus.
-Suggesting that while there is bilateral difference in the S1 pathways from controls, there is little differentiation between the segments closer to the S1 cortex.
+We found that that while the left (affected) S1 RD classifier had higher accuracy (87\%) compare to the right S1 RD classifier (83\%), the left-verus-right classifier windows showed more overlap in subdivisions close to the thalamus.
+This suggests that while there is bilateral difference in the S1 pathways from controls, the difference between the segments mitigates as these reach closer to the S1 cortex.
 
 The classifier windows showed a large variation of size.
-For example, amongst CN V classifiers (Figure \ref{fig:GPfigure3}), right MD consisted of 18 subdivisions, while right FA consisted of 7; The left classifiers showed a variety of sizes, yet the accuracy derived from each classifier were about 80\%.
+For example, amongst CN V classifiers (Figure \ref{fig:GPfigure3}), right MD consisted of 18 subdivisions, while right FA consisted of 7; The left classifiers showed a variety of sizes, yet the accuracy derived from each classifier were near 80\%.
 In the same manner, we observed variabilities in TPT and S1 classifiers window sizes.
@@
 
-FA classifiers consistently achieved the highest accuracy in the different comparisons, with the exception of S1.
+FA classifiers consistently achieved the highest accuracy in CNV and TPT, with the exception of S1.
 This might be due to the fact that it is known to have trouble in regions of heavy crossing-fibers.
-The TPT pathway also situations in regions of complex crossing fibers.
-One explanation is that because the TPT courses through the narrow passage of pontine/hypothalamic regions, the fiber structures are highly contrained.
-Therefore FA still remains as a sensitive region for the TPT.
+However, the TPT pathway also situations in regions of complex crossing fibers.
+One explanation for FA classifier in TPT is that the pathway courses through the narrow passage of pontine/thalamic regions, the fiber structures are highly constrained.
+Therefore FA still remains as a sensitive diffusivity metric for the TPT.
 
@@
 
-Overall the GP classifiers windows revealed areas in the TN pathways that may very well have been missed with singular diffusivity metric analysis.
-This suggests that cross-metric features may have a greater value and should be considered in future studies.
+Overall the GP classifiers windows revealed areas in the TN pathways that would not have been captured using singular diffusivity metric analysis.
+Cross-metric features therefore should be considered in future studies.
 
@@
 
-The diffusivity metrics of the deeper white matter structures remains difficult to interpret, due to the complex crossing fibers in these regions.
-Therefore, it is imperative that interpretable diffusivity measures for crossing regions be established and adopted by the neuroimaging tractography field for future studies.
+The DWI images were acquired at 0.9375$\times$0.9375$\times$3 $mm\textsuperscript{3}$ voxel resolution.
+The acquisition was a compromise to allow better in-plane resolution to image the trigeminal nerve, due to the limits of the 3T GE HDx MRI with 8-channels head coil.
+Our analysis is internally consistent, and SAGIT was previously validated on a identical sequence \cite{Chen2016,Chen2016a}, therefore we believe the approach is valid.
 

### 2018-02-28 thesis/contents/reviews/machine-learning.tex
Updates from ShareLaTeX

@@
 Strategies and algorithms for constructing various models overlap with the fields of artificial intelligence, computational statistics, pattern recognition, and neuroscience.
-Advanced in modern machine learning, especially in deep neural networks \cite{LeCun2015}, in recent years have led to wide spread adaptation of machine learning based online services in translation \cite{bahdanau2014neural}, image search \cite{simonyan2014very}, and speech recognition \cite{graves2013speech}.
+Advanced in modern machine learning, especially in deep neural networks \cite{LeCun2015}, in recent years have led to wide spread adaptation of machine learning based online services in translation \cite{Bahdanau2014a}, image search \cite{simonyan2014very}, and speech recognition \cite{graves2013speech}.
 There is also surging interest in deploy machine learning systems in radiology and disease diagnosis in medicine\cite{Kamnitsas2016,Suk2014}.

### 2018-02-28 thesis/contents/sagit.tex
Updates from ShareLaTeX

@@
 \chaptermark{Study III}
+\label{section:study3}
 
@@
 \end{equation}
-where $n$ was the number of bins; $v_0$ was the number of voxels where $s \textgreater 0$;  $v_i$ was the number of voxels with $ 0 \textless s \ge \frac{i}{n}$; this study used $n=10$.
+where $n$ was the number of bins; $v_0$ was the number of voxels where $s > 0$;  $v_i$ was the number of voxels with $ 0 < s \ge \frac{i}{n}$; this study used $n=10$.
 Examples of NOS score behaviour are available in Figure 4.

### 2018-02-28 thesis/contents/tn_sagit.tex
- sync tn-gp-sagit

@@
 The tractography and diffusivity parameters of the trigeminal pathways more central to the brainstem have not been studied, primarily due to methodological limitations.
-However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain.
+However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status \cite{Hung2017} and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain \cite{Miller2009}.
 It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibers.
@@
 This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm \cite{Neher2015,Wilkins2015}.
-In white matter structures such as CN V as well as other small pathways, the structure diameter may be fairly small (less than 3mm) and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
+In white matter structures such as CN V as well as other small pathways, the structure diameter may be fairly small (less than 3 mm) and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.
 
@@
 A more focused ROI based delineation approach is therefore preferred in these circumstances.
-We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework \cite{Chen2016} to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning.
-This technique permits that tracts to be merged into a template space using non-linear deformations.
+We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework \cite{Chen2016} to automate group-wise tractography delineation, which ensures consistent region-of-interest (ROI) projection, tractography seeding sampling scheme, and tractography parameter tuning.
+This technique permits tracts to be merged into a template space using non-linear deformations.
 A centroid can be directly defined on the merged tractography bundle, and the diffusivity metrics from multiple subjects can be pooled onto the centroid directly.
-The extract diffusivity data can then be analyzed by machine learning.
+The extact diffusivity data can then be analyzed by machine learning.
 
@@
 Unlike a time-series however, the measurements represent physical locations along a pathway, thus exhibiting bi-directional dependency.
-Posing the study as a supervised machine learning problem, the feature vector would be a N dimensional vector representing N measures at each division.
+Posing the study as a supervised machine learning problem, the feature vector would be an N-dimensional vector representing N measures at each division.
 The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output.
@@
 		m(x) &= \mathbb{E}[f(x)], \\
-		k(x,x') &= \mathbb{E}[(f(x)-m(x)(f(x')-m(x'))],
+		k(x,x') &= \mathbb{E}[(f(x)-m(x))(f(x')-m(x'))],
 	\end{split}
 \end{equation}
-And the Gaussain processed is defined as
+and the Gaussain processed is defined as
 \begin{equation}
@@
 
-The covariance function is termed called kernel function.
+The covariance function is also called the kernel function.
 The radial-basis function kernel (RBF kernel) is widely used in machine learning tasks.
@@
 \includegraphics[width=\linewidth]{figure-method.pdf}
-\caption{Diagram of the processing steps}
+\caption{Diagram of the overall processing steps.
+The details of SAGIT processing steps can be found in \protect\cite{Chen2016}.
+Details of the centroid identification can be found in Figure \ref{fig:GPfigure-method-centroid}.
+The details for machine learning process can be found in Figure \ref{fig:GPfigure-gp-train}.
+}
 \label{fig:GPMethods}
@@
 \subsubsection{Acquisitions}
-We studied 36 TN subjects with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).
+We studied 36 TN subjects with classic TN (28 females, 8 males; age 60.1$\pm$16.0 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12.0 years).
 All TN subjects underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment.
 MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions.
-T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256.
-DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.
+T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375 mm, with slice spacing = 1 mm, TE = 5.052 ms, TR = 11.956 ms, flip angle = $20^\circ$, FOV = 240 mm, and matrix = 256x256.
+DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3 mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, where b0 = 1000 s/$mm^2$, TE = 88.6 ms, TR = 17000 ms, flip angle = $90^\circ$, field of view (FOV) = 120 mm, and matrix = 128x128.
 
@@
 DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}.
-Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains \cite{Chen2016}.
+ROI for tractography seeding were manually created on a normalized template brain from 42 healthy brains \cite{Chen2016}.
 The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}.
@@
 Deterministic CST was applied to delineate CN V and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy.
-The parameters for each were as follows: Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration.
-Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = 45$\deg$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum lengh = 80 mm, initial cut-off = 0.15, cut-off = 0.15.
-Probabilistic tractography for TPT project was performed with step size = 1mm, and streamline count = 2000, and maximal length = 60 mm.
+The parameters for each were as follows: Deterministic CST was performed with step size = 0.3 mm, angle threshold = $45^\circ$, streamline count = 200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration.
+Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = $45^\circ$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum length = 80 mm, initial cut-off = 0.15, cut-off = 0.15.
+Probabilistic tractography for TPT project was performed with step size = 1 mm, and streamline count = 2000, and maximal length = 60 mm.
 Tractography was first performed in the native DWI space for each subject.
@@
 GP models were initiated with Matern kernel (length scale = 1).
-The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed.
+The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windows.
 Classification accuracy greater than 70\% were accepted, and the window with the greatest accuracy was then determined.
@@
 Only models with at least 70\% accuracy were accepted.
-Table is sorted by anatomical region, classifier type and accuracy.
+Table is sorted by anatomical region, classifier type (R denotes right (affected) side versus controls; L denotes left (unaffected) side versus controls; LVR denotes left-versus-right ) and accuracy.
 Start and end denote the beginning and end of the window subdivision positions along the centroid.
@@
 We have demonstrated the use of a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy strictly from white-matter measures.
-Our work is in agreement with previous findings of TN diffusivity changes along the CN V, yet the present technique offers much higher regional specificity.
-We also examined TPT and S1 projections of the trigeminal sensory pathway.
+Our work is in agreement with previous findings of TN diffusivity changes along the CN V, yet the present technique offers much higher regional specificity, as well as examining the TPT and S1 projections of the trigeminal sensory pathway.
 The study is the first instance of machine learning discrimination of TN vs.
@@
 
-The TPT left-verus-right FA/RD/GFA classifiers demonstrated very high accuracy (>90\%), with overlapping windows in the ascending segment.
+The TPT left-verus-right FA/RD/GFA classifiers demonstrated very high accuracy (\textgreater 90\%), with overlapping windows in the ascending segment.
 The TPT merged tractography delineations showed that decussation pathway tractography is possibly asymmetric.
@@
 The acquisition was a compromise to allow better in-plane resolution to image the trigeminal nerve, due to the limits of the 3T GE HDx MRI with 8-channels head coil.
-Our analysis is internally consistent, and SAGIT was previously validated on a identical sequence \cite{Chen2016,Chen2016a}, therefore we believe the approach is valid.
+Our analysis is internally consistent, and SAGIT was previously validated on a identical sequence \cite{Chen2016,Chen2016a}, highlighting the validity of this approach.
 

### 2018-02-28 thesis/contents/ap_paper.tex
checkpoint on edit
Updates from ShareLaTeX

@@
 \chaptermark{Study II}
+\label{section:study2}
+
+\bibentry{Chen2018ap}
+
 \section{Abstract}
@@
 
-B0, ADWI, MDWI, FA, GFA and AP images were each generated (Figures 1, 2) as follows: B0, ADWI and MDWI images were created using FSL (Smith et al., 2004); FA images were created in 3D Slicer by calculating the scalars following diffusion tensor estimation \cite{Pieper2006}; GFA and AP images were created with Dipy software library \cite{Descoteaux2007a,Garyfallidis2014}.
+B0, ADWI, MDWI, FA, GFA and AP images were each generated (Figures \ref{fig:APfigure1}, \ref{fig:APfigure2}) as follows: B0, ADWI and MDWI images were created using FSL (Smith et al., 2004); FA images were created in 3D Slicer by calculating the scalars following diffusion tensor estimation \cite{Pieper2006}; GFA and AP images were created with Dipy software library \cite{Descoteaux2007a,Garyfallidis2014}.
 
@@
 Pair-wise log joint histogram of the DWI derived scalar images was plotted \cite{Hunter2007} to evaluate image intensity orthogonality.
+
+\begin{figure}[ht]
+\centering
+\begin{tabular}{c}
+    \includegraphics[width=0.8\linewidth]{figure1}
+\end{tabular}
+\caption{Illustrated overview of the processing steps to measure DWI to T1 registration similarities.}
+\label{fig:APfigure1}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=\linewidth]{figure2}
+\caption{DWI derived scalar images (B0, MDWI, FA, GFA, and AP) comparing to T1 (top left).}
+\centering
+\label{fig:APfigure2}
+\end{figure}
 
@@
 \subsection{Results}
-Pair-wise joint log histogram (Figure 3) of the DWI derived scalar images revealed distinct correlation patterns in some pairs of images.
+Pair-wise joint log histogram (Figure \ref{fig:APfigure3}) of the DWI derived scalar images revealed distinct correlation patterns in some pairs of images.
 FA–AP and FA–GFA showed distinct diagonal correlations patterns with each other; while B0 showed no clear pattern of correlation with other image types, with the exception of MDWI.
@@
 
+\begin{figure}[p]
+\includegraphics[width=\linewidth]{figure3}
+\caption{Pair-wise joint log histogram of the DWI derived scalar images.}
+\centering
+\label{fig:APfigure3}
+\end{figure}
+
+\begin{figure}[p]
+\includegraphics[width=\linewidth]{figure4}
+\caption{Visual comparisons of MDWI to T1 co-registrations. 
+Shown are examples of MDWI images from one subject deformed using different registration transforms derived from the different image intermediates. 
+T1 is overlaid as thresholded outlines in red. 
+Primary differences in registration quality can be found in the brainstem (left and middle columns), and in the lateral ventricles, especially in the ventricular-caudate boundary (right column). }
+\centering
+\label{fig:APfigure4}
+\end{figure}
+
 \subsubsection{Differences in initial similarity scores affect registration accuracy}
-Visual overlay of MDWI to T1 co-registrations (Figure 4) suggests that DWI derived scalar image with higher initial CC and MI scores (Figures 5) appear to result in more accurate registrations in the brainstem, insula, temporal cortices, and lateral ventricles (especially around the ventricular-caudate boundary where CSF/grey/white-matter are found in close proximity) compared to the other image types.
-Visually, B0, FA and GFA show registration inaccuracies fitting the areas surrounding the brainstem and the lateral ventricles (Figure 4), while MDWI and AP result in improved registration in these key areas.
+Visual overlay of MDWI to T1 co-registrations (Figure \ref{fig:APfigure4}) suggests that DWI derived scalar image with higher initial CC and MI scores (Figures \ref{fig:APfigure5}) appear to result in more accurate registrations in the brainstem, insula, temporal cortices, and lateral ventricles (especially around the ventricular-caudate boundary where CSF/grey/white-matter are found in close proximity) compared to the other image types.
+Visually, B0, FA and GFA show registration inaccuracies fitting the areas surrounding the brainstem and the lateral ventricles (Figure \ref{fig:APfigure4}), while MDWI and AP result in improved registration in these key areas.
 The results of MDWI and AP are visually very similar.
@@
 The initial T1–intermediate CC and MI similarity scores without any applied transformation ($T_{\text{Identity}}$) showed varied initial values.
-MDWI showed the highest initial CC scores, with FA images showing lowest initial CC (Figure 5); AP showed similar initial CC score with B0 and GFA.
+MDWI showed the highest initial CC scores, with FA images showing lowest initial CC (Figure \ref{fig:APfigure5}); AP showed similar initial CC score with B0 and GFA.
 For MI scores (it is important to note that a more negative MI score implies higher similarity in this study), AP showed very similar initial MI score when compared to MDWI (best), while GFA resulted in the highest (worst) initial MI score.
@@
 \subsubsection{Registration improvements are not uniformly transferred when applied to other image types}
-The progression of similarity scores following $T_{\text{Affine}}$ and $T_{\text{Affine}}$+SyN for each of the scalar images (B0, ADWI, MDWI, FA, GFA, and AP) showed similar trends between CC and MI cost-functions (Figures 5,  Supplementary 1; top rows).
+The progression of similarity scores following $T_{\text{Affine}}$ and $T_{\text{Affine}}$+SyN for each of the scalar images (B0, ADWI, MDWI, FA, GFA, and AP) showed similar trends between CC and MI cost-functions (Figures \ref{fig:APfigure5},  Figure \ref{fig:APSup1}; top rows).
 In general, the similarities of the images between T1 and transformed DWI scalar images increased marginally after $T_{\text{Affine}}$, and substantially after $T_{\text{Affine}}$+SyN for most of the co-registration transforms.
-An exception is FA transforms, where CC scores decreased when the transform was applied to other images (Figure 5; rows 1-2, column 4), with the exception of AP.
+An exception is FA transforms, where CC scores decreased when the transform was applied to other images (Figure \ref{fig:APfigure5}; rows 1-2, column 4), with the exception of AP.
 
-Improvement in CC and MI measures were not uniform for all DWI derived scalar images; this was better illustrated by percent changes (Figures 5; bottom rows).
-Different intermediate transforms differentially affected the scores in accordance with the degree of intensity correlations as measured in Figure 3.
+Improvement in CC and MI measures were not uniform for all DWI derived scalar images; this was better illustrated by percent changes (Figures \ref{fig:APfigure5}; bottom rows).
+Different intermediate transforms differentially affected the scores in accordance with the degree of intensity correlations as measured in Figure \ref{fig:APfigure3}.
 Therefore, DSG transforms preferentially improved DSG, and ASG transforms improved ASG images.
-FA transforms, however, are notable, where CC scores across all DWI derived scalar images diverged for both CC and MI cost functions (Figures 5, Supplementary 2; column 4).
+FA transforms, however, are notable, where CC scores across all DWI derived scalar images diverged for both CC and MI cost functions (Figures \ref{fig:APfigure5}, Figure \ref{fig:APSup2}; column 4).
 In fact, FA transforms substantially decreased DSG CC scores, and also negatively impacted DSG MI scores as well.
@@
 GFA behaved more like DSG images, where B0 and MDWI all favorably affected its CC and MI scores, while at the same time it was strongly affected by AP as well.
-However GFA seem to behave differently under autodeformation, as evident by its distinct MI changes (Figures 5; column 5).
+However GFA seem to behave differently under autodeformation, as evident by its distinct MI changes (Figures \ref{fig:APfigure5}; column 5).
 AP transforms showed the least amount of variability when applied to DWI derived scalar images, and consistently improved all the image scores.
@@
 Between-group statistical comparisons of only $T_{\text{Affine}}$+SyN percent change revealed significant findings.
-Under the CC cost-function (Figure 6), percent change in $D_{\text{auto}}$ (Figure 6, bottom panel, column 3) showed that FA and GFA have the highest percent change, while AP has the lowest percent change.
-With MI scores, $D_{\text{auto}}$ showed that only GFA percent changes are significantly greater than others (Figure 6, bottom panel, column 4).
-GFA, however, also shows the greatest variability in distribution (Figure 6, top panel, column 4).
-$D_{\text{all}}$ in contrast, showed that FA is an outlier where its percentage change is significantly lower than all other images; it is also notable that it shows a distinct bimodal distribution (Figure 6, top panel, column 1); GFA was shown to be significantly higher than B0, MDWI and FA, but not AP.
-MI percent changes are not significantly different from each other (Figure 6, bottom panel, column 2).
-MI cost-function results were similar to that of the CC cost-function (Supplementary 2).
+Under the CC cost-function (Figure \ref{fig:APfigure6}), percent change in $D_{\text{auto}}$ (Figure \ref{fig:APfigure6}, bottom panel, column 3) showed that FA and GFA have the highest percent change, while AP has the lowest percent change.
+With MI scores, $D_{\text{auto}}$ showed that only GFA percent changes are significantly greater than others (Figure \ref{fig:APfigure6}, bottom panel, column 4).
+GFA, however, also shows the greatest variability in distribution (Figure \ref{fig:APfigure6}, top panel, column 4).
+$D_{\text{all}}$ in contrast, showed that FA is an outlier where its percentage change is significantly lower than all other images; it is also notable that it shows a distinct bimodal distribution (Figure \ref{fig:APfigure6}, top panel, column 1); GFA was shown to be significantly higher than B0, MDWI and FA, but not AP.
+MI percent changes are not significantly different from each other (Figure \ref{fig:APfigure6}, bottom panel, column 2).
+MI cost-function results were similar to that of the CC cost-function (Figure \ref{fig:APSup2}).
 
@@
 
-The log joint histogram showed that there is a degree of intensity orthogonality between FA, AP, and GFA as a group (ASG) compared to B0, ADWI and MDWI (DSG), suggesting that there is some connection between co-registration performance and intensity similarity (Figure 3).
+The log joint histogram showed that there is a degree of intensity orthogonality between FA, AP, and GFA as a group (ASG) compared to B0, ADWI and MDWI (DSG), suggesting that there is some connection between co-registration performance and intensity similarity (Figure \ref{fig:APfigure3}).
 The greater the intensity similarities between the two images, the greater the effect of the same set of registration transforms.
-Intensity similarity is not the only factor, however, as evident by the exception to this rule in FA and GFA deformation results (Figure 5).
+Intensity similarity is not the only factor, however, as evident by the exception to this rule in FA and GFA deformation results (Figure \ref{fig:APfigure5}).
 This suggests that intensity orthogonality is not a limiting factor in registration transfer across these scalar images.
@@
 The optimal T1-DWI co-registration performance of an intermediate image could be identified by the image’s deformation by its own transform.
-However, we show that only comparing the autodeformation results is not ideal (Figure 7, 8).
+However, we show that only comparing the autodeformation results is not ideal (Figure \ref{fig:APSup1}, \ref{fig:APSup2}).
 Specifically, while the FA autodeformation (using $T_{\text{Affine}}$+SyN) performed better compared to all other autodeformations, registrations of other scalar images based on FA transformations clearly resulted in the worst results.
@@
 
-\begin{figure}[H]
-\includegraphics[width=\linewidth]{figure1}
-\caption{Illustrated overview of the processing steps to measure DWI to T1 registration similarities.}
-\centering
-\label{fig:APfigure1}
-\end{figure}
+\afterpage{
 
- \begin{figure}[H]
- \includegraphics[width=\linewidth]{figure2}
- \caption{DWI derived scalar images (B0, MDWI, FA, GFA, and AP) comparing to T1 (top left).}
- \centering
- \label{fig:APfigure2}
- \end{figure}
-
-\begin{figure}[H]
-\includegraphics[width=\linewidth]{figure3}
-\caption{Pair-wise joint log histogram of the DWI derived scalar images.}
-\centering
-\label{fig:APfigure3}
-\end{figure}
-
-\begin{figure}[H]
-\includegraphics[width=\linewidth]{figure4}
-\caption{Visual comparisons of MDWI to T1 co-registrations. 
-Shown are examples of MDWI images from one subject deformed using different registration transforms derived from the different image intermediates. 
-T1 is overlaid as thresholded outlines in red. 
-Primary differences in registration quality can be found in the brainstem (left and middle columns), and in the lateral ventricles, especially in the ventricular-caudate boundary (right column). }
-\centering
-\label{fig:APfigure4}
-\end{figure}
-
-\begin{figure}[H]
+    \begin{figure}[p]
     \includegraphics[width=\linewidth]{figure5}
@@
 
-\begin{figure}[H]
+    \clearpage
+
+    \thispagestyle{empty}% empty page style (?)
+    \begin{landscape}% Landscape page
+
+        \begin{figure}[p]
         \includegraphics[width=\linewidth]{figure6}
@@
         \end{figure}
+    \end{landscape}
 
-\begin{figure}[H]
+    \clearpage
+
+    \begin{figure}[p]
     \includegraphics[width=\linewidth]{sup1}
@@
     \centering
-\label{fig:APfigure7}
+    \label{fig:APSup1}
     \end{figure}
 
-\begin{figure}[H]
+    \clearpage
+
+    \thispagestyle{empty}% empty page style (?)
+    \begin{landscape}% Landscape page
+        \begin{figure}[p]
         \includegraphics[width=\linewidth]{sup2}
@@
         \centering
-\label{fig:APfigure8}
+        \label{fig:APSup2}
         \end{figure}
+    \end{landscape}
+}

### 2018-02-28 thesis/contents/discussions.tex

@@
 Moving away from clinical studies alone, DTI is able to derive metrics that provide important information on the microstructure of the nerve which can then be related to specific processes such as changes in myelination or the microstructure of the axons.
-Thus DTI provides an important objective measure in the study of pain, which was unavailable up unto this point.
-The earliest work on the feasibility of accurate reconstruction of cranial nerve with DTI tractography \cite{Hodaie2010} in 2010.
+Thus DTI provides an important objective measure in the study of pain, through a previously unavailable approach.
+The earliest work on the feasibility of accurate reconstruction of cranial nerve with DTI tractography \cite{Hodaie2010} dates to 2010.
 This study was able to image CN II, III, and importantly CN V.

### 2018-02-28 thesis/contents/reviews/dmri.tex
Updates from ShareLaTeX

@@
 \begin{tabular}{c}
-\includegraphics[width=3.5in]{diffusion.pdf}
+\includegraphics[width=0.8\textwidth]{diffusion.pdf}
 \end{tabular}
@@
 
-FA has been found to correlate with a wide spectrum of neuromorphological changes, including but not limited to multiple sclerosis \cite{Cercignani2001,Ciccarelli2003d,Miron2012}, vestibular schwannomas \cite{Chen2011b,Taoka2006,Wai2009}, tumors \cite{Mori2002,Schonberg2006,Nimsky2005}, schizophrenia \cite{Rametti2009,Fitzsimmons2009}, traumatic brain injury \cite{Bendlin2008,Kinnunen2011c,Hellyer2012}.
-RD is found to positively correlate with degree of demyelination in multiple sclerosis \cite{Song2005,Song2002,Klawiter2011,Janve2013}, and well as stages of white-matter development and aging \cite{Counsell2006,Sala2010}, including alzheimer's disease \cite{Acosta-Cabronero2012}.
-AD is correlated positively with axonal integrity \cite{Song2003}, and is found to be a correlate of axonal injury \cite{Budde2009} such as those found in traumatic brain injury \cite{Kinnunen2011c} and Gulf War illness \cite{Rayhan2013}.
-MD is correlated with tissue inflammation and edema \cite{Senda2012g}, and is found to correlate with diffusivity changes as a result of gliomas \cite{Stadlbauer2007}, multiple sclerosis \cite{Cercignani2001,Senda2012g,DeGroot2013}, amyotrophic lateral sclerosis \cite{Sharma2012}, and aging \cite{Kantarci2011}.
+FA has been found to correlate with a wide spectrum of neuromorphological changes, including but not limited to multiple sclerosis \shortcite{Cercignani2001,Ciccarelli2003d,Miron2012}, vestibular schwannomas \shortcite{Chen2011b,Taoka2006,Wai2009}, tumors \shortcite{Mori2002,Schonberg2006,Nimsky2005}, schizophrenia \shortcite{Rametti2009,Fitzsimmons2009}, traumatic brain injury \shortcite{Bendlin2008,Kinnunen2011c,Hellyer2012}.
+RD is found to positively correlate with degree of demyelination in multiple sclerosis \shortcite{Song2005,Song2002,Klawiter2011,Janve2013}, and well as stages of white-matter development and aging \shortcite{Counsell2006,Sala2010}, including alzheimer's disease \shortcite{Acosta-Cabronero2012}.
+AD is correlated positively with axonal integrity \shortcite{Song2003}, and is found to be a correlate of axonal injury \shortcite{Budde2009} such as those found in traumatic brain injury \shortcite{Kinnunen2011c} and Gulf War illness \shortcite{Rayhan2013}.
+MD is correlated with tissue inflammation and edema \shortcite{Senda2012g}, and is found to correlate with diffusivity changes as a result of gliomas \shortcite{Stadlbauer2007}, multiple sclerosis \shortcite{Cercignani2001,Senda2012g,DeGroot2013}, amyotrophic lateral sclerosis \shortcite{Sharma2012}, and aging \shortcite{Kantarci2011}.
 
@@
 \begin{center}
-\begin{tabular}{c}
-\includegraphics[width=100mm]{harmonics.pdf}
-\end{tabular}
-\caption{Visualization of the spherical harmonic functions up to l=3.}
+\includegraphics[width=\textwidth]{harmonics.pdf}
+\caption{Visualization of the spherical harmonic functions up to l=3. Modified from \protect\cite{sh-image-wikipedia}}
 \label{fig:harmonics}

### 2018-02-28 thesis/contents/reviews/multiplesclerosis.tex
Updates from ShareLaTeX

@@
 Both central and peripheral pain are reports in people of MS.
-Central pain in MS is reported to occur 27-58\% of the subjects
-\cite{OConnor2008}. The most common pain in MS include extremity pain and TN. TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population.
+Central pain in MS is reported to occur 27-58\% of the subjects \cite{OConnor2008}.
 
-TN secondary to MS, or symptomatic TN (MS-TN), is frequently indistinguishable from TN in terms of clinical presentation.
-It does however present with a demonstrable structural lesion other than vascular compression of CN V.
-MS-TN has a prevalence rate of about 2-4\% and in some studies, there are reports of prevalence rate as high as 6\%.
-Unilateral MS-TN is the most common pain type, however there is up to 30\% occurrences of bilateral pain.
+The most common pain in MS includes extremity pain and TN (MS-TN), which is also termed symptomatic TN.
+MS-TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population \cite{Cruccu2009}.
+The clinical presentation of the pain is similar to classic TN.
+MS-TN has a prevalence rate of about 2-4\%, and in some studies, there are reports of prevalence rate as high as 6\%.
+Unilateral MS-TN is the most common pain type.
+However, there are up to 30\% occurrences of bilateral pain.
+MS-TN pathophysiology is believed to be different from classic TN, as it is often accompanied by the presentation of pontine MS lesions \cite{Cruccu2009}.
+The treatment of MS-TN via ablative interventions is also less effective than in classic TN \cite{Nurmikko2009}.
+The exact pathophysiological differences between MS-TN and classic TN, however, are not known.
+While it shares the feature of vascular compression with classic TN, its relevance to MS-TN pain has not been precisely determined \cite{Nurmikko2009}.

### 2018-02-28 contents/glossary.tex
Updates from ShareLaTeX

\newacronym{tn}{TN}{Trigeminal Neuralgia}
\newacronym{dwi}{DWI}{Diffusion Weighted Imaging}
\newacronym{dti}{DTI}{Diffusion Tensor Imaging}
\newacronym{hardi}{HARDI}{High angular resolution diffusion imaging}
\newacronym{mri}{MRI}{Magnetic Resonance Imaging}
\newacronym{fmri}{fMRI}{Functional Magnetic Resonaonce Imaging}
\newacronym{dsi}{DSI}{Diffusion Spectrum Imaging}
\newacronym{sh}{SH}{Spherical Harmonics}
\newacronym{fod}{FOD}{Fiber orientation distribution}
\newacronym{csd}{CSD}{Constrained spherical deconvolution}
\newacronym{fa}{FA}{Fractional Anisotropy}
\newacronym{rd}{RD}{Radial Anisotropy}
\newacronym{ad}{AD}{Axial Anisotropy}
\newacronym{md}{MD}{Mean Anisotropy}
\newacronym{adc}{ADC}{Apperant diffusion coefficient}
\newacronym{ms}{MS}{Multiple Sclerosis}
\newacronym{mstn}{MS-TN}{Symptomatic Trigeminal Neuralgia Secondary To Multiple Sclerosis}
\newacronym{trg}{TRG}{Trigeminal Root Ganglion}
\newacronym{gp}{GP}{Gaussian Process}
\newacronym{svm}{SVM}{Support Vector Machine}
\newacronym{cnv}{CN V}{Cranial Nerve 5, the trigeminal nerve}
\newacronym{dmri}{dMRI}{Diffusion Magnetic Resonance Imaging}
\newacronym{nvc}{NVC}{Neurovascular Compression}
\newacronym{rez}{REZ}{Root Entry Zone}
\newacronym{tpt}{TPT}{Trigeminopontothalamic tract}
\newacronym{s1}{S1}{primary sensory cortex}
\newacronym{s2}{S2}{secondary sensory cortex}
\newacronym{acc}{ACC}{anterior cingulate cortex}
\newacronym{ofc}{OFC}{orbitofrontal cortex}
\newacronym{pfc}{PFC}{prefrontal cortex}
\newacronym{nac}{NAc}{nucleus accumbens}
\newacronym{bold}{BOLD}{Blood-Oxygen-Level Dependent}
\newacronym{vmpo}{VMpo}{ventralmedial posterior thalamus}
\newacronym{te}{TE}{Echo Time}
\newacronym{tr}{TR}{Repetition Time}
\newacronym{gre}{GRE}{Gradient Echo}
\newacronym{dnn}{DNN}{Deep neural network}
\newacronym{eeg}{EEG}{Electroencephalography}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{pca}{PCA}{Principal components analysis}
\newacronym{epi}{EPI}{Echo Planar Imaging}
\newacronym{ap}{AP}{Anisotropic Power Image}
\newacronym{sagit}{SAGIT}{Selective automated group integrated tractography}
\newacronym{nos}{NOS}{Normalized overlap score}
\newacronym{xst}{XST}{eXtended Streamline Tractography}
\newacronym{flair}{FLAIR}{Fluid-attenuated inversion recovery imaging}
\newacronym{mprage}{MP-RAGE}{Magnetization-prepared 180 degrees radio-frequency pulses and rapid gradient-echo imaging}
\newacronym{mdwi}{MDWI}{Mean diffusion weighted image}
\newacronym{cc}{CC}{Cross correlation}
\newacronym{MI}{MI}{Mutual information}

\glsaddall

### 2018-02-28 contents/reviews/dmri.tex
Updates from ShareLaTeX

--- 2018-02-28 thesis/contents/reviews/dmri.tex
@@
 The most wide-spread strategy is to equally distribute gradient directions as vertices of a triangle-tessellated geodesic sphere \cite{Frank2001}.
-Alternatively, it is also possible to distribute gradient directions in a rectilinear grid, which forms the basis of diffusion spectral imaging (DSI) \cite{Wedeen2005}.
+Alternatively, it is also possible to distribute gradient directions in a rectilinear grid, which forms the basis of diffusion spectrum imaging (DSI) \cite{Wedeen2005}.
 Additional acquisition strategies such as distributing gradient directions across multiple concentric spherical shells are also actively explored \cite{Jeurissen2014a}.

### 2018-02-28 contents/reviews/mri.tex
Updates from ShareLaTeX

--- 2018-02-02 thesis/contents/reviews/mri.tex
@@
 T1 contrast is adjusted by tuning repetition time (TR), which is the length of relaxation interval between excitation pulses.
-When TR is long (> 1500 ms), $M_z$ recovers totally, and therefore subsequent excitation will again retain maximal signal strength.
-When the TR is short (< 600 ms), the $M_z$ has not entirely recovered, and the next excitation pulse will not be able to regain signal strength.
+When TR is long (\textgreater 1500 ms), $M_z$ recovers totally, and therefore subsequent excitation will again retain maximal signal strength.
+When the TR is short (
+\textless 600 ms), the $M_z$ has not entirely recovered, and the next excitation pulse will not be able to regain signal strength.
 Different tissues have different intrinsic T1 differences, and will show strong contrast under short TR.

### 2018-02-28 contents/tn_sagit.tex
- sync tn-sagit changes

--- 2018-02-28 thesis/contents/tn_sagit.tex
@@
 TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}.
-The pathophysiology of classic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although classic TN can also occur without evidence of vascular compression \cite{Lee2014}.
+The pathophysiology of classic TN is thought to be vascular compression at the root of the CN V nerve entry zone (REZ) \cite{Linn2011,Love2001}, although classic TN can also occur without evidence of vascular compression \cite{Lee2014}.
 While the focus of the study of TN has been primarily the peripheral segments of the trigeminal nerve, we recently compared TN and MS-TN diffusivity differences in four segments of the CN V, including at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.
@@
 MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions.
-T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375 mm, with slice spacing = 1 mm, TE = 5.052 ms, TR = 11.956 ms, flip angle = $20^\circ$, FOV = 240 mm, and matrix = 256x256.
-DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3 mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, where b0 = 1000 s/$mm^2$, TE = 88.6 ms, TR = 17000 ms, flip angle = $90^\circ$, field of view (FOV) = 120 mm, and matrix = 128x128.
+T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375 mm, with slice spacing = 1 mm, TE = 5.052 ms, TR = 11.956 ms, flip angle = $20^\circ$, field of view (FOV) = 240 mm, and matrix = 256x256.
+DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3 mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, where b0 = 1000 s/$mm^2$, TE = 88.6 ms, TR = 17000 ms, flip angle = $90^\circ$, FOV = 120 mm, and matrix = 128x128.
 

### 2018-03-01 contents/discussions.tex
Updates from ShareLaTeX

--- 2018-02-28 thesis/contents/discussions.tex
@@
 The earliest work on the feasibility of accurate reconstruction of cranial nerve with DTI tractography \cite{Hodaie2010} dates to 2010.
-This study was able to image CN II, III, and importantly CN V.
-Similarly, we were able to establish the methods to image the TRG and the cisternal CN V with manual delineation, followed by visualization of the CN V postganglionic GK target, prior and after surgery \cite{Hodaie2012g}.
-We found alternations in both the CN V reconstruction, as well as FA in the surgical target location, that improved our understanding beyond what is visible with post-surgical gadolinium enhancements.
-This has now led our team to study whether CN V GK target diffusivity at 6 months post-surgery can predict long-term pain relief in TN \cite{Tohyama2018}.
-Similarly whether the diffusivity changes are specific enough that they can serve as predictors of outcome \cite{Hung2017}.
+This study was able to image the optic nerve (CN II), oculomotor nerve (CN III), and importantly CN V.
+Similarly, we were able to establish the methods to image the trigeminal ganglion and the cisternal CN V with manual delineation, followed by visualization of the CN V area of radiosurgery target, prior and after treatment \cite{Hodaie2012g}.
+We found alternations in both the CN V reconstruction, as well as FA in the surgical target location, that improved our understanding beyond what is visible with post-surgical MRI gadolinium enhancements.
+This has now led our team to study whether CN V GK target diffusivity at 6 months post-surgery can predict long-term pain relief in TN \cite{Tohyama2018}, and similarly, whether the diffusivity changes are specific enough that they can serve as predictors of outcome \cite{Hung2017}.
 The tractography study of TN has been directly linked with the study of gray and white matter abnormalities of the TN brain \cite{Desouza2013c}.
@@
 While TN is associated with compression of the REZ and possible associated focal central demyelination \cite{Devor2002a,Peker2006}, MS is associated with inflammation and myelin changes resulting in CNS plaques.
-While both diseases involve the same clinical expression of pain, the pathophysiology of pain differs, but this had not been shown directly using DTI based imaging techniques.
+While both diseases share a similar clinical expression of pain, the pathophysiology of pain differs, however this had not been shown directly using DTI based imaging techniques.
 From a technical perspective, the identification of the CN V nerve inside the brainstem from MR images had been previously highly inconsistent, and the precision of multiple ROI placements could not be maintained.
 Multi-tensor tractography was instrumental in the feasibility of the method.
-The study opened the possibility that multi-tensor tractograpy aided analysis added a measure of objective assessment in examining pain conditions, and may facilitate similar studies that provide insight into the fundamental changes that occur at the level of white matter.
+The study opened the possibility that multi-tensor tractography aided analysis added a measure of objective assessment in examining pain conditions, and may facilitate similar studies that provide insight into the fundamental changes that occur at the level of white matter.
 
 As a result of the methodology developed in Study I, a recent publication details the investigation of whether presurgical CN V segment diffusivity values can predict response to TN surgery \cite{Hung2017}.
-The important study found that lowered cistern MD and AD pre-surgery was correlated with positive surgical outcome, whereas lowered REZ FA and increased pontine AD was correlated with the lack of response from surgery.
+This important study found that lowered cistern MD and AD pre-surgery was correlated with positive surgical outcome, whereas lowered REZ FA and increased pontine AD was correlated with the lack of response from surgery.
 This provided further evidence of the role of cisternal CN V diffusivities in the study of TN.
-The fact that the novel finding of pontine diffusivity in classic TN favours a non-responder outcome points to the role of CNS alterations in classic TN.
+The fact that the novel finding of the role of pontine diffusivity in classic TN favours a non-responder outcome points to the role of CNS alterations in classic TN.
 
-Study IV (Chapter \ref{section:study4}), applied group tractography to the study of the trigeminal sensory pathway, and highlighted the observation that the unaffected CN V diffusivity in TN differs from that of controls.
+Study IV (Chapter \ref{section:study4}), applied group tractography to the study of the trigeminal sensory pathway, and highlighted the observation that the unaffected CN V diffusivity in TN differs significantly from that of controls.
 There have been other studies suggesting bilateral abnormalities in TN.
 A group level diffusivity study of TN was conducted by DeSouza et al.
-in 2013 \cite{Desouza2013} presented similar findings.
-DeSouza manually defined ROI in the trigeminal REZ, as well as performing full-brain Tract-Based Spatial Statistics.
+in 2013 \cite{Desouza2013} presented similar findings, with significant differences between affected and unaffected sides, but also between unaffected sides and controls.
+DeSouza manually defined the ROI in the trigeminal REZ, as well as performing full-brain Tract-Based Spatial Statistics.
 Bilateral differences in the REZ between TN patients and controls were found in MD, RD, and AD.
-Wide-spread white matter differences were also found, especially in the corpus callosum, cingulum, and longitudinal fasciculus.
-DeSouza had also found increased cortical and grey-matter volume changes in TN patients \cite{Desouza2013c}, including increase in contralateral S1 and thalamus volumes.
+Widespread white matter differences were also found, especially in the corpus callosum, cingulum, and longitudinal fasciculus.
+
 Using SAGIT in study IV, the ROIs were projected to each individual subject using an automated pipeline.
-The GP classifier produced results by exhaustively examine moving windows at all position and sizes, and the results are auto-discovered by the algorithm, in contrast to manual-placement of the measurement ROI in previous studies.
+The GP classifier produced results by exhaustively examining moving windows at all position and sizes, and the results are auto-discovered by the algorithm, in contrast to manual-placement of the measurement ROI in previous studies.
 There is therefore consistency in CN V findings in the location and diffusivity across both manual and automated methods.
 
-The observation of bilateral differences has important implications in TN pain.
+The observation of bilateral differences has important implications in our understanding of TN pain.
 From previous studies, it is clear that altered diffusivities in the cisternal and pontine segments of CN V are correlated with TN pain in the affected side.
-However there are now a number of studies that point to bilateral REZ/pontine CN V diffusivities being altered in unilateral TN, although each point to significant differences between affected and unaffected sides as well.
+However, there are now a number of studies that point to bilateral REZ/pontine CN V diffusivities being altered in unilateral TN, although each point to significant differences between affected and unaffected sides as well.
 The exact nature of bilateral diffusivity alterations is not yet clear.
@@
 
-The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain.
-In many studies it has also been used as a basis of comparison.
+Importantly, the unaffected side has been previously presumed to be similar to healthy controls due to the absence of pain.
+In many studies, it has also been used as a basis of comparison.
 With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between pathological and healthy tissue.
-Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what point does the clinical expression of pain start.
+Additionally, the logical next question is to determine that if the unaffected nerve differs from controls, at what point does the clinical expression of pain start.
 Investigations into bilateral TN pain, and vascular compression in the absence of pain, may reveal additional insight into the relationship between white matter changes and TN pain in general.
@@
 Prior to this, the ROI identification was done visually, and sometimes with simple heuristics methods based on distance from surrounding landmarks.
-Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be always assessed.
+Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not always be assessed.
 Improvements in tractography of the brainstem trigeminal nerve allowed the shift towards more algorithmic driven and automated segmentation and ROI placement.
 
-Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter remains a limitation of diffusion tractography.
-The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure therefore poses a fundamental challenge for neuroimaging.
+Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter remains a limitation.
+The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure, therefore, poses a fundamental challenge for neuroimaging.
 Techniques to improve visualization of white matter fibers consist of probabilistic tractography, which uses tissue priors to sample a large number of directions at each step stochastically, and combines the final result as a volumetric probability map.
@@
 Reduction in acquisition time for these algorithms is currently an active area of research.
+
 The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), are a significant development.
@@
 CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions.
-Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.
+Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with at least 50 gradient directions.
 
@@
 
-As part of our focusing on ???
-and for the purposes of imaging the trigeminal pathways, where a significant number of crossing fibers exist, we divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
+For the purposes of imaging the trigeminal pathways, we divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern.
 The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure.
-In theory the number of divisions could be infinitely large, given that a centroid curve was identified.
+In theory, the number of divisions could be infinitely large, given that a centroid curve was identified.
 Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths are variable.
 White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales.
-A generalized algorithm that can reliably calculate a centroid for these structure is of great value and yet still a topic of research.
+A generalized algorithm that can reliably calculate a centroid for these structures is of great value and yet still a topic of research.
 Merged tractography simplified this task by allowing a single curve to be defined manually in the template tractography space.
-This greatly simplied the task, since centroids no longer need to be defined repeatedly for each subject.
+This greatly simplified the task, since centroids no longer need to be separately defined  for each subject.
 Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.
@@
 
-The findings in S1 in Study IV (Chapter \ref{section:study4}) showed TN diffusivity differences in regions surrounding the thalamus, as well as in the mid-segments of the S1 pathways.
-These findings were in agreement with pain literature, where thalamus and S1/S2 activations are often reported as part of the pain neuromatrix.
+The findings relating to S1 in Study IV (Chapter \ref{section:study4}) showed TN diffusivity differences in regions surrounding the thalamus, as well as in the mid-segments of the S1 pathways.
+These findings were in agreement with previously published literature in this field, where thalamus and S1/S2 activations are often reported as part of the pain neuromatrix \cite{Melzack2008,Moayedi2012a}.
 The TPT pathway resulted in differences near the TGN and the thalamus, areas that are expected to be involved in TN pain.
@@
 We demonstrated in Study III, that different tractography algorithms exhibit different geometric point distributions.
-In other words, given two tracts generated from identical ROI definitions, with similar cross-sectional area, but of different algorithm, the underlying spacial distribution of the points are different.
-It can be argued that we average the point-metrics during along-the-tract analysis, which may be considered a form of smoothing/down-sampling, and therefore the spatial distribution of the tractography algorithm might not be important.
+In other words, given two tracts generated from identical ROI definitions, with similar cross-sectional area, but of different algorithm, the underlying spatial distribution of the points may be different.
+It can be argued that we average the point-metrics during along-the-tract analysis, which may be considered a form of smoothing/down-sampling, and therefore the spatial distribution of the tractography algorithm might not be as important.
 However, until the exact influence of spatial distribution have on the measured distribution of diffusivity metrics can be understood, comparing between similar tracts from two different tractography algorithms should be avoided.
@@
 This lead to some challenges in the scalability of the analytical process, but also opportunities in leveraging more expressive machine learning approaches.
-The need for data to improve diagnosis and treat of disease, particularly in the emergence of artificial intelligence in medical imaging is increasingly been recognized.
+The need for data to improve diagnosis and treat of disease, particularly in the emergence of artificial intelligence in medical imaging is increasingly being recognized.
 Data is now the foundation from which new techniques and discoveries are possible.
@@
 The need to include large amount of variables, patient attributes, study measures, time sequences, can now provide valuable insight that is beyond the mean of traditional statistics.
-In this thesis we developed an machine learning approach that is scalable, and one that can quantify uncertainty as demanded by the medical need.
-The data processing and machine learning is end-to-end, meaning that the system can be practically deployed in a clinical setting.
+In this thesis, we developed a machine learning approach that is scalable, and one that can quantify uncertainty as demanded by the medical need.
+The data processing and machine learning pipelineis end-to-end, meaning that the system can be practically deployed in a clinical setting.
 
@@
 The analysis of the trigeminal nerve in TN versus controls agreed with our earlier findings, and therefore validates the study's GP classification method.
-We performed traditional multiple comparison statistical analysis for our model, but found the result hard to interpret.
-This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics.
+Traditional multiple comparison statistical analysis in this setting becomes difficult to interpret.
 We therefore decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which may not fit with frequentist assumptions.
+
 The GP classifier was chosen over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency.
-2) SVM results are hard to interpret, whereas GP is a bayesian method, and therefore it is much more intuitive.
+2) SVM results are hard to interpret, whereas GP is a Bayesian method, and therefore it is much more intuitive.
 More specifically, the streamline measurements resemble time-series data, in that it exhibits positional dependency between measurements.
 Unlike a time-series however, the streamline measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency, as opposed to uni-directionally dependent in time.
-Posing the study as a supervised machine learning problem, the feature vector would be a N dimension vector representing N measures at each division, and a classification label.
-SVM would assume the features to be points in a N-dimensional space, and each measurement would assume independence.
+Posing the study as a supervised machine learning problem, the feature vector would be an N-dimension vector representing N measures at each division, and a classification label.
+SVM would assume the features to be points in an N-dimensional space, and each measurement would assume independence.
 This makes the result of the SVM output difficult to interpret.
@@
 Therefore the GP model is able to take the positional dependency into account.
-Changing the GP classification model into a regression one is very simple.
-This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a single patient's CN V, what is the probability that is resembles a pathological population as opposed to a healthy population?
+Changing the GP classification model into a regression one is relatively simple.
+This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a single patient's CN V, what is the probability that the measurement distribution resembles a pathological population as opposed to a healthy population?
 It is possible to answer these questions trivially with the GP model using regression.
@@
 Unlike major white matter bundles, cranial nerves diameters are in the millimetre range, and therefore voxel sizes greater than 2 mm may result in partial volume averaging.
-Achieving 2mm isovoxel on a 3T scanner requires 32 channel head coil.
+Achieving 2 mm isovoxel on a 3T scanner requires a 32 channel head coil.
 Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable.
@@
 Neurosurgical visualization however requires three dimensional representation of the fibers, and therefore streamline-based tractography visualizations are of broader appeal.
-In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions.
+In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater than 50 gradient directions.
 
-An important technical point, but rarely mentioned in literally, is the tractography seeding pattern.
-Therefore I will address it here: When tracing the streamlines, an algorithm will need to generate a number of starting points in a defined volume.
+An important technical point consists in the tractography seeding pattern.
+When tracing the streamlines, an algorithm will need to generate a number of starting points in a defined volume.
 Common patterns in prepackaged software include in a regularly spaced grid and random sampling.
@@
 Another consideration is the reproducibility of the desired anatomy, as randomly seeded bundles cannot be reproduced deterministically.
-Grid seeding may result in sparse spacing of the tract bundles when viewed from a certain angle, while random seeding can reduce this kind of visual artefact.
+Grid seeding may result in sparse spacing of the tract bundles when viewed from a certain angle, while random seeding can reduce this kind of visual artifact.
 A good balance might be to have a reproducible sampling scheme that also is robust to different viewing directions, for example a phyllotaxis pattern.
@@
 Tractography is a powerful methodology that permits the visualization of white matter fibers in three dimensions.
-Its primary use however has been limited to case studies and small-scale anatomy comparative analysis.
+Its primary use, however, has been limited to case studies and small-scale anatomy comparative analysis.
 The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of larger groups.
@@
 Efforts in MRI medical image registration have paved the way to this end, especially in the fields of diffeomorphic registration.
-The difficulty in adapting registration to tractography is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods, as they are computationally untenable.
-To date there is no direct non-linear 3D deformation methods for tractography.
+The difficulty in adapting registration to tractography is that tractography data has 3D geometry, and therefore is extremely difficult for analytical methods.
+To date, there are no direct non-linear 3D deformation methods for tractography.
 The best efforts to date apply affine transformation for between-tract registrations \cite{Garyfallidis2015}.
-Our approach is to recast tractography as a brain segmentation method while also make use of its orientation information.
-Tractography then can be considered a point-based sub-sampling method that is represented in the same coordinate frame as the native DWI space.
+Our approach is to recast tractography as a brain segmentation method while also making use of its orientation information.
+Thus tractography can be considered a point-based sub-sampling method that is represented in the same coordinate frame as the native DWI space.
 Any transforms applied to this DWI space is then applicable also to any tractography models that share its coordinate frame.
@@
 Any registration errors in these two steps will become multiplicative in the end result.
-T1 registration is well studied, however T1---DWI registration has few literature.
+T1 registration is well studied, however T1---DWI registration has limited literature.
 This demonstrates the need to investigate the best method to minimize registration errors that were compatible with our image data format.
@@
 Despite this, certain regions such as the TPT could not be reliably reconstructed in all of the subjects.
-Anatomy featuring intricate but complex cross-sections, such as the thalamoinsular projections, which involves the claustrum, and medullary decussations of the spinal trigeminal fibers will remain out of reach of the current algorithms without substantial improvement in DWI resolution.
+Anatomy featuring intricate but complex cross-sections, such as the thalamoinsular projections which include the claustrum, and medullary decussations of the spinal trigeminal fibers will remain out of reach of the current algorithms without substantial improvement in DWI resolution.
 Fortunately, SAGIT can adopt MR images of any resolution, so future studies in higher isovoxel resolution can be immediately processed as soon as they become available.
@@
 
-The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers,  are not limited to the application of TN.
+The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers, are not limited to the application to the study of TN.
 The methods can generalize to any white matter anatomy where consistent tracts can be obtained at a group level.
@@
 
-I) Keep pushing the limit of anatomical delineations towards other areas of the trigeminal system.
-This include the spinal trigeminal pathways, which include the spinal trigeminal ganglion and its sub-ganglions, and its medullary decussations.
+I) Advancing the limit of anatomical delineations towards other connections of the trigeminal system.
+This includes the spinal trigeminal pathways, which include the spinal trigeminal ganglion and its sub-ganglions, and its medullary decussations.
 Re-examine TPT tracts, both to establish its diffusivity baseline, and to confirm its asymmetry.
 Image the medial pain pathways, and descending projections to the PAG.
-These tasks are currently limited by the resolution of availabile DWI images.
+These tasks are currently limited by the resolution of available DWI images.
 Immediate gains may be obtained by using SAGIT to delineate these structure on the Human Connectome Project (HCP) dataset.
@@
 II) Re-examine TN and other pain in the MS population with S1/S2 and TPT pathways.
-The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal additional insights in the MS population.
-Where focal lesion groups can be correlated with disease symptoms such as pain and motor disabilities.
-Comparitive studies with other trigeminal pain syndromes such as Herpetic neuralgia and trigeminal deafferentation pain would also be a valuable direction.
+The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal additional insights in the MS population, where focal lesion groups can be correlated with disease symptoms such as pain and motor disabilities.
+Comparative studies with other trigeminal pain syndromes such as post herpetic neuralgia and trigeminal deafferentation pain would be a valuable direction.
 
@@
 Investigate how to diagnose and provide treatment targets from MR images with minimal turn around time.
-This can involve apply GP regression to the prediction of patient disease, and multi-modal learning, possibly with deep neural networks.
+This can involve applying GP regression to the prediction of patient disease, and multi-modal learning, possibly with deep neural networks.
 Gaussian Process can be used as a regressor to model tissue parameters along merged white-matter bundles.
@@
 
-We observed that merged tractography can delineate more complete pathway delineations than from any single subject.
-This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high quality templates.
-Therefore merged tractography may serve to produce high quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals.
+We observed that merged tractography could delineate more complete pathways than any single subject.
+This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high-quality templates.
+Therefore merged tractography may serve to produce high-quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals.
 We have explored the use of SAGIT in generating the first neonatal porcine tractography template \cite{Zhong2016a}.
 Such template does not exist for humans to the best of our knowledge, and can be readily created from Human Connectome Project data.
-In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low resolution clinical acquisitions, these template tracts may provide a standard population distribution for GP regression, and can used as proxy to enable advanced analysis.
+In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low-resolution clinical acquisitions, these template tracts may provide a standard population distribution for GP regression, and can be used as a proxy to enable advanced analysis.
 
 The pain neuromatrix, as studied by various functional and structural connectivity studies, distributes over the full range of neocortical areas.
-These areas are associated with complex white-matter pathways that extends to all regions of the brain.
+These areas are associated with complex white-matter pathways that extend to all regions of the brain.
 These include sharp curvatures of the anterior pain pathways of the ACC, the distant projections of the temporal cortices.
-Some pathways are just beyond the feasibility of current methods, these include the projections of the insula and claustrum, small decending pathways to PAG, and the spinal trigeminal pathways in the brainstem.
-Detailed examination of the pain matrix, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest.
+Some pathways are just beyond the feasibility of current methods.
+These include the projections of the insula and claustrum, small descending pathways to PAG, and the spinal trigeminal pathways in the brainstem.
+Detailed examination of the pain matrix with the inclusion of these pathways, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest.
 Multi-modal and multi-variable study with functional connectivity with this technique would also be possible.
 
-Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D.
+Tractography, in conjunction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise difficult to visualize directly, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualize in 2D.
 We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}.
-Another contribution is to visualize tissue changes from surgical interventions that can otherwise be invisible.
-There are cases where post-operative GammaKnife ablation target on the trigeminal nerve did not result in dicernable gadonenium enhancement in standard MR, but the nerve FA changes can be seen in the delineated tractography.
+Another contribution is to visualize tissue changes from surgical interventions.
+There are cases where post-operative GammaKnife ablation target on the trigeminal nerve did not result in discernible gadolinium enhancement in standard MR, but the nerve FA changes can be seen in the delineated tractography.
 Therefore tractography can provide both spatial, as well as tissue-based insights to the surgeons to aid in assessment.
@@
 Therefore advances in 3D printing can be easily adopted in neurosurgery to allow surgeons to physically examine tractography reconstructions.
-On the otherhand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making.
+On the other hand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making.
 Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.

### 2018-03-01 contents/paper1.tex
Updates from ShareLaTeX

--- 2017-12-02 thesis/contents/paper1.tex
@@
-\graphicspath{{images/mspaper/drafts/}}
+\graphicspath{{images/mspaper/}}
 
@@
 
-This study was published in Multiple Sclerosis Journal \textcopyright 2016 Chen et al. :
+This study was published in Multiple Sclerosis Journal \textcopyright 2016:
 
@@
 Classic trigeminal neuralgia is presumed to occur as a result of neurovascular compression at the root entry zone (REZ) of CN V (also referred as ‘idiopathic’; henceforth referred to simply as TN)(M Hodaie \& Coello, 2013).
-TN pain symptoms can also occur secondary to multiple sclerosis (MS-TN), where MS patients have a 20-fold risk of developing TN pain than the general population (up to 0.1\%)\cite{VanHecke2014}.
-While experientially similar, the pathophysiology of MS-TN differs from TN and involves CNS demyelination\cite{lazar1979trigeminal,Nurmikko2009}.
+TN pain symptoms can also occur secondary to multiple sclerosis (MS-TN), where MS patients have a 20-fold risk of developing TN pain than the general population (up to 0.1\%) \cite{VanHecke2014}.
+While experientially similar, the pathophysiology of MS-TN differs from TN and involves CNS demyelination \cite{lazar1979trigeminal,Nurmikko2009}.
 The manner in which CNS demyelination results in pain is unclear.
 Brainstem plaques can be common in MS, but not all MS patient with brainstem plaques suffer from MS-TN.
-Furthermore, the presence of bilateral plaques does not clearly correlate with bilateral MS-TN pain\cite{DeSanti2011a,Love2001a}, highlighting the complex and uncertain relationship between the presence of plaques and pain\cite{DaSilva2005}.
-Importantly, this also highlights the fact that conventional MR plaque visualization does not fully correlate with MS clinical symptoms or their severity, and so improved neuroimaging methods are needed\cite{Seixas2014}.
-As classic TN is most commonly associated with vascular compression at the trigeminal nerve root entry zone, it is possible that in some MS patients neurovascular compression can also be present, adding further complexity to the diagnosis and treatment of trigeminal pain\cite{Sandell2010e}.
+Furthermore, the presence of bilateral plaques does not clearly correlate with bilateral MS-TN pain \cite{DeSanti2011a,Love2001a}, highlighting the complex and uncertain relationship between the presence of plaques and pain\cite{DaSilva2005}.
+Importantly, this also highlights the fact that conventional MR plaque visualization does not fully correlate with MS clinical symptoms or their severity, and so improved neuroimaging methods are needed \cite{Seixas2014}.
+As classic TN is most commonly associated with vascular compression at the trigeminal nerve root entry zone, it is possible that in some MS patients neurovascular compression can also be present, adding further complexity to the diagnosis and treatment of trigeminal pain \cite{Sandell2010e}.
 
 Understanding the microstructural anatomy of the trigeminal fibres is essential to determine how MS affects CN V brainstem fibres and how alterations to CN V in MS are related to the clinical presentation of pain.
-Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure 1A-B)\cite{Cruccu2009,DaSilva2005}.
+Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure 1A-B) \cite{Cruccu2009,DaSilva2005}.
 Conventional MR imaging cannot visualize in isolation the brainstem fibres of CN V.
-Furthermore conventional tractography cannot resolve the brainstem crossing fibres\cite{Farquharson2013}, which leads to unreliable CN V delineations within the brainstem, and inaccuracies in diffusivity measurements.
+Furthermore conventional tractography cannot resolve the brainstem crossing fibres \cite{Farquharson2013}, which leads to unreliable CN V delineations within the brainstem, and inaccuracies in diffusivity measurements.
 
@@
 Metrics including fractional anisotropy (FA), and radial (RD), axial (AD), and mean diffusivities (MD), reflect diffusivity change measured as a parameter of the estimated Gaussian diffusion tensor eigenvectors.
-FA reflects the shape of the ellipsoid Gaussian tensor as a measure of anisotropy, and is correlated with both structural and cognitive changes that follow adverse brain changes such as traumatic brain injury, aging and Alzheimer’s disease\cite{Bendlin2008,Charlton2006,Kinnunen2011c,Mielke2012}.
+FA reflects the shape of the ellipsoid Gaussian tensor as a measure of anisotropy, and is correlated with both structural and cognitive changes that follow adverse brain changes such as traumatic brain injury, aging and Alzheimer’s disease \cite{Bendlin2008,Charlton2006,Kinnunen2011c,Mielke2012}.
 RD is a measure of the cross-sectional diffusion as an average of the second and third diffusion eigenvalues, and has been shown to positively correlate with axonal demyelination \cite{Song2002,Song2005}.
-AD is the greatest diffusion eigenvalue and has been positively correlated with axonal degeneration\cite{Budde2009,Song2002}, and MD is the mean of all three diffusion eigenvalues and is related to the increase in tissue inflammation and edema\cite{Beaulieu2002}.
+AD is the greatest diffusion eigenvalue and has been positively correlated with axonal degeneration \cite{Budde2009,Song2002}, and MD is the mean of all three diffusion eigenvalues and is related to the increase in tissue inflammation and edema \cite{Beaulieu2002}.
 In MS studies, FA, RD and MD appear abnormal in or near MS lesions.
-The changes in diffusivity alter the Gaussian tensor profile and results in decreased FA, increased RD and increased MD\cite{Braley2012,Janve2013,Klawiter2011,Lin2007,Senda2012g}.
-While AD is generally not correlated with demyelination\cite{Song2002}, there have been findings associating decreases in AD with the presence of neurological damage in MS\cite{Budde2008,Kim2006}.
+The changes in diffusivity alter the Gaussian tensor profile and results in decreased FA, increased RD and increased MD \cite{Braley2012,Janve2013,Klawiter2011,Lin2007,Senda2012g}.
+While AD is generally not correlated with demyelination \cite{Song2002}, there have been findings associating decreases in AD with the presence of neurological damage in MS \cite{Budde2008,Kim2006}.
 
-The major technical limitation of diffusion tractography to date has been the inability of resolving crossing fibres using a single tensor Gaussian diffusion model\cite{Farquharson2013}, which has proved unsuitable for the visualization of brainstem tracts.
+The major technical limitation of diffusion tractography to date has been the inability of resolving crossing fibres using a single tensor Gaussian diffusion model \cite{Farquharson2013}, which has proved unsuitable for the visualization of brainstem tracts.
 However, techniques that use high angular resolution imaging  with an increased number of gradient directions acquired during diffusion weighted imaging (DWI), and advances in multi-tensor tractography (MTT) have improved the ability to delineate fibres in brain regions that have dense fibre crossings, such as the brainstem \cite{DellAcqua2007,Descoteaux2009c,Fillard2011,Qazi2009}.
-The microstructure of CN V in TN has recently been shown to be abnormal in terms of DTI metrics in the trigeminal REZ\cite{Desouza2013}.
+The microstructure of CN V in TN has recently been shown to be abnormal in terms of DTI metrics in the trigeminal REZ \cite{Desouza2013}.
 DTI metrics, as correlates of microstructural anatomy, are also important in identifying the nerve in pain in TN\cite{Hodaie2009a}.
@@
 
-Images were analyzed with 3D Slicer (NA-MIC©, http://www.slicer.org)\cite{Pieper2004}.
+Images were analyzed with 3D Slicer (NA-MIC©, http://www.slicer.org) \cite{Pieper2004}.
 DWI sequences were corrected for eddy-current and motion distortions with appropriate corrections to gradient vectors, and imported into 3D Slicer for single-tensor diffusion tractography (SDT), where tracts were generated with initial seed spacing=0.5 mm, initial seeding FA threshold=0.2, stopping FA threshold=0.1, curvature threshold=0.8 and integration distance=0.1.
@@
 Patient group is considered an independent and between-subject factor; symptomatic and asymptomatic sides within the same patient are considered within-subject measures; in controls, the mean measures of left/right sides are used for comparison against other groups.
-A linear mixed model\cite{Pinheiro2011} is used to determine the statistical significance of within- and between-subject effects.
-Post hoc analyses were performed using pair-wise Student’s t-test, where multiple comparisons are corrected with false discovery rate correction\cite{Hothorn2008}.
-Statistical results were plotted\cite{Hunter2007} as diffusivity difference estimate matrices between anatomical regions and different subject groups.
+A linear mixed model \cite{Pinheiro2011} is used to determine the statistical significance of within- and between-subject effects.
+Post hoc analyses were performed using pair-wise Student’s t-test, where multiple comparisons are corrected with false discovery rate correction \cite{Hothorn2008}.
+Statistical results were plotted \cite{Hunter2007} as diffusivity difference estimate matrices between anatomical regions and different subject groups.
 Values are mapped as red/blue based on mixed linear regression estimates.
@@
 The coefficient matrices (Figures 4-7) revealed specific patterns demonstrating TN-specific diffusivity changes in the cistern and MS-TN-specific changes in the peri-lesional regions.
-In the TN REZ, FA changes are similar to the results found by DeSouza et al (Desouza, Moayedi, Chen, Davis, \& Hodaie, 2013; DeSouza et al., 2014), where the FA of the TN(+; pain-affected side) CN V is lower than that of TN(-; non-affected side).
+In the TN REZ, FA changes are similar to the results found by DeSouza et al \cite{Desouza2013,Desouza2013c}, where the FA of the TN(+; pain-affected side) CN V is lower than that of TN(-; non-affected side).
 Additionally, while RD and MD changes in the TN REZ showed significant differences, no AD differences in the REZ were detected, suggesting that the nature of the diffusivity changes may be due to demyelination of the central myelin within the REZ.
@@
 Diffusivity disruptions of CN V in MS-TN patients are likely due to MS plaques at the regions proximal to the main sensory nucleus.
-While AD has been associated with neurological disruptions in MS\cite{Budde2009,Kim2006}, the present study showed a lack of change in AD.
+While AD has been associated with neurological disruptions in MS \cite{Budde2009,Kim2006}, the present study showed a lack of change in AD.
 This may relate to the observation that we did not select subjects based on their disease severity, but based on their pain.

### 2018-04-16 contents/abstract.tex
Updates from ShareLaTeX

--- 2017-12-02 thesis/contents/abstract.tex
@@
+\chapter{Abstract}
+
 Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches.
 TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear.
-Single-tensor diffusion tensor neuroimaging (DTI) studies of TN revealed diffusivity changes in the cistern CNV.
+Single-tensor diffusion tensor neuroimaging (DTI) studies of CN V revealed diffusivity changes in the cistern segment.
 However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers.
@@
 There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments.
-2) to apply the method developed to the analysis of TN in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further.
+2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further.
 The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway.
 Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve.
-In Study II, we discover the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image.
+In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image.
 In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) software framework.
-Finally, in Study IV, we present an end-to-end machine-learning TN classification study that auto-discovers diffusivity disruptions in the cistern/REZ CNV and thalamocortico S1 pathway.
+Finally, in Study IV, we deploy end-to-end machine-learning TN classification to auto-discovers diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway.
 In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification.
-The application of this system to TN revealed specific diffusivity changes in deeper white matter structures, and pin-points the exact location of the diffusivity disruption at millimetre level.
+The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.

### 2018-04-16 contents/acknowledgements.tex
Updates from ShareLaTeX

--- 2017-11-24 thesis/contents/acknowledgements.tex
@@
 First I thank my parents.
-They have sacrificed their careers and endured much hardship by immigrating to Canada, to provide me with a better future.
+They have sacrificed their careers and endured much hardships by immigrating to Canada to provide me with a better future.
 All my achievements are only possible after I stand on their shoulders.
 
-I met my amazing wife Jidan Zhong in 2012, at Human Brain Mapping conference in Beijing.
+I met my amazing wife Jidan Zhong in 2012, at the Human Brain Mapping conference in Beijing.
 Little did I know that I had found my kindred spirit.
@@
 She took a chance on me when I had zero experience in neuroscience, and graciously supported me while I was having a hard time being accepted into the graduate program.
-I also thank her incredible patience and support for during my thesis writing.
+I also thank her incredible patience and support during my thesis writing.
 Dr.
 Hodaie is the most accomplished and hard-working person I have ever known.
-She is a world-class neurosurgeon, while at the same time balances her time between graduate mentoring, grant writing, international neurosurgical education, and at the same time multiple projects in scientific research.
+She is a world-class neurosurgeon, while at the same time balances her time between graduate mentoring, grant writing, international neurosurgical education, and multiple projects in scientific research.
 Every time I feel dejected from the workload, I always remind myself that Moji is doing more and that I have no excuses.
-It is truly my fortune to have her as my PI, and words cannot express how thankful I am.
+It is truly my fortune to have her as my supervisor, and words cannot express how thankful I am.
 
@@
 Karen Davis and Dr.
-David Mikulis.
-Whose extensible experience and insights were invaluable for guiding me through this process.
+David Mikulis, whose extensive experience and insights were invaluable for guiding me through this process.
 Dr.
@@
 
-I would also like to thank the past and present graduate and post-graduate members of the Hodaie Lab, whom I have shared an unforgettable period of my life: Danielle DeSouza, Allie Rosen, David Hayes, Brendan Behan, Matt Walker and Peter Hung.
+I would also like to thank the past and present graduate and post-graduate members of the Hodaie Lab, with whom I have shared an unforgettable period of my life: Danielle DeSouza, Allie Rosen, David Hayes, Brendan Behan, Matt Walker and Peter Hung.
 I also thank Jessica Quan, who introduced me to Dr.
-Hodaie and spark my journey in Neuroscience.
+Hodaie and sparked my journey in Neuroscience.
 
-I was fortunate and thankful to receive support from Multiple Sclerosis Society of Canada, James F. Crother's Family Fellowship in peripheral nerve damage, and Unilever/Lipton graduate fellowships.
+I was fortunate and thankful to receive support from the Multiple Sclerosis Society of Canada, James F.
+Crother's Family Fellowship in peripheral nerve damage, and Unilever/Lipton graduate fellowships.
 
-I thank all the research assistants over the years that helped me to perform the thankless job of download and organize patient data: Philip Yee, Udi Blankstein, Tessa Burkholder, Kevin Liang, and Erika Wharton-Shukster.
+I thank all the research assistants over the years that helped me to perform the thankless job of downloading and organizing data: Philip Yee, Udi Blankstein, Tessa Burkholder, Kevin Liang, and Erika Wharton-Shukster.
 

### 2018-04-16 contents/discussions.tex
Updates from ShareLaTeX

@@
 
-The contribution of this thesis are as follows:
+The contributions of this thesis are as follows:
 
@@
 
-In sum, this thesis provides novel evidence on the diffusivity tissue disruptions of the CN V in TN, and that there is evidence of bilateral CN V tissue changes in TN patients when compared to controls.
-The examination was also extended to the thalamocortical S1 pathway and have demonstrated localized white matter changes near the white-grey matter boundary.
-These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies as a whole.
-These results demonstrate that TN involves wide-spread changes at all levels of CNS white matter, and that the methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.
+In sum, this thesis provides novel evidence for diffusivity tissue disruptions of the CN V in TN, and bilateral CN V tissue abnormalities in TN patients when compared to controls.
+The study of the trigeminal pontine decussation (TPT) and thalamocortical S1 pathway demonstrated white matter diffusivity changes at all levels of the trigeminal sensory pathway.
+These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies.
+These results demonstrate that TN involves wide-spread changes at all levels of CNS white matter.
+The methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.

### 2018-04-16 contents/hypotheses.tex
Updates from ShareLaTeX

--- 2018-02-02 thesis/contents/hypotheses.tex
@@
 This necessitates detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls.
+
 With an increasing number of subjects, single subject analysis becomes cumbersome.
@@
 These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy.
+
 Lastly, these methodological advances will be applied to the large-scale study of TN versus healthy control.
-The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, to identify specific diffusivity pattern that differentiates between TN and non-TN.
+The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, in order to identify specific diffusivity pattern that differentiates between TN and non-TN.
 The roadmap and rationale for the specific studies are described below.
@@
 \section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
-MS-TN is thought to relate pathophysiologically to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons.
+MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons.
 However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals.
@@
 \subsection{Main Aim} 
-Establish the feasibility of multi-tensor diffusion tractography in delineating the cisternal, root entry zone and brainstem segments of CN V and how these changes correlate with pain in MS-TN.
+Establish the feasibility of multi-tensor diffusion tractography to delineating the cisternal, root entry zone and brainstem segments of CN V and how diffusivity changes in these sub-regions correlate with pain in MS-TN.
 
@@
     \item Use multi-tensor tractography to delineate CN V into the the pontine brainstem, in order to quantify tissue diffusivity measurements in targeted sub-regions.
-    \item Differentiate TN and MS-TN based on the obtained CN V diffusivity measurements.
+    \item Differentiate TN from MS-TN based on CN V diffusivity measurements.
 \end{itemize}
@@
 \section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
-The construction of an automated group tractography software framework depends on the establishment of the feasiblity of CN V brainstem delineation using multi-tensor tractography, and the optimal DWI-T1 co-registration strategy.
-The software aims to automate the tractography process on a large-scale, and also aims to streamline research workflow by providing region-of-interest, and tractography parameter management.
-It also provides the ability to report the tractography output at a group level both visually, as well as quantifying the output by an original scalar score called Normalized Overlap Score (NOS) score.
+The construction of an automated group tractography software framework depends on CN V brainstem delineation using multi-tensor tractography and the optimal DWI-T1 co-registration strategy.
+The aim is to create and validate a software framework that automates the tractography process on a large-scale, and also streamline research workflow by providing region-of-interest, and tractography parameter management.
+The software also provides the ability to report the tractography output at a group level both visually, as well as quantifying the consistency of group tractography output by developing a rating score called Normalized Overlap Score (NOS) score.
 
@@
 The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population.
-The tasks are performed  at a scale that would be prohibitive with manual methods.
+The validations are performed  at a scale that would be prohibitive with manual methods.
 The comparisons determine the best tractography methods to use for different types of neuroanatomy.
@@
 \begin{itemize}
-    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD in 42 subjects. 
-    \item Use the above techniques to study a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
-    \item Validate NOS scores by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
+    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD. 
+    \item Use the above techniques to delineate a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
+    \item Validate normalized overlapping scores (NOS) by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
 \end{itemize}
@@
 The grouped tracts should however show consistent regional anatomical overlap across individuals.
-    \item By merging the tracts of many individuals, the tractography delinations can complement each other, and this may improve the anatomical fidelity over individual cases.
+    \item By merging the tracts of many individuals, the tractography delineations can complement each other, and this may improve the anatomical fidelity over individual cases.
     \item Specific tractography algorithms may be preferentially suited for the visualization of specific areas in the brain.
@@
 The method used is now thoroughly tested, and the best-suited tractography methods determined from the previous study.
-The study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group.
+This study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group.
 The pathway is divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment.
-With the increased datapoints, more detailed along-the-tract analysis can be carried out on each segment.
-Following this, we use machine learning algorithms to auto-determine relevant segments that maximally differentiates TN from controls.
+Along-the-tract analysis are carried out by dividing each segment into 30 subdivisions.
+Following this, we use machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.
 
 \subsection{Main Aim}
-The aim of this project is the reconstruction of the trigeminal pathways from the cisternal segment to the level of S1 sensory cortex, and examination using Gaussian Process classification.
+The aim of this project is to reconstruct the trigeminal pathways from the cisternal segment to the level of S1 sensory cortex, and to examination these pathways using Gaussian Process classification to determine the subdivisions along each pathways that maximally differentiates TN from controls.
 
@@
     \item Delineate and visualize the three segments of the trigeminal sensory pathway: CN V, trigeminopontothalamic and  thalamocortical S1 segment. 
-    \item At each segment, we will look for microstructural features to distinguish the affected vs. unaffected sides.
+    \item At each segment, identify microstructural features to distinguish the affected vs. unaffected sides.
 \end{itemize}

### 2018-04-16 contents/paper1.tex
Updates from ShareLaTeX

@@
 Red arrows denote the starting point of tract generation for both SDT and MTT methods.
-Panel B shows that MTT tractography visualized CN V as it coursed through the brainstem towards the trigeminal main sensory nuclei (yellow arrows).}
+Panel B shows that MTT tractography visualized CN V as it coursed through the brainstem towards the trigeminal main sensory nuclei.}
 \centering
@@
 Panel B shows the TN tractography delineations, where there is no such FA decrease in the brainstem segment.
-Please refer to Figure 2 for explanations of color scales and figure annotations.}
+Please refer to Figure \protect\ref{fig:MSfigure2} for explanations of color scales and figure annotations.}
 \centering

### 2018-04-16 contents/reviews/dmri.tex
Updates from ShareLaTeX

@@
 
-Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain tissue in-vivo, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.
+Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibers \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.
 
@@
 
-The basis of dMRI is based on the observation that the location probability distribution of water molecules, which undergoes Brownian motion during diffusion, is a function of its suspension medium.
+DMRI is based on the observation that the location probability distribution of water molecules, which undergoes Brownian motion during diffusion, is a function of its suspension medium.
 The rate of diffusion can be formulated as \cite{Mukherjee2008}:
@@
 where $<r^2>$ is the mean square displacement of the molecule after time $t$ with diffusion constant $D$.
-In an unhindered medium, such as in liquid water, its diffusion probability is can be described as an spherical volume, in other words it undergoes \textit{isotropic} diffusion.
-In biological tissue such as the neuron axonal bundle, its diffusion probability is then constrained to be more likely in a direction along the bundle, rather than parallel to the cross-section, its diffusion state is then described as \textit{anisotropic} (\ref{fig:diffusion}).
+In an unhindered medium, such as in liquid water, the water molecule diffusion probability is can be described as an spherical volume, in other words it undergoes \textit{isotropic} diffusion.
+In biological tissue such as the neuron axonal bundle, water molecule diffusion probability is then constrained to be more likely in a direction along the bundle, rather than parallel to the cross-section, its diffusion state is then described as \textit{anisotropic} (\ref{fig:diffusion}).
 
@@
 
-The image contrast is determined by the amount molecular motion between the two gradients, where motion along the gradient axis results in a loss of image intensity. The contrast is defined as: 
+The image contrast is determined by the amount molecular motion between the two diffusion gradients, where motion along the gradient axis results in a loss of signal intensity.
+The contrast is defined as:
 \begin{equation}
@@
 $ADC_i$ being the apparent diffusion coefficient in the $i$ direction.
-$b$ is the diffusion weighted factor that is a function is the duration, strength and temporal spacing of the gradients:
+$b$ is the diffusion weighted factor that is a function of the duration, strength and temporal spacing of the gradients:
 \begin{equation}
@@
 Clinical DWI sequences typically use a diffusion time of 10-50 ms, or a travel of about 10$\mu$m.
-Typical clinical $b$ values range from 600 to 1500, to balance signal contrast with acquisition time.
+Typical clinical $b$ values range from 600 to 1500 $s/mm^2$, to balance signal contrast with acquisition time.
 
@@
 
-FA has been found to correlate with a wide spectrum of neuromorphological changes, including but not limited to multiple sclerosis \shortcite{Cercignani2001,Ciccarelli2003d,Miron2012}, vestibular schwannomas \shortcite{Chen2011b,Taoka2006,Wai2009}, tumors \shortcite{Mori2002,Schonberg2006,Nimsky2005}, schizophrenia \shortcite{Rametti2009,Fitzsimmons2009}, traumatic brain injury \shortcite{Bendlin2008,Kinnunen2011c,Hellyer2012}.
+Changes in white matter FA has been found to correlate with a wide spectrum of neuromorphological changes, including but not limited to multiple sclerosis \shortcite{Cercignani2001,Ciccarelli2003d,Miron2012}, vestibular schwannomas \shortcite{Chen2011b,Taoka2006,Wai2009}, tumors \shortcite{Mori2002,Schonberg2006,Nimsky2005}, schizophrenia \shortcite{Rametti2009,Fitzsimmons2009}, traumatic brain injury \shortcite{Bendlin2008,Kinnunen2011c,Hellyer2012}.
 RD is found to positively correlate with degree of demyelination in multiple sclerosis \shortcite{Song2005,Song2002,Klawiter2011,Janve2013}, and well as stages of white-matter development and aging \shortcite{Counsell2006,Sala2010}, including alzheimer's disease \shortcite{Acosta-Cabronero2012}.
@@
 
-Modifications to B-value needs to be considered for HARDI.
+Modifications to b-value needs to be considered for HARDI.
 The measured diffusion contrasts are dependent on the diffusion compartment variances.
-Therefore higher B-values will increase the measured signals variance contrasts, and allow better separation between crossing diffusion directions.
-Different HARDI algorithms have different requirements on B-values.
-For example Q-ball imaging needs B=3000 $s/mm^2$ \cite{Descoteaux2007a}, and multi-shell acquisitions will require stepped values of B values ranging from 500  $s/mm^2 $  to greater than 3500  $s/mm^2 $ \cite{Tournier2013}.
+Therefore higher b-values will increase the measured signals variance contrasts, and allow better separation between crossing diffusion directions.
+Different HARDI algorithms have different requirements on b-values.
+For example Q-ball imaging needs b=3000 $s/mm^2$ \cite{Descoteaux2007a}, and multi-shell acquisitions will require stepped values of b-values ranging from 500  $s/mm^2 $  to greater than 3500  $s/mm^2 $ \cite{Tournier2013}.
 
-The determination of gradient sampling pattern and B-value directly affect acquisition time and image quality.
-More complex gradient sampling and higher B-values can improve tractography quality.
+The determination of gradient sampling pattern and b-value directly affect acquisition time and image quality.
+More complex gradient sampling and higher b-values can improve tractography quality.
 However the acquisition time can become prohibitively long in a clinical setting.
@@
 
-Spherical decovolution is considered to be the action of a set of rotations on a function over a sphere.
-Given $F^l $ as a vector of $l$th order spherical harmonic decomposition of  $F(\theta,\phi)$, with vector length $(2l + 1)$, and $R^l$ as the $l$th order rotational harmonic decomposite of $R(\theta)$, with vector length $(2l+1)^2$.
+Spherical covolution is considered to be the action of a set of rotations on a function over a sphere.
+Given $F^l $ as a vector of $l$th order spherical harmonic decomposition of  $F(\theta,\phi)$, with vector length $(2l + 1)$, and $R^l$ as the $l$th order rotational harmonic decomposite of $R(\theta)$, with matrix size $(2l+1)^2$.
 The spherical harmonic representation of $S(\theta,\phi)$ is then:
@@
 \begin{equation}
-\|{Af - b}^2\|^2 + \lambda^2 \| l(f - f')\|^2
+\|{Af - b}\|^2 + \lambda^2 \| l(f - f')\|^2
 \end{equation}
@@
 \end{equation}
-Where $Q$ is the SH coefficient to DW intensity mapping, and $R$ is the spherical deconvolution. 
+Where $Q$ is the SH coefficient to diffusion weighted intensity mapping, and $R$ is the spherical deconvolution. 
 The regularization term is the constrained matrix $L$ multiplying the norm of $f$ relative to an initial estimate $f'$.
@@
 \subsection{Tractography}
-Tractography is defined as the construction, or propagation of streamline from an underlying field of orientation information, in order to visualize continuous structures within a volume.
+
+Tractography is a computer graphics technique used by neuroscientists to delineate neural pathways with DWI imaging.
+Applying pathway tracking to DWI imaging in the brain was first explored by Mori \cite{Mori2002b} and Basser\cite{Basser2000}.
+Tractography in essence is the construction, or propagation of streamline from an underlying field of orientation information, in order to visualize continuous structures within a volume.
 The orientation field is generally provided as a 3D matrix of directional information obtained from DTI, CSD, or other methods.
@@
 
-Applying pathway tracking to DWI imaging was first explored by Mori \cite{Mori2002b} and Basser\cite{Basser2000}.
 The family of methods that propagates by deterministic algorithms based on local orientation data is collectively called deterministic tractography.
@@
 The integrated tractography approaches are now standard practice for deterministic tractography \cite{Tournier2004,Pieper2004}.
-
 Deterministic tractography lends itself naturally to anatomical interpretation and further geometric analysis.
 It was traditionally paired with DTI, and so was thought to be limited in the case of crossing fibers.
+
 Other methods attempted to circumvent the limitation with probabilistic approaches that include white matter tissue priors in the fiber propagation model.
@@
 \subsubsection{Applications}
-Tractography permits the mapping of white matter structures with computer graphic and statistic methods.
+Tractography permits the mapping of white matter pathway structures with computer graphic and statistic methods.
 It permits the study of brain structural connectivity at individual level \cite{Boorman2007} and group level \cite{Meskaldji2013},  circuitry mapping \cite{Kirsch2015} and macro-structural insights between the brain hemispheres \cite{Kucyi2012a}.
-It is often paired with functional connectivity for studies in multiple sclerosis \cite{Zhong2016,Rocca2007a,Kucyi2015}, chronic pain \cite{Mansour2013,Seifert2011,Wiech2014}, and other areas of study such as empathy \cite{Bernhardt2014}, and aphasia \cite{Catani2013c}.
+It is often paired with functional connectivity for studies in multiple sclerosis \cite{Zhong2016,Rocca2007a}, chronic pain \cite{Mansour2013,Seifert2011,Wiech2014}, and other areas of study such as empathy \cite{Bernhardt2014}, and aphasia \cite{Catani2013c}.
 
@@
 \end{tabular}
-\caption{Example of cranial nerve reconstruction with deterministic tractography.} 
+\caption{Example of cranial nerve reconstruction with deterministic tractography.
+The image represents a saggital view of a vestibular schwannoma (green) displacing the trigeminal nerve (blue) and the facial/vestibular nerve complex (red) at the level of the brainstem.
+}
 \label{fig:tract1}

### 2018-04-16 contents/reviews/machine-learning.tex
Updates from ShareLaTeX

--- 2018-02-28 thesis/contents/reviews/machine-learning.tex
@@
 Strategies and algorithms for constructing various models overlap with the fields of artificial intelligence, computational statistics, pattern recognition, and neuroscience.
-Advanced in modern machine learning, especially in deep neural networks \cite{LeCun2015}, in recent years have led to wide spread adaptation of machine learning based online services in translation \cite{Bahdanau2014a}, image search \cite{simonyan2014very}, and speech recognition \cite{graves2013speech}.
-There is also surging interest in deploy machine learning systems in radiology and disease diagnosis in medicine\cite{Kamnitsas2016,Suk2014}.
+Advances in modern machine learning, especially in deep neural networks \cite{LeCun2015}, in recent years have led to wide-spread adoption of machine learning based online services in translation \cite{Bahdanau2014a}, image search \cite{simonyan2014very}, and speech recognition \cite{graves2013speech}.
+There is also surging interest in applying machine learning systems in medicine, particularly in radiology to assist in and disease diagnosis \cite{Kamnitsas2016,Suk2014}.
 
 Machine learning training strategies can be categorized into Supervised Learning, Unsupervised Learning, and Reinforcement Learning.
-Supervised learning involves training data that are pre-labelled.
-For example, the data supplied can be a list of patients' age and gender, and the label can be whether they or not they are in pain.
-Unsupervised learning involves data without any prior labels, and the algorithm attempts to discover patterns embedded within, and is often used in data mining.
+Supervised learning uses data that are pre-labelled according to some measure, in order to train a model.
+For example, the data supplied can be a list of patients' age and gender, and the label can be whether or not they are in pain.
+Unsupervised learning is a strategy that uses data without any prior labels, and the algorithm attempts to discover patterns embedded.
+Thus this approach is often used in data mining.
 Reinforcement learning involves continuously learning with a reward function.
-It is generally used to train agents that have a continuous stream of data and needs to interact with an external environment and is widely used in robotics.
+It is generally used to train autonomous agents using continuous streams of data, and is widely used in robotics.
 
@@
 Neuroimaging studies often attempt to correlate physiological and image-based measures with neuro-activities with specific experimental variables or conditions.
-Neuro-activities may be measured from EGG sensors, or as voxels in functional or structural MR images.
-Traditionally data from each measured feature component was assumed to be independent.
-These components maybe individual sensor readings, or voxel-based measures.
+Neuro-activities may be measured from EEG sensors, or as voxels in functional or structural MR images.
+Traditionally, data from each measured feature component was assumed to be independent.
+These components may be individual sensor readings or voxel-based measures.
 Correlative analysis were performed using general linear models, with post-hoc false-positive corrections.
-This type of univariate analysis do not have the ability to discover key relationships between feature components.
+This type of univariate analysis does not have the ability to discover key relationships between feature components.
 Therefore univariate models cannot take into account spatial or temporally correlative dynamics.
 
-Increasing complexity in neuroimaging data has lead to widespread applications of multi-variable machine learning methods in neuroimaging.
-Application of ML in pain neuroimaging became controversial when Tor Wager \cite{Wager2013} demonstrated a Lasso-PCA model that classifies nociceptive pain with sensitivity and specificity of 90\% solely from fMRI imaging data.
+Increasing complexity in neuroimaging data has led to widespread applications of multi-variable machine learning methods in neuroimaging.
+Application of ML in pain neuroimaging became controversial when Tor Wager \cite{Wager2013} demonstrated a Lasso-PCA model that classified acute experimental pain with sensitivity and specificity of up to 90\% solely from fMRI imaging data.
 Additional studies showed consistently good performance of ML-based nociceptive pain discrimination.
-Discrimination of chronic pain from nociceptive pain proved to be more challenging, due to the high degree of similarity in pain-matrix activation patterns.
-The improvement in chronic pain detection lead to debates of the validity of presenting such methods as evidences in courtroom, where Davis et al argued against such actions \cite{Davis2012a}.
-While it is commonly accepted that pain activity in the brain involves multiple regions of the pain-matrix, evidences in some studies suggested that the posterior-insula is heavily involved in chronic pain.
-Segerdale et al claimed that the posterior insula is a primary pain region \cite{Segerdahl2015a}, thereby ignited renewed debates surrounding regional specific brain centers for chronic pain \cite{Davis2015}.
+Discrimination of chronic pain from nociceptive pain proved to be more challenging, due to the high degree of similarity in activation patterns.
+The development of pain detection through brain imaging led to debates of the validity of presenting such methods as evidences in courtroom, where Davis et al argued against such actions \cite{Davis2012a}.
+While it is commonly accepted that pain evokes activity in the brain in multiple regions of the dynamic pain connectome, some studies suggest that the dorsal posterior-insula is heavily involved in chronic pain.
+Segerdale et al claimed that the dorsal posterior insula is a primary pain region \cite{Segerdahl2015a}, and this proposition ignited renewed debates surrounding regional specific brain centers for chronic pain \cite{Davis2015}.
 
 Despite the controversies, interest in machine learning in medical imaging prognosis have sparked both public and private efforts across the globe.
-Classification studies with using multi-modal data with features in functional connectivity and structural connectivity in pain syndromes such as trigeminal neuralgia continue to improve.
+Classification studies using multi-modal data with features involving functional connectivity and structural connectivity in pain syndromes such as TN continue to improve.
 It is clear that machine learning is here to stay in pain neuroimaging research.

### 2018-04-16 contents/reviews/mri.tex
Updates from ShareLaTeX

@@
 
-Protons aligned in the parallel alignment are more preferred as it has a lower energy state than in anti-parallel orientation.
+Protons aligned in the parallel alignment are more preferred as they have a lower energy state than protons in anti-parallel orientation.
 The energy E of a magnetic moment $\vec{\mu}$ when in a magnetic field $B_0$ is given by: $E=-\vec{\mu}\cdot \textbf{B}_0=-\mu_z B_0=-\gamma m \hbar B_0$.
@@
 Immediate after excitation, the phase precessions are synchronous, and therefore are in \textit{phase coherence}.  
-However due to various reasons, the various spins begin to lose coherence, and their total sum magnetization $M_xy$ diminishes.
+However due to the following reasons, the various spins begin to lose coherence, and their total sum magnetization $M_xy$ diminishes.
 
@@
 
-\item There are intrinsic inhomogeneities by the magnetic field generator, and the scanning subject.
+\item There are intrinsic inhomogeneities by the main static magnetic field generator, and the scanning subject.
 These intrinsic properties contribute to the dephasing, and has a shorter time decay than T2.
@@
 \item Proton density: the number of excitable spins per unit volume, which determines the maximum amount of signal that can be measured from the tissue.
-It can be measured by minimizing T1 and T2 times.
+It can be measured by minimizing T1 and maximizing T2 times.
 \item T1 time: the amount time for $M_xy$ to recover to $M_z$, to be ready for re-excitation.
@@
 \end{itemize}
-The contrast in brain tissues of the resulting MR image can be tuned by adjusting these three parameters in pulse sequence design.
+The contrast in brain tissues of the resulting MR image can be tuned by adjusting T1  and T2 parameters in pulse sequence design.
 
@@
 \paragraph{Echo Time}
-MR sequences have a gradient field which is used to control field inhomogeneities and to encode spatial origin.
+MR sequences have a gradient field which is used to encode spatial origin.
 The gradients also contribute to spin dephasing.
-Therefore a refocusing pulse is necessary to measure MR signals adequately.
-The time delay between refocusing and the measured signal under phase coherence (spin echo) is known as echo time (TE).
-When TE is short ($<$ 30 ms), there are little signal differences between tissue types; when TE is long ( $\geq$ 60 ms ), the intrinsic tissue differences will allow good contrast in the resulting image.
-Therefore long TE is said to be T2 weighted.
+In spin echo imaging, a 90 deg pulse and a 180 deg refocusing pulse are needed to measure MR signals adequately.
+The time delay between the first pulse and the measured signal under phase coherence (spin echo) is known as echo time (TE).
+When TE is short ($<$ 30 ms), and TR is short, the image is said to be T1 weighted; when TE is long ( $\geq$ 60 ms ), and TR is long, the image is said to be T2-weighted.
 
@@
 \caption{Illustration of the process of spin-echo.
-Initially, the spin precesses about the B0 field under steady-state.
-When a 90deg RF pulse is applied, the spins orients towards the xy-plane.
-When the RF pulse is switched off, the spins now relaxes towards the z-axis, and the spin undergoes phase decoherence.
-A 180 degrees flip angle pulse is applied to prevent spin saturation, and then a refocusing pulse is applied to bring spins back into coherence.
+Initially, the spins orients to wards B0 field under steady-state.
+When a 90deg RF pulse is applied, the spins orients to the xy-plane.
+When the RF pulse is switched off, the spins undergoe phase decoherence.
+A 180 degrees refocusing pulse is applied to bring spins back into coherence, at which point in time the peak of an "echo" occurs.
 } \label{fig:spin-echo}
@@
 
-Typically, T1-weighted spin echo (SE)  sequence is acquired with TR/TE at 340/13 ms; T2-weighted fast spin echo (FSE) at TR/TE of 3500/120 ms. 
+Typically, T1-weighted spin echo (SE)  sequence is acquired with TR \textless 800 ms, TE \textless 20 ms; T2-weighted fast spin echo (FSE) at TR \textgreater 2000 ms, TE \textgreater 60 ms. 
 In the case of T1-weighted MRI, images are created with short TE and short TR times.
@@
 
-The T2-weighted sequence uses a long TR and long TE. The T2-weighted sequence can be employed as a dual echo sequence. Like the T1-weighted scan, fat is
-differentiated from water, but with fat dark and water bright (Figure \ref{fig:T1T2PD}, Middle).
+The T2-weighted sequence uses a long TR and long TE. The T2-weighted sequence can be employed as a multi-echo sequence. Like the T1-weighted scan, fat is
+differentiated from water in single echo sequences, but with fat dark and water bright (Figure \ref{fig:T1T2PD}, Middle).
 
-The first or shorter echo ($T_E<30msec$) is proton density (PD) weighted or a mixture of T1 and T2.
-This image can provide good contrast of tissue and CSF, thus very helpful for evaluating periventricular pathology (Figure \ref{fig:T1T2PD}, Right).
-$T_2^*$-weighted scans use a gradient echo (GRE) sequence, with long TE and long TR.
+Another type of image with shorter echo ($T_E<30msec$) is proton density (PD) (Figure \ref{fig:T1T2PD}, right) weighted.
+This image provides poor contrast of tissue and CSF, but good bright contrast against lesions, thus very helpful for evaluating periventricular pathology .
+$T_2^*$-weighted scans use a gradient echo (GRE) sequence, which uses only one RF pulse and gradient reversal.
 It is prone to susceptibility losses at air/tissue boundaries, but can increase contrast for certain types of tissue, such as venous blood.
@@
 \item Phase encoding: The phase encoding gradient is applied in the y-direction.
-It alters the Larmor frequency of the spins based the location of the y-axis, and results in phase shifts of the spins relative to each other.
+It temporarily alters the Larmor frequency of the spins based their location of the y-axis, and results in phase shifts of the spins relative to each other.
 The phase shift is determined by the duration and amplitude of the gradient, when the gradient is switched off, all spins return to their initial rate of precession, but maintains the relative phase shift.
@@
 Each step in $k_y$ represents a single phase encoding step.
-The x-axis origin $k_x=0$ is the gradient iso-center, and the frequency distribution is symmetric about the x-origin.
+The x-axis origin $k_x=0$ is the gradient iso-center, where the gradient magnetic field equals to zero, and the frequency distribution is symmetric about the x-origin.
 The 2D-FT operation is applied to the k-space to transform the frequency domain information into an image.

### 2018-04-16 contents/reviews/multiplesclerosis.tex
Updates from ShareLaTeX

--- 2018-02-28 thesis/contents/reviews/multiplesclerosis.tex
@@
 \section{Multiple Sclerosis}
-This chapter will review multiple sclerosis (MS) as a disorder, given the importance of MS-related TN as an independent TN subtype.
+This chapter will review multiple sclerosis (MS) as a disorder, given the importance of MS-related TN is an independent TN subtype.
 MS is a chronic autoimmune demyelination disease of the central nervous system.
-It is a progressive disease characterized by wide-spread lesion, or plaques, within the brain, spinal cord, and optic nerve.
+It is a progressive disease characterized by wide-spread lesions, or plaques, within the brain, spinal cord, and optic nerve.
 Its pathogenesis is related to the focal inflammation and damage of the central myelin, where the disruptions of the myelin sheath negatively affect signal transduction of the central white-matter, and thereby result in severe disability.
@@
 Currently there is no cure for MS.
-An estimated 2,300,000 people world wide are affected by MS, and 50\% of the patient will require motor assistance 15 years after disease onset \cite{Goldenberg2012}.
+An estimated 2,300,000 people world wide are affected by MS, and 50\% of the patients will require motor assistance 15 years after disease onset \cite{Goldenberg2012}.
 The course of MS can be highly variable due to the distinct plaque distribution pattern in each individual, and therefor MS is difficult to diagnose and treat.
@@
 
-Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different episodes must occur in the disease course (time dissemination); and C) Chronic inflammation of the CNS.
+Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different MS exacerbation episodes must occur in the disease course (time dissemination); and C) Chronic inflammation of the CNS.
 One or more of these criteria result in a general diagnosis of MS.
@@
 
-MS is defined by the existence of focal demyelination in the CNS that manifests as lesions or plaques.
+MS is identified by the existence of focal demyelination in the CNS that manifests as lesions or plaques.
 Ultrastructural lesion analysis shows demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells.
@@
 
-The MS lesions or plaques are characterized as focal hyperintensities in T2-weighted or proton-density images.
+The MS lesions or plaques are characterized as focal hyperintensities (bright spots) in T2-weighted or proton-density images.
 Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium.
@@
 They are also found densely dispersed in paraventricular white matter.
-The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered for the diagnosis.
+The presence of MR hyperintensities is not unique to MS, and therefore the location, shape and time dissemination must be considered for the diagnosis.
 The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}.
@@
 
-Diffusion MRI (dMRI) is found to be sensitive to tissue changes such as edema, axonal injury and demyelination, and is becoming an important MS tissue quantification strategy.
-A number of diffusion metrics (Mean diffusivity (MD), radial diffusivity (RD), and axial diffusivity (AD)) have become important in MS tissue quantification and analysis.
+Diffusion MRI (dMRI) has been found to be sensitive to tissue changes such as edema, axonal injury and demyelination, and is becoming an important MS tissue quantification strategy.
+A number of diffusion metrics (mean diffusivity (MD), radial diffusivity (RD), and axial diffusivity (AD)) have become important in MS tissue quantification and analysis.
 They are particularly useful when study normal-appearing white matter.
-Normal-appearing white matter (NAWM) is microscopically abnormal white matter tissue that is visually indistinguishable from normal white matter in T2 MRI images.
+Normal-appearing white matter (NAWM) is white matter tissue in close proximity to MS lesions, and is visually indistinguishable from normal white matter on all conventional MRI sequences.
 They are generally defined to be at least 1 cm away from a lesions.
@@
 Only about 28 \% of NAWM is found to be normal, and there is a 12-42 \% reduction in axonal density in these tissues.
-DMRI study of NAWM have found decreased fractional anisotropy and increased mean diffusivity \cite{Ciccarelli2003d,DeGroot2013}.
+DMRI study of NAWM have found decreased FA and increased MD \cite{Ciccarelli2003d,DeGroot2013}.
 
@@
 Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}.
-Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as trigeminal neuralgia.
+Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as TN.
 91\% of the patients report pain at the time of evaluation, and most experience daily pain.

### 2018-04-16 contents/reviews/pain.tex
Updates from ShareLaTeX

--- 2018-02-02 thesis/contents/reviews/pain.tex
@@
 \begin{figure}[ht]
-\includegraphics[width=0.3\textwidth]{descartes-reflex.JPG}
+\includegraphics[width=0.5\textwidth]{descartes-reflex.JPG}
 \centering
-\caption{Illustration of Descartes' pain pathway from the Treatise of Man}
+\caption{Illustration of Descartes' pain pathway from the Treatise of Man \protect\cite{descartes-wikipedia}}
 \label{fig:descarte}
@@
 
-The gate-control theory \cite{Melzack1965a} (\ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations.
+The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations.
 The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission.
@@
  \begin{figure}[ht]
- \includegraphics[width=0.8\textwidth]{gate-control.jpg}
+ \includegraphics[width=\textwidth]{gate-control-diagram.pdf}
  \centering
@@
 
-The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory.
+The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory.
 The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes the necessary information for pain.
@@
 Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn (laminae V), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex.
-A central concept, which was also introduced by Melzacks, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}.
+A central concept, which was also introduced by Melzack, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}.
 (See section \ref{section:neuromatrix})
@@
 \begin{figure}[ht]
-\includegraphics[width=0.8\textwidth]{pain-drg.jpg}
+\includegraphics[width=0.8\textwidth]{dorsal-lamina.pdf}
 \centering
@@
 The interneurons in the dorsal horn plays an important part in somatosensory integration as the sensory afferent pass from peripheral into the central nervous system.
-For example, the thermal grill illusion, discovered by Thunberg in 1896, involved the paradoxical feeling of heat or burning pain when adjacent skin regions are exposed to innocuous warm or cold stimuli \cite{Craig1994}.
-The mechanism is through the antagonistic interactions between innocuous cold A-fibers, cold heat pinch (CHP) C-fibers and warm C-fibers.
-Whereby the warm stimulus activates C-fibers that inhibit adjacent A-fiber activation, thereby unmasks the CHP C-fiber activations into the central nervous system.
+For example, the wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI.
+These neurons respond to both innocuous and noxious stimuli, and are decidedly non-specific.
+They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing evidence that WDR neurons are responsive to prolonged nociceptive pain.
 
-Wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI.
-These neurons respond to both innocuous and painful stimuli, and are decidedly non-specific.
-They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing conclusive evidence that WDR neurons are sufficient for the percept of prolonged nociceptive pain.
-
-In each spinal afferent level, the dorsal horn neurons project noceciptive signals to the brain via five distinct paths (\ref{fig:pain-pathways}).
+In each spinal afferent level, the dorsal horn neurons project noceciptive signals to the brain via five distinct paths (Figure \ref{fig:pain-pathways}).
 Lamina I, V-VII axons crosses the mid-line to form the \textit{spinothalamic} tract that ascend in the contralateral anterolateral column of the spinal cord, and project to thalamic nuclei.
 Laminae VII and VIII neurons from the \textit{spinorecticular} tract ascend via the anterolateral quadrant of the spinal cord, without crossing the midline, to reach the reticular formation and thalamus.
-The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, periaqueductual gray.
+The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, including the periaqueductual gray.
 The \textit{cervicothalamic} tract include only the white matters of the upper two cervical segments.
@@
 Before reaching the neocortex, several important second order pain afferents arrive at the thalamus.
-The spinothalamic tracts subserving face and body sensory information terminate in the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for face and body respectively.
+The spinothalamic tracts subserving face and body sensory information terminate in several regions of the thalamus.
+A major termination site is the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for the face and body respectively.
 The VP complex sends third-order afferents to the primary sensory cortex (S1).
@@
 
-Another termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse.
+Another spinothalamic termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse.
 The MD nucleus projects third order afferents to the anterior cingulate cortex (ACC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.
 
-An important nucleus has thought to be unique in humans is the ventromedial-posterior nucleus (VMpo)\cite{Willis2002,Craig2014}.
-It receives afferents primarily from dorsal horn lamina I, and can be considered to be contiguous with the basal ventral medial nucleus (VMb), that forms a relay for homeostatic afferents \cite{Craig2003}.
+An important nucleus thought to be unique in humans is the ventromedial-posterior nucleus (VMpo)\cite{Willis2002,Craig2014}.
+It receives inputs primarily from dorsal horn lamina I, and can be considered to be contiguous with the basal ventral medial nucleus (VMb), that forms a relay for homeostatic afferents \cite{Craig2003}.
 The VMPo projects to the dorsal posterior insular cortex, which is thought to involve body interoception of pain.
@@
  Important cortical regions involved in pain include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1).
-The role of these cortical regions in pain were discovered through neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during painful stimulation.
-Regions such as ACC, PFC and insula are also shown to be active in social pain during social exclusion and emotional stress associated with PTSD revealed changes in ACC and insula grey matter (Section \ref{fig:pain-matrix}).
+The role of these cortical regions in pain were studied in neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during noxious stimulation.
+(Figure \ref{fig:pain-matrix}).
 
@@
  \centering
- \caption{Cortical and sub-cortical regions of the pain matrix.} 
+ \caption{ Cortical and sub-cortical regions of the pain matrix. \protect\cite{Tracey2007a}} 
  \label{fig:pain-matrix}
@@
 Thus, the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}.
-The dynamic rhythms of activities that attune to and away from saliency, form a complex pain connectome \cite{Kucyi2015}.
-Conditions such as chronic pain therefore show a differentiated tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable nociceptive stimuli \cite{Baliki2008,Legrain2009}.
+The dynamic rhythms of activities that attune to and away from saliency, form a dynamic pain connectome \cite{Kucyi2015}.
+Chronic pain therefore demonstrated differential tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable noxious stimuli \cite{Baliki2008,Legrain2009}.
 
 \paragraph{Sensory-Discrimination}
-Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the primary sensory (S1) and secondary sensory (S2) cortices.
+Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the S1 ana S2 cortices.
 The S1 cortex is responsible for the discrimination of fine-touch sensations.
@@
 S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies \cite{Wager2013,Brodersen2012}.
-The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}.
-Therefore it is believe to plays an important role in pain sensory-discrimination.
+The posterior insula also receives direct nociceptive inputs from the VMpo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}.
+Therefore the insula, as well as S1, are believed to plays an important role in sensory-discrimination aspect of pain.
 
@@
 The cortical regions that are involved in affective-motivational pain processing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions.
-The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus.
+The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the anterior insula receives input from the VMpo thalamus.
 The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain.
@@
 Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is a strong supportive finding of a homologue region in humans.
-Critics of this finding raised issues with the low number of subjects, long sampling duration and the incompatibility of this interpretation with long standing bodies of neuroimaging literature \cite{Davis2015}.
+Critics of this finding raised issues with the low number of subjects, long sampling duration and the incompatibility of this interpretation with long standing bodies of neuroimaging and neurophysiological literature \cite{Davis2015}.
 
@@
 
-Chronic pain is pain that persists or recurrent past the healing phase following an injury \cite{Apkarian2009,Merskey2012}.
-It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months, other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}.
+Chronic pain is pain that persists past the healing phase following an injury \cite{Apkarian2009,Merskey2012}.
+It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months.
+Other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}.
 The new ICD-11 revision of chronic pain classification include a new category for chronic primary pain.
@@
 
-The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission and processing circuitry. 
+The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission, processing, and modulatory circuitry. 
 Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex.

### 2018-04-16 contents/reviews/trigeminal.tex
Updates from ShareLaTeX

--- 2018-02-02 thesis/contents/reviews/trigeminal.tex
@@
 
-The sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messner's, Pacinian and Ruffini corpuscles.
+Trigeminal sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messiner's, Pacinian and Ruffini corpuscles.
 
@@
 V3 contains sensory and motor branches.
-It innervates skins of the jaw, areas above the ear, lips, gum, and teeth.
+It innervates skin of the jaw, areas above the ear, lips, gum, and teeth.
 The V3 division carries the mastication efferent to the four muscles recruited for mastications: the masseter, temporal, medial and lateral pterygoids.
@@
 The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons.
-The CNS myelin can be several millimetres beyond the surface of the pons, before it gives away to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}.
-Study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm).
+The CNS myelin can be several millimetres beyond the surface of the pons, before giving away to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}.
+A study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm).
 The transition zone can be up to 25\% of the nerve length \cite{Peker2006}.
@@
  \begin{figure}[ht]
- \includegraphics[width=100mm]{trig-pathway.png}
+ \includegraphics[width=\textwidth]{trig-pathway-wilson.jpg}
  \centering
- \caption{Overview of the trigeminal afferent pathways}
+ \caption{Overview of the trigeminal afferent pathways \protect\cite{wilson2010cranial}}
  \label{fig:trig-pathway}
@@
 CN V passes into the pons (\ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem.
-It then curve towards the medial line before synapses onto the trigeminal nucleus in the basis of the pontine brainstem.
-The trigeminal nucleus is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei.
+It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem.
+The trigeminal brainstem sensory nucleus complex is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei.
 The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.
@@
 
-The main sensory nucleus receives discriminative somatosensation from the oralfacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract.
+The main sensory nucleus receives discriminative input from the orofacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract.
 A separate branch responsible for oral cavity, however, projects to the ipsilateral VC via the dorsal trigeminothalamic tract.
@@
 Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms \cite{Haller2016,Love2001}.
-However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the clinical expression of pain \cite{Desouza2013,Hodaie2013}.
+However NVC cannot be assumed to be directly causal of TN, since presence of vascular compression does not necessarily result in pain \cite{Desouza2013,Hodaie2013}.
 In approximately 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.
@@
 This permitted high fidelity visualization and targeted characterization of CN V diffusivity in TN, demonstrating significant reduction in fractional anisotropy (FA) post-surgery at the site of the GK target \cite{Hodaie2009a}.
-Study of CN V diffusivity in conjunction with deep white matter using TBSS, again revealed reduced FA on the affected CN V, with also bilateral increased in radial, axial, and mean diffusivity.
+Study of CN V diffusivity in conjunction with deep white matter again revealed reduced FA on the affected CN V, with also bilateral increased in radial, axial, and mean diffusivity.
 Disruptions were also found in the cognitive-affective, attention, and motor regions of the cortex, consistent with cortical patterns in chronic pain \cite{Desouza2013}.
 Lowered FA was reported in multiple TN studies \cite{Herweh2007, Lutz2011}.
-Study in temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}, thereby demonstrating that CN V FA is related to nociception.
-The TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.
+A Study in patients with temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}.
+This TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.

### 2018-04-23 contents/abstract.tex
Updates from ShareLaTeX

@@
 \chapter{Abstract}
-
+% \begin{abstract}
+\doublespacing
 Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches.
@@
 The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.
+% \end{abstract}

### 2018-04-23 contents/hypotheses.tex
Updates from ShareLaTeX

@@
 \begin{itemize}
-    \item TN is known to relate to vascular compression of the cistern REZ, therefore we expect its diffusivity signature to occur in the cisternal region.
-    \item MS-TN should be related to brainstem MS lesions.
-Therefore CN V segments close to the lesions are expected to show more significant changes in tissue diffusivity, whereas segments far away from the lesions will not.
-    \item As a demyelinating disease, we expect to find a pattern of change in diffusivity signatures that reflects demyelination, including increase in RD, and decrease in FA. 
+    \item TN is known to relate to vascular compression of the cistern REZ, therefore its diffusivity signature will occur in the cisternal region.
+    \item MS-TN is related to brainstem MS lesions.
+Therefore CN V segments close to the lesions will show more significant changes in tissue diffusivity, whereas segments far away from the lesions will not.
+    \item As a demyelinating disease, the pattern of change for MS-TN diffusivity signatures will reflects demyelination, including increase in RD, and decrease in FA. 
 \end{itemize}
@@
     \item Symmetric-diffeomorphic registration will result in the best accuracy, but with limited asymptotic improvement over affine registration alone.
-    \item Averaged DWI should be the best scalar image, as it provides a balanced contrast between white and grey matter.
+    \item Averaged DWI will provide good registration quality, as it provides a balanced contrast between white and grey matter.
     \item Anisotropic-power (AP) image will show more substantial improvement over the existing scalar images.
@@
     \item DTI is known to not able to delineate cross-fibers regions.
-If the group tractography works consistently, then its output should reflect such known algorithmic limitations, and DTI group tracts should not be able to perform well on more complex anatomies.
-The grouped tracts should however show consistent regional anatomical overlap across individuals.
-    \item By merging the tracts of many individuals, the tractography delineations can complement each other, and this may improve the anatomical fidelity over individual cases.
-    \item Specific tractography algorithms may be preferentially suited for the visualization of specific areas in the brain.
-For examples, deterministic tractography should preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
-    \item NOS score should correlate with expert ratings, and show improved rating consistency over human rater variability.
+If the group tractography works consistently, then its output will reflect such known algorithmic limitations, and DTI group tracts will not be able to perform well on more complex anatomies.
+The grouped tracts will however show consistent regional anatomical overlap across individuals.
+    \item Specific tractography algorithms will be preferentially suited for the visualization of specific areas in the brain.
+For examples, deterministic tractography will preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
+    \item NOS score will correlate with expert ratings, and show improved rating consistency over human rater variability.
 \end{itemize}
@@
 \begin{itemize}
-    \item To validate the study, CN V diffusivity signature should match findings from Study I.
-    \item At the level of CN V, the affected cistern segment should show differentiation from the unaffected side.
-    \item The affected S1 segment should differ from the unaffected side.
-    \item The affected TPT segment might show more differentiation near the end points, due to possible differences in trigeminal nucleus and thalamus diffusivity.
-However, its mid-segment diffusivity might be inconclusive, as the diffusivity metrics such as FA and RD might not be adequate in describing the complex crossings at the pontine decussation.
+    \item To validate the study, CN V diffusivity signature will match findings from Study I.
+    \item At the level of CN V, the affected cistern segment will show differentiation from the unaffected side.
+    \item The affected S1 segment will differ from the unaffected side.
+    \item The affected TPT segment will show more differentiation near the end points, due to possible differences in trigeminal nucleus and thalamus diffusivity.
+However, its mid-segment diffusivity willht be inconclusive, as the diffusivity metrics such as FA and RD might not be adequate in describing the complex crossings at the pontine decussation.
 \end{itemize}

### 2018-04-23 contents/reviews/pain.tex

@@
 148]{BookCottingham1991v3}.
-In his \textit{Treatise of Man}, published in 1664, he stated: "Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch.
+In his \textit{Treatise of Man}, published in 1664, he stated:
+\begin{quote}"Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch.
 In this way they pull the tiny fibre which you see attach to it, and simultaneously open the entrance to the pore, located opposite the point where this fibre terminates -- just as when you pull one of a string, you cause a bell hanging at the other end to ring at the same time.
 When the entrance to the pore or small tube is opened in this way, the animal spirits from cavity enter and are carried through it...the fibres cause a movement in the brain which gives occasion for the soul...to have the sensation of pain" \cite[p.101-103]{BookCottingham1991v1}.
+\end{quote}
 
@@
 
-The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations.
+The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that relays touch or pain sensations.
 The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission.
@@
  \centering
- \caption{Melzack's gate control theory diagram.}
+ \caption{Melzack's gate control theory diagram. Modified from \protect\cite{Melzack1965a}}
  \label{fig:gate-theory} 
@@
 
-The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory.
-The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes the necessary information for pain.
-That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities.
-This entire pattern of neural activities constitute the sensory component of the pain experience.
-Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn (laminae V), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex.
-A central concept, which was also introduced by Melzack, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}.
+An important contribution of the gate-control theory was to emphasize central nervous system mechanisms \cite{Melzack1999}, and the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory.
+Contemporary views consider pain to be complex and multi-dimensional, where the processing of ascending nociceptive signals, in conjunction with descending modulatory pathways, shape the pain experience.
+Melzack and Casey further proposed that pain consists of sensory-discriminative, affective-motivational, and cognitive-evaluative components \cite{Melzack1968}.
+Specifically, the sensory-discriminative component consists of peripheral nociceptive afferents transmit to the spinal dorsal horn (laminae V), whereby nociceptive specific (NS) and wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex.
+A central concept, which was also introduced by Melzack, is the  neuromatrix theory of pain, which dictates that the pain experience is produced by a "neurosignature" pattern, that is generated from widely distributed neural network, which Melzack refers as the "body-self neuromatrix".
+\cite{Melzack1999}.
+The concept is now more commonly referred to as the "pain matrix" \cite{tracey2007a}.
 (See section \ref{section:neuromatrix})
@@
 
-\textit{Nociceptors}, the receptors in the peripheral nervous system that detect tissue damage can respond to a range of electrical or chemical signals released by noxious stimuli on the tissue, such as physical trauma or heat/cold temperatures exceeding the pain threshold.
-These receptors broadly separate into two categories, differentiated by the type of axonal innervation they receive.
-Acute pain receptors are innervated by $A\delta$ fibres, which permits fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma.
-Dull pain is triggered by receptors that are innervated by slow conducting C fibres ($<1$ m/s).
+\textit{Nociceptors}, the peripheral nociceptors respond to noxious stimuli.
+These receptors have free nerve endings, and can respond to thermal, mechanical, and/or chemical stimuli, such as physical trauma or heat/cold temperatures.
+These receptors broadly separate into two categories, differentiated by the fibre type.
+Acute nociceptors are associated with $A\delta$ fibres, which permit fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma.
+Dull pain is triggered by receptors that are associated with slow conducting C fibres ($<1$ m/s).
 $A\delta$ and C fibres can be co-activated in the presence of noxious stimuli.
+C-fibre nociceptors can be further differentiated into mechano-heat, silent, and polymodal subtypes.
+$A\delta$-fibre nociceptors can be divided into high-threshold mechano-heat, low-threshold mechano-heat, and noxious cold subtypes.
 
@@
 
-The peripheral nociception signals are conducted from the axonal endings of the nociciption neurons, whose cell bodies are located in the spinal cord dorsal root gangions, to the dorsal horn of the spinal cord (\ref{fig:drg}).
+The peripheral nociception signals are conducted from the axonal endings of the nociception neurons, whose cell bodies are located in the spinal cord dorsal root gangions, to the dorsal horn of the spinal cord (\ref{fig:drg}).
 The $A\delta$ and C afferents project directly to lamina I of the dorsal horn.
@@
 Another spinothalamic termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse.
-The MD nucleus projects third order afferents to the anterior cingulate cortex (ACC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.
+The MD nucleus projects third order afferents to the anterior/middle cingulate cortex (ACC/MCC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.
 
@@
 
-Pain percept, defined as the totality of the experience of pain, is thought to involve distributed regions of the brain, coined as the “pain neuromatrix”\cite{Melzack1999a}.
+Pain percept, defined as the subjective experience of pain, is thought to involve distributed regions of the brain, coined as the pain “neuromatrix”\cite{Melzack1999a}, or more commonly the "pain matrix".
 The regions are not functionally independent, but rather thought to be dynamically inter-related.
 
- Important cortical regions involved in pain include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1).
+ Important cortical regions involved in pain include anterior and middle cingulate cortices (ACC/MCC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1).
 The role of these cortical regions in pain were studied in neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during noxious stimulation.
@@
 The unpleasantness of pain has been thought to be a motivational force \cite{Auvray2010}, that biases actions towards avoidance.
-There is evidence that the motor cortex, which is involved in action planning, and the ACC are activated during pain \cite{Erpelding2013}.
+There is evidence that the motor cortex, which is involved in action planning, and the ACC/MCC are activated during pain \cite{Parker2016,Burns2016,Erpelding2013}.
 The cortical excitability of the motor cortex is also shown to be altered in chronic pain \cite{Parker2016}.
@@
 The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation.
-Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}.
-DMN activation has been correlated with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.
+Regions such as vmPFC, dMPFC and the anterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}.
+Pain related DMN deactivation have been found to be attenuated  in mind wandering away from pain, and may reflect antinociceptive processes \cite{Kucyi2013}, and that the DMN may undergo functional reorganization under chronic pain \cite{Baliki2014}.
 
@@
 
-The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission, processing, and modulatory circuitry. 
+The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission, processing, and descending modulatory circuitry. 
 Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex.
@@
 Trauma and other disruptions to the spinal pathways can affect noxious signal transmission.
-White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis, such that the synchronicity of action potential conduction between cortical and subcortical structures are disrupted, and thereby give rise to neuropathic pain.
+White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis.
 It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures.

### 2018-04-23 contents/reviews/trigeminal.tex

@@
 
-The afferent branches converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion), which then forms the trunk or cisternal component of CN V.
+The afferent branches converge and form the trigeminal root ganglion (TRG) (also known as Gasserian, or semilunar ganglion), which then leads to the trunk or cisternal component of CN V.
 The TRG contains all the trigeminal sensory cell bodies.
@@
 The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons.
-The CNS myelin can be several millimetres beyond the surface of the pons, before giving away to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}.
+The CNS myelin can be several millimetres beyond the surface of the pons, before leading to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}.
 A study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm).
@@
 
-CN V passes into the pons (\ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem.
+CN V passes into the pons (Figure \ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem.
 It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem.

### 2018-04-24 contents/reviews/pain.tex
Updates from ShareLaTeX

@@
 Another spinothalamic termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse.
-The MD nucleus projects third order afferents to the anterior/middle cingulate cortex (ACC/MCC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.
+The MD nucleus projects third order afferents to the anterior/middle cingulate cortex (ACC/MCC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain \cite{Vogt2016}.
 
@@
 \paragraph{Sensory-Discrimination}
-Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the S1 ana S2 cortices.
+Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the S1 and S2 cortices.
 The S1 cortex is responsible for the discrimination of fine-touch sensations.
@@
 The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}.
-The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation.
-Regions such as vmPFC, dMPFC and the anterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}.
+The regions have been shown to consistently decrease in activity during attention-demanding, and non-self-referential tasks,when compared with activities during relaxed non-task states \cite{Raichle2015}.
+The DMN is believed to instantiates processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}.
+The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation.
+Regions such as vmPFC and dMPFC overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}.
 Pain related DMN deactivation have been found to be attenuated  in mind wandering away from pain, and may reflect antinociceptive processes \cite{Kucyi2013}, and that the DMN may undergo functional reorganization under chronic pain \cite{Baliki2014}.
@@
 Craig postulates that the insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}.
+The anteior insula has been shown by functional connectivity analysis to have interactions with anteior MCC and middle insula with posterior MCC \cite{Taylor2009}.
+Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring \cite{Vogt2016}.
 
@@
 Trauma and other disruptions to the spinal pathways can affect noxious signal transmission.
+Where repeated and sustained noxious stimuli may cause the central nervous system to adapt and amplify responses to stimulation.
+These can involve activation-dependence plasticity in neurons of the dorsal horn, which result in windup of action potential discharge of these neurons.
+Additionally, long-term depression may also result in the inhibitory mechanism in interneurons of the substanita gelatinosa \cite{Woolf2000}.
 White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis.

### 2018-04-24 contents/reviews/pain.tex
Updates from ShareLaTeX

@@
 \cite{Melzack1999}.
-The concept is now more commonly referred to as the "pain matrix" \cite{tracey2007a}.
+The concept is now more commonly referred to as the "pain matrix" \cite{Tracey2007a}.
 (See section \ref{section:neuromatrix})
@@
 These regions include the posterior insula, lateral thalamus and dorsal striatum.
-In chronic pain conditions however, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Lorenz2005}.
+In chronic pain conditions however, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Jensen2016,Alomar2016,Lorenz2005}.
 In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}.
 Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and medial prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}.
-A shift in forebrain activities is observed in neuropathic versus acute pain, abnormal activities in anterior cingulate cortex and prefrontal cortex is commonly observed in these studies \cite{Davis2006}.
+ Additionally, evidence of a "pain memory" circuitry involving the MCC \cite{Vogt2016,Nevian2017}, and abnormalities in the reward circuitry \cite{Tracey2017} have been implicated in chronic pain.
 
@@
 Differentiating neuropathic and acute pain however is more difficult.
-Functional connectivity based classification on differentiating fibromyalgia, rheumatoid arthritis and controls was able to achieve a diagnositc accuracy of 69\%, and classification accuracy of 78\% \cite{Sundermann2014}.
+Functional connectivity based classification on differentiating fibromyalgia, rheumatoid arthritis and controls was only able to achieve a diagnositc accuracy of 69\%, and classification accuracy of 78\% \cite{Sundermann2014}.

### 2018-04-25 contents/abstract.tex
Updates from ShareLaTeX

--- 2018-04-16 contents/abstract.tex
@@
-\chapter{Abstract}
-
+\begin{abstract}
+\addcontentsline{toc}{chapter}{Abstract}
 Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches.
@@
 The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.
+\end{abstract}

### 2018-04-25 contents/ap_paper.tex
Updates from ShareLaTeX

--- 2018-02-28 thesis/contents/ap_paper.tex
@@
 
-\chapter{Study II: Diffusion Weighted Image to T1 Co-registration Without Reverse-blip: Investigation of Best Practices}
+\chapter{Study II: Diffusion Weighted Image Co-registration: Investigation of Best Practices}
 \chaptermark{Study II}
@@
 \includegraphics[width=\linewidth]{figure4}
-\caption{Visual comparisons of MDWI to T1 co-registrations. 
+\caption[Visual comparisons of MDWI to T1 co-registrations.]{Visual comparisons of MDWI to T1 co-registrations. 
 Shown are examples of MDWI images from one subject deformed using different registration transforms derived from the different image intermediates. 
@@
     \includegraphics[width=\linewidth]{figure5}
-    \caption{Similarity scores between deformed dMRI scalar images and T1 following registrations with CC as cost function for SyN. 
+    \caption[Similarity scores between deformed dMRI scalar images and T1 following registrations with CC as cost function for SyN.]{Similarity scores between deformed dMRI scalar images and T1 following registrations with CC as cost function for SyN.
     Three steps of the registration process are plotted: Initial identity transformation (none), affine transformation (affine) and affine+SyN transformation (SyN).
@@
         \includegraphics[width=\linewidth]{figure6}
-        \caption{
+        \caption[Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with CC as cost function.]{
         Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with CC as cost function.
@@
     \includegraphics[width=\linewidth]{sup1}
-    \caption{Similarity scores between deformed dMRI scalar images and T1 following registrations with MI as cost function for SyN. 
+    \caption[Similarity scores between deformed dMRI scalar images and T1 following registrations with MI as cost function for SyN.
+]{Similarity scores between deformed dMRI scalar images and T1 following registrations with MI as cost function for SyN.
     Three steps of the registration process are plotted: Initial identity transformation (none), affine transformation (affine) and affine+SyN transformation (SyN).
@@
         \includegraphics[width=\linewidth]{sup2}
-        \caption{Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with MI as cost function.
+        \caption[Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with MI as cost function.]{Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with MI as cost function.
 Top row shows the violin plot distribution of CC and MI percent change under All Deformations and Autodeformations.

### 2018-04-25 contents/contributions.tex
Updates from ShareLaTeX

\chapter*{Statement of Contributions}
\addcontentsline{toc}{chapter}{Statement of Contributions}

{\parskip=12pt
\noindent
Dr. David Qixiang Chen (author) - all aspect of the work, including some healthy subject enrollment and data collection, experiment design, data analysis, methods development, data interpretation, and thesis preparation.

\noindent
Dr. Mojgan Hodaie (supervisor) - mentorship, assistance with experiment design, assistance with data interpretation, and assistance with thes
is preparation.

\noindent
Dr. Karen D. Davis (committee member) - mentorship, assistance with thesis preparation.

\noindent
Dr. David Mikulis (commitee member) - mentorship, assistance with thesis preparation.

\noindent
Dr. Jidan Zhong - assistance with experiment design, assistance with data interpretation, assitance with method development

\noindent
Dr. Dave J. Hayes - assistance with experiment design, assistance with data interpretation

\noindent
Powell Chu - assistance with data collection, assistance with method development
}

### 2018-04-25 contents/paper1.tex
Updates from ShareLaTeX

@@
 \includegraphics[width=\textwidth]{figure1.png}
-\caption{Figure shows the MR images of a typical MS-TN patient at the level of pons.
+\caption[Figure shows the MR images of a typical MS-TN patient at the level of pons.]{Figure shows the MR images of a typical MS-TN patient at the level of pons.
 The presence of MS plaque in the CNS at the level of trigeminal nerve is characteristic of MS-TN (panel A).
@@
 \includegraphics[width=\textwidth]{figure2.png}
-\caption{Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views.
+\caption[Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views.]{Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views.
 The reconstructed tracts are overlaid onto the axial DTI/anatomical images.
@@
 \includegraphics[width=\textwidth]{figure3.png}
-\caption{Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method.
+\caption[Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method.]{Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method.
 Images are shown in superior axial view.
@@
 \includegraphics[width=\textwidth]{figure4.png}
-\caption{Top row: Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls.
+\caption[Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls.]{Top row: Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls.
 Colored squares indicate statistically significant differences (p\textless 0.05) after correcting for multiple comparisons with false discovery rate.

### 2018-04-25 contents/reviews/pain.tex
Updates from ShareLaTeX

@@
 Dull pain is triggered by receptors that are associated with slow conducting C fibres ($<1$ m/s).
-$A\delta$ and C fibres can be co-activated in the presence of noxious stimuli.
 C-fibre nociceptors can be further differentiated into mechano-heat, silent, and polymodal subtypes.
-$A\delta$-fibre nociceptors can be divided into high-threshold mechano-heat, low-threshold mechano-heat, and noxious cold subtypes.
+$A\delta$-fibre nociceptors can be divided into high-threshold mechano-heat, low-threshold mechano-heat, and noxious cold subtypes \cite{Julius2013}.
 
@@
 
-The peripheral nociception signals are conducted from the axonal endings of the nociception neurons, whose cell bodies are located in the spinal cord dorsal root gangions, to the dorsal horn of the spinal cord (\ref{fig:drg}).
+The peripheral nociception signals are conducted from the axonal endings of the nociceptive neurons, whose cell bodies are located in the spinal cord dorsal root gangions, to the dorsal horn of the spinal cord (\ref{fig:drg}).
 The $A\delta$ and C afferents project directly to lamina I of the dorsal horn.
@@
 It takes inputs from laminae III and IV, and projects to the thalamus via the medial lemniscus.
-The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions.
+The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions \cite{kandel2000principles}.
 
@@
 Peripheral neuropathic pains can arise when there are disruptions in peripheral signal transmission.
-The disruption can be a result of injury to the first order peripheral afferent neurons, which result in increased sensitivity to stimuli and via ectopic action potentials.
+The disruption can be a result of injury to the first order peripheral afferent neurons, which result in increased sensitivity to stimuli and via ectopic action potentials \cite{Costigan2009}.
 Changes in peripheral neurotropic factors as a result of an injury can cause increased membrane excitability or spurious sprouting of the synaptic terminals in the dorsal horn of the spinal cord.

### 2018-04-25 contents/sagit.tex
- reorg folder
Updates from ShareLaTeX

--- 2018-02-28 thesis/contents/sagit.tex
@@
-\graphicspath{{images/sagit/drafts/}}
+\graphicspath{{images/sagit/}}
 
@@
 
-This study was published in Frontiers In Neuroanatomy \textcopyright 2016 Chen et al. :
+This study was published in Frontiers In Neuroanatomy \textcopyright 2016:
 
@@
 \centering
-\includegraphics[width=140mm]{figure1.png}
+\includegraphics[width=140mm]{figure1.pdf}
 \caption{Flow diagram of the SAGIT group tractography framework. The SAGIT framework is designed to be fully configurable and extensible.}
@@
 \includegraphics[width=\linewidth]{figure2.png}
-\caption{Example of the auto-generated visual panels.
+\caption[Example of the auto-generated visual panels.]{Example of the auto-generated visual panels.
 The panels that best represent the resulting anatomy are compiled.
@@
 \includegraphics[width=\linewidth]{figure3.png}
-\caption{Example of merged tractography models for each of the techniques and anatomy.
+\caption[Example of merged tractography models for each of the techniques and anatomy.]{Example of merged tractography models for each of the techniques and anatomy.
 The fornix is colored to distinguish tracts from each of the 42 subjects, in order to demonstrate the effect of combining the group data.
@@
 	\centering
-	\includegraphics[width=\textwidth]{figure5.png}
-	\caption{Normalized anatomical scores as rated by experts.
+	\includegraphics[width=\textwidth]{figure5.pdf}
+	\caption[Normalized anatomical scores as rated by experts.]{Normalized anatomical scores as rated by experts.
 The error bars denote standard deviations of the ratings.
@@
 \centering
-\includegraphics[width=\textwidth]{figure6.png}
+\includegraphics[width=\textwidth]{figure6.pdf}
 \caption{NOS scores for each anatomy. It can be noted that having ROI filters severely reduces reproducibility in deterministic methods.}
@@
 \centering
-\includegraphics[width=\textwidth]{figure7.png}
+\includegraphics[width=\textwidth]{figure7.pdf}
 \caption{Correlation between rating scores and NOS. Filtered tractography NOS is significantly (p\textless 0.05) correlated with ratings. However unfiltered tractography show no significance.}

### 2018-04-25 contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 \includegraphics[width=\linewidth]{figure-path-analysis.pdf}
-\caption{Diagram for the along-the-tract data processing.
+\caption[Diagram for the along-the-tract data processing.]{Diagram for the along-the-tract data processing.
 Process starts with 2 specific goals: 1) Derive the spatial structure that best represent the merged streamlines, 2) Deriving sub-regional tissue information from these.
@@
 \includegraphics[width=\linewidth]{figure-GP-training.pdf}
-\caption{Diagram for generating the best performing classifier for a particular window size.
+\caption[Diagram for generating the best performing classifier for a particular window size.]{Diagram for generating the best performing classifier for a particular window size.
 The data was derived from Figure \ref{fig:GPfigure-method-centroid}, where the top bar represents data from each centroid division, and each position represents the anatomical region of the trigeminal pathways within it.
@@
 \includegraphics[width=\linewidth]{view-decussation.png}
-\caption{Focused view of the TPT and its decussation, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
+\caption[Focused view of the TPT and its decussation, prior to the generation of centroids.]{Focused view of the TPT and its decussation, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
 Colours were chosen arbitrarily to denote left and right pathways, however the colour intensity represents FA values as depicted by FA scale, top right corner.
@@
 \includegraphics[width=\linewidth]{view-left-right.png}
-\caption{Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.
+\caption[Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.]{Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.
 Images obtained prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
@@
 \includegraphics[width=1.2\textwidth,center]{figure-GP-CNV.pdf}
-\caption{CN V classifier performances.
+\caption[CN V classifier performances.]{CN V classifier performances.
 This figure consists of 3 panels.

### 2018-04-26 contents/ap_paper.tex
Updates from ShareLaTeX

@@
 An alternative to FA is to use the generalized fractional anisotropy (GFA) image.
-Since FA shows low intensity in cross-fiber regions, it may adversely affect registration in those areas.
-GFA can properly characterize crossing-fiber intensities, and therefore may be a better candidate than FA.
+Since FA shows low intensity in cross-fibre regions, it may adversely affect registration in those areas.
+GFA can properly characterize crossing-fibre intensities, and therefore may be a better candidate than FA.
 The GFA is computed from the SH coefficients of spherical harmonics based methods such as Q-ball, where the diffusion orientation distribution function (ODF) is denoted as  $\Psi(u)$ \cite{Tuch2004}, such that:
@@
 AP have similar visual intensity to T$_1$, and can be computed from HARDI acquisitions without any acquisition modifications.
-Since AP are derived from the SH model, it is also capable of proper cross-fiber region characterization.
+Since AP are derived from the SH model, it is also capable of proper cross-fibre region characterization.
 AP images are derived from the even harmonic orders ($l$) of SH coefficients \cite{Descoteaux2006,Frank2002} and are defined as

### 2018-04-26 contents/discussions.tex
Updates from ShareLaTeX

@@
 \item Imaging of the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography.
-Prior to this work, CN V tractography was limited to single-tensor DTI models and trigeminal fibers could not be visualized as they enter into the pons.
+Prior to this work, CN V tractography was limited to single-tensor DTI models and trigeminal fibres could not be visualized as they enter into the pons.
 Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increasing the range of study of trigeminal-related pathology that can be studied with neuroimaging, such as MS-TN.
@@
 
-\section{Tractography of the trigeminal nerve and brainstem fibers}
+\section{Tractography of the trigeminal nerve and brainstem fibres}
 
 An important motivation for the use of tractography to study the trigeminal nerve was the ability to overcome the significant limitations of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus.
-Prior to the use of tractography, the fibers of the trigeminal nerve and connected pathways could only be inferred from anatomical studies or postmortem dissections.
+Prior to the use of tractography, the fibres of the trigeminal nerve and connected pathways could only be inferred from anatomical studies or postmortem dissections.
 
@@
 
-Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter remains a limitation.
-The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure, therefore, poses a fundamental challenge for neuroimaging.
-Techniques to improve visualization of white matter fibers consist of probabilistic tractography, which uses tissue priors to sample a large number of directions at each step stochastically, and combines the final result as a volumetric probability map.
+Diffusion tensor imaging revolutionized how we examine brain white matter fibres, yet resolving cross-fibres of the cerebral white matter remains a limitation.
+The brain white matter consists of complex crossing fibre crossings, and the inability for DTI to reliably image these structure, therefore, poses a fundamental challenge for neuroimaging.
+Techniques to improve visualization of white matter fibres consist of probabilistic tractography, which uses tissue priors to sample a large number of directions at each step stochastically, and combines the final result as a volumetric probability map.
 This kind of image-based probabilistic tractography has proved useful for whole-brain white matter studies.
@@
 Moreover, the algorithm depends on pre-generated tissue-priors, which excludes cranial nerves.
-Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempt to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions where required) in the diffusion acquisition.
+Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempt to model the fibre orientation distribution by dramatically increasing the number of gradient directions (256--512 directions where required) in the diffusion acquisition.
 The acquisition time of these sequences can be impractical for the clinical environment.
@@
 The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), are a significant development.
-SH-based methods model the fiber orientation distribution as a decomposition of SH basis functions.
+SH-based methods model the fibre orientation distribution as a decomposition of SH basis functions.
 This means that these methods are more robust to noise, and therefore require less number of gradient acquisitions.
 CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions.
-Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with at least 50 gradient directions.
+Another family of tractography (XST) algorithm constrained its model assumption to two fibre crossings, also proved to provide good results with at least 50 gradient directions.
 
-These algorithmic developments brought imaging of crossing-fibers using tractography to the realm of possibility.
+These algorithmic developments brought imaging of crossing-fibres using tractography to the realm of possibility.
 We explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST.
 The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images.
-2) it courses through simple (cistern) and complex (pons) fiber environments; 3) it is a prominent cranial nerve, and therefore a good benchmark for imaging small anatomy.
+2) it courses through simple (cistern) and complex (pons) fibre environments; 3) it is a prominent cranial nerve, and therefore a good benchmark for imaging small anatomy.
 
@@
 
-\subsection{Delineating the brainstem fibers of the trigeminal nerve}
+\subsection{Delineating the brainstem fibres of the trigeminal nerve}
 
@@
 While the acquisition is not ideal, it sacrifices resolution only in the z-axis, however the tractography visual fidelity is improved.
-Imaging the brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers.
+Imaging the brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibres.
 A number of methods have been devised, including probabilistic tractography (FSL)\cite{Behrens2007}, which generates image-based results that represent the propagation probability of a seed region.
-Neurosurgical visualization however requires three dimensional representation of the fibers, and therefore streamline-based tractography visualizations are of broader appeal.
+Neurosurgical visualization however requires three dimensional representation of the fibres, and therefore streamline-based tractography visualizations are of broader appeal.
 In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater than 50 gradient directions.
@@
 \subsection{Tractography group registration}
-Tractography is a powerful methodology that permits the visualization of white matter fibers in three dimensions.
+Tractography is a powerful methodology that permits the visualization of white matter fibres in three dimensions.
 Its primary use, however, has been limited to case studies and small-scale anatomy comparative analysis.
@@
 Despite this, certain regions such as the TPT could not be reliably reconstructed in all of the subjects.
-Anatomy featuring intricate but complex cross-sections, such as the thalamoinsular projections which include the claustrum, and medullary decussations of the spinal trigeminal fibers will remain out of reach of the current algorithms without substantial improvement in DWI resolution.
+Anatomy featuring intricate but complex cross-sections, such as the thalamoinsular projections which include the claustrum, and medullary decussations of the spinal trigeminal fibres will remain out of reach of the current algorithms without substantial improvement in DWI resolution.
 Fortunately, SAGIT can adopt MR images of any resolution, so future studies in higher isovoxel resolution can be immediately processed as soon as they become available.
@@
 
-\section{Future Directions}
+\chapter{Conclusions}
+
+In sum, this thesis provides novel evidence for diffusivity tissue disruptions of the CN V in TN, and bilateral CN V tissue abnormalities in TN patients when compared to controls.
+The study of the trigeminal pontine decussation (TPT) and thalamocortical S1 pathway demonstrated white matter diffusivity changes at all levels of the trigeminal sensory pathway.
+These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies.
+These results demonstrate that TN involves wide-spread changes at all levels of CNS white matter.
+The methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.
+
+\chapter{Future Directions}
 
@@
 Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.
-
-\section{Conclusions}
-
-In sum, this thesis provides novel evidence for diffusivity tissue disruptions of the CN V in TN, and bilateral CN V tissue abnormalities in TN patients when compared to controls.
-The study of the trigeminal pontine decussation (TPT) and thalamocortical S1 pathway demonstrated white matter diffusivity changes at all levels of the trigeminal sensory pathway.
-These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies.
-These results demonstrate that TN involves wide-spread changes at all levels of CNS white matter.
-The methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.

### 2018-04-26 contents/hypotheses.tex
Updates from ShareLaTeX

@@
 
-The general aim of this thesis is to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease.
-This necessitates detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls.
+This thesis consists of two components: 1) to develop and validate methodology to study specific white matter in individuals and in groups of individuals, and 2) to apply the developed methods to gain insight into the changes in the trigeminal system in TN.
+The methodology developed for group analysis lays the foundation for future clinical applications.
 
-With an increasing number of subjects, single subject analysis becomes cumbersome.
-Critical methodological advances are necessary for the feasibility of the general aim.
+The general goal was to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease.
+This necessitated detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls.
+
+With an increasing number of subjects, single subject analysis can become cumbersome.
+Critical methodological advances were necessary for the feasibility of the general aim.
 These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy.
 
-Lastly, these methodological advances will be applied to the large-scale study of TN versus healthy control.
-The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, in order to identify specific diffusivity pattern that differentiates between TN and non-TN.
+Lastly, the methodological advances of this thesis were applied to the large-scale study of TN versus healthy control.
+The considerable amount of data output from the methodology  enabled the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, and in order to identify specific diffusivity pattern differentiating TN from healthy controls.
 The roadmap and rationale for the specific studies are described below.
@@
 \section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
-MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons.
+MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibres in the pons.
 However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals.
@@
 \subsection{Main Aim} 
-Establish the feasibility of multi-tensor diffusion tractography to delineating the cisternal, root entry zone and brainstem segments of CN V and how diffusivity changes in these sub-regions correlate with pain in MS-TN.
+Establish the feasibility of multi-tensor diffusion tractography to delineate the cisternal, root entry zone and brainstem segments of CN V and how diffusivity alterations in these sub-regions correlate with pain in MS-TN.
 
@@
 \begin{itemize}
-    \item TN is known to relate to vascular compression of the cistern REZ, therefore its diffusivity signature will occur in the cisternal region.
-    \item MS-TN is related to brainstem MS lesions.
-Therefore CN V segments close to the lesions will show more significant changes in tissue diffusivity, whereas segments far away from the lesions will not.
+    \item MS-TN is pathophysiologcially related to brainstem MS lesions. Therefore CN V segments close to these lesions will demonstrate diffusivity alterations that correlate with pain laterality.
     \item As a demyelinating disease, the pattern of change for MS-TN diffusivity signatures will reflects demyelination, including increase in RD, and decrease in FA. 
@@
 
-\subsection{Main Aim}
+\subsection{Main Objective}
 The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population.
@@
 
-\subsection{Specific Aim}
+\subsection{Specific Objectives}
 \begin{itemize}
     \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD.
+
     \item Use the above techniques to delineate a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
+
+    \item Determine if specific tractography algorithms are preferentially suited to delineate specific pathways in the brain.
+For example, deterministic tractography may show more specificity in delineating cranial nerves, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
+
     \item Validate normalized overlapping scores (NOS) by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
-\end{itemize}
+Demonstrate that NOS score will correlate with expert ratings, and show improved rating consistency over human rater variability.
 
-\subsection{Specific Hypothesis}
-\begin{itemize}
-    \item DTI is known to not able to delineate cross-fibers regions.
-If the group tractography works consistently, then its output will reflect such known algorithmic limitations, and DTI group tracts will not be able to perform well on more complex anatomies.
-The grouped tracts will however show consistent regional anatomical overlap across individuals.
-    \item Specific tractography algorithms will be preferentially suited for the visualization of specific areas in the brain.
-For examples, deterministic tractography will preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
-    \item NOS score will correlate with expert ratings, and show improved rating consistency over human rater variability.
 \end{itemize}
@@
 \section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
-The developed methodology is applied to the study of the trigeminal sensory pathway.
-The method used is now thoroughly tested, and the best-suited tractography methods determined from the previous study.
-This study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group.
-The pathway is divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment.
-Along-the-tract analysis are carried out by dividing each segment into 30 subdivisions.
-Following this, we use machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.
+The methodology developed in the previous study was applied to the study of the trigeminal sensory pathway.
+The aim of this study was to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group.
+The pathway was divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment.
+Along-the-tract analysis were carried out by dividing each segment into 30 subdivisions.
+Following this, we used machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.
 

### 2018-04-26 contents/paper1.tex
Updates from ShareLaTeX

@@
 Panels C and D show the placements of the ROIs along the CN V nerve bilaterally in order to measure diffusivity statistics, overlaid with the axial anatomical/ DTI image.
-The colors of the DTI image are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations.
+The colours of the DTI image are rendered as colour-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations.
 Diffusion statistics were measured from four groups of regions: cisternal segment, root entry zone (REZ), pontine, and peri-lesional, as illustrated.
@@
 The reconstructed tracts are overlaid onto the axial DTI/anatomical images.
-The colors of the underlying scan are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations.
-The colors of the tractography fibers represent the spectrum values of FA (0 to 1), as shown by the legend at the top right corner.
-Panel A shows that SDT tractography delineated the cerebellar peduncle fibers, but cannot distinguish the brainstem trigeminal fibers (yellow arrows).
+The colours of the underlying scan are rendered as colour-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations.
+The colours of the tractography fibres represent the spectrum values of FA (0 to 1), as shown by the legend at the top right corner.
+Panel A shows that SDT tractography delineated the cerebellar peduncle fibres, but cannot distinguish the brainstem trigeminal fibres (yellow arrows).
 Red arrows denote the starting point of tract generation for both SDT and MTT methods.
@@
 Panel B shows the TN tractography delineations, where there is no such FA decrease in the brainstem segment.
-Please refer to Figure \protect\ref{fig:MSfigure2} for explanations of color scales and figure annotations.}
+Please refer to Figure \protect\ref{fig:MSfigure2} for explanations of colour scales and figure annotations.}
 \centering
@@
 \caption[Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls.]{Top row: Diffusivity difference estimate matrices of the Cisternal ROIs demonstrating significant differences in regional diffusivities between MS-TN patients, TN patients and controls.
-Colored squares indicate statistically significant differences (p\textless 0.05) after correcting for multiple comparisons with false discovery rate.
+coloured squares indicate statistically significant differences (p\textless 0.05) after correcting for multiple comparisons with false discovery rate.
 The (+/-) sign indicate symptomatic (+) and asymptomatic (-) sides of the subject group.
@@
 Similar findings can be seen in the pontine region, where only the MS-TN AD measures are distinctly different from the other groups.
-However while AD is commonly correlated with axonal integrity, the high concentration of crossing fibers in the pontine region may introduce confounds into these correlations.
+However while AD is commonly correlated with axonal integrity, the high concentration of crossing fibres in the pontine region may introduce confounds into these correlations.
 Improvements in diffusivity metric measurements in such tissues are needed to further understand these findings.

### 2018-04-26 contents/reviews/dmri.tex
Updates from ShareLaTeX

@@
 
-Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibers \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.
+Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibres \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.
 
@@
 DTI is based on the Gaussian model of water diffusion, and therefore imposes an important limitation.
-The Gaussian tensor can derive only one diffusion direction at a given voxel, and therefore is incapable of modelling cross-fiber diffusion of two or more directions.
+The Gaussian tensor can derive only one diffusion direction at a given voxel, and therefore is incapable of modelling cross-fibre diffusion of two or more directions.
 By extension, its derived metrics (FA, AD, RD) can only characterize a single diffusion direction as well.
-Therefore the interpretation of these metrics in white matter that contains cross-fiber configurations is tenuous.
+Therefore the interpretation of these metrics in white matter that contains cross-fibre configurations is tenuous.
 
@@
 An end-to-end spherical deconvolution diffusion model was first proposed by Tournier et al. \cite{Tournier2004}. 
-Consider a fiber population with distinct fiber components. The measured diffusion signal $S(\theta,\phi)$ can be written as:
+Consider a fibre population with distinct fibre components. The measured diffusion signal $S(\theta,\phi)$ can be written as:
 
@@
 \end{equation}
-where $\theta$ is the elevation angle,  $\phi$ the azimuth angle in spherical coordinates; $f_i$ is the volume fraction of the $i$-th fiber component, $\hat{A}$ is the rotation operator that orients fiber, and $R(\theta)$ is an axially symmetric response function.
+where $\theta$ is the elevation angle,  $\phi$ the azimuth angle in spherical coordinates; $f_i$ is the volume fraction of the $i$-th fibre component, $\hat{A}$ is the rotation operator that orients fibre, and $R(\theta)$ is an axially symmetric response function.
 
-$S(\theta,\phi) $ can be rewritten as a convolution over the unit sphere of $R(\theta)$ with a fiber orientation density function (FOD) $F(\theta, \phi)$, such that:
+$S(\theta,\phi) $ can be rewritten as a convolution over the unit sphere of $R(\theta)$ with a fibre orientation density function (FOD) $F(\theta, \phi)$, such that:
 \begin{equation}
@@
 Spherical harmonics are orthonormal basis functions over a sphere, and therefore can approximate any symmetric diffusion geometry in the spherical q-space.
-The rotational harmonics similarly form the orthonormal basis over all possible rotations, and can represent any symmetric fiber configurations.
+The rotational harmonics similarly form the orthonormal basis over all possible rotations, and can represent any symmetric fibre configurations.
 The harmonics have two important parameters: the harmonic order $l$ and phase factor $m$, with $ -l \leq m \leq l $ for each harmonic order level $l$.
@@
 To reduce the number of parameters to be computed, we can make the further assumption that the diffusion process is symmetric, and therefore the odd $l$ harmonic orders can be ignored, with only the even harmonics considered.
-The response function $R(\theta)$ can be directly estimated from the dataset by supplying DWI region with high anisotropy from a single fiber bundle, such as the body of the corpus callosum.
+The response function $R(\theta)$ can be directly estimated from the dataset by supplying DWI region with high anisotropy from a single fibre bundle, such as the body of the corpus callosum.
 
@@
 Deterministic tractography lends itself naturally to anatomical interpretation and further geometric analysis.
-It was traditionally paired with DTI, and so was thought to be limited in the case of crossing fibers.
+It was traditionally paired with DTI, and so was thought to be limited in the case of crossing fibres.
 
-Other methods attempted to circumvent the limitation with probabilistic approaches that include white matter tissue priors in the fiber propagation model.
-The most popular of the probabilistic approach is the FSL software's probabilistic tractography module (FDT) \cite{Behrens2007}, which also includes a model-based ball-and-stick diffusion model that can resolve 2-fiber crossings.
+Other methods attempted to circumvent the limitation with probabilistic approaches that include white matter tissue priors in the fibre propagation model.
+The most popular of the probabilistic approach is the FSL software's probabilistic tractography module (FDT) \cite{Behrens2007}, which also includes a model-based ball-and-stick diffusion model that can resolve 2-fibre crossings.
 The FDT software produces visitation images of its probabilistic streams that collate the probability of tract propagation at each voxel.
@@
 
-The availability of higher order HARDI methods, especially in CSD allows the reinterpretation of FOD as local fiber orientation probability distribution \cite{Tournier2010,Jeurissen2011b}.
+The availability of higher order HARDI methods, especially in CSD allows the reinterpretation of FOD as local fibre orientation probability distribution \cite{Tournier2010,Jeurissen2011b}.
 Development of a probabilistic streamline propagator based on CSD can now combine the advantage of probabilistic models with the geometric interpretability of deterministic tractography.

### 2018-04-26 contents/reviews/pain.tex
Updates from ShareLaTeX

@@
 The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that relays touch or pain sensations.
-The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission.
+The transmission control depends on the large and small diameter primary afferents, where large fibres inhibit, while small fibres facilitate the transmission.
 The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.
@@
 The regions have been shown to consistently decrease in activity during attention-demanding, and non-self-referential tasks,when compared with activities during relaxed non-task states \cite{Raichle2015}.
-The DMN is believed to instantiates processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}.
+The DMN is believed to instantiate processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}.
 The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation.
@@
 Craig postulates that the insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}.
-The anteior insula has been shown by functional connectivity analysis to have interactions with anteior MCC and middle insula with posterior MCC \cite{Taylor2009}.
+The anterior insula has been shown by functional connectivity analysis to have interactions with anterior MCC and middle insula with posterior MCC \cite{Taylor2009}.
 Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring \cite{Vogt2016}.
@@
 These can involve activation-dependence plasticity in neurons of the dorsal horn, which result in windup of action potential discharge of these neurons.
-Additionally, long-term depression may also result in the inhibitory mechanism in interneurons of the substanita gelatinosa \cite{Woolf2000}.
+Additionally, long-term depression may also result in the inhibitory mechanism in interneurons of the substantia gelatinosa \cite{Woolf2000}.
 White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis.
-It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures.
+It is believed that injury to white matter fibres can affect action potential conduction between cortical and subcortical structures.
 This may give rise or contribute to neuropathic pain \cite{Geha2008}.

### 2018-04-26 contents/reviews/trigeminal.tex
Updates from ShareLaTeX

@@
 
-The trigeminal nerve, or the fifth cranial nerve (CN V) is the primary afferent for both facial somatosensation and pain (\ref{fig:trig-branches}).
+The trigeminal nerve, or the fifth cranial nerve (CN V) is the primary afferent for both facial somatosensation and pain (Figure \ref{fig:trig-branches}).
 It consists of, from superior to inferior, the ophthalmic (V1), maxillary (V2) and mandibular (V3) branches.
@@
 
-Trigeminal sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messiner's, Pacinian and Ruffini corpuscles.
+Trigeminal sensory afferents include the myelinated A and unmyelinated C-fibres that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messiner's, Pacinian and Ruffini corpuscles.
 
@@
 
-CN V passes into the pons (Figure \ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem.
+CN V passes into the pons (Figure \ref{fig:trig-pathway}), through the transverse pontine fibres and enters the tegmentum of the pontine brainstem.
 It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem.
@@
 
-The main sensory nucleus receives discriminative input from the orofacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract.
+The main sensory nucleus receives discriminative input from the orofacial region, and along with the spinal trigeminal nucleus fibres, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract.
 A separate branch responsible for oral cavity, however, projects to the ipsilateral VC via the dorsal trigeminothalamic tract.
@@
 Manual segmentation of T1 and T2 MRI requires great effort, with mixed results \cite{Miller2008a}.
-Diffusion MRI provided fiber orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}.
+Diffusion MRI provided fibre orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}.
 Additionally, DTI showed strong ability to visualize spatial relationships in white matter anatomy, for example in delineating fornix sub-regions \cite{Chen2015c}, and in visualizing the distortion of CN VII/VIII by vestibular schwannomas \cite{Chen2011b}.

### 2018-04-26 contents/sagit.tex
Updates from ShareLaTeX

@@
 
-\textbf{Results:} At a group level, merged tractography demonstrated that different methods have different fiber distribution characteristics.
+\textbf{Results:} At a group level, merged tractography demonstrated that different methods have different fibre distribution characteristics.
 CST$_{prob}$ is prone to false-positives, and thereby suitable in anatomy with strong priors.
@@
 Diffusion magnetic resonance imaging (dMRI) tractography is an imaging analysis technique that permits non-invasive visualization of white matter anatomy $in$ $vivo$ \cite{Basser2000}.
-It is based on the observation that the Brownian motion of water molecules within white matter fibers is constrained by the axonal bundles, and therefore such anisotropic water diffusion can be used to probe tissue microstructure \cite{Bihan2001a}.
+It is based on the observation that the Brownian motion of water molecules within white matter fibres is constrained by the axonal bundles, and therefore such anisotropic water diffusion can be used to probe tissue microstructure \cite{Bihan2001a}.
 By scanning in multiple angular directions using diffusion weighted imaging (DWI) sequence, a model of diffusion can be estimated for each voxel of the scanned image volume.
@@
 Using the tensor information, tractography algorithms \cite{Mori2002b} can be used to trace out structures within the DTI volume.
-The limitation of DTI is that there is insufficient information to resolve areas with crossing fibers with one tensor per voxel.
-Improvements over the limits of the DTI model, particularly in increasing angular resolution to improve crossing fibers resolution has been a subject of great interest \cite{Fritzsche2010d,Jeurissen2012d,Tuch2002f}.
+The limitation of DTI is that there is insufficient information to resolve areas with crossing fibres with one tensor per voxel.
+Improvements over the limits of the DTI model, particularly in increasing angular resolution to improve crossing fibres resolution has been a subject of great interest \cite{Fritzsche2010d,Jeurissen2012d,Tuch2002f}.
 
-To improve the crossing fiber information per voxel, significant changes in DWI acquisition strategies are required.
+To improve the crossing fibre information per voxel, significant changes in DWI acquisition strategies are required.
 These methods include diffusion spectrum imaging (DSI)\cite{Wedeen2005,Wedeen2008} which samples the Q-space in a Cartesian grid, and Q-ball imaging that is based on the Funk-Radon transform \cite{Cho2008,Descoteaux2007a} that samples the Q-space in a spherical shell.
@@
 Alternatively, methods based on spherical deconvolution (SD) can be used on HARDI scans of lesser angular resolution.
-The approach is to model the HARDI signal as a convolution of the fiber orientations.
-The resulting deconvoluted fiber orientation distribution (FOD) shows better discrimination of fiber directions compared to Q-ball under similar scanning parameters, but is susceptible to image noise\cite{Anderson2005,DellAcqua2007,Tournier2004}.
-Alternatively, multi-tensor model-based approaches estimate the fiber configuration with the assumption that there are no more than two or three crossings in a given voxel.
+The approach is to model the HARDI signal as a convolution of the fibre orientations.
+The resulting deconvoluted fibre orientation distribution (FOD) shows better discrimination of fibre directions compared to Q-ball under similar scanning parameters, but is susceptible to image noise\cite{Anderson2005,DellAcqua2007,Tournier2004}.
+Alternatively, multi-tensor model-based approaches estimate the fibre configuration with the assumption that there are no more than two or three crossings in a given voxel.
 These include eXtended Streamline Tractography (XST), which can delineate lateral projections of the corticospinal tract in the motor cortex, in DWI scans with 50 gradient directions \cite{Qazi2009}.
@@
 Finally, we deploy our system to the evaluation of various tractography methods in their ability to delineate neuroanatomy that are highly specific but difficult to image.
-These include supra and infratentorial structures such as: small white matter bundles in heavy crossing fibers (cranial nerves), pontine decussation (rubrocerebellar pathway) and curved central pathways (fornix, visual and auditory radiations).
+These include supra and infratentorial structures such as: small white matter bundles in heavy crossing fibres (cranial nerves), pontine decussation (rubrocerebellar pathway) and curved central pathways (fornix, visual and auditory radiations).
 We recruited 5 neuroanatomists to evaluate and rate the resulting visual reports generated by SAGIT.
@@
 The overlaps were then visualized using concentric isosurfaces created from step-wise (10\%) thresholds of the underlying overlap volume using the MayaVI data visualization library \cite{Ramachandran2011}.
-A color lookup table with even visual brightness falloff \cite{VanderWalt2015} was chosen to avoid visual judgment biases from the color presentation.
+A colour lookup table with even visual brightness falloff \cite{VanderWalt2015} was chosen to avoid visual judgment biases from the colour presentation.
 Different viewpoints (axial, coronal, sagittal, and perspective) were created for each visualization, and then composed together to create the visual report.
@@
 CST$_{prob}$ however also delineated much of the cerebellum.
-DTT by Slicer was not able to delineate the brainstem portions of the fiber and therefore was scored lower than the other methods.
+DTT by Slicer was not able to delineate the brainstem portions of the fibre and therefore was scored lower than the other methods.
 These differences in delineation can be more clearly seen in the merged tracts (Figure 3), where the XST also showed further lateral delineations of CN VII/VIII.
@@
 For anatomical reference, the best representative anatomy image slice is composed with the tract conjunction map.
-The color scale of the conjunction image represents the percentage reproducible for each region, where 1=100\%.
+The colour scale of the conjunction image represents the percentage reproducible for each region, where 1=100\%.
 For each algorithm (rows), different visual perspectives are created.
@@
 \caption[Example of merged tractography models for each of the techniques and anatomy.]{Example of merged tractography models for each of the techniques and anatomy.
-The fornix is colored to distinguish tracts from each of the 42 subjects, in order to demonstrate the effect of combining the group data.
+The fornix is coloured to distinguish tracts from each of the 42 subjects, in order to demonstrate the effect of combining the group data.
 The other anatomies are coloured by FA intensity.
@@
 It is possible that the resulting average delineations can be used to generate population-specific anatomy atlases.
-It's also possible to use fiber density images rather than binary mask for conjunction.
-However the result of an average fiber density image is difficult to interpret, and therefore should be explored in future studies.
+It's also possible to use fibre density images rather than binary mask for conjunction.
+However the result of an average fibre density image is difficult to interpret, and therefore should be explored in future studies.
 
@@
 While CST$_{prob}$ decussation delineation shows the highest reproducibility, it also produces wider projection coverage that’s harder to interpret.
-The low DTT score is not surprising, as the single-tensor's inability to resolve crossing-fibers is well known.
+The low DTT score is not surprising, as the single-tensor's inability to resolve crossing-fibres is well known.
 Based on the result of the study, we recommend the use of CST$_{prob}$ for tasks in which the target anatomy is well defined and false positively can be easily recognized , or when interhemispheric and long projection distances are desired.

### 2018-04-26 contents/tn_sagit.tex
Updates from ShareLaTeX

@@
 The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus.
-The pain pathway also projects to the medial nucleus of the thalamus, after which it courses towards the fibers of the cingulate cortex.
+The pain pathway also projects to the medial nucleus of the thalamus, after which it courses towards the fibres of the cingulate cortex.
 
@@
 However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status \cite{Hung2017} and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain \cite{Miller2009}.
-It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibers.
+It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibres.
 
@@
 \subsubsection{Tractography}
-ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (specifications: ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibers must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary and exclude the brainstem).
+ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (specifications: ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibres must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary and exclude the brainstem).
 
@@
 These include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise.
-TPT streamlines that did not decussate at the level of the pons, as well as fibers that stray into the contralateral TGN were removed.
+TPT streamlines that did not decussate at the level of the pons, as well as fibres that stray into the contralateral TGN were removed.
 S1 streamlines that stray medially towards other regions such as the corpus callosum were also removed.
@@
 \caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
-The expected increase in FA as the fibers enter the brainstem can be visualized, and denotes good registration between subjects.
+The expected increase in FA as the fibres enter the brainstem can be visualized, and denotes good registration between subjects.
 }
@@
 The first pathway appears to head into the spine, while the second pathway consists of the TPT.
-Following the decussation, the fibers proceed to the area of the sensory thalamus and from there to the cortical sensory area S1.}
+Following the decussation, the fibres proceed to the area of the sensory thalamus and from there to the cortical sensory area S1.}
 \label{fig:GPfigure3}
@@
 The left (unaffected) classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA).
-The right side showed greater overlap of segment ranges between positions 14--24, which corresponds to the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4, within the substance of the trigeminal fibers in the brainstem.
+The right side showed greater overlap of segment ranges between positions 14--24, which corresponds to the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4, within the substance of the trigeminal fibres in the brainstem.
 Overall, the windows of the right side classifiers extend further into the distal subdivisions, into the cistern and the area of the ganglion, while the left side extended more into the brainstem.
@@
 The asymmetry of this decussation has not been demonstrated before.
-It is not clear whether this represents the anatomical arrangement of these fibers, or whether decussations are always purely symmetrical (akin to a perfectly shuffled deck of cards).
+It is not clear whether this represents the anatomical arrangement of these fibres, or whether decussations are always purely symmetrical (akin to a perfectly shuffled deck of cards).
 Further assessment of other decussation points will be helpful to understand this process.
 
-We were able to successfully delineate the course of the fibers from the thalamus to S1 given the use of SAGIT and incorporation of Freesurfer cortical segmentation maps.
+We were able to successfully delineate the course of the fibres from the thalamus to S1 given the use of SAGIT and incorporation of Freesurfer cortical segmentation maps.
 Previous studies have suggested that S1 is involvement in pain location discrimination \cite{bushnell1999pain}, and have been found to exhibit grey-matter abnormalities in TN pain \cite{Desouza2013c}.
@@
 FA classifiers consistently achieved the highest accuracy in CNV and TPT, with the exception of S1.
-This might be due to the fact that it is known to have trouble in regions of heavy crossing-fibers.
-However, the TPT pathway also situations in regions of complex crossing fibers.
-One explanation for FA classifier in TPT is that the pathway courses through the narrow passage of pontine/thalamic regions, the fiber structures are highly constrained.
+This might be due to the fact that it is known to have trouble in regions of heavy crossing-fibres.
+However, the TPT pathway also situations in regions of complex crossing fibres.
+One explanation for FA classifier in TPT is that the pathway courses through the narrow passage of pontine/thalamic regions, the fibre structures are highly constrained.
 Therefore FA still remains as a sensitive diffusivity metric for the TPT.
 
-We included GFA to investigate its potential for measuring cross-fiber white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric.
-Although metrics such as FA are popular in current neuroimaging literature, they nevertheless are derived from the Gaussian tensor model, which cannot adequately describe cross-fiber regional dynamics.
-For example, in the crossing-fiber region of healthy tissue, where both bundles have equal myelin and axonal integrity, will have a low FA score due to a spherical tensor; its AD and RD will be roughly equal, and therefore MD will be higher.
+We included GFA to investigate its potential for measuring cross-fibre white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric.
+Although metrics such as FA are popular in current neuroimaging literature, they nevertheless are derived from the Gaussian tensor model, which cannot adequately describe cross-fibre regional dynamics.
+For example, in the crossing-fibre region of healthy tissue, where both bundles have equal myelin and axonal integrity, will have a low FA score due to a spherical tensor; its AD and RD will be roughly equal, and therefore MD will be higher.
 In a pathological cross region, where only one bundle is healthy, the FA increases due to the diffusivity bias in one direction; the RD will be lower, and MD will be lower as well.
@@
 The merged CN V delineations of all the subjects showed caudal projections towards the spinal cord.
-The consistent observation of diverging streamlines at the levels of the medulla strongly suggests that these fibers correspond to trigeminal spinal decussations.
+The consistent observation of diverging streamlines at the levels of the medulla strongly suggests that these fibres correspond to trigeminal spinal decussations.
 However the caudal projection were not consistently seen.
-This is likely due to the limited size of voxel resolution that in turn does not permit adequate resolution of medullary and spinal cord fiber details.
+This is likely due to the limited size of voxel resolution that in turn does not permit adequate resolution of medullary and spinal cord fibre details.
 

### 2018-04-26 contents/tn_sagit_abstract.tex
Updates from ShareLaTeX

--- 2018-02-26 thesis/contents/tn_sagit_abstract.tex
@@
 
-\textbf{Methods:} We used SAGIT based group merged tractography to study 36 sex-matched TN subjects (right-sided pain) and 36 controls, examining the following trigeminal related white matter pathways: trigeminal nerve (CN V), pontine decussation (TPT), and thalamocortical fibers (S1).
+\textbf{Methods:} We used SAGIT based group merged tractography to study 36 sex-matched TN subjects (right-sided pain) and 36 controls, examining the following trigeminal related white matter pathways: trigeminal nerve (CN V), pontine decussation (TPT), and thalamocortical fibres (S1).
 GP classifiers were trained by scrolling a moving window over CN V, TPT, and S1 tractography centroids.
-FA, GFA, RD, AD, and MD metrics were evaluated for both groups, analyzing TN vs.
+Fractional anisotropy (FA), generalized FA (GFA), radial diffusivity (RD), axial diffusivity (AD), and mean diffusivity (MD) metrics were evaluated for both groups, analyzing TN vs.
 control groups and affected vs.
@@
 
-\textbf{Conclusions:} This is the first study in TN that combines group-wise merged tractography, machine learning classification using GP, and analysis of the complete trigeminal pathways from the the trigeminal peripheral fibers to S1 cortex.
+\textbf{Conclusions:} This is the first study in TN that combines group-wise merged tractography, machine learning classification using GP, and analysis of the complete trigeminal pathways from the the trigeminal peripheral fibres to S1 cortex.
 This analysis demonstrates that TN is characterized by bilateral changes throughout the trigeminal pathway compared with healthy controls, as well as changes between clinically affected and unaffected sides.

### 2018-04-26 contents/reviews/pain.tex
Updates from ShareLaTeX

@@
 \centering
-\caption{The axonal projections of the fast conduction ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn.}
+\caption{The axonal projections of the fast conduction ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn. Modified from \protect\cite{DMello2008}}
 \label{fig:drg}
@@
  \centering
- \caption{Afferent pain pathways \protect\cite{kandel2000principles}}
+ \caption{Three of the five afferent pain pathways \protect\cite{kandel2000principles}. Reproduced with permission from Copyright Clearance Center (License ID: 4336310447247)}
  \label{fig:pain-pathways}
@@
  \centering
- \caption{ Cortical and sub-cortical regions of the pain matrix. \protect\cite{Tracey2007a}} 
+ \caption{ Cortical and sub-cortical regions of the pain matrix \protect\cite{Tracey2007a}. Reproduced with permission from Copyright Clearance Center (License ID: 4316760598837).}
  \label{fig:pain-matrix}

### 2018-06-29 contents/abstract.tex
final submission

@@
 In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image.
-In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) software framework.
-Finally, in Study IV, we deploy end-to-end machine-learning TN classification to auto-discovers diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway.
+In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) processing pipeline framework.
+Finally, in Study IV, we deploy end-to-end machine-learning TN classification to automatically discover diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway.
 In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification.

### 2018-06-29 contents/ap_paper.tex
Updates from ShareLaTeX
final submission

@@
 \begin{figure}[p]
-\includegraphics[width=\linewidth]{figure4}
+\centering
+\includegraphics[width=0.9\linewidth]{figure4}
 \caption[Visual comparisons of MDWI to T1 co-registrations.]{Visual comparisons of MDWI to T1 co-registrations. 
@@
 Primary differences in registration quality can be found in the brainstem (left and middle columns), and in the lateral ventricles, especially in the ventricular-caudate boundary (right column). }
-\centering
 \label{fig:APfigure4}
@@
 Under Autodeformations (\protect $D_{\text{auto}}$), CC percent change of FA and GFA are significantly higher than others, while AP has lower percent change.
-MI shows little percentage between all the images; GFA shows significantly differences under \protect $D_{\text{auto}}$, but also shows high variability.}
+MI shows little percentage changes between all the images; GFA shows significantly differences under \protect $D_{\text{auto}}$, but also shows high variability.}
         \centering

### 2018-06-29 contents/reviews/dmri.tex
Updates from ShareLaTeX
final submission

@@
 
-Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibres \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.
+Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibres \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques \cite{Johansen-Berg2013}.
 
@@
 \end{tabular}
-\caption{ Comparison of angle discrimination ability and CSD FOD. B-value=3000 \protect $s/mm^2$. Taken from \protect \cite{Tournier2007b} } 
+\caption{ Comparison of angle discrimination ability and CSD FOD.
+B-value=3000 \protect $s/mm^2$.
+Taken from \protect \cite{Tournier2007b}.
+Reproduced with permission from Copyright Clearance Center (License ID: 4316770133531) }
 \label{fig:csd}
@@
 Applying pathway tracking to DWI imaging in the brain was first explored by Mori \cite{Mori2002b} and Basser\cite{Basser2000}.
-Tractography in essence is the construction, or propagation of streamline from an underlying field of orientation information, in order to visualize continuous structures within a volume.
+Tractography in essence is the construction, or propagation of streamlines from an underlying field of orientation information, in order to visualize continuous structures within a volume.
 The orientation field is generally provided as a 3D matrix of directional information obtained from DTI, CSD, or other methods.

### 2018-06-29 contents/reviews/machine-learning.tex
final submission

@@
 Additional studies showed consistently good performance of ML-based nociceptive pain discrimination.
-Discrimination of chronic pain from nociceptive pain proved to be more challenging, due to the high degree of similarity in activation patterns.
-The development of pain detection through brain imaging led to debates of the validity of presenting such methods as evidences in courtroom, where Davis et al argued against such actions \cite{Davis2012a}.
+Detecting chronic pain proved to be more challenging, due to the high degree of similarity between acute and chronic pain fMRI activation patterns.
+The development of pain detection through brain imaging led to debates of the validity of presenting such methods as evidences in courtroom, where Davis et al argued against such actions \cite{Davis2012a,Davis2017}.
 While it is commonly accepted that pain evokes activity in the brain in multiple regions of the dynamic pain connectome, some studies suggest that the dorsal posterior-insula is heavily involved in chronic pain.

### 2018-06-29 contents/reviews/mri.tex
Updates from ShareLaTeX
final submission

@@
 This causes the nuclei to produce NMR signals detectable by the MRI scanner.
-Based on the precession frequency decay, the signals are recorded to construct an image of the scanned area based on the contrasts in water and fat concentrations.
+Based on the precession frequency decay, the signals are recorded to construct an image of the scanned area based on the contrasts in water and fat concentrations \cite{Plewes2012,Johansen-Berg2013}.
 
@@
 \caption{Illustration of the process of transverse relaxation.
-Initially, the spin precesses about the B0 field under steady-state.
-When a 90deg RF pulse is applied, the spin orients towards the xy-plane.
+Initially, the net magnetization orientes to the B0 field under steady-state.
+When a 90 deg RF pulse is applied, the spin orients towards the xy-plane.
 When the RF pulse is switched off, the spin now relaxes towards the z-axis.
@@
 \begin{tabular}{c}
-\includegraphics[width=\textwidth]{mri/spin-echo.pdf}
+\includegraphics[width=\textwidth]{mri/spin-echo-2.pdf}
 \end{tabular}

### 2018-06-29 contents/reviews/multiplesclerosis.tex
final submission

@@
 Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as TN.
-91\% of the patients report pain at the time of evaluation, and most experience daily pain.
-Both central and peripheral pain are reports in people of MS.
+91\% of the patients report pain at the time of evaluation, and most experience daily pain, where both central and peripheral pain are reported in people of MS.
 Central pain in MS is reported to occur 27-58\% of the subjects \cite{OConnor2008}.
@@
 The most common pain in MS includes extremity pain and TN (MS-TN), which is also termed symptomatic TN.
-MS-TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population \cite{Cruccu2009}.
+MS-TN has a prevalence 2\% in the MS population, which is 20 times higher than in a general population \cite{Cruccu2009}.
 The clinical presentation of the pain is similar to classic TN.

### 2018-06-29 contents/reviews/pain.tex
final submission

@@
 
-The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that relays touch or pain sensations.
-The transmission control depends on the large and small diameter primary afferents, where large fibres inhibit, while small fibres facilitate the transmission.
+The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which contributes to the modern theory of pain, postulates that cells of the substantia gelatinosa selectively inhibits signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that integrate and relay somatosensation and pain.
+The transmission control depends on the large ($A\delta$ fibres)  and small diameter (C-fibres) primary afferents, where large fibres inhibit, while small fibres excite the the transmission cell.
 The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.
@@
 
-An important contribution of the gate-control theory was to emphasize central nervous system mechanisms \cite{Melzack1999}, and the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory.
-Contemporary views consider pain to be complex and multi-dimensional, where the processing of ascending nociceptive signals, in conjunction with descending modulatory pathways, shape the pain experience.
+An important contribution of the gate-control theory was to emphasize central nervous system mechanisms \cite{Melzack1999}, and the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferents with pattern theory.
 Melzack and Casey further proposed that pain consists of sensory-discriminative, affective-motivational, and cognitive-evaluative components \cite{Melzack1968}.
-Specifically, the sensory-discriminative component consists of peripheral nociceptive afferents transmit to the spinal dorsal horn (laminae V), whereby nociceptive specific (NS) and wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex.
-A central concept, which was also introduced by Melzack, is the  neuromatrix theory of pain, which dictates that the pain experience is produced by a "neurosignature" pattern, that is generated from widely distributed neural network, which Melzack refers as the "body-self neuromatrix".
+The sensory-discriminative component includes peripheral nociceptive afferents transmitting to the spinal dorsal horn (laminae V), whereby nociceptive specific (NS) and wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and eventually the primary sensory cortex.
+A central concept, also introduced by Melzack, is the  neuromatrix theory of pain, which dictates that the pain experience is produced by a "neurosignature" pattern, that is generated from widely distributed neural network, which Melzack refers as the "body-self neuromatrix".
 \cite{Melzack1999}.
-The concept is now more commonly referred to as the "pain matrix" \cite{Tracey2007a}.
-(See section \ref{section:neuromatrix})
+The concept is now more commonly referred to as the "pain matrix" \cite{Tracey2007a}, although the term itself is controversial \cite{iannetti2010neuromatrix} (See section \ref{section:neuromatrix}).
+Contemporary views consider pain to be complex and multi-dimensional, where the processing of ascending nociceptive signals, in conjunction with descending modulatory pathways, the default mode and salience network, along with temporal changes in these networks \cite{Kucyi2015}, shape the pain experience.
 
@@
 
-\textit{Nociceptors}, the peripheral nociceptors respond to noxious stimuli.
+Nociceptors are peripheral receptors that respond to noxious stimuli.
 These receptors have free nerve endings, and can respond to thermal, mechanical, and/or chemical stimuli, such as physical trauma or heat/cold temperatures.
-These receptors broadly separate into two categories, differentiated by the fibre type.
-Acute nociceptors are associated with $A\delta$ fibres, which permit fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma.
+These broadly separate into two categories, differentiated by the fibre type.
+Nociceptors responsible for acute pain are associated with $A\delta$ fibres, which permit fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma.
 Dull pain is triggered by receptors that are associated with slow conducting C fibres ($<1$ m/s).
-C-fibre nociceptors can be further differentiated into mechano-heat, silent, and polymodal subtypes.
-$A\delta$-fibre nociceptors can be divided into high-threshold mechano-heat, low-threshold mechano-heat, and noxious cold subtypes \cite{Julius2013}.
+\cite{Julius2013}.
 
@@
 \centering
-\caption{The axonal projections of the fast conduction ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn. Modified from \protect\cite{DMello2008}}
+\caption{The axonal projections of the fast conducting ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn. Modified from \protect\cite{DMello2008}}
 \label{fig:drg}
@@
 For example, the wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI.
-These neurons respond to both innocuous and noxious stimuli, and are decidedly non-specific.
+These neurons appear to respond to both innocuous and noxious stimuli, and are non-specific.
 They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing evidence that WDR neurons are responsive to prolonged nociceptive pain.
@@
 The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, including the periaqueductual gray.
-The \textit{cervicothalamic} tract include only the white matters of the upper two cervical segments.
-It takes inputs from laminae III and IV, and projects to the thalamus via the medial lemniscus.
 The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions \cite{kandel2000principles}.
@@
 Thus, the VP representation is somatotopically organized with the face region in the medial aspect, and lower limbs towards the lateral aspect.
-The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.
+The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect \cite{kandel2000principles}.
 
@@
 
-Pain percept, defined as the subjective experience of pain, is thought to involve distributed regions of the brain, coined as the pain “neuromatrix”\cite{Melzack1999a}, or more commonly the "pain matrix".
-The regions are not functionally independent, but rather thought to be dynamically inter-related.
+Pain percept, which is defined as the subjective experience of pain, is thought to involve distributed regions of the brain that are commonly referred to as the pain “neuromatrix”\cite{Melzack1999a}, or more commonly the "pain matrix".
+The regions are not functionally independent, but rather thought to be dynamically inter-related \cite{Kucyi2015}.
+As such, multiple neuroanatomical areas as well as networks are implied in the experience of pain.
 
- Important cortical regions involved in pain include anterior and middle cingulate cortices (ACC/MCC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1).
-The role of these cortical regions in pain were studied in neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during noxious stimulation.
+ Important cortical regions involved in pain, which demonstrated consistent BOLD co-activations during noxious stimulation, include anterior and middle cingulate cortices (ACC/MCC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1).
+The role of these cortical regions in pain have been further investigated using neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}.
 (Figure \ref{fig:pain-matrix}).
@@
 A recent study suggests that pain is correlated to the dynamically changing patterns of BOLD activities over time in the regions involved.
-Thus, the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}.
-The dynamic rhythms of activities that attune to and away from saliency, form a dynamic pain connectome \cite{Kucyi2015}.
-Chronic pain therefore demonstrated differential tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable noxious stimuli \cite{Baliki2008,Legrain2009}.
+Thus, the focus of the pain percept requires a switch away from the default mode network \cite{Kucyi2013}.
+The dynamic rhythms of activities that attune to and away from pain, form a dynamic pain connectome \cite{Kucyi2015}.
+Chronic pain demonstrated differential functional activation of the default mode network and attention patterns of the brain \cite{Baliki2008,Legrain2009}.
 
@@
 The posterior insula also receives direct nociceptive inputs from the VMpo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}.
-Therefore the insula, as well as S1, are believed to plays an important role in sensory-discrimination aspect of pain.
+Therefore the posterior insula, as well as S1, are believed to plays an important role in sensory-discrimination aspect of pain.
 
@@
 \paragraph{Default mode network}
-The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}.
+The default mode network (DMN) are regions of the brain which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}.
+The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation.
 The regions have been shown to consistently decrease in activity during attention-demanding, and non-self-referential tasks,when compared with activities during relaxed non-task states \cite{Raichle2015}.
 The DMN is believed to instantiate processes that support self-referential activities, emotional processing, and recollection \cite{Raichle2015}.
-The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), dorsal medial prefrontal cortex (dMPFC), posterior cingulate, inferior parietal, lateral temporal cortex, and hippocampal formation.
 Regions such as vmPFC and dMPFC overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}.
@@
 These gyri are not found consistently in the general population, but rather show individual variability \cite{Rosen2015}.
-Craig postulates that the insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}.
+Craig postulates that the anterior insula is a key region for homeostasis and interoception \cite{Craig2003b}, and is necessary for pain discriminative sensations via VMpo projections in the middle insula in humans \cite{Craig2003a}.
 The anterior insula has been shown by functional connectivity analysis to have interactions with anterior MCC and middle insula with posterior MCC \cite{Taylor2009}.
-Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring \cite{Vogt2016}.
+Further suggest that the anterior insula and MCC are involved in an emotional/salience network for interoception and emotional integration and monitoring, and posterior insula are involved in pain discrimination \cite{Vogt2016}.
 
@@
 
-Neuropathic pain can also arise from changes in the central neuroaxis.
-Trauma and other disruptions to the spinal pathways can affect noxious signal transmission.
-Where repeated and sustained noxious stimuli may cause the central nervous system to adapt and amplify responses to stimulation.
+Neuropathic pain can also arise from changes in the central nervous system.
+Trauma and other disruptions to the periphery or the spinal pathways can affect noxious signal transmission, where repeated and sustained noxious stimuli may cause the central nervous system to adapt and amplify responses to stimulation.
 These can involve activation-dependence plasticity in neurons of the dorsal horn, which result in windup of action potential discharge of these neurons.
@@
 White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis.
-It is believed that injury to white matter fibres can affect action potential conduction between cortical and subcortical structures.
 This may give rise or contribute to neuropathic pain \cite{Geha2008}.
@@
  Neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research.
-Brain regions that are involved in the somatosensation of pain process are involved in both acute and chronic pain.
-These regions include the posterior insula, lateral thalamus and dorsal striatum.
-In chronic pain conditions however, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Jensen2016,Alomar2016,Lorenz2005}.
+In chronic pain conditions, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Jensen2016,Alomar2016}.
 In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}.
-Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and medial prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}.
+Evidence also suggest that back pain chronification is associated with increase activation in the emotional regions (amygada and medial prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}.
  Additionally, evidence of a "pain memory" circuitry involving the MCC \cite{Vogt2016,Nevian2017}, and abnormalities in the reward circuitry \cite{Tracey2017} have been implicated in chronic pain.

### 2018-06-29 contents/reviews/trigeminal.tex
Updates from ShareLaTeX
final submission

@@
 The injury can also be intentional, due to interventions such as rhizotomy or gangliolysis; these result in numbness and loss of sensation of the face and are termed trigeminal deafferentation pain.
-Symptomatic TN, also termed MS-TN, is due to trigeminal nerve injury relating to multiple sclerosis related central white matter inflammation.
+Symptomatic TN, includes MS-TN, an important form of TN due to trigeminal injury relating to multiple sclerosis related central white matter inflammation and plaques.
 Postherpetic TN is pain resulting from the infection of the herpes zoster virus, and can involve burning pain and deafferentation.
@@
 
-The afferent branches converge and form the trigeminal root ganglion (TRG) (also known as Gasserian, or semilunar ganglion), which then leads to the trunk or cisternal component of CN V.
+The afferent branches converge and form the trigeminal ganglion (TRG) (also known as Gasserian, or semilunar ganglion), which then leads to the trunk or cisternal component of CN V.
 The TRG contains all the trigeminal sensory cell bodies.
@@
  \begin{figure}[ht]
- \includegraphics[width=\textwidth]{trig-pathway-wilson.jpg}
+ \includegraphics[width=110mm]{noback_2005_10_dot_6.jpg}
  \centering
- \caption{Overview of the trigeminal afferent pathways \protect\cite{wilson2010cranial}}
+ \caption{Overview of the trigeminal afferent pathways \protect\cite{noback2005human}. Reproduced with permission from Copyright Clearance Center (License ID: 4336271012841).}
  \label{fig:trig-pathway}
@@
 The region of disruption may include the peripheral-central myelin transition zone, and extending into central myelin regions.The TRG is also postulated to be responsible for neuralgia \cite{Rappaport1994}.
-It is believed that ganglionic ectopic autonomic firing can be triggered by mechanical disturbances as a result of nerve compression.
-Additionally, ganglionic cross aftercharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding neurons.
+It is believed that ganglionic ectopic firing can be triggered by mechanical disturbances as a result of nerve compression.
+Additionally, ganglionic cross action potential discharge may occur as a result of repeated stimulations in the nerve injury site, which ignites surrounding neurons.
 Histological findings by Devor et al.
@@
 GKRS treatments result in 75\% initial pain relief \cite{Tuleasca2015,Kondziolka2010}, about 50-80\% remain pain-free at 1 year, 45-65\% 2 years and 30-40\% 3 years after surgery \cite{Kondziolka2010,Sheehan2005}.
-More favourable treatment outcomes were related to no other accompanying symptoms, no prior surgery and shorter term pain history of fewer than three years.
+Better treatment outcomes were related to factors such as having no other accompanying symptoms, no prior surgery and shorter term pain history of fewer than three years.
 
@@
 
-Application of tractography permitted visual reconstructions of human cranial nerves in vivo \cite{Hodaie2010}.
+Tractography allows high quality \textit{in-vivo} reconstructions of human cranial nerves \cite{Hodaie2010}.
 Cisternal CN V could be reliably delineated with Gaussian tensors with just 25 MR gradient directions.
@@
 Lowered FA was reported in multiple TN studies \cite{Herweh2007, Lutz2011}.
-A Study in patients with temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}.
+A study in patients with temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}.
 This TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.

### 2018-06-29 contents/tn_sagit.tex
final submission

@@
 \subsubsection{Acquisitions}
-We studied 36 TN subjects with classic TN (28 females, 8 males; age 60.1$\pm$16.0 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12.0 years).
+We studied 36 TN subjects with classic TN (28 females, 8 males; age 60.1$\pm$SD 16.0 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$SD 12.0 years).
 All TN subjects underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment.
@@
 \includegraphics[width=\linewidth]{cnv-inferior-view-annotated.png}
-\caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
+\caption{Merged trigeminal nerve () reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
 The expected increase in FA as the fibres enter the brainstem can be visualized, and denotes good registration between subjects.
@@
 \includegraphics[width=\linewidth]{view-decussation.png}
-\caption[Focused view of the TPT and its decussation, prior to the generation of centroids.]{Focused view of the TPT and its decussation, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
+\caption[Focused view of the trigeminothalamic pathway (TPT) and its decussation, prior to the generation of centroids.]{Focused view of the TPT and its decussation, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).
 Colours were chosen arbitrarily to denote left and right pathways, however the colour intensity represents FA values as depicted by FA scale, top right corner.
@@
 \includegraphics[width=\linewidth]{view-left-right.png}
-\caption[Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.]{Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.
+\caption[Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the primary sensory (S1) cortex.]{Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the S1 cortex.
 Images obtained prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A).

### 2018-06-29 contents/tn_sagit_abstract.tex
final submission

@@
 \textbf{Introduction:}  Imaging of trigeminal neuralgia (TN) has demonstrated key DTI based diffusivity alterations in the trigeminal nerve but imaging has primarily focused on the peripheral nerve segment, since technical limitations have  prevented us from assessing the trigeminal nerve and its connections in the central nervous system.
-We apply fully-automated group white-matter tractography to image the trigeminal sensory pathway and machine learning classification using a Gaussian Process (GP) classifier to pinpoint key white-matter diffusivity changes that maximally differentiate TN subjects from healthy controls.
+We apply fully-automated group white-matter tractography to image the trigeminal sensory pathway and machine learning classification using a Gaussian Process (GP) classifier to pinpoint key white-matter diffusivity abnormalities that maximally differentiate TN subjects from healthy controls.
 
@@
 
-\textbf{Results:} GP classifier consistently demonstrated bilateral trigeminal changes, differentiating them from controls with an accuracy of 80\%, even though the clinical expression of pain was strictly unilateral.
+\textbf{Results:} GP classifier consistently demonstrated bilateral trigeminal differences, differentiating them from controls with an accuracy of 80\%, even though the clinical expression of pain was strictly unilateral.
 Affected and unaffected sides could also be differentiated from each other with an accuracy of 75\%.

### 2018-07-23 contents/abstract.tex
final sync, 2-sided print correction

@@
 \begin{abstract}
-\addcontentsline{toc}{chapter}{Abstract}
+% \addcontentsline{toc}{chapter}{Abstract}
 Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches.
 TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear.
-Single-tensor diffusion tensor neuroimaging (DTI) studies of CN V revealed diffusivity changes in the cistern segment.
-However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers.
-Thereby the study of TN using diffusion imaging is time-consuming and error-prone due to the complete manual involvement in all steps of data process, placement of region-of-interest (ROI), and tractography parameter tuning, and data analysis.
-There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments.
-2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further.
-The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway.
-Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve.
-In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image.
+Single-tensor diffusion tensor neuroimaging (DTI) studies of CNV revealed diffusivity changes in the cistern segment.
+However, the portion of the nerve within the brainstem remained elusive due to DTI limits.
+Diffusion imaging of TN is also error-prone due to manual data processing and analysis.
+
+This thesis aims to: 1) develop a fully-automated software framework to analyze diffusion tractography to reduce error and increase speed of experiments.
+2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art methods to advance the study of TN further.
+
+The specific aims are: a) establish the feasibility of multi-tensor tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway.
+Towards these goals, in Study I, we establish the feasibility of applying multi-tensor tractography to delineate the full course of CNV and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN by disruptions in the brainstem course of the nerve.
+In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image.
 In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) processing pipeline framework.
 Finally, in Study IV, we deploy end-to-end machine-learning TN classification to automatically discover diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway.
-In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification.
-The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.
+
+In sum, this thesis presents a detailed road-map of the development and application of end-to-end diffusion tractography machine learning classification.
+The application to TN revealed specific diffusivity changes in trigeminal CN V, pontine, and S1 white matter pathways, and pin-points the locations of the diffusivity disruptions at millimetre level.
 \end{abstract}

### 2018-07-23 contents/contributions.tex
final sync, 2-sided print correction

@@
-\chapter*{Statement of Contributions}
-\addcontentsline{toc}{chapter}{Statement of Contributions}
+\chapter{Statement of Contributions}
+% \addcontentsline{toc}{chapter}{Statement of Contributions}
 
@@
 \noindent
-Dr. Mojgan Hodaie (supervisor) - mentorship, assistance with experiment design, assistance with data interpretation, and assistance with thes
-is preparation.
+Dr. Mojgan Hodaie (supervisor) - mentorship, assistance with experiment design, assistance with data interpretation, and assistance with thesis preparation.
 
@@
 \noindent
-Dr. Jidan Zhong - assistance with experiment design, assistance with data interpretation, assitance with method development
+Dr. Jidan Zhong - assistance with experiment design, assistance with data interpretation, assistance with method development
 

### 2018-07-23 contents/hypotheses.tex
final sync, 2-sided print correction

@@
 
-\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
+\section[Study I]{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
 MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibres in the pons.
@@
 
-\section{Study II: Diffusion Weighted Image to T1 Co-registration Without Reversed-blip: Investigation of Best Practices}
+\section[Study II]{Study II: Diffusion Weighted Image to T1 Co-registration Without Reversed-blip: Investigation of Best Practices}
 The task of establishing inter-subject anatomical correspondences is imperative for extending tractography to multiple subjects.
@@
 
-\section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
+\section[Study III]{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
 The construction of an automated group tractography software framework depends on CN V brainstem delineation using multi-tensor tractography and the optimal DWI-T1 co-registration strategy.
@@
 
-\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
+\section[Study IV]{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
 The methodology developed in the previous study was applied to the study of the trigeminal sensory pathway.

### 2018-07-23 contents/paper1.tex
figure layout fixes
final sync, 2-sided print correction

@@
 Understanding the microstructural anatomy of the trigeminal fibres is essential to determine how MS affects CN V brainstem fibres and how alterations to CN V in MS are related to the clinical presentation of pain.
-Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure 1A-B) \cite{Cruccu2009,DaSilva2005}.
+Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure \ref{fig:MSfigure1}A-B) \cite{Cruccu2009,DaSilva2005}.
 Conventional MR imaging cannot visualize in isolation the brainstem fibres of CN V.
@@
 MTT allowed for the visualization of brainstem CN V fibres in isolation, by distinguishing them from the surrounding crossing cerebellar peduncular fibres.
-The SDT method could not adequately delineate the nerve (Figure 2).
+The SDT method could not adequately delineate the nerve (Figure \ref{fig:MSfigure2}).
 Comparison of tractography results between SDT and MTT showed that SDT was unable to differentiate fibres crossing the cerebellar peduncle fibres, while MTT delineated the brainstem course of CN V into the region of the trigeminal main sensory nucleus.
-Delineation of MS-TN CN V by MTT revealed that the tracts became visibly sparse as they came into proximity of the MS plaque (Figure 3A).
+Delineation of MS-TN CN V by MTT revealed that the tracts became visibly sparse as they came into proximity of the MS plaque (Figure \ref{fig:MSfigure3}A).
 FA decreases in proximity to the MS plaque could also be visualized by colour shift in the tractography model with more orange/yellow colour presented.
-The fibres of the unaffected CN V did not exhibit this change, and extended caudally into the trigeminal nuclei (Figure 3B).
+The fibres of the unaffected CN V did not exhibit this change, and extended caudally into the trigeminal nuclei (Figure \ref{fig:MSfigure3}B).
 
@@
 \subsubsection{Cisternal changes are unique in TN}
-In the cistern segment (Figure 4), statistically significant differences are found only between the TN group and other groups.
+In the cistern segment (Figure \ref{fig:MSfigure4}), statistically significant differences are found only between the TN group and other groups.
 No significant differences were found in between segments of other subject groups.
@@
 \subsubsection{REZ changes are found predominantly in TN}
-In the REZ (Figure 5), statistically significant results are found within, but not unique to TN.
+In the REZ (Figure \ref{fig:MSfigure5}), statistically significant results are found within, but not unique to TN.
 TN(+) FA is shown to be significantly lower than TN(-) FA, while TN(+) RD and MD are shown to be significantly greater than TN(-); TN(+) MD and RD are shown to be significantly greater compared to all other groups and sides; while TN AD showed no significant differences.
@@
 \subsubsection{Pontine diffusivities showed no significant changes in all groups}
-In the pontine tissue (Figure 6), there are no significant intra-group or inter-group differences in FA, RD and MD for all groups and sides.
+In the pontine tissue (Figure \ref{fig:MSfigure6}), there are no significant intra-group or inter-group differences in FA, RD and MD for all groups and sides.
 However MS-TN(-) AD is significantly higher than TN(+/-) and controls.
@@
 
-In the peri-lesional region (Figure 7), the patterns shows significant intra-group and inter-group differences in MS-TN.
+In the peri-lesional region (Figure \ref{fig:MSfigure7}), the patterns shows significant intra-group and inter-group differences in MS-TN.
 MS-TN(+) FA is significantly lower than that of MS-TN(-), as well as significantly lower than TN(+/-)and Controls, while MS-TN(-) FA showed no inter-group differences with TN and Controls.
@@
 Moreover TN+ shows significant different from MS-TN+ as well.
-Please refer to Figure 4 for explanations of layout and annotations.}
+Please refer to Figure \protect\ref{fig:MSfigure4} for explanations of layout and annotations.}
 \centering
@@
 However MS-TN- shows significantly different AD than TN- and controls.
-Please refer to Figure 4 for explanations of layout and annotations.}
+Please refer to Figure \protect\ref{fig:MSfigure4} for explanations of layout and annotations.}
 \centering
@@
 
-\begin{figure}[p]
+\begin{figure}[h]
 \includegraphics[width=\textwidth]{figure7.png}
@@
 RD of MS-TN- is also significantly lower than that of MS-TN+.
-Please refer to Figure 4 for explanations of layout and annotations.
-http://msj.sagepu}
+Please refer to Figure \protect\ref{fig:MSfigure4} for explanations of layout and annotations.}
 \centering
@@
 These changes highlight the differences in diffusivities between MS-TN, TN and healthy controls, thereby permitting their use as neuroimaging signatures for distinguishing these conditions.
-Previous studies of the CN V with DTI tractography have been met with limitations of the single-tensor Gaussian diffusion model, and were incapable of resolving brainstem trigeminal tracts (Figure 2A).
-Recent improvements in multi-tensor imaging are now capable of more complex diffusion profile of crossing fibres in neural tissue, thereby permitting reliable reconstruction of the brainstem CN V (Figure 2B), and subsequently the study of CN V changes in MS patients (Figure 3).
+Previous studies of the CN V with DTI tractography have been met with limitations of the single-tensor Gaussian diffusion model, and were incapable of resolving brainstem trigeminal tracts (Figure \ref{fig:MSfigure2}A).
+Recent improvements in multi-tensor imaging are now capable of more complex diffusion profile of crossing fibres in neural tissue, thereby permitting reliable reconstruction of the brainstem CN V (Figure \ref{fig:MSfigure2}B), and subsequently the study of CN V changes in MS patients (Figure \ref{fig:MSfigure3}).
 

### 2018-07-23 contents/sagit.tex
Updates from ShareLaTeX
final sync, 2-sided print correction

@@
 \subsection{Preprocessing}
-In order to automate the generation of tractography using multiple methods and minimize user-bias, the SAGIT framework is created (Figure 1).
+In order to automate the generation of tractography using multiple methods and minimize user-bias, the SAGIT framework is created (Figure \ref{fig:SGfigure1}).
 DWI sequences were corrected for eddy-current and motion distortions with appropriate rotational corrections to gradient vectors \cite{Leemans2009}.
@@
 where $n$ was the number of bins; $v_0$ was the number of voxels where $s > 0$;  $v_i$ was the number of voxels with $ 0 < s \ge \frac{i}{n}$; this study used $n=10$.
-Examples of NOS score behaviour are available in Figure 4.
+Examples of NOS score behaviour are available in Figure \ref{fig:SGfigure4}.
 The rater scores were then normalized to each anatomy, and the normalized rater scores were linearly regressed against the NOS scores for correlations.
@@
 \centering
-\includegraphics[width=140mm]{figure1.pdf}
+\includegraphics[width=0.8\linewidth]{figure1.pdf}
 \caption{Flow diagram of the SAGIT group tractography framework. The SAGIT framework is designed to be fully configurable and extensible.}
@@
 The fornix was shown to be consistent across the different algorithms.
-The visual report (Figure 2; the complete visual report can be found in supplemental) suggested that the subregions of the fornix were consistently delineated.
-The merged fornix tracts (Figure 3) showed different patterns of streamline distributions.
+The visual report (Figure \ref{fig:SGfigure2}; the complete visual report can be found in supplemental) suggested that the subregions of the fornix were consistently delineated.
+The merged fornix tracts (Figure \ref{fig:SGfigure3}) showed different patterns of streamline distributions.
 CST$_{prob}$ and XST both showed wide spread streamline dispersions, while CST$_{det}$ and particularly DTT were limited to the region of the fornix.
-The fornix ratings (Figure 5) showed essentially no variability between raters.
-The fornix NOS scores (Figure 6) similarly were consistent across the algorithms with little variability.
+The fornix ratings (Figure \ref{fig:SGfigure5}) showed essentially no variability between raters.
+The fornix NOS scores (Figure \ref{fig:SGfigure6}) similarly were consistent across the algorithms with little variability.
 Both filtered and unfiltered ratings were identical.
 
-For the CN VII/VIII, CST$_{prob}$, CST$_{det}$ and XST were all able to delineate the cranial as well as brainstem segments of the nerve (Figure 2).
+For the CN VII/VIII, CST$_{prob}$, CST$_{det}$ and XST were all able to delineate the cranial as well as brainstem segments of the nerve (Figure \ref{fig:SGfigure2}).
 CST$_{prob}$ however also delineated much of the cerebellum.
 DTT by Slicer was not able to delineate the brainstem portions of the fibre and therefore was scored lower than the other methods.
-These differences in delineation can be more clearly seen in the merged tracts (Figure 3), where the XST also showed further lateral delineations of CN VII/VIII.
+These differences in delineation can be more clearly seen in the merged tracts (Figure \ref{fig:SGfigure3}), where the XST also showed further lateral delineations of CN VII/VIII.
 In unfiltered tractography, there was much more inter-rater variability for all except CST$_{prob}$.
-CST$_{det}$ and XST also showed lower average rating when unfiltered (Figure 5).
-The NOS showed similar trend where DTT resulted in the lowest score (Figure 6).
+CST$_{det}$ and XST also showed lower average rating when unfiltered (Figure \ref{fig:SGfigure5}).
+The NOS showed similar trend where DTT resulted in the lowest score (Figure \ref{fig:SGfigure6}).
 The NOS also showed higher score for CST$_{det}$ when comparing to XST.
@@
 For CN X, CST$_{prob}$ was rated the highest with no inter-rater variability.
-CST$_{det}$ and XST resulted in more rating variability and a lower average score (Figure 5).
-CST$_{prob}$ visually showed delineations that extended into the ipsilateral higher brain regions even with filters (Figures 2, 3).
+CST$_{det}$ and XST resulted in more rating variability and a lower average score (Figure \ref{fig:SGfigure5}).
+CST$_{prob}$ visually showed delineations that extended into the ipsilateral higher brain regions even with filters (Figures \ref{fig:SGfigure2}, \ref{fig:SGfigure3}).
 DTT showed the worst reproducibility, although its anatomy was recognizable in a small number of individuals.
@@
 DTT also showed improved ratings when unfiltered.
-NOS scores (Figure 6) for both filtered and unfiltered CN X tractography similarly showed an overall increase in score with unfiltered tractography.
+NOS scores (Figure \ref{fig:SGfigure6}) for both filtered and unfiltered CN X tractography similarly showed an overall increase in score with unfiltered tractography.
 
-For the red nucleus projections, visually CST$_{prob}$ showed the most consistent reproducibility of the decussation (Figure 2, Figure 3), with XST showing higher reproducibility over CST$_{det}$.
+For the red nucleus projections, visually CST$_{prob}$ showed the most consistent reproducibility of the decussation (Figure \ref{fig:SGfigure2}, Figure \ref{fig:SGfigure3}), with XST showing higher reproducibility over CST$_{det}$.
 DTT failed to delineate the pathway and instead delineated an erroneous path that decussated at the corpus callosum.
 CST$_{prob}$, CST$_{det}$ and XST resulted similar ratings, with unfiltered tracts showing lower rating and higher variability.
-DTT had the lowest rating, and very high variability (Figure 5).
-NOS score of the red nucleus projections highlighted the stronger performance of the CST$_{prob}$ over the other algorithms, and XST result was scored higher than CST$_{det}$ (Figure 6).
+DTT had the lowest rating, and very high variability (Figure \ref{fig:SGfigure5}).
+NOS score of the red nucleus projections highlighted the stronger performance of the CST$_{prob}$ over the other algorithms, and XST result was scored higher than CST$_{det}$ (Figure \ref{fig:SGfigure6}).
 The NOS scores of unfiltered tracts were notably higher for all algorithms, as the resulting tracts seemed to delineate much of the cortical spinal projections.
 
- The optic radiation showed the greatest reproducibility with CST$_{prob}$, where the Meyer’s loop could be seen in the result (Figure 2, Figure 3).
+ The optic radiation showed the greatest reproducibility with CST$_{prob}$, where the Meyer’s loop could be seen in the result (Figure \ref{fig:SGfigure2}, Figure \ref{fig:SGfigure3}).
 CST$_{det}$ and XST showed low reproducibility of less than 20\%, and DTT failed to delineate any structure on the right side (see supplemental).
-The ratings reflected the observation (Figure 5), with CST$_{prob}$ having the full rating.
+The ratings reflected the observation (Figure \ref{fig:SGfigure5}), with CST$_{prob}$ having the full rating.
 CST$_{det}$ was rated higher in filtered tractography, while XST was rated higher when unfiltered.
 DTT also showed higher ratings when unfiltered.
-NOS score characterized the drop-off in reproducibility in filtered tractography for the deterministic methods (Figure 6), and the trend of rating under filtered tractography matched well with the ratings.
+NOS score characterized the drop-off in reproducibility in filtered tractography for the deterministic methods (Figure \ref{fig:SGfigure6}), and the trend of rating under filtered tractography matched well with the ratings.
 NOS scores for unfiltered tractography however showed little match with ratings.
@@
 The auditory radiation was the least reproducible region.
-In filtered tractography, although CST$_{prob}$ showed high reproducibility in tractography (Figure 2, Figure 3), it was not clear if its delineation was correct.
+In filtered tractography, although CST$_{prob}$ showed high reproducibility in tractography (Figure \ref{fig:SGfigure2}, Figure \ref{fig:SGfigure3}), it was not clear if its delineation was correct.
 Visually there were wide spread area of false positives towards the occipital and anterior temporal lobe.
@@
 DTT was completely unable to produce any structures at all.
-For filtered ratings,  CST$_{prob}$ was the highest, and with XST rated higher than CST$_{det}$ (Figure 5).
+For filtered ratings,  CST$_{prob}$ was the highest, and with XST rated higher than CST$_{det}$ (Figure \ref{fig:SGfigure5}).
 DTT scored zero due to the lack of delineated structures.
@@
 Unfiltered ratings were over all higher than filtered ratings.
-NOS scores’ trends for filtered tractography matched well with ratings, but also highlighted the large differences in reproducibility (Figure 6).
+NOS scores’ trends for filtered tractography matched well with ratings, but also highlighted the large differences in reproducibility (Figure \ref{fig:SGfigure6}).
 While NOS scores under unfiltered tractography did not correlate with ratings.
 
-Linear regression of rater scores against NOS (Figure 7) showed significant (p\textless0.05) correlation of the two sets of scores in filtered tractography.
+Linear regression of rater scores against NOS (Figure \ref{fig:SGfigure7}) showed significant (p\textless0.05) correlation of the two sets of scores in filtered tractography.
 However correlations were not significant (p\textgreater0.05) for unfiltered tractography.
@@
 \begin{figure}[p]
-\includegraphics[width=\linewidth]{figure3.png}
+\centering
+\includegraphics[width=0.9\linewidth]{figure3.png}
 \caption[Example of merged tractography models for each of the techniques and anatomy.]{Example of merged tractography models for each of the techniques and anatomy.
@@
 The Red Nuclei shows different metrics (FA and AD) depending on lateralization, in order to highlight the pontine decussation.}
-\centering
 \label{fig:SGfigure3}
@@
 	\centering
-	\includegraphics[width=\textwidth]{figure5.pdf}
+	\includegraphics[width=\linewidth]{figure5.pdf}
 	\caption[Normalized anatomical scores as rated by experts.]{Normalized anatomical scores as rated by experts.
@@
 \centering
-\includegraphics[width=\textwidth]{figure6.pdf}
+\includegraphics[width=\linewidth]{figure6.pdf}
 \caption{NOS scores for each anatomy. It can be noted that having ROI filters severely reduces reproducibility in deterministic methods.}
@@
 \centering
-\includegraphics[width=\textwidth]{figure7.pdf}
+\includegraphics[width=\linewidth]{figure7.pdf}
 \caption{Correlation between rating scores and NOS. Filtered tractography NOS is significantly (p\textless 0.05) correlated with ratings. However unfiltered tractography show no significance.}
@@
 When attempting to judge tractography delineation quality, human experts often are able to assess a particular delineation by visually identifying known anatomical priors.
-Direct tractography visualizations suggest that wide streamline dispersions often obscure anatomical details in visualizations, making anatomical assessment difficult (figure 3).
+Direct tractography visualizations suggest that wide streamline dispersions often obscure anatomical details in visualizations, making anatomical assessment difficult (Figure \ref{fig:SGfigure3}).
 Therefore, image conjunctions offer better ability to assess reproducibility than direct tractography renderings alone.
@@
 This is because in practice, incorrect tract path are often the result of imaging anomalies or algorithm limit, and are unstable across a population.
-This can be observed as a general trend in Figure 3, for example where DTI delineation of the rubrocerebellar tract is clearly incorrect, and results in low NOS score.
+This can be observed as a general trend in Figure \ref{fig:SGfigure3}, for example where DTI delineation of the rubrocerebellar tract is clearly incorrect, and results in low NOS score.
 Therefore, NOS score is capable of characterizing the uncertainty as a result of low algorithmic performance.
